---
title: "Humanized mice EAE analysis - hashed brain nuclei (Cerebellum / Olfactory Bulb / Striatum / Cortex)"
author: "Mattia Zaghi"
date: "2026-09-15"
output:
  html_document: default
  word_document: default
  pdf_document: default
---

## What is this vignette for?

This vignette processes the **hashed brain nuclei** samples defined in
`Preprocessing/config/config_hashing.yaml`. Four brain regions were pooled into
each 10x channel and separated afterwards by TotalSeq-B anti-Nuclear-Pore-Complex
hashtags:

| Hashtag | Antibody     | Region         |
|---------|--------------|----------------|
| AB1     | TotalSeq-B0451 | Cerebellum     |
| AB2     | TotalSeq-B0452 | Olfactory Bulb |
| AB3     | TotalSeq-B0453 | Striatum       |
| AB4     | TotalSeq-B0454 | Cortex         |

The design differs from the other datasets in this repository in three ways:

* **Region is a within-library variable, not a library.** All four regions share a
  channel, so region is recovered by demultiplexing rather than read off the sample
  sheet. A region-specific effect is therefore free of library batch.
* **`rep1` and `rep2` are technical replicates of the same animal** - one nuclei
  suspension split over two 10x channels. They carry no independent biological
  information and are pooled into a single per-animal profile for every statistical
  test; `rep` is retained only as a batch/QC covariate. The units of replication are
  `CFA` (n = 3 animals), `EAE-NonSymp` (n = 3) and `EAE-Symp` (**n = 1**).
* **These libraries are substantially more contaminated than the first-run brain
  nuclei.** Section 3 quantifies this against the five unhashed whole-hemisphere
  libraries of `analysis.Rmd` (`CFA_brain_1-2`, `EAE_brain_1-3`), which are single
  nuclei suspensions from the same protocol, the same tissue and the same Snakemake
  rules - hashing is the only step that differs. That makes them the right baseline:
  any gap between the two datasets is attributable to the hashing prep rather than to
  nuclei versus whole cells. The headline is that CellBender's *cell calling* fails
  on the hashed libraries even though its *denoising* works well, which is why section
  4 takes the cell definition from Cell Ranger and the counts from CellBender.

As in `analysis_CD45.Rmd`, the single symptomatic animal is reserved from testing:
with one biological replicate there is no within-group variance to estimate, so a
nominal p-value would reflect only within-animal cell-to-cell noise. `EAE-Symp` is
embedded, clustered, annotated and plotted alongside the rest and reported as effect
sizes without p-values (section 11).

Everything else mirrors the established pipeline: per-sample QC, per-sample
DoubletFinder, merge, SCTransform, clustering, per-region reclustering, and
pseudobulk DESeq2 within each cluster.

##1. Load packages and configure directories

```{r setup, include=FALSE}
library(Seurat)
library(future)
library(tidyverse)
library(dplyr)
library(ggthemes)
library(scales)
library(cowplot)
library(RColorBrewer)
library(openxlsx)
library(purrr)
library(forcats)
library(hdf5r)
library(patchwork)
library(pals)
library(knitr)
library(DoubletFinder)
library(ggpubr)
library(ggridges)
library(pheatmap)
library(DESeq2)
library(ggrepel)
library(yaml)
opts_knit$set(root.dir = "/ws/gcb/MZ/Humanized_mice_OD/hashing/") # change accordingly
```

```{r}
base_dir <- "/ws/gcb/MZ/Humanized_mice_OD/hashing/"

# The first-run brain libraries, used in section 3 only, as the same-protocol
# baseline for the contamination diagnostics: single nuclei suspensions from whole
# half-hemispheres, 2 CFA and 3 EAE, unhashed. These live one level up from the
# hashing directory, each in its own sample folder.
ref_dir <- "/ws/gcb/MZ/Humanized_mice_OD/"

# Their CellBender output sits in epochs_300/, not epochs/.
ref_epochs <- "epochs_300"

# knitr's root.dir points at the data directory, not the repository, so the
# config is addressed absolutely. Adjust repo_dir if the checkout moves.
repo_dir <- "/home/mattia/Humanized_mice_EAE"

res_dir     <- "/ws/gcb/MZ/Humanized_mice_OD/Results"
obj_dir     <- file.path(res_dir, "objects")
markers_dir <- file.path(res_dir, "cluster_markers")
plots_dir   <- file.path(res_dir, "Plots")
for (d in c(res_dir, obj_dir, markers_dir, plots_dir)) {
  dir.create(d, recursive = TRUE, showWarnings = FALSE)
}
```

##2. Sample sheet

Sample names are read from the preprocessing config rather than retyped, so this
vignette cannot drift out of sync with what Cell Ranger actually processed.
Condition, animal and technical replicate are parsed from the sample name
(`<condition>_Brain_<animal>_rep<n>`). `animal` is shared by `rep1` and `rep2` and
is the unit of replication for every condition contrast.

Region names come from the feature reference that Cell Ranger was given, for the
same reason. Seurat rewrites underscores in feature names to dashes, so the
hashtag names are sanitised here once and used consistently from this point on
(`Olfactory_Bulb` -> `Olfactory-Bulb`).

```{r}
config_path  <- file.path(repo_dir, "Preprocessing/config/config_hashing.yaml")
feature_path <- file.path(repo_dir,
                          "Preprocessing/config/feature_reference_TotalSeqB_NPC_hashtags.csv")
stopifnot(file.exists(config_path), file.exists(feature_path))

cfg             <- yaml::read_yaml(config_path)
hashing_samples <- names(cfg$samples)

feature_ref   <- read.csv(feature_path, stringsAsFactors = FALSE)
region_levels <- gsub("_", "-", feature_ref$name)
hto_ids       <- feature_ref$id
names(region_levels) <- hto_ids
region_levels

sample_sheet <- data.frame(
  sample = hashing_samples,
  stringsAsFactors = FALSE
)
# <condition>_Brain_<animal>_rep<n>
sample_sheet$condition <- sub("_Brain_[0-9]+_rep[0-9]+$", "", sample_sheet$sample)
sample_sheet$animal_n  <- sub("^.*_Brain_([0-9]+)_rep[0-9]+$", "\\1", sample_sheet$sample)
sample_sheet$rep       <- sub("^.*_(rep[0-9]+)$", "\\1", sample_sheet$sample)
# rep1 and rep2 are the same mouse, so the animal id drops the replicate.
sample_sheet$animal    <- paste(sample_sheet$condition, sample_sheet$animal_n, sep = "_")

sample_sheet$condition <- factor(sample_sheet$condition,
                                 levels = c("CFA", "EAE-NonSymp", "EAE-Symp"))
sample_sheet$rep       <- factor(sample_sheet$rep)

# Sanity checks: two channels per animal, and how many animals per condition
with(sample_sheet, table(condition, rep))
with(sample_sheet, tapply(animal, condition, function(x) length(unique(x))))
sample_sheet
```

##3. Audit of the Cell Ranger and CellBender runs

This section is pure file parsing - it reads the metrics that the pipeline already
wrote and puts the 14 hashing libraries next to the 5 first-run brain libraries
(`CFA_brain_1-2`, `EAE_brain_1-3`). Those are the right comparison: single nuclei
suspensions from whole half-hemispheres of the same model, prepared by the same
protocol and processed by the same Snakemake rules with the same Cell Ranger 9.0.1,
the same GRCh38-and-mm10 reference and the same CellBender settings (`--fpr 0.01`).
The only difference is the hashing step. A gap between the two datasets therefore
isolates the hashing prep, where a comparison against sorted whole cells would
confound it with the nuclei-versus-cells difference.

This runs before any data is loaded, because its result determines how the data is
loaded in section 4.

```{r}
# Cell Ranger writes percentages and thousands separators into metrics_summary.csv.
parse_metric <- function(x) {
  x <- gsub(",", "", as.character(x))
  pct <- grepl("%", x, fixed = TRUE)
  v <- suppressWarnings(as.numeric(gsub("%", "", x)))
  v[pct] <- v[pct] / 100
  v
}

read_cellranger_metrics <- function(dir, samples) {
  do.call(rbind, lapply(samples, function(s) {
    f <- file.path(dir, s, "RNA_AAAAGGGG/cellranger/outs/metrics_summary.csv")
    if (!file.exists(f)) return(NULL)
    m <- read.csv(f, check.names = FALSE, stringsAsFactors = FALSE)
    get1 <- function(col) if (col %in% colnames(m)) parse_metric(m[[col]][1]) else NA_real_
    data.frame(
      sample         = s,
      cr_cells       = get1("Estimated Number of Cells"),
      mean_reads     = get1("Mean Reads per Cell"),
      saturation     = get1("Sequencing Saturation"),
      pct_txome      = get1("Reads Mapped Confidently to Transcriptome"),
      frac_in_cells  = get1("Fraction Reads in Cells"),
      hg38_cells     = get1("GRCh38 Estimated Number of Cell Partitions"),
      mm10_cells     = get1("mm10 Estimated Number of Cell Partitions"),
      barnyard_mult  = get1("Fraction GEMs with >1 Cell"),
      ab_reads       = get1("Antibody: Number of Reads"),
      ab_usable      = get1("Antibody: Fraction Antibody Reads Usable"),
      ab_unrecog     = get1("Antibody: Fraction Unrecognized Antibody"),
      ab_median_umi  = get1("Antibody: Median UMIs per Cell (summed over all recognized antibody barcodes)"),
      stringsAsFactors = FALSE
    )
  }))
}

read_cellbender_metrics <- function(dir, samples, epochs_dir = "epochs") {
  do.call(rbind, lapply(samples, function(s) {
    ep <- file.path(dir, s, "RNA_AAAAGGGG", epochs_dir)
    f  <- file.path(ep, "cellbender_output_metrics.csv")
    if (!file.exists(f)) return(NULL)
    m <- read.csv(f, header = FALSE, stringsAsFactors = FALSE)
    v <- setNames(as.numeric(m[[2]]), m[[1]])

    # The cell and empty count priors are only in the log, and their ratio is the
    # single most informative contamination diagnostic: it is how well separated
    # the cell and empty-droplet count distributions are.
    lg <- file.path(ep, "cellbender_output.log")
    grab <- function(pat) {
      if (!file.exists(lg)) return(NA_real_)
      hit <- grep(pat, readLines(lg, warn = FALSE), value = TRUE)
      if (length(hit) == 0) return(NA_real_)
      as.numeric(sub(".*?([0-9]+)\\s*$", "\\1", tail(hit, 1)))
    }
    data.frame(
      sample          = s,
      cb_cells        = v[["found_cells"]],
      cb_expected     = v[["expected_cells"]],
      cb_ratio        = v[["ratio_of_found_cells_to_expected_cells"]],
      cb_empties      = v[["found_empties"]],
      frac_nonempty   = v[["fraction_of_analyzed_droplets_that_are_nonempty"]],
      frac_removed    = v[["fraction_counts_removed"]],
      counts_out      = v[["output_average_counts_per_cell"]],
      convergence     = v[["convergence_indicator"]],
      prior_cells     = grab("Prior on counts for cells is"),
      prior_empty     = grab("Prior on counts for empty droplets is"),
      largest_empty   = grab("Largest surely-empty droplet has"),
      stringsAsFactors = FALSE
    )
  }))
}
```

```{r}
qc_hash <- left_join(read_cellranger_metrics(base_dir, hashing_samples),
                     read_cellbender_metrics(base_dir, hashing_samples),
                     by = "sample") %>%
  left_join(sample_sheet, by = "sample") %>%
  mutate(dataset          = "hashing (nuclei)",
         prior_ratio      = prior_cells / prior_empty,
         cb_over_cr       = cb_cells / cr_cells,
         frac_human_cells = hg38_cells / (hg38_cells + mm10_cells))

# Baseline sample names come from the first run's own config, filtered to brain, so
# the spleen libraries of that run are excluded and nothing is retyped.
ref_config  <- file.path(repo_dir, "Preprocessing/config/config.yaml")
stopifnot(file.exists(ref_config))
ref_samples <- grep("_brain_", names(yaml::read_yaml(ref_config)$samples),
                    value = TRUE)
ref_samples <- ref_samples[file.exists(file.path(
  ref_dir, ref_samples, "RNA_AAAAGGGG", ref_epochs,
  "cellbender_output_metrics.csv"))]
cat("Baseline libraries:", paste(ref_samples, collapse = ", "), "\n")

qc_ref <- left_join(read_cellranger_metrics(ref_dir, ref_samples),
                    read_cellbender_metrics(ref_dir, ref_samples,
                                            epochs_dir = ref_epochs),
                    by = "sample") %>%
  mutate(dataset          = "first run (unhashed nuclei)",
         condition        = factor(sub("_brain_[0-9]+$", "", sample)),
         prior_ratio      = prior_cells / prior_empty,
         cb_over_cr       = cb_cells / cr_cells,
         frac_human_cells = hg38_cells / (hg38_cells + mm10_cells))

qc_hash %>%
  select(sample, condition, rep, cr_cells, mean_reads, saturation, pct_txome,
         frac_in_cells, mm10_cells, hg38_cells, frac_human_cells) %>%
  arrange(sample)
```

The contamination comparison. Every row is a diagnostic where the two datasets
should look alike, since they are the same tissue, the same nuclei protocol and the
same pipeline.

```{r}
# Common columns only, so the two sample sheets (which differ) do not clash.
shared_cols <- intersect(names(qc_hash), names(qc_ref))
qc_both     <- bind_rows(select(qc_hash, all_of(shared_cols)) %>%
                           mutate(condition = as.character(condition)),
                         select(qc_ref, all_of(shared_cols)) %>%
                           mutate(condition = as.character(condition)))

qc_compare <- qc_both %>%
  group_by(dataset) %>%
  summarise(
    n_libraries            = n(),
    `Fraction reads in cells` = sprintf("%.2f - %.2f (med %.2f)",
                                        min(frac_in_cells), max(frac_in_cells),
                                        median(frac_in_cells)),
    `Cell:empty count prior`  = sprintf("%.0fx - %.0fx", min(prior_ratio), max(prior_ratio)),
    `Fraction counts removed` = sprintf("%.3f - %.3f", min(frac_removed), max(frac_removed)),
    `CellBender/CellRanger cells` = sprintf("%.1fx - %.1fx", min(cb_over_cr), max(cb_over_cr)),
    `Counts per cell after denoising` = sprintf("%.0f - %.0f", min(counts_out), max(counts_out)),
    .groups = "drop"
  )
knitr::kable(t(qc_compare))
```

Everything follows from the cell:empty count-prior ratio - how far apart CellBender
puts the count distributions of cells and of empty droplets. The first-run nuclei
separate them by roughly an order of magnitude; the hashed libraries by about
fourfold. Both are far from the clean separation seen in sorted whole cells, which
is expected for nuclei, but the hashed libraries are about three times worse than
our own unhashed nuclei, and the consequences scale with it: three to four times
more counts removed as ambient, less than half the counts per cell surviving
denoising, and - the point that matters for section 4 - a complete breakdown of cell
calling that is absent from the baseline.

The second panel is the decisive one. On the unhashed libraries CellBender and Cell
Ranger agree on the number of cells to within a few percent, so either cell
definition could be used. On the hashed libraries CellBender calls four to six times
as many, because with the cell and empty distributions overlapping it has no basis
for the boundary. That is a failure of *cell calling* specifically, not of
denoising, and it is what motivates taking barcodes from one tool and counts from
the other.

```{r, fig.height=5, fig.width=14}
qc_both %>%
  ggplot(aes(x = prior_ratio, y = frac_removed, colour = dataset)) +
  geom_point(size = 3, alpha = 0.85) +
  scale_x_log10() +
  labs(x = "CellBender cell:empty count prior ratio (log scale)",
       y = "Fraction of counts removed as ambient", colour = NULL) +
  theme_classic(base_size = 13) -> g1

qc_both %>%
  ggplot(aes(x = cr_cells, y = cb_cells, colour = dataset)) +
  geom_abline(slope = 1, linetype = "dashed", colour = "grey50") +
  geom_point(size = 3, alpha = 0.85) +
  labs(x = "Cell Ranger estimated cells",
       y = "CellBender called cells", colour = NULL,
       subtitle = "Dashed line: agreement") +
  theme_classic(base_size = 13) -> g2

g1 + g2
```

Per-library detail behind the two panels, hashing libraries first:

```{r}
qc_both %>%
  select(dataset, sample, cr_cells, cb_cells, cb_over_cr, prior_cells, prior_empty,
         prior_ratio, largest_empty, frac_in_cells, frac_removed, counts_out) %>%
  arrange(dataset, desc(cb_over_cr))
```

```{r, fig.height=5, fig.width=12}
ggplot(qc_hash, aes(x = reorder(sample, frac_removed), y = frac_removed,
                    fill = condition)) +
  geom_col() +
  coord_flip() +
  scale_y_continuous(labels = percent) +
  labs(x = NULL, y = "Fraction of counts removed as ambient by CellBender") +
  theme_classic(base_size = 12)
```

Libraries to keep an eye on. Two criteria are used, and **both** must fire, because
neither is decisive alone:

* `convergence_indicator` well above zero means the ELBO was still moving when
  training stopped. On its own this says little - the baseline libraries reach
  comparable values without any sign of trouble, so it is a noisy indicator rather
  than a verdict. The baseline range is printed below for exactly this reason.
* A `found_empties` count in the low hundreds together with
  `fraction_of_analyzed_droplets_that_are_nonempty` near 1 means CellBender found
  almost no empty-droplet population in the range it analysed, so the ambient profile
  it subtracted is essentially unconstrained. This one *is* specific: the baseline
  libraries find thousands of empties and classify only 64-85% of analysed droplets
  as non-empty.

A library flagged on both counts should not be the one carrying a cluster on its own;
section 8 plots composition against `is_flagged` to check that.

```{r}
cat("Baseline convergence_indicator:",
    sprintf("%.2f - %.2f", min(qc_ref$convergence), max(qc_ref$convergence)), "\n")
cat("Baseline found_empties:",
    sprintf("%.0f - %.0f", min(qc_ref$cb_empties), max(qc_ref$cb_empties)),
    "| fraction non-empty:",
    sprintf("%.2f - %.2f", min(qc_ref$frac_nonempty), max(qc_ref$frac_nonempty)), "\n")

qc_flags <- qc_hash %>%
  mutate(flag_convergence = convergence > 0.8,
         flag_no_empties  = cb_empties < 500 & frac_nonempty > 0.97) %>%
  select(sample, convergence, cb_empties, frac_nonempty,
         flag_convergence, flag_no_empties) %>%
  arrange(desc(flag_convergence & flag_no_empties), desc(convergence))
qc_flags

flagged_samples <- qc_flags$sample[qc_flags$flag_convergence & qc_flags$flag_no_empties]
cat("Flagged on both criteria:", paste(flagged_samples, collapse = ", "), "\n")
```

Two further points that change how the standard metrics should be read.

**The barnyard multiplet estimate is not usable here.** These are host-mouse brains
carrying a human oligodendrocyte graft, so one genome dominates completely.
`Fraction GEMs with >1 Cell` is derived from human-mouse collisions, and with the
human cell partitions in the tens against thousands of mouse partitions it is
estimated from a handful of events. The baseline libraries are included below to show
this is a property of the model and not of the hashing prep - they have the same
handful of human partitions. It is reported for completeness and is *not* used to set
the doublet rate; section 7 uses the hashtags for that instead.

```{r, fig.height=5, fig.width=12}
qc_both %>%
  select(dataset, sample, mm10_cells, hg38_cells, barnyard_mult) %>%
  mutate(pct_human = 100 * hg38_cells / (hg38_cells + mm10_cells)) %>%
  arrange(dataset, desc(pct_human))
```

**The hashtag library is very deeply sequenced but only a small fraction of its
reads land in cells**, which is the same ambient story seen in the transcriptome.
The baseline has no counterpart for this table - it was not hashed - so it is
hashing-only. What matters for demultiplexing is the per-cell hashtag depth, and that
is ample (roughly a thousand UMI per cell and up).

```{r, fig.height=5, fig.width=12}
qc_hash %>%
  select(sample, ab_reads, ab_usable, ab_unrecog, ab_median_umi) %>%
  arrange(ab_usable)
```

##4. Create the Seurat objects

The audit above leads to one departure from `analysis.Rmd` and `analysis_CD45.Rmd`,
and it is worth being explicit about why, because it is the step that determines
which cells the rest of the vignette analyses.

CellBender does two separate jobs: it decides **which droplets are cells**, and it
decides **what the counts in those droplets should be** once ambient RNA is
subtracted. On these libraries the second job is done well and the first is not.

This is specific to the hashed libraries, which is why `analysis.Rmd` did not need
it: on the first-run brain nuclei CellBender's call set agrees with Cell Ranger's to
within a few percent, so taking either one gives the same cells. Here they disagree
by four- to sixfold, so the choice has to be made deliberately.

*Why the cell calling fails.* CellBender separates cells from empties using the
total-count distribution, and section 3 showed those two distributions barely
separate here (prior ratio around 4, surely-empty droplets holding hundreds of UMI).
The result is that thousands of ambient-only droplets cross the cell-probability
threshold: for `CFA_Brain_1_rep1` CellBender calls 22,539 cells whose *median*
content is about 108 UMI in 80 genes. Cell Ranger's EmptyDrops-style call is more
conservative because it tests each barcode's expression *profile* against the
ambient profile instead of thresholding total counts, so it is not fooled the same
way - it calls 3,718 cells for the same library.

*Why `min.features = 200` does not rescue it.* That filter selects on the exact
quantity ambient inflates. Applied to the CellBender call set of the same library it
still leaves 7,906 barcodes, more than twice Cell Ranger's count, so the ambient-only
droplets are simply being admitted at a higher threshold.

*Why the two can be combined cleanly.* `cellbender_output_seurat.h5` is indexed by
the full raw barcode list (about 1.13 M barcodes), not by CellBender's call set, so
any column subset can be taken from it. Cell Ranger's called barcodes are
essentially a subset of CellBender's (3,712 of 3,718 for this library), which means
every Cell Ranger cell was already treated as a cell by CellBender's model - the
denoised counts for those columns are the model's cell-mode output, not an
extrapolation onto something it considered empty. So the intersection keeps
CellBender's ambient correction and swaps in Cell Ranger's more conservative cell
definition. Barcode strings carry the same `-1` suffix in both files, so they join
directly.

The handful of Cell Ranger cells that CellBender did not call cells are reported per
library below; their counts are heavily stripped and they generally fall out on
`min.features` anyway.

The feature-name normalisation from `analysis_CD45.Rmd` is applied **to the gene
expression block only**. Applying it to the hashtags as well would rewrite
`Olfactory_Bulb` to `Olfactory-Bulb` inside the matrix while the sanitised names
from section 2 were derived separately, and the two must not be allowed to drift
apart. Keeping the gene normalisation byte-identical to the other vignettes is what
lets gene names (`mm10-Gabra6`, `GRCh38-...`) match the already-annotated brain
object used for label transfer in section 8.

```{r}
mt_cutoff <- 5   # <- provisional; revisited in section 6 with the distribution

load_hashed_sample <- function(sample) {
  cb_h5 <- file.path(base_dir, sample, "RNA_AAAAGGGG/epochs",
                     "cellbender_output_seurat.h5")
  cr_h5 <- file.path(base_dir, sample, "RNA_AAAAGGGG/cellranger/outs",
                     "filtered_feature_bc_matrix.h5")
  if (!file.exists(cb_h5) || !file.exists(cr_h5)) {
    warning("Missing input for ", sample)
    return(NULL)
  }

  cb <- Read10X_h5(filename = cb_h5, use.names = TRUE)
  cr <- Read10X_h5(filename = cr_h5, use.names = TRUE)
  if (!is.list(cb) || !"Antibody Capture" %in% names(cb)) {
    stop("No Antibody Capture block in ", cb_h5)
  }
  cb_gex <- cb[["Gene Expression"]]
  cb_hto <- cb[["Antibody Capture"]]
  cr_gex <- if (is.list(cr)) cr[["Gene Expression"]] else cr

  # Cell definition from Cell Ranger, counts from CellBender
  keep <- intersect(colnames(cr_gex), colnames(cb_gex))
  message(sample, ": ", ncol(cr_gex), " Cell Ranger cells, ",
          ncol(cr_gex) - length(keep), " absent from the CellBender output")

  cb_cells <- readLines(file.path(base_dir, sample, "RNA_AAAAGGGG/epochs",
                                  "cellbender_output_cell_barcodes.csv"))
  message("  of those, ", sum(!keep %in% cb_cells),
          " were not called cells by CellBender")

  gex <- cb_gex[, keep, drop = FALSE]
  hto <- cb_hto[, keep, drop = FALSE]

  # Gene names only: underscores -> dash, collapse repeated dashes
  fn <- rownames(gex)
  fn <- gsub("_+", "-", fn)
  fn <- gsub("-{2,}", "-", fn)
  rownames(gex) <- fn

  # Hashtags get the sanitised region names resolved in section 2, matched by the
  # feature reference id where possible so the order cannot silently shift.
  hto_rows <- rownames(hto)
  mapped <- region_levels[match(hto_rows, hto_ids)]
  if (any(is.na(mapped))) mapped <- gsub("_", "-", hto_rows)
  rownames(hto) <- mapped

  colnames(gex) <- paste0(sample, "_", colnames(gex))
  colnames(hto) <- colnames(gex)

  obj <- CreateSeuratObject(counts = gex, project = sample,
                            min.cells = 3, min.features = 200)
  obj[["HTO"]] <- CreateAssayObject(counts = hto[, colnames(obj), drop = FALSE])

  i <- match(sample, sample_sheet$sample)
  obj$sample    <- sample
  obj$condition <- as.character(sample_sheet$condition[i])
  obj$animal    <- sample_sheet$animal[i]
  obj$rep       <- as.character(sample_sheet$rep[i])
  obj
}

seurat_list <- list()
for (sample in sample_sheet$sample) {
  message("Loading ", sample)
  obj <- load_hashed_sample(sample)
  if (!is.null(obj)) seurat_list[[sample]] <- obj
}

data.frame(sample = names(seurat_list),
           cells  = vapply(seurat_list, ncol, numeric(1)),
           row.names = NULL)
```

##5. Hashtag demultiplexing: assigning a brain region to every nucleus

Hashtag counts are CLR-normalised **per feature** (`margin = 2`). This is not the
default and it matters here: section 3 showed the four tags differ in total
abundance by up to about fourfold, and a per-cell normalisation would leave that
imbalance in place and bias every call toward the most abundant tag.

`HTODemux` is the primary caller. Because CellBender has already removed most of the
ambient hashtag signal, the per-cell distribution is close to binary, which is an
easier problem than `HTODemux` was designed for but also an unusual input for its
background fit - so `MULTIseqDemux` and a plain dominant-fraction rule are run
alongside it and the three are compared. If they agree, the call is not
caller-dependent and `HTODemux` stands.

```{r}
dominant_call <- function(counts_mat, min_frac = 0.6, min_total = 50) {
  tot  <- Matrix::colSums(counts_mat)
  frac <- sweep(as.matrix(counts_mat), 2, pmax(tot, 1), "/")
  best <- rownames(counts_mat)[apply(frac, 2, which.max)]
  out  <- ifelse(tot < min_total, "Negative",
                 ifelse(apply(frac, 2, max) >= min_frac, best, "Doublet"))
  setNames(out, colnames(counts_mat))
}

for (sample in names(seurat_list)) {
  message("Demultiplexing ", sample)
  obj <- seurat_list[[sample]]

  obj <- NormalizeData(obj, assay = "HTO", normalization.method = "CLR",
                       margin = 2, verbose = FALSE)
  obj <- HTODemux(obj, assay = "HTO", positive.quantile = 0.99, verbose = FALSE)
  obj$hto_demux <- as.character(obj$hash.ID)

  obj <- MULTIseqDemux(obj, assay = "HTO", autoThresh = TRUE, verbose = FALSE)
  obj$hto_multiseq <- as.character(obj$MULTI_ID)

  obj$hto_dominant <- dominant_call(
    GetAssayData(obj, assay = "HTO", layer = "counts")
  )[colnames(obj)]

  cat("\n---", sample, "---\n")
  print(table(HTODemux = obj$hto_demux, MULTIseq = obj$hto_multiseq))
  cat("HTODemux vs dominant-fraction agreement on singlets: ",
      round(mean(obj$hto_demux[obj$hto_demux %in% region_levels] ==
                 obj$hto_dominant[obj$hto_demux %in% region_levels]), 3), "\n")

  seurat_list[[sample]] <- obj
}
```

Diagnostics for each library: how cleanly the four tags separate, and how many
nuclei each region contributed.

```{r, fig.height=10, fig.width=14}
for (sample in names(seurat_list)) {
  obj <- seurat_list[[sample]]
  Idents(obj) <- "hto_demux"
  print(
    RidgePlot(obj, assay = "HTO", features = rownames(obj[["HTO"]]), ncol = 2) +
      patchwork::plot_annotation(title = paste(sample, "- hashtag CLR by call"))
  )
  print(HTOHeatmap(obj, assay = "HTO") + ggtitle(sample))
}
```

```{r, fig.height=6, fig.width=14}
# Read straight out of @meta.data and force a plain unnamed character vector, so
# neither `$.Seurat` nor the cell names can leak into the data frame. The column is
# called hto_call rather than `call`, which would shadow base::call.
hto_summary <- dplyr::bind_rows(lapply(names(seurat_list), function(s) {
  md <- seurat_list[[s]]@meta.data
  data.frame(sample   = s,
             hto_call = unname(as.character(md$hto_demux)),
             row.names = NULL,
             stringsAsFactors = FALSE)
})) %>%
  dplyr::count(sample, hto_call, name = "n") %>%
  left_join(sample_sheet, by = "sample")

hto_summary %>%
  mutate(class = ifelse(hto_call %in% region_levels, "Singlet", hto_call)) %>%
  dplyr::count(sample, class, wt = n, name = "n") %>%
  pivot_wider(names_from = class, values_from = n, values_fill = 0) %>%
  mutate(pct_singlet = round(100 * Singlet / (Singlet + Doublet + Negative), 1))

ggplot(filter(hto_summary, hto_call %in% region_levels),
       aes(x = sample, y = n, fill = factor(hto_call, levels = region_levels))) +
  geom_col(position = "stack") +
  coord_flip() +
  scale_fill_manual(values = as.vector(pals::tableau20(4))) +
  labs(x = NULL, y = "Singlet nuclei", fill = "Region") +
  theme_classic(base_size = 12)

ggplot(filter(hto_summary, hto_call %in% region_levels),
       aes(x = sample, y = n, fill = factor(hto_call, levels = region_levels))) +
  geom_col(position = "fill") +
  coord_flip() +
  scale_y_continuous(labels = percent) +
  scale_fill_manual(values = as.vector(pals::tableau20(4))) +
  labs(x = NULL, y = "Fraction of singlet nuclei", fill = "Region") +
  theme_classic(base_size = 12)
```

Keep the singlets, record the region, and store the hashtag doublet rate per library
- section 7 uses it to set the DoubletFinder expectation.

```{r}
hto_rates <- list()

for (sample in names(seurat_list)) {
  obj <- seurat_list[[sample]]
  call <- obj$hto_demux

  # Rates are computed over singlets + doublets; Negative cells are excluded
  # because they are a detection failure, not a droplet class.
  n_singlet <- sum(call %in% region_levels)
  n_doublet <- sum(call == "Doublet")
  p <- prop.table(table(factor(call[call %in% region_levels], levels = region_levels)))

  hto_rates[[sample]] <- list(
    f_obs     = n_doublet / (n_singlet + n_doublet),
    sum_p2    = sum(as.numeric(p)^2),
    n_singlet = n_singlet,
    n_doublet = n_doublet,
    n_neg     = sum(call == "Negative")
  )

  keep <- colnames(obj)[call %in% region_levels]
  obj  <- subset(obj, cells = keep)
  obj$region <- factor(as.character(obj$hto_demux), levels = region_levels)
  seurat_list[[sample]] <- obj
}

do.call(rbind, lapply(names(hto_rates), function(s) {
  r <- hto_rates[[s]]
  data.frame(sample = s, singlets = r$n_singlet, hto_doublets = r$n_doublet,
             negatives = r$n_neg,
             hto_doublet_rate = round(r$f_obs, 4),
             sum_p2 = round(r$sum_p2, 3))
}))
```

##6. Cell quality control

Mitochondrial content is computed against both halves of the joint
GRCh38-and-mm10 reference and summed, as in `analysis.Rmd` and
`analysis_CD45.Rmd`.

For **nuclei** this metric carries extra weight. A nucleus should contain almost no
mitochondrial transcript, so whatever mitochondrial signal is present is
cytoplasmic carry-over or ambient - exactly the contamination section 3 quantified.
`analysis.Rmd` used `percent.mt < 5` on comparable nuclei, so 5 is the established
value for this tissue and keeping it preserves comparability. But section 3 showed
three to four times more ambient here than in those libraries, so 5 may well be too
loose to do any work. The distribution is plotted first, the retention under 1 %,
2 % and 5 % is tabulated, and `mt_cutoff` is set from that below: if the bulk of
nuclei sit under 1-2 % then 5 is passing everything through and should be tightened.

```{r}
for (sample in names(seurat_list)) {
  obj <- seurat_list[[sample]]
  obj[["percent.mt.mm10"]] <- PercentageFeatureSet(obj, pattern = "mt-")
  obj[["percent.mt.hg38"]] <- PercentageFeatureSet(obj, pattern = "MT-")
  obj[["percent.mt"]]      <- obj[["percent.mt.mm10"]] + obj[["percent.mt.hg38"]]
  seurat_list[[sample]] <- obj
}

qc_cells <- do.call(rbind, lapply(names(seurat_list), function(s) {
  md <- seurat_list[[s]]@meta.data
  data.frame(sample = s, region = md$region, percent.mt = md$percent.mt,
             nFeature_RNA = md$nFeature_RNA, nCount_RNA = md$nCount_RNA)
}))
```

```{r, fig.height=10, fig.width=14}
p_mt <- ggplot(qc_cells, aes(x = sample, y = percent.mt)) +
  geom_violin(fill = "grey85", scale = "width") +
  geom_hline(yintercept = c(1, 2, 5), linetype = "dashed",
             colour = c("firebrick", "darkorange", "grey40")) +
  coord_flip() +
  labs(x = NULL, y = "percent.mt (mm10 + hg38)",
       subtitle = "Dashed: 1 % / 2 % / 5 %") +
  theme_classic(base_size = 12)

p_feat <- ggplot(qc_cells, aes(x = sample, y = nFeature_RNA)) +
  geom_violin(fill = "grey85", scale = "width") +
  scale_y_log10() + coord_flip() +
  labs(x = NULL, y = "Genes detected (log scale)") +
  theme_classic(base_size = 12)

p_count <- ggplot(qc_cells, aes(x = sample, y = nCount_RNA)) +
  geom_violin(fill = "grey85", scale = "width") +
  scale_y_log10() + coord_flip() +
  labs(x = NULL, y = "UMI after CellBender (log scale)") +
  theme_classic(base_size = 12)

p_mt + p_feat + p_count

# Fraction of nuclei that each candidate cutoff would retain
qc_cells %>%
  group_by(sample) %>%
  summarise(`<1%` = mean(percent.mt < 1), `<2%` = mean(percent.mt < 2),
            `<5%` = mean(percent.mt < 5), .groups = "drop") %>%
  mutate(across(-sample, ~ round(100 * .x, 1)))
```

```{r}
mt_cutoff <- 3   # <- set from the distribution above

for (sample in names(seurat_list)) {
  seurat_list[[sample]] <- subset(seurat_list[[sample]],
                                  subset = percent.mt < mt_cutoff)
}
data.frame(sample = names(seurat_list),
           cells  = vapply(seurat_list, ncol, numeric(1)), row.names = NULL)
```

Human graft versus mouse host. The species call uses the majority of `GRCh38-` over
`mm10-` counts with a 0.2-0.8 ambiguous band, the same rule as `analysis.Rmd`.

Section 3 showed Cell Ranger placing the human cell partitions in the tens against
thousands of mouse partitions, so the graft is a small minority of the nuclei in
these libraries. That is a real constraint on what can be said about the human cells,
and the per-library counts below are the number any downstream human-specific claim
rests on. Note also that with roughly a quarter of counts being ambient, the
`mixed` band here is as much a contamination readout as a genuine doublet signal.

```{r}
for (sample in names(seurat_list)) {
  obj <- seurat_list[[sample]]
  DefaultAssay(obj) <- "RNA"
  cm <- GetAssayData(obj, assay = "RNA", layer = "counts")

  human_umis <- Matrix::colSums(cm[grep("^GRCh38-", rownames(cm)), , drop = FALSE])
  mouse_umis <- Matrix::colSums(cm[grep("^mm10-",   rownames(cm)), , drop = FALSE])
  ratio      <- human_umis / (human_umis + mouse_umis)

  obj$human_frac <- ratio
  obj$species <- ifelse(human_umis > mouse_umis, "human", "mouse")
  obj$species[ratio > 0.2 & ratio < 0.8] <- "mixed"

  seurat_list[[sample]] <- obj
}

# table() requires all its arguments to be the same length, so the sample name is
# added as a column afterwards rather than passed in as a one-element factor.
species_tab <- dplyr::bind_rows(lapply(names(seurat_list), function(s) {
  tb <- table(as.character(seurat_list[[s]]@meta.data$species))
  data.frame(sample  = s,
             species = names(tb),
             Freq    = as.integer(tb),
             row.names = NULL,
             stringsAsFactors = FALSE)
})) %>%
  pivot_wider(names_from = species, values_from = Freq, values_fill = 0)
species_tab
```

```{r, fig.height=5, fig.width=12}
human_frac_df <- do.call(rbind, lapply(names(seurat_list), function(s) {
  data.frame(sample = s, human_frac = seurat_list[[s]]$human_frac,
             region = seurat_list[[s]]$region)
}))

ggplot(human_frac_df, aes(x = human_frac)) +
  geom_histogram(bins = 60, fill = "grey40") +
  geom_vline(xintercept = c(0.2, 0.8), linetype = "dashed", colour = "firebrick") +
  scale_y_log10() +
  facet_wrap(~ region) +
  labs(x = "Human fraction of UMI per nucleus", y = "Nuclei (log scale)",
       subtitle = "Dashed: bounds of the 'mixed' band") +
  theme_classic(base_size = 12)
```

##7. Doublet removal

Hashing gives an **empirical** doublet rate for each library, which is better than
both alternatives: the 10x lookup table is a generic function of recovered cells,
and the barnyard estimate is unusable here for the reasons given in section 3.

The arithmetic. With four tags at observed singlet proportions `p`, a doublet is
only visible to `HTODemux` if its two nuclei carried *different* tags, which happens
with probability `1 - sum(p^2)`. So if `f_obs` is the observed hashtag-doublet
fraction, the total doublet rate is `d = f_obs / (1 - sum(p^2))`. Section 5 already
removed the inter-region doublets, so what remains in the object is the same-tag
fraction, `d * sum(p^2)`, spread over the `1 - f_obs` of cells that were kept:

```
rate_remaining = d * sum(p^2) / (1 - f_obs)
```

That is what `nExp` is set from. `pK` is still estimated per library with
`paramSweep`/`find.pK`, and the homotypic proportion is still applied **downward**
(`nExp * (1 - homotypic)`), both following `analysis_CD45.Rmd`. Because the
remaining doublets are by construction within-region, they are disproportionately
homotypic, which is precisely the population `modelHomotypic` is meant to discount.

`doublet_rate_10x()` is carried over unchanged and printed next to the hashtag
estimate as a comparison, not used.

```{r}
# Expected multiplet rate as a function of recovered cells, interpolated from
# the 10x Chromium 3' table (~0.8% per 1,000 cells, near-linear in this range).
doublet_rate_10x <- function(n_cells) {
  ref_cells <- c(  500,  1000,  2000,  3000,  4000,  5000,
                  6000,  7000,  8000,  9000, 10000)
  ref_rate  <- c(0.004, 0.008, 0.016, 0.023, 0.031, 0.039,
                 0.046, 0.054, 0.061, 0.069, 0.076)
  as.numeric(approx(ref_cells, ref_rate, xout = n_cells, rule = 2)$y)
}

hto_remaining_rate <- function(r) {
  d <- r$f_obs / (1 - r$sum_p2)          # total doublet rate in the library
  min(max(d * r$sum_p2 / (1 - r$f_obs), 0), 0.25)
}

rate_compare <- do.call(rbind, lapply(names(seurat_list), function(s) {
  r <- hto_rates[[s]]
  data.frame(sample = s,
             cells          = ncol(seurat_list[[s]]),
             hto_total_rate = round(r$f_obs / (1 - r$sum_p2), 4),
             rate_remaining = round(hto_remaining_rate(r), 4),
             rate_10x       = round(doublet_rate_10x(ncol(seurat_list[[s]])), 4))
}))
rate_compare
```

```{r}
for (sample in names(seurat_list)) {
  message("Doublet detection: ", sample)
  obj <- seurat_list[[sample]]

  # Light preprocessing for doublet detection (per sample, SCT)
  DefaultAssay(obj) <- "RNA"
  obj <- SCTransform(obj, assay = "RNA", verbose = FALSE)
  obj <- RunPCA(obj, verbose = FALSE)
  obj <- FindNeighbors(obj, dims = 1:10, verbose = FALSE)
  obj <- FindClusters(obj, resolution = 0.1, verbose = FALSE)
  obj <- RunUMAP(obj, dims = 1:10, verbose = FALSE)

  umap_before <- DimPlot(obj, reduction = "umap", group.by = "seurat_clusters") +
    ggtitle(paste(sample, "before doublet removal"))

  # ---- pK estimation (per sample) ----
  sweep_res   <- paramSweep(obj, PCs = 1:10, sct = TRUE)
  sweep_stats <- summarizeSweep(sweep_res, GT = FALSE)
  bcmvn       <- find.pK(sweep_stats)
  pK <- as.numeric(as.character(bcmvn$pK[which.max(bcmvn$BCmetric)]))
  if (is.na(pK) || length(pK) == 0) pK <- 0.09  # fall back to the fixed value
  message("  pK = ", pK)
  pN <- 0.25

  # ---- expected number of doublets, from the hashtags ----
  rate <- hto_remaining_rate(hto_rates[[sample]])

  # modelHomotypic needs a clustering fine enough to separate cell types; the
  # res = 0.1 used for the QC UMAP is too coarse and inflates the homotypic
  # estimate. A finer clustering is computed here for this purpose only.
  obj <- FindClusters(obj, resolution = 0.8,
                      cluster.name = "clusters_for_homotypic", verbose = FALSE)
  homotypic <- modelHomotypic(obj$clusters_for_homotypic)

  nExp_poi <- round(rate * ncol(obj))
  nExp_adj <- round(nExp_poi * (1 - homotypic))
  message("  cells = ", ncol(obj),
          ", hashtag rate = ", signif(rate, 3),
          " (10x table would give ", signif(doublet_rate_10x(ncol(obj)), 3), ")",
          ", homotypic = ", signif(homotypic, 3),
          ", nExp = ", nExp_poi, " -> ", nExp_adj)

  if (nExp_adj < 1) {
    message("  nExp_adj < 1, no DoubletFinder doublets to call")
    obj$doublet_class <- "Singlet"
    obj$doublet_pANN  <- NA_real_
  } else {
    obj <- doubletFinder(obj, PCs = 1:10, pN = pN, pK = pK,
                         nExp = nExp_poi, reuse.pANN = NULL, sct = TRUE)

    df_col <- grep("^pANN_", colnames(obj@meta.data), value = TRUE)[1]
    if (is.na(df_col)) stop("DoubletFinder pANN metadata not found")

    obj <- doubletFinder(obj, PCs = 1:10, pN = pN, pK = pK,
                         nExp = nExp_adj, reuse.pANN = df_col, sct = TRUE)

    # The adjusted classification is the one keyed to nExp_adj; select it by name
    # rather than by position so the choice does not depend on column ordering.
    df_class_col <- grep(paste0("^DF.classifications_.*_", nExp_adj, "$"),
                         colnames(obj@meta.data), value = TRUE)
    if (length(df_class_col) == 0) {
      df_class_col <- tail(grep("^DF.classifications_",
                                colnames(obj@meta.data), value = TRUE), 1)
    }
    if (length(df_class_col) == 0) stop("DoubletFinder classification not found")
    message("  using ", df_class_col)

    obj$doublet_class <- obj@meta.data[[df_class_col]]
    obj$doublet_pANN  <- obj@meta.data[[df_col]]
  }

  # Diagnostic: where do the called doublets sit? A cluster that is almost
  # entirely doublet-called is a real doublet cluster; doublets scattered
  # evenly across all clusters suggest the rate is set too high.
  print(DimPlot(obj, reduction = "umap", group.by = "doublet_class") +
          ggtitle(paste(sample, "doublet calls")))
  print(table(cluster = obj$seurat_clusters, call = obj$doublet_class))
  print(VlnPlot(obj, features = c("nFeature_RNA", "nCount_RNA"),
                group.by = "doublet_class", pt.size = 0) +
          patchwork::plot_annotation(title = paste(sample, "complexity by call")))

  sel_singlets <- obj$doublet_class == "Singlet"
  if (sum(sel_singlets, na.rm = TRUE) == 0) {
    stop("No cells annotated as Singlet in ", sample)
  }
  obj <- subset(obj, cells = colnames(obj)[which(sel_singlets)])

  # Recompute after doublet removal
  DefaultAssay(obj) <- "RNA"
  obj[["SCT"]] <- NULL
  obj <- SCTransform(obj, assay = "RNA", verbose = FALSE)
  obj <- RunPCA(obj, verbose = FALSE)
  obj <- FindNeighbors(obj, dims = 1:10, verbose = FALSE)
  obj <- FindClusters(obj, resolution = 0.1, verbose = FALSE)
  obj <- RunUMAP(obj, dims = 1:10, verbose = FALSE)

  umap_after <- DimPlot(obj, reduction = "umap", group.by = "seurat_clusters") +
    ggtitle(paste(sample, "after doublet removal"))

  print(umap_before + umap_after)

  seurat_list[[sample]] <- obj
}
```

Record how many nuclei each library lost, so an outlying sample is visible before it
influences the merged object. The `rep1`/`rep2` pairs should track each other - they
are the same suspension.

```{r, fig.height=6, fig.width=12}
doublet_summary <- do.call(rbind, lapply(names(seurat_list), function(s) {
  data.frame(sample = s, cells_retained = ncol(seurat_list[[s]]))
}))
doublet_summary <- left_join(doublet_summary, sample_sheet, by = "sample")
doublet_summary

ggplot(doublet_summary, aes(x = animal, y = cells_retained, fill = rep)) +
  geom_col(position = "dodge") +
  coord_flip() +
  facet_wrap(~ condition, scales = "free_y", ncol = 1) +
  labs(x = NULL, y = "Nuclei retained after doublet removal", fill = NULL) +
  theme_classic(base_size = 12)
```

##8. Merge and cluster: the global object

```{r}
humanized_mouse_hashing <- Reduce(function(x, y) merge(x, y), seurat_list)
rm(seurat_list); gc()

# Remove per-sample doublet-detection metadata, which is meaningless once the
# libraries are merged (the columns are keyed to each sample's own nExp), along
# with the per-library HTO classifications superseded by `region`.
md <- humanized_mouse_hashing@meta.data
drop_cols <- grep(paste0("^(pANN_|DF.classifications_|nCount_SCT|nFeature_SCT|",
                         "SCT_snn_res|seurat_clusters|clusters_for_homotypic|",
                         "HTO_|hash.ID|MULTI_)"),
                  colnames(md), value = TRUE)
for (cc in drop_cols) humanized_mouse_hashing@meta.data[[cc]] <- NULL

humanized_mouse_hashing <- subset(humanized_mouse_hashing,
                                  subset = percent.mt < mt_cutoff)
humanized_mouse_hashing

table(humanized_mouse_hashing$region, humanized_mouse_hashing$condition)
```

Normalisation follows the other vignettes: a log-normalised RNA branch (kept for
marker plots and for the pseudobulk step, which reads raw counts) and an SCT branch
used for dimensionality reduction and clustering.

```{r}
options(future.globals.maxSize = 100e+09)

DefaultAssay(humanized_mouse_hashing) <- "RNA"

humanized_mouse_hashing[["SCT"]] <- NULL

# Log-normalized analysis on RNA
humanized_mouse_hashing <- NormalizeData(humanized_mouse_hashing, assay = "RNA")
humanized_mouse_hashing <- FindVariableFeatures(humanized_mouse_hashing, assay = "RNA")
humanized_mouse_hashing <- ScaleData(humanized_mouse_hashing, assay = "RNA")
humanized_mouse_hashing <- RunPCA(humanized_mouse_hashing, assay = "RNA",
                                  reduction.name = "pca_lognorm")

# SCT analysis on SCT assay
humanized_mouse_hashing <- SCTransform(humanized_mouse_hashing, assay = "RNA",
                                       verbose = TRUE)
humanized_mouse_hashing <- FindVariableFeatures(humanized_mouse_hashing, assay = "SCT")
humanized_mouse_hashing <- RunPCA(humanized_mouse_hashing, assay = "SCT",
                                  reduction.name = "pca_SCT")
```

Read the elbow before fixing `n_dims`. These are whole-brain nuclei spanning four
regions, so the closest precedent is the unhashed whole-hemisphere analysis in
`analysis.Rmd` rather than the sorted CD45+ populations; expect a similar elbow to
the former.

```{r}
ElbowPlot(humanized_mouse_hashing, reduction = "pca_SCT", ndims = 50)
```

```{r}
n_dims <- 40  # <- set from the elbow plot above

humanized_mouse_hashing <- FindNeighbors(humanized_mouse_hashing, dims = 1:n_dims,
                                         reduction = "pca_SCT")
humanized_mouse_hashing <- FindClusters(humanized_mouse_hashing, resolution = 0.1,
                                        cluster.name = "unintegrated_clusters")
humanized_mouse_hashing <- RunUMAP(humanized_mouse_hashing, dims = 1:n_dims,
                                   reduction = "pca_SCT",
                                   reduction.name = "umap.unintegrated")
```

```{r, fig.height=12, fig.width=18}
my_colors <- as.vector(pals::polychrome(36))

p1 <- DimPlot(humanized_mouse_hashing, reduction = "umap.unintegrated",
              group.by = "unintegrated_clusters", cols = my_colors,
              label = TRUE) + labs(x = "UMAP 1", y = "UMAP 2")
p2 <- DimPlot(humanized_mouse_hashing, reduction = "umap.unintegrated",
              group.by = "region") + labs(x = "UMAP 1", y = "UMAP 2")
p3 <- DimPlot(humanized_mouse_hashing, reduction = "umap.unintegrated",
              group.by = "condition") + labs(x = "UMAP 1", y = "UMAP 2")
p4 <- DimPlot(humanized_mouse_hashing, reduction = "umap.unintegrated",
              group.by = "species") + labs(x = "UMAP 1", y = "UMAP 2")
(p1 + p2) / (p3 + p4)
```

Cluster composition. Four panels, each answering a different question:

* **by region** - does the demultiplexing produce the regional structure it should?
  Cerebellar granule cells confined to Cerebellum is the positive control.
* **by animal** - a cluster carried by a single animal is a batch effect candidate,
  not a biological population.
* **by rep** - `rep1` and `rep2` are the same suspension, so any cluster split
  between them is a channel artefact.
* **by library, restricted to the flagged samples from section 3** - a cluster
  carried by the libraries whose ambient profile is least constrained is an ambient
  artefact candidate.

```{r, fig.height=14, fig.width=16}
comp <- function(var, lab) {
  df <- as.data.frame(table(cluster = humanized_mouse_hashing$unintegrated_clusters,
                            grp = humanized_mouse_hashing@meta.data[[var]]))
  ggplot(df, aes(x = cluster, y = Freq, fill = grp)) +
    geom_col(position = "fill") + scale_y_continuous(labels = percent) +
    labs(x = "Cluster", y = "Fraction of nuclei", fill = lab) +
    theme_classic(base_size = 13)
}

humanized_mouse_hashing$is_flagged <- ifelse(
  humanized_mouse_hashing$sample %in% flagged_samples,
  "weak ambient fit", "other")

comp("region", "Region") / comp("animal", "Animal") /
  comp("rep", "Replicate") / comp("is_flagged", "Section 3 flag")
```

Cluster markers.

```{r}
DefaultAssay(humanized_mouse_hashing) <- "SCT"
humanized_mouse_hashing <- PrepSCTFindMarkers(humanized_mouse_hashing)
hashing_markers <- FindAllMarkers(humanized_mouse_hashing, only.pos = TRUE)

write.xlsx(hashing_markers,
           file.path(markers_dir, "hashing_global_markers.xlsx"),
           overwrite = TRUE)
hashing_markers %>% group_by(cluster) %>% slice_max(avg_log2FC, n = 10)
```

Because these are whole-brain nuclei from the same model as the already-annotated
object built in `analysis.Rmd`, the labels there are a legitimate starting point
rather than a guess. Transfer them and read the result as a confusion matrix against
the clustering above. The gene-name normalisation in section 4 was kept
byte-identical to that vignette precisely so this join works.

This is a **starting point, not the annotation**: a transferred label inherits the
reference's mistakes and has nothing to say about a population absent from it. The
markers written above are the evidence; the named vector below is where the call
gets made.

```{r}
ref_candidates <- file.path(obj_dir, c("humanized_mouse_brain_whole_doublets.rds",
                                       "humanized_mouse_brain.rds"))
ref_path <- ref_candidates[file.exists(ref_candidates)][1]

if (!is.na(ref_path)) {
  message("Transferring labels from ", ref_path)
  ref <- readRDS(ref_path)

  label_col <- intersect(c("broad_clustering", "broad_clusters", "fine_clusters"),
                         colnames(ref@meta.data))[1]
  if (is.na(label_col)) {
    warning("No known label column in the reference; skipping transfer")
  } else {
    DefaultAssay(ref) <- "SCT"
    anchors <- FindTransferAnchors(
      reference = ref, query = humanized_mouse_hashing,
      normalization.method = "SCT", reference.reduction = "pca",
      dims = 1:n_dims
    )
    pred <- TransferData(anchorset = anchors,
                         refdata = as.character(ref@meta.data[[label_col]]),
                         dims = 1:n_dims)
    humanized_mouse_hashing$transferred_label <- pred$predicted.id
    humanized_mouse_hashing$transfer_score    <- pred$prediction.score.max

    print(table(cluster = humanized_mouse_hashing$unintegrated_clusters,
                transferred = humanized_mouse_hashing$transferred_label))
  }
  rm(ref); gc()
} else {
  warning("No reference object found under ", obj_dir, "; skipping label transfer")
}
```

```{r, fig.height=6, fig.width=16}
if ("transferred_label" %in% colnames(humanized_mouse_hashing@meta.data)) {
  t1 <- DimPlot(humanized_mouse_hashing, reduction = "umap.unintegrated",
                group.by = "transferred_label", cols = my_colors, label = TRUE,
                repel = TRUE) + labs(x = "UMAP 1", y = "UMAP 2")
  t2 <- FeaturePlot(humanized_mouse_hashing, reduction = "umap.unintegrated",
                    features = "transfer_score") +
    labs(x = "UMAP 1", y = "UMAP 2")
  t1 + t2
}
```

Annotation. Fill the vectors below from the marker table and the transferred labels,
then drop any cluster that turns out to be ambient contamination rather than a
population - the equivalent of the cluster dropped in `analysis_CD45.Rmd`. With
about a quarter of counts ambient in these libraries, expect at least one such
cluster: it will have low `nFeature_RNA`, no specific positive markers, a flat
profile resembling the library average, and spread evenly across all four regions.

```{r, fig.height=6, fig.width=14}
# Evidence for the ambient-contamination call, read alongside the marker table
VlnPlot(humanized_mouse_hashing, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"),
        group.by = "unintegrated_clusters", pt.size = 0, ncol = 1)
```

```{r}
# ---- Fill these in before running the rest of the vignette ----------------
# Clusters judged to be ambient contamination / low-quality, to be removed.
ambient_clusters <- c()   # e.g. c("13")

if (length(ambient_clusters) > 0) {
  Idents(humanized_mouse_hashing) <- "unintegrated_clusters"
  keep_ids <- setdiff(levels(humanized_mouse_hashing$unintegrated_clusters),
                      ambient_clusters)
  humanized_mouse_hashing <- subset(humanized_mouse_hashing, idents = keep_ids)
  humanized_mouse_hashing$unintegrated_clusters <-
    droplevels(humanized_mouse_hashing$unintegrated_clusters)
}

Idents(humanized_mouse_hashing) <- "unintegrated_clusters"

# Fine annotation: one entry per remaining cluster id.
fine_global <- c(
  # "0" = "Cerebellar granule cells",
  # "1" = "Excitatory cortical neurons",
  # ...
)

# Broad / lineage annotation: one entry per fine label.
lineage_global <- c(
  # "Cerebellar granule cells"    = "Neuron",
  # "Excitatory cortical neurons" = "Neuron",
  # ...
)

if (length(fine_global) > 0) {
  stopifnot(setequal(names(fine_global),
                     levels(humanized_mouse_hashing$unintegrated_clusters)))
  humanized_mouse_hashing <- RenameIdents(humanized_mouse_hashing, fine_global)
  humanized_mouse_hashing$fine_clusters <- Idents(humanized_mouse_hashing)

  if (length(lineage_global) > 0) {
    stopifnot(setequal(names(lineage_global), unname(fine_global)))
    humanized_mouse_hashing <- RenameIdents(humanized_mouse_hashing, lineage_global)
    humanized_mouse_hashing$broad_clusters <- Idents(humanized_mouse_hashing)
  }
  table(humanized_mouse_hashing$fine_clusters)
} else {
  message("fine_global is empty - the vignette will use unintegrated_clusters ",
          "as the cluster label throughout.")
}

# The cluster column every downstream section keys on.
cluster_col_global <- if ("fine_clusters" %in% colnames(humanized_mouse_hashing@meta.data)) {
  "fine_clusters"
} else {
  "unintegrated_clusters"
}
cluster_col_global

saveRDS(humanized_mouse_hashing,
        file.path(obj_dir, "humanized_mouse_hashing_SCT.rds"))
```

Export UMAP coordinates and metadata.

```{r}
wb <- createWorkbook()
md <- humanized_mouse_hashing@meta.data
md$barcode <- colnames(humanized_mouse_hashing)
df <- data.frame(cell = rownames(md), md, row.names = NULL,
                 check.names = FALSE, stringsAsFactors = FALSE)
addWorksheet(wb, "hashing")
writeData(wb, "hashing", df)
saveWorkbook(wb, file.path(res_dir, "hashing_metadata.xlsx"), overwrite = TRUE)
cat(sprintf("  hashing : %d nuclei x %d metadata columns\n", nrow(df), ncol(df) - 1))

wb <- createWorkbook()
emb <- humanized_mouse_hashing@reductions$umap.unintegrated@cell.embeddings
df  <- data.frame(cell = rownames(emb), emb, row.names = NULL,
                  check.names = FALSE, stringsAsFactors = FALSE)
addWorksheet(wb, "hashing")
writeData(wb, "hashing", df)
saveWorkbook(wb, file.path(res_dir, "hashing_UMAP_coordinates.xlsx"),
             overwrite = TRUE)
```

Save the full count matrix for future use.

```{r}
library("zellkonverter")
library(SingleCellExperiment)
sce <- as.SingleCellExperiment(humanized_mouse_hashing, assay = "RNA")
writeH5AD(sce, file.path(res_dir, "humanized_mouse_hashing.h5ad"))

library(Matrix); library(data.table)

obj <- humanized_mouse_hashing
DefaultAssay(obj) <- "RNA"
obj <- JoinLayers(obj)
mat <- GetAssayData(obj, assay = "RNA", layer = "data")   # or layer = "counts" for raw

outfile <- file.path(res_dir, "hashing_Count_Matrix.csv.gz")
dir.create(dirname(outfile), recursive = TRUE, showWarnings = FALSE)

chunk <- 2000
idx   <- split(seq_len(ncol(mat)), ceiling(seq_len(ncol(mat)) / chunk))

for (i in seq_along(idx)) {
  blk <- as.matrix(t(mat[, idx[[i]], drop = FALSE]))    # cells x genes
  dt  <- data.table(cell = rownames(blk), blk)
  fwrite(dt, outfile, append = i > 1, col.names = i == 1, compress = "gzip")
  message("chunk ", i, "/", length(idx))
}
rm(obj, mat); gc()
```

##9. Reclustering within each brain region

Same loop as the per-tissue reclustering in `analysis_CD45.Rmd`, over the four
hashtag-defined regions. Each region gets its own normalisation, embedding and
clustering, so a population that is a minor subset globally can resolve into
substructure where it actually lives. The global object from section 8 is kept
intact alongside these.

```{r}
regions <- levels(droplevels(factor(humanized_mouse_hashing$region)))

region_objects <- list()

for (bc in regions) {
  obj_name <- make.names(bc)
  message("Processing region: ", bc, " --> ", obj_name)

  RS <- subset(humanized_mouse_hashing, subset = region == bc)

  DefaultAssay(RS) <- "RNA"
  RS[["SCT"]] <- NULL
  RS <- SCTransform(RS, assay = "RNA", vst.flavor = "v2", verbose = FALSE)
  RS <- RunPCA(RS, verbose = FALSE,
               reduction.name = paste0("pca", obj_name, "_SCT"))
  RS <- FindNeighbors(RS, reduction = paste0("pca", obj_name, "_SCT"),
                      dims = 1:n_dims,
                      graph.name = c(paste0(obj_name, "_nn"),
                                     paste0(obj_name, "_snn")))
  RS <- FindClusters(RS, resolution = 0.5,
                     graph.name = paste0(obj_name, "_snn"),
                     cluster.name = paste0(obj_name, "_clusters"))
  RS <- RunUMAP(RS, reduction = paste0("pca", obj_name, "_SCT"), dims = 1:n_dims,
                reduction.name = paste0("umap_", obj_name))

  region_objects[[obj_name]] <- RS
  saveRDS(RS, file = file.path(obj_dir, paste0(obj_name, "_hashing_SCT.rds")))

  # markers
  DefaultAssay(RS) <- "SCT"
  RS <- PrepSCTFindMarkers(RS, assay = "SCT", verbose = FALSE)
  RS_markers <- FindAllMarkers(RS, only.pos = TRUE)

  wb <- createWorkbook()
  for (cluster_value in unique(RS_markers$cluster)) {
    cluster_data <- subset(RS_markers, cluster == cluster_value)
    sheet_name <- paste0("Cluster_", cluster_value)
    addWorksheet(wb, sheet_name)
    writeData(wb, sheet_name, cluster_data)
  }
  saveWorkbook(wb, file.path(markers_dir, paste0(obj_name, "_hashing_markers.xlsx")),
               overwrite = TRUE)
}
```

```{r, fig.height=12, fig.width=16}
plots <- lapply(names(region_objects), function(mk) {
  DimPlot(region_objects[[mk]], reduction = paste0("umap_", mk),
          group.by = paste0(mk, "_clusters"), cols = my_colors, label = TRUE) +
    ggtitle(mk) + labs(x = "UMAP 1", y = "UMAP 2")
})
patchwork::wrap_plots(plots, ncol = 2)
```

```{r, fig.height=12, fig.width=16}
plots <- lapply(names(region_objects), function(mk) {
  DimPlot(region_objects[[mk]], reduction = paste0("umap_", mk),
          group.by = "condition") +
    ggtitle(mk) + labs(x = "UMAP 1", y = "UMAP 2")
})
patchwork::wrap_plots(plots, ncol = 2)
```

Save UMAP coordinates and metadata for the per-region objects.

```{r}
cat("regions:", paste(names(region_objects), collapse = ", "), "\n")

wb <- createWorkbook()
for (mk in names(region_objects)) {
  md <- region_objects[[mk]]@meta.data
  md$barcode <- colnames(region_objects[[mk]])
  df <- data.frame(cell = rownames(md), md, row.names = NULL,
                   check.names = FALSE, stringsAsFactors = FALSE)
  addWorksheet(wb, mk)
  writeData(wb, mk, df)
  cat(sprintf("  %-16s : %d nuclei x %d metadata columns\n", mk, nrow(df), ncol(df) - 1))
}
saveWorkbook(wb, file.path(res_dir, "Reclustering_hashing_metadata.xlsx"),
             overwrite = TRUE)

wb <- createWorkbook()
for (mk in names(region_objects)) {
  emb <- region_objects[[mk]]@reductions[[paste0("umap_", mk)]]@cell.embeddings
  df  <- data.frame(cell = rownames(emb), emb, row.names = NULL,
                    check.names = FALSE, stringsAsFactors = FALSE)
  addWorksheet(wb, mk)
  writeData(wb, mk, df)
}
saveWorkbook(wb, file.path(res_dir, "Reclustering_hashing_UMAP.xlsx"),
             overwrite = TRUE)
```

##10. Differential gene expression: EAE-NonSymp vs CFA

Pseudobulk DESeq2 within each cluster, with the **animal** as the unit of
replication. Three features of this design need handling:

* **`rep1` and `rep2` are the same animal.** The pseudobulk key is
  `animal x region`, not `sample x region`, so the two channels are summed into one
  profile. Treating them as separate columns would double the apparent sample size
  while adding no biological information, and would make every p-value
  anticonservative.
* **Each animal contributed four regions.** Aggregating per animal x region and
  fitting `~ region + condition` blocks the region effect while keeping the
  condition term estimated across animals. Region is dropped automatically when a
  cluster is confined to one region, which would otherwise leave the model
  rank-deficient.
* **`EAE-Symp` has n = 1** and is excluded from testing. It is summarised separately
  in section 11 without p-values.

CFA is the reference level, so a positive log2FC means higher in EAE.

```{r}
run_pseudobulk_dge_hashing <- function(seurat_obj,
                                       cluster_col,
                                       cluster_id,
                                       test_conditions  = c("CFA", "EAE-NonSymp"),
                                       animal_col       = "animal",
                                       region_col       = "region",
                                       condition_col    = "condition",
                                       use_region       = TRUE,
                                       min_cells        = 10,
                                       min_cells_per_pb = 10,
                                       min_counts       = 10) {

  idx <- seurat_obj@meta.data[[cluster_col]] == cluster_id &
    as.character(seurat_obj@meta.data[[condition_col]]) %in% test_conditions
  if (sum(idx, na.rm = TRUE) < min_cells) {
    message("Skipping cluster ", cluster_id, ": only ", sum(idx, na.rm = TRUE),
            " testable nuclei")
    return(NULL)
  }

  sub_obj <- seurat_obj[, which(idx)]
  DefaultAssay(sub_obj) <- "RNA"
  sub_obj <- JoinLayers(sub_obj)
  counts_mat <- GetAssayData(sub_obj, assay = "RNA", layer = "counts")
  meta <- sub_obj@meta.data

  # One pseudobulk profile per animal x region. rep1 and rep2 are the same
  # animal, so they are summed here rather than kept apart.
  pb_group <- if (use_region) {
    paste(meta[[animal_col]], meta[[region_col]], sep = "__")
  } else {
    as.character(meta[[animal_col]])
  }

  keep_groups <- names(which(table(pb_group) >= min_cells_per_pb))
  if (length(keep_groups) < 4) {
    message("Skipping cluster ", cluster_id,
            ": fewer than 4 usable pseudobulk groups")
    return(NULL)
  }

  pb_mat <- vapply(keep_groups, function(g) {
    as.numeric(Matrix::rowSums(counts_mat[, pb_group == g, drop = FALSE]))
  }, numeric(nrow(counts_mat)))
  rownames(pb_mat) <- rownames(counts_mat)

  col_meta <- data.frame(
    group     = keep_groups,
    animal    = if (use_region) sub("__.*$", "", keep_groups) else keep_groups,
    row.names = keep_groups,
    stringsAsFactors = FALSE
  )
  if (use_region) {
    col_meta$region <- factor(sub("^.*__", "", keep_groups))
  }
  col_meta$condition <- factor(
    vapply(col_meta$animal, function(a)
      unique(as.character(meta[[condition_col]][meta[[animal_col]] == a]))[1],
      character(1)),
    levels = test_conditions
  )

  # Require at least two animals per condition, or the dispersion estimate is
  # not identifiable and the test is not interpretable.
  n_animals <- tapply(col_meta$animal, col_meta$condition,
                      function(x) length(unique(x)))
  if (any(is.na(n_animals)) || any(n_animals < 2)) {
    message("Skipping cluster ", cluster_id,
            ": fewer than 2 animals in one condition")
    return(NULL)
  }

  if (use_region) {
    col_meta$region <- droplevels(col_meta$region)
    ok_region <- nlevels(col_meta$region) > 1 &&
      !any(rowSums(table(col_meta$region, col_meta$condition) > 0) < 2)
    design <- if (ok_region) ~ region + condition else ~ condition
  } else {
    design <- ~ condition
  }
  message("Cluster ", cluster_id, ": ", nrow(col_meta),
          " pseudobulk groups, design ", deparse(design))

  pb_mat <- pb_mat[rowSums(pb_mat) >= min_counts, , drop = FALSE]
  if (nrow(pb_mat) < 10) {
    message("Skipping cluster ", cluster_id, ": too few expressed genes")
    return(NULL)
  }

  tryCatch({
    dds <- DESeqDataSetFromMatrix(countData = round(pb_mat),
                                  colData   = col_meta,
                                  design    = design)
    dds <- DESeq(dds, quiet = TRUE)
    res <- results(dds, contrast = c("condition", test_conditions[2],
                                     test_conditions[1]))
    res_df <- as.data.frame(res)
    res_df$gene    <- rownames(res_df)
    res_df$cluster <- as.character(cluster_id)
    res_df$design  <- deparse(design)
    res_df$n_pb    <- nrow(col_meta)
    res_df[order(res_df$padj, na.last = TRUE), ]
  }, error = function(e) {
    message("DESeq2 error - cluster ", cluster_id, ": ", conditionMessage(e))
    NULL
  })
}

# Volcano plot helper, as in analysis.Rmd
plot_volcano <- function(res_df, cluster_label,
                         padj_thresh = 0.05, lfc_thresh = 0.5, top_n = 15) {
  df <- res_df %>%
    filter(!is.na(padj), !is.na(log2FoldChange)) %>%
    mutate(
      sig = case_when(
        padj < padj_thresh & log2FoldChange >  lfc_thresh ~ "Up in EAE",
        padj < padj_thresh & log2FoldChange < -lfc_thresh ~ "Up in CFA",
        TRUE ~ "NS"
      )
    ) %>%
    mutate(label = ifelse(sig != "NS" & rank(padj, ties.method = "first") <= top_n,
                          gene, NA_character_))

  n_up <- sum(df$sig == "Up in EAE")
  n_dn <- sum(df$sig == "Up in CFA")

  ggplot(df, aes(x = log2FoldChange, y = -log10(padj), color = sig)) +
    geom_point(alpha = 0.6, size = 1.5) +
    ggrepel::geom_text_repel(aes(label = label), size = 3, max.overlaps = 20,
                             na.rm = TRUE) +
    scale_color_manual(
      values = c("Up in EAE" = "firebrick", "Up in CFA" = "steelblue", "NS" = "grey70")
    ) +
    geom_vline(xintercept = c(-lfc_thresh, lfc_thresh), linetype = "dashed",
               colour = "grey40") +
    geom_hline(yintercept = -log10(padj_thresh), linetype = "dashed",
               colour = "grey40") +
    annotate("text", x = Inf,  y = Inf, hjust = 1.1, vjust = 1.0,
             label = paste0("Up in EAE: ", n_up), color = "firebrick",
             size = 4.5, fontface = "bold") +
    annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 1.0,
             label = paste0("Up in CFA: ", n_dn), color = "steelblue",
             size = 4.5, fontface = "bold") +
    labs(
      title = paste("EAE-NonSymp vs CFA -", cluster_label),
      x     = "log2 Fold Change (EAE / CFA)",
      y     = "-log10(adjusted p-value)",
      color = NULL
    ) +
    theme_classic() +
    theme(
      plot.title  = element_text(hjust = 0.5, size = 16, face = "bold"),
      axis.text   = element_text(size = 12),
      axis.title  = element_text(size = 14),
      legend.text = element_text(size = 12)
    )
}
```

###10.1 Pooled across regions, blocking on region

Maximum power per cluster: all four regions contribute, and the region term absorbs
the regional baseline differences. A region-restricted effect will be diluted here,
which is what section 10.2 is for.

```{r}
cluster_ids_global <- levels(droplevels(factor(
  humanized_mouse_hashing@meta.data[[cluster_col_global]])))

dge_pooled <- lapply(cluster_ids_global, function(cl) {
  message("[pooled] Running DGE for cluster: ", cl)
  tryCatch(
    run_pseudobulk_dge_hashing(
      seurat_obj  = humanized_mouse_hashing,
      cluster_col = cluster_col_global,
      cluster_id  = cl,
      use_region  = TRUE
    ),
    error = function(e) { message("  skipped: ", conditionMessage(e)); NULL }
  )
})
names(dge_pooled) <- cluster_ids_global

wb_pooled <- createWorkbook()
for (cl in names(dge_pooled)) {
  res <- dge_pooled[[cl]]
  if (is.null(res) || nrow(res) == 0) next
  sheet_name <- substr(gsub("[\\[\\]:*?/\\\\]", "_", paste0("Cluster_", cl)), 1, 31)
  addWorksheet(wb_pooled, sheet_name)
  writeData(wb_pooled, sheet_name, res)
}
if (length(names(wb_pooled)) > 0) {
  saveWorkbook(wb_pooled,
    file.path(res_dir, "DGE_EAE-NonSymp_vs_CFA_hashing_global_clusters.xlsx"),
    overwrite = TRUE)
  cat("Saved pooled DGE results.\n")
} else {
  message("No pooled results - nothing saved.")
}
```

```{r, fig.height=6, fig.width=8}
for (cl in names(dge_pooled)) {
  if (!is.null(dge_pooled[[cl]]) && sum(!is.na(dge_pooled[[cl]]$padj)) > 0) {
    p <- plot_volcano(dge_pooled[[cl]], cl)
    ggplot2::ggsave(
      file.path(plots_dir,
                paste0("volcano_DGE_hashing_pooled_", make.names(cl), ".png")),
      p, width = 8, height = 6, dpi = 300
    )
    print(p)
  }
}
```

###10.2 Stratified per region

Subset to each region, then test within each of that region's own clusters, 3 vs 3
animals. This is the comparison to report when a cluster behaves differently in
Cerebellum, Olfactory Bulb, Striatum and Cortex. `use_region = FALSE` because there
is only one region left inside each subset.

```{r}
dge_by_region <- list()
for (mk in names(region_objects)) {
  obj_reg  <- region_objects[[mk]]
  clus_col <- paste0(mk, "_clusters")
  for (cl in levels(droplevels(factor(obj_reg@meta.data[[clus_col]])))) {
    key <- paste(mk, cl, sep = "_cluster")
    message("Running DGE for ", key)
    dge_by_region[[key]] <- tryCatch(
      run_pseudobulk_dge_hashing(
        seurat_obj  = obj_reg,
        cluster_col = clus_col,
        cluster_id  = cl,
        use_region  = FALSE
      ),
      error = function(e) { message("  skipped: ", conditionMessage(e)); NULL }
    )
  }
}

wb_region <- createWorkbook()
for (key in names(dge_by_region)) {
  if (!is.null(dge_by_region[[key]])) {
    sheet_name <- substr(gsub("[\\[\\]:*?/\\\\]", "_", key), 1, 31)
    addWorksheet(wb_region, sheet_name)
    writeData(wb_region, sheet_name, dge_by_region[[key]])
  }
}
if (length(names(wb_region)) > 0) {
  saveWorkbook(wb_region,
               file.path(res_dir, "DGE_EAE-NonSymp_vs_CFA_hashing_by_region.xlsx"),
               overwrite = TRUE)
  cat("Saved per-region DGE results.\n")
} else {
  message("No per-region results - nothing saved.")
}
```

```{r, fig.height=6, fig.width=8}
for (key in names(dge_by_region)) {
  if (!is.null(dge_by_region[[key]]) && sum(!is.na(dge_by_region[[key]]$padj)) > 0) {
    p <- plot_volcano(dge_by_region[[key]], key)
    ggplot2::ggsave(
      file.path(plots_dir, paste0("volcano_DGE_hashing_", make.names(key), ".png")),
      p, width = 8, height = 6, dpi = 300
    )
    print(p)
  }
}
```

###10.3 Summary and pseudobulk QC

Number of significant genes per cluster, both analyses side by side.

```{r}
count_sig <- function(lst, label) {
  do.call(rbind, lapply(names(lst), function(k) {
    r <- lst[[k]]
    if (is.null(r)) return(data.frame(key = k, analysis = label,
                                      up = NA_integer_, down = NA_integer_,
                                      n_pb = NA_integer_))
    data.frame(key = k, analysis = label,
               up   = sum(r$padj < 0.05 & r$log2FoldChange >  0.5, na.rm = TRUE),
               down = sum(r$padj < 0.05 & r$log2FoldChange < -0.5, na.rm = TRUE),
               n_pb = r$n_pb[1])
  }))
}

dge_summary <- rbind(count_sig(dge_pooled, "pooled"),
                     count_sig(dge_by_region, "per-region"))
dge_summary

wb <- createWorkbook()
addWorksheet(wb, "summary")
writeData(wb, "summary", dge_summary)
saveWorkbook(wb, file.path(res_dir, "DGE_EAE-NonSymp_vs_CFA_hashing_summary.xlsx"),
             overwrite = TRUE)
```

Pseudobulk PCA per region, computed **blind** to the design so it is an honest QC
plot. `animal x region` is the aggregation unit, matching the DGE above, and PC1 is
correlated against sequencing depth: if PC1 is a depth axis rather than a condition
axis, the contrast is being driven by library size and not biology.

```{r, fig.height=6, fig.width=9}
for (mk in names(region_objects)) {
  obj <- region_objects[[mk]]
  DefaultAssay(obj) <- "RNA"
  obj <- JoinLayers(obj)
  cts <- GetAssayData(obj, assay = "RNA", layer = "counts")

  # sparse aggregation: genes x animals (avoids densifying the full matrix)
  g  <- factor(obj$animal)
  mm <- Matrix::sparse.model.matrix(~ 0 + g); colnames(mm) <- levels(g)
  pb <- as.matrix(cts %*% mm)

  cd <- data.frame(
    animal    = colnames(pb),
    condition = factor(sub("_[0-9]+$", "", colnames(pb)),
                       levels = c("CFA", "EAE-NonSymp", "EAE-Symp")),
    row.names = colnames(pb)
  )

  dds <- DESeqDataSetFromMatrix(round(pb), cd, ~ condition)
  dds <- dds[rowSums(counts(dds)) >= 10, ]
  vsd <- vst(dds, blind = TRUE)          # blind = TRUE: ignore the design, honest QC

  pdat <- plotPCA(vsd, intgroup = "condition", returnData = TRUE)
  pv   <- round(100 * attr(pdat, "percentVar"))

  print(
    ggplot(pdat, aes(PC1, PC2, colour = condition, label = name)) +
      geom_point(size = 4) + geom_text_repel(size = 3, show.legend = FALSE) +
      labs(x = paste0("PC1: ", pv[1], "% variance"),
           y = paste0("PC2: ", pv[2], "% variance"), colour = NULL,
           title = paste0(mk, " pseudobulk, ", ncol(pb), " animals")) +
      coord_fixed() + theme_bw()
  )

  cat(mk, "- cor(PC1, depth) =", round(cor(pdat$PC1, colSums(pb)[pdat$name]), 3),
      " cor(PC1, nGene) =", round(cor(pdat$PC1, colSums(pb > 0)[pdat$name]), 3), "\n")
  rm(obj, cts, pb); gc()
}
```

##11. The symptomatic animal, described rather than tested

`EAE-Symp_Brain_1` is a single animal (two technical replicates), so there is no
valid test to run on it and none is run here. It is still informative, and two
things can be reported honestly: how its cluster composition compares to the groups
that do have replication, and how large its expression shift is for genes already
identified in the replicated `EAE-NonSymp` vs `CFA` contrast. No p-values are
produced in this section by design.

```{r, fig.height=9, fig.width=12}
prop_hash <- humanized_mouse_hashing@meta.data %>%
  mutate(cluster = .data[[cluster_col_global]]) %>%
  count(animal, condition, region, cluster, name = "n") %>%
  group_by(animal, region) %>%
  mutate(frac = n / sum(n)) %>%
  ungroup()

ggplot(prop_hash, aes(x = cluster, y = frac, colour = condition)) +
  geom_point(position = position_dodge(width = 0.6), size = 2) +
  facet_wrap(~ region, ncol = 1) +
  scale_y_continuous(labels = percent) +
  labs(x = "Cluster", y = "Fraction of nuclei in animal x region",
       colour = "Condition") +
  theme_classic(base_size = 13) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Region-collapsed version of the same comparison, which is easier to read when a
cluster is present in all four regions:

```{r, fig.height=5, fig.width=11}
prop_hash %>%
  group_by(animal, condition, cluster) %>%
  summarise(n = sum(n), .groups = "drop_last") %>%
  group_by(animal) %>%
  mutate(frac = n / sum(n)) %>%
  ungroup() %>%
  ggplot(aes(x = cluster, y = frac, colour = condition)) +
  geom_point(position = position_dodge(width = 0.6), size = 2.5) +
  scale_y_continuous(labels = percent) +
  labs(x = "Cluster", y = "Fraction of nuclei in animal", colour = "Condition") +
  theme_classic(base_size = 13) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Effect size for the symptomatic animal on the replicated hits: pseudobulk CPM per
`animal x region`, the same aggregation unit the DGE used, so the symptomatic value
can be read directly against the spread of the three CFA and three non-symptomatic
animals.

```{r}
pseudobulk_cpm <- function(seurat_obj, cluster_col, cluster_id,
                           min_cells_per_pb = 10) {
  idx <- seurat_obj@meta.data[[cluster_col]] == cluster_id
  if (sum(idx, na.rm = TRUE) < min_cells_per_pb) return(NULL)
  sub_obj <- seurat_obj[, which(idx)]
  DefaultAssay(sub_obj) <- "RNA"
  sub_obj <- JoinLayers(sub_obj)
  cm   <- GetAssayData(sub_obj, assay = "RNA", layer = "counts")
  meta <- sub_obj@meta.data
  grp  <- paste(meta$animal, meta$region, sep = "__")
  keep <- names(which(table(grp) >= min_cells_per_pb))
  if (length(keep) == 0) return(NULL)
  mat <- vapply(keep, function(g)
    as.numeric(Matrix::rowSums(cm[, grp == g, drop = FALSE])),
    numeric(nrow(cm)))
  rownames(mat) <- rownames(cm)
  cpm <- sweep(mat, 2, colSums(mat), "/") * 1e6
  as.data.frame(cpm) %>%
    mutate(gene = rownames(cpm)) %>%
    pivot_longer(-gene, names_to = "group", values_to = "cpm") %>%
    mutate(animal  = sub("__.*$", "", group),
           region  = sub("^.*__", "", group),
           cluster = as.character(cluster_id))
}

# Example: the top replicated hits of one cluster, with the symptomatic animal
# shown alongside. Change `cl` to inspect other clusters.
cl <- cluster_ids_global[1]
if (!is.null(dge_pooled[[cl]])) {
  top_genes <- head(dge_pooled[[cl]]$gene[which(dge_pooled[[cl]]$padj < 0.05)], 12)

  if (length(top_genes) > 0) {
    cpm_long <- pseudobulk_cpm(humanized_mouse_hashing, cluster_col_global, cl) %>%
      filter(gene %in% top_genes) %>%
      left_join(distinct(sample_sheet, animal, condition), by = "animal")

    print(
      ggplot(cpm_long, aes(x = condition, y = cpm, colour = condition)) +
        geom_point(position = position_jitter(width = 0.15), size = 2) +
        facet_wrap(~ gene, scales = "free_y") +
        labs(x = NULL, y = "Pseudobulk CPM",
             title = paste("Cluster", cl,
                           "- replicated hits, all animals shown")) +
        theme_classic(base_size = 12) +
        theme(axis.text.x = element_text(angle = 45, hjust = 1),
              legend.position = "none")
    )
  }
}
```

The same view split by region, which is where a symptomatic-specific effect would be
expected to be regionally restricted rather than uniform:

```{r, fig.height=7, fig.width=12}
if (exists("cpm_long") && !is.null(cpm_long) && nrow(cpm_long) > 0) {
  print(
    ggplot(cpm_long, aes(x = region, y = cpm, colour = condition)) +
      geom_point(position = position_jitterdodge(jitter.width = 0.15,
                                                 dodge.width = 0.6), size = 2) +
      facet_wrap(~ gene, scales = "free_y") +
      labs(x = NULL, y = "Pseudobulk CPM", colour = "Condition",
           title = paste("Cluster", cl, "- replicated hits by region")) +
      theme_classic(base_size = 12) +
      theme(axis.text.x = element_text(angle = 45, hjust = 1))
  )
}
```

Export the symptomatic animal's pseudobulk CPM alongside the replicated groups for
every cluster, so the descriptive comparison is available outside this document.

```{r}
cpm_all <- do.call(rbind, lapply(cluster_ids_global, function(cl) {
  out <- pseudobulk_cpm(humanized_mouse_hashing, cluster_col_global, cl)
  if (is.null(out)) return(NULL)
  keep_genes <- if (!is.null(dge_pooled[[cl]])) {
    dge_pooled[[cl]]$gene[which(dge_pooled[[cl]]$padj < 0.05)]
  } else character(0)
  if (length(keep_genes) == 0) return(NULL)
  out[out$gene %in% keep_genes, , drop = FALSE]
}))

if (!is.null(cpm_all) && nrow(cpm_all) > 0) {
  cpm_all <- left_join(cpm_all, distinct(sample_sheet, animal, condition),
                       by = "animal")
  wb <- createWorkbook()
  addWorksheet(wb, "pseudobulk_CPM")
  writeData(wb, "pseudobulk_CPM", cpm_all)
  saveWorkbook(wb, file.path(res_dir, "hashing_EAE-Symp_descriptive_CPM.xlsx"),
               overwrite = TRUE)
  cat("Saved descriptive CPM table for", length(unique(cpm_all$cluster)),
      "clusters.\n")
} else {
  message("No significant genes to describe - nothing saved.")
}
```

##12. Session info

```{r}
sessionInfo()
```
