---
title: "Humanized mice EAE analysis Brain"
author: "Mattia Zaghi"
date: "2026-03-06"
output:
  word_document: default
  html_document: default
  pdf_document: default
---

## What is this vignette for?
This Vignette contains all the necessary data analysis steps followed to analyzed the single cell RNA-seq of Mature humanized_mouse_braindendrocytes treated with FGF and Tomato controls including the code to generate the different figures.


##1.Load packages, configure directories and create the Seurat object

This section of the code is necessary to load all the necessary packages for the analysis, configure the directories and create the Seurat object that is going to be used for the rest of the analysis.

##1.1 Load packages, configure directories

1.First load all the necessary packages in order to run all the subsequent steps & set the working directory where all the output of this analysis will be saved and stored

```{r setup, include=FALSE}
library(Seurat)
library(future)
library(tidyverse)
library(dplyr)
library(ggthemes)
library(scales)
library(cowplot)
library(RColorBrewer)
library(openxlsx)
library(homologene)
library(clusterProfiler)
library(org.Mm.eg.db)
library(org.Hs.eg.db)
library(purrr)
library(forcats)
library(readxl)
library(hdf5r)
library(patchwork)
library(pals)
library(knitr)
library(DoubletFinder)
library(ggpubr)
library(rstatix)
library(pheatmap)
library(yaml)
opts_knit$set(root.dir = "/date/gcb/gcb_MZ/Humanized_mice_OD/")# change accordingly

#source("/home/mattia/FGF_signaling_induces_disease-associated_humanized_mouse_braindendrocytes/functions.R")
```

##1.2 Helper functions

The same plot themes, Excel exports and enrichment steps are needed in every
section below. They are defined once here so that a change to a theme or to an
export format propagates to every figure and table in the vignette, instead of
being retyped at each call site.

```{r}
# ---- Output locations ----------------------------------------------------
results_dir <- "/date/gcb/gcb_MZ/Humanized_mice_OD/Results"
plot_dir    <- file.path(results_dir, "Plots")

# ---- Plot themes ---------------------------------------------------------
# Theme shared by every UMAP: no grid, no tick labels, large centred title.
umap_theme <- function(legend_title_face = "bold") {
  list(
    labs(x = "UMAP 1", y = "UMAP 2"),
    theme_minimal(),
    theme(
      plot.title       = element_text(hjust = 0.5, size = 24, face = "bold"),
      axis.title.x     = element_text(size = 20),
      axis.title.y     = element_text(size = 20),
      legend.text      = element_text(size = 16),
      legend.title     = element_text(size = 24, face = legend_title_face),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      axis.text.x      = element_blank(),
      axis.text.y      = element_blank()
    )
  )
}

# Theme shared by every marker DotPlot.
dotplot_theme <- function() {
  theme(
    axis.text.x  = element_text(hjust = 1, vjust = 0.5, family = "Arial", size = 24, angle = 90),
    axis.text.y  = element_text(family = "Arial", size = 20),
    axis.title.y = element_text(family = "Arial", size = 24),
    legend.text  = element_text(size = 20, family = "Arial"),
    legend.title = element_text(size = 24, family = "Arial")
  )
}

# Theme shared by every GO / KEGG enrichment dot plot.
enrichment_theme <- function() {
  theme(
    axis.text.x  = element_text(hjust = 1, vjust = 0.5, family = "Arial", size = 24, angle = 45),
    axis.text.y  = element_text(family = "Arial", size = 24),
    axis.title.y = element_text(family = "Arial", size = 28, face = "bold"),
    axis.title.x = element_text(family = "Arial", size = 28, face = "bold"),
    legend.text  = element_text(size = 24, family = "Arial"),
    legend.title = element_text(size = 32, family = "Arial", face = "bold")
  )
}

# ---- Saving --------------------------------------------------------------
# Write a named list of ggplots to plot_dir as "<name><suffix>.png".
save_plots <- function(plots, suffix = "", dir = plot_dir,
                       width = 14, height = 16, dpi = 300) {
  dir.create(dir, recursive = TRUE, showWarnings = FALSE)
  for (nm in names(plots)) {
    ggplot2::ggsave(file.path(dir, paste0(nm, suffix, ".png")),
                    plots[[nm]], width = width, height = height, dpi = dpi)
  }
  invisible(plots)
}

# One worksheet per cluster of a FindAllMarkers-style data.frame.
write_cluster_workbook <- function(df, file, cluster_col = "cluster",
                                   prefix = "Cluster_") {
  wb <- createWorkbook()
  for (cluster_value in unique(df[[cluster_col]])) {
    sheet_name <- substr(paste0(prefix, cluster_value), 1, 31)
    addWorksheet(wb, sheet_name)
    writeData(wb, sheet_name, df[df[[cluster_col]] == cluster_value, ])
  }
  dir.create(dirname(file), recursive = TRUE, showWarnings = FALSE)
  saveWorkbook(wb, file, overwrite = TRUE)
  invisible(file)
}

# One worksheet per entry of a named list of enrichResult objects. Sheet names
# are truncated to Excel's 31-character limit; `keep_suffix` preserves a
# trailing tag such as _EAE_up when the cluster name is long.
write_enrichment_workbook <- function(enrich_list, file, label = "GO terms",
                                      keep_suffix = FALSE) {
  wb <- createWorkbook()
  for (cluster_name in names(enrich_list)) {
    res <- enrich_list[[cluster_name]]
    if (is.null(res) || nrow(as.data.frame(res)) == 0) {
      cat(sprintf("No significant %s for cluster: %s\n", label, cluster_name))
      next
    }
    res_df <- as.data.frame(res)
    res_df$mLog10Padj <- -log10(res_df$p.adjust)

    if (keep_suffix) {
      suffix     <- ifelse(grepl("_EAE_up$", cluster_name), "_EAE_up", "_CFA_up")
      prefix     <- sub("_(EAE|CFA)_up$", "", cluster_name)
      sheet_name <- paste0(substr(prefix, 1, 31 - nchar(suffix)), suffix)
    } else {
      sheet_name <- substr(cluster_name, 1, 31)
    }

    addWorksheet(wb, sheet_name)
    writeData(wb, sheet_name, res_df)
    cat(sprintf("Added sheet for cluster: %s (%d %s)\n",
                cluster_name, nrow(res_df), label))
  }
  dir.create(dirname(file), recursive = TRUE, showWarnings = FALSE)
  saveWorkbook(wb, file, overwrite = TRUE)
  cat(sprintf("\n%s results saved to %s\n", label, file))
  invisible(file)
}

# ---- Quality control plots ----------------------------------------------
# Mean and median of `y_var`, annotated above the violin of each sample.
add_mean_median_annotation <- function(plot, data, x_var, y_var) {
  for (sample in unique(data[[x_var]])) {
    sample_data  <- data[data[[x_var]] == sample, ]
    mean_value   <- mean(sample_data[[y_var]], na.rm = TRUE)
    median_value <- median(sample_data[[y_var]], na.rm = TRUE)
    plot <- plot +
      annotate("text", x = sample, y = max(sample_data[[y_var]], na.rm = TRUE) * 1.05,
               label = paste("Mean:", round(mean_value, 2)),
               color = "blue", size = 5, fontface = "bold") +
      annotate("text", x = sample, y = max(sample_data[[y_var]], na.rm = TRUE) * 1.50,
               label = paste("Median:", round(median_value, 2)),
               color = "red", size = 5, fontface = "bold")
  }
  plot
}

sample_palette <- c(
  "#1f77b4",  # blue
  "#ff7f0e",  # orange
  "#2ca02c",  # green
  "#d62728",  # red
  "#9467bd"   # purple
)

# Violin + boxplot of one QC metric per sample, annotated with mean and median.
qc_violin <- function(data, y_var, y_lab, x_var = "sample",
                      palette = sample_palette) {
  p <- ggplot(data) +
    aes(x = .data[[x_var]], y = .data[[y_var]],
        fill = .data[[x_var]], color = .data[[x_var]]) +
    geom_violin(trim = TRUE, color = "#000000") +
    geom_boxplot(width = 0.1, color = "#000000", fill = "#ffffff", outlier.shape = NA) +
    scale_fill_manual(values = palette) +
    scale_color_manual(values = palette) +
    ggthemes::theme_base() +
    xlab("") +
    ylab(y_lab) +
    theme(panel.border = element_rect()) +
    theme_classic() + theme(legend.position = "none") +
    theme(axis.text.x  = element_text(size = 18, family = "Arial", angle = 45, hjust = 1),
          axis.text.y  = element_text(size = 18, family = "Arial"),
          axis.title.y = element_text(size = 18, family = "Arial"),
          axis.line    = element_line(size = 1))
  add_mean_median_annotation(p, data, x_var, y_var)
}

# Stacked bar of the per-animal composition of every cluster.
cluster_composition_plot <- function(cluster_vec, sample_vec, colors_vector) {
  t  <- table(Cluster = cluster_vec, Batch = sample_vec)
  t  <- t[, rev(names(colors_vector))]
  df <- as.data.frame(t) %>%
    group_by(Cluster) %>%
    mutate(Percent = Freq / sum(Freq) * 100)

  ggplot(df, aes(x = Cluster, y = Percent, fill = Batch)) +
    geom_bar(stat = "identity", position = "fill") +
    scale_y_continuous(labels = percent_format(), expand = c(0, 0)) +
    scale_fill_manual(values = rev(colors_vector)) +
    theme_classic() +
    theme(
      axis.text.x  = element_text(hjust = 1, vjust = 0.5, family = "Arial", size = 16),
      axis.text.y  = element_text(family = "Arial", size = 16),
      axis.title.x = element_text(family = "Arial", size = 18),
      axis.title.y = element_text(family = "Arial", size = 18),
      legend.text  = element_text(size = 16, family = "Arial"),
      legend.title = element_text(size = 18, family = "Arial")
    ) +
    labs(y = "Percentage of cells", x = "Cluster", fill = "Animal") +
    guides(fill = guide_legend(ncol = 3, title.position = "top", title.hjust = 0.5))
}

# ---- Subset reclustering -------------------------------------------------
# Renormalise a subset on its own cells and recluster it. Every reduction and
# graph is named after the subset, so several subsets can be carried in the
# same session without overwriting one another.
recluster_subset <- function(obj, name, dims = 1:15, resolution = 0.1,
                             pca_name     = paste0("pca_", name, "_SCT"),
                             graph_prefix = name,
                             cluster_name = paste0(name, "_clusters"),
                             umap_name    = paste0("umap_", name),
                             verbose      = FALSE) {
  DefaultAssay(obj) <- "RNA"
  obj[["SCT"]] <- NULL
  obj <- SCTransform(obj, vst.flavor = "v2", verbose = verbose)
  obj <- RunPCA(obj, verbose = verbose, reduction.name = pca_name)
  obj <- FindNeighbors(obj, reduction = pca_name, dims = dims,
                       graph.name = c(paste0(graph_prefix, "_nn"),
                                      paste0(graph_prefix, "_snn")))
  obj <- FindClusters(obj, resolution = resolution,
                      graph.name   = paste0(graph_prefix, "_snn"),
                      cluster.name = cluster_name)
  RunUMAP(obj, reduction = pca_name, dims = dims, reduction.name = umap_name)
}

# Cluster markers on the SCT assay, written to one worksheet per cluster.
subset_markers <- function(obj, file) {
  DefaultAssay(obj) <- "SCT"
  obj <- PrepSCTFindMarkers(obj, assay = "SCT", verbose = FALSE)
  markers <- FindAllMarkers(obj, only.pos = TRUE)
  write_cluster_workbook(markers, file)
  markers
}

# ---- Functional enrichment ----------------------------------------------
# Marker symbols -> ENTREZ ids, one vector per cluster. The mm10- prefix that
# the joint reference adds to mouse genes is stripped before mapping.
entrez_by_cluster <- function(markers_df, cluster_col = "cluster",
                              gene_col = "gene") {
  marker_list     <- split(markers_df[[gene_col]], markers_df[[cluster_col]])
  all_genes_clean <- sub("^mm10-", "", unique(unlist(marker_list)))
  gene_map <- bitr(all_genes_clean, fromType = "SYMBOL", toType = "ENTREZID",
                   OrgDb = org.Mm.eg.db)
  sym2ent  <- setNames(gene_map$ENTREZID, gene_map$SYMBOL)
  lapply(marker_list, function(x) na.omit(unname(sym2ent[sub("^mm10-", "", x)])))
}

go_bp_by_cluster <- function(entrez_list) {
  ego <- lapply(names(entrez_list), function(cl) {
    enrichGO(gene          = entrez_list[[cl]],
             OrgDb         = org.Mm.eg.db,
             keyType       = "ENTREZID",
             ont           = "BP",
             pAdjustMethod = "BH",
             pvalueCutoff  = 0.05,
             qvalueCutoff  = 0.2,
             readable      = TRUE)
  })
  setNames(ego, names(entrez_list))
}

kegg_by_cluster <- function(entrez_list) {
  ekegg <- lapply(names(entrez_list), function(cl) {
    enrichKEGG(gene = entrez_list[[cl]], organism = "mmu", pvalueCutoff = 0.05)
  })
  setNames(ekegg, names(entrez_list))
}

# Flatten a named list of enrichResult objects into one data.frame.
enrichment_to_df <- function(enrich_list) {
  map2_df(names(enrich_list), enrich_list, ~{
    if (is.null(.y) || nrow(as.data.frame(.y)) == 0) return(NULL)
    df <- as.data.frame(.y)
    df$cluster    <- .x
    df$mLog10Padj <- -log10(df$p.adjust)
    df
  })
}

# Top `top_n` terms per cluster, as a dot plot sized by gene count.
plot_enrichment_dotplot <- function(enrich_df, title, x_lab = "Cluster",
                                    y_lab = "Term", top_n = 10) {
  plot_df <- enrich_df %>%
    group_by(cluster) %>%
    slice_max(order_by = mLog10Padj, n = top_n, with_ties = FALSE) %>%
    ungroup() %>%
    mutate(Description = fct_reorder(Description, mLog10Padj))

  if (nrow(plot_df) == 0) return(NULL)

  ggplot(plot_df, aes(x = cluster, y = Description, size = Count, color = mLog10Padj)) +
    geom_point() +
    scale_color_gradient(low = "gold", high = "red") +
    labs(title = title, x = x_lab, y = y_lab, color = "-log10(FDR)") +
    theme_bw(base_size = 12) + theme_classic() +
    enrichment_theme()
}

# ---- Label transfer ------------------------------------------------------
# Transfer `refdata` from a reference onto the query and tag the prediction
# columns with `suffix`, so several references can be carried side by side.
transfer_labels <- function(reference, query, features, refdata, suffix,
                            dims = 1:50) {
  anchors <- FindTransferAnchors(
    reference       = reference,
    query           = query,
    reference.assay = "RNA",
    query.assay     = "RNA",
    reduction       = "cca",
    dims            = dims,
    features        = features
  )
  query <- MapQuery(anchorset = anchors, reference = reference,
                    query = query, refdata = refdata)

  cn <- colnames(query@meta.data)
  cn[cn == "predicted.id"]       <- paste0("predicted.id_", suffix)
  cn[cn == "predicted.id.score"] <- paste0("predicted.id.score_", suffix)
  colnames(query@meta.data) <- cn
  query
}

# Distribution of the per-reference similarity scores, one panel per cluster.
prediction_score_violins <- function(obj, cluster_col, assay = "prediction.score.id") {
  assay_data <- GetAssayData(obj, assay = assay, layer = "data")

  for (cluster_id in unique(obj@meta.data[[cluster_col]])) {
    cluster_cells <- colnames(obj)[obj@meta.data[[cluster_col]] == cluster_id]

    all_scores <- do.call(rbind, lapply(rownames(assay_data), function(ref_celltype) {
      data.frame(score              = as.numeric(assay_data[ref_celltype, cluster_cells]),
                 reference_celltype = ref_celltype,
                 stringsAsFactors   = FALSE)
    }))

    print(
      ggplot(all_scores, aes(x = reference_celltype, y = score, fill = reference_celltype)) +
        geom_violin(trim = TRUE, scale = "width") +
        geom_boxplot(width = 0.1, fill = "white", outlier.shape = NA) +
        ggtitle(paste("Cluster:", cluster_id)) +
        xlab("Reference Cell Type") +
        ylab("Similarity Score") +
        ylim(0, 1) +
        theme_classic() +
        theme(
          plot.title   = element_text(hjust = 0.5, size = 36, face = "bold"),
          axis.text.x  = element_text(hjust = 1, vjust = 0.5, family = "Arial", size = 14, angle = 90),
          axis.text.y  = element_text(family = "Arial", size = 20),
          axis.title.x = element_text(family = "Arial", size = 24, face = "bold"),
          axis.title.y = element_text(family = "Arial", size = 24, face = "bold"),
          legend.position = "none"
        )
    )
  }
}
```

##2. Sample sheet

The sample names are read from the preprocessing config rather than retyped, so
this vignette cannot drift out of sync with what Cell Ranger actually
processed. Only the brain libraries are kept here. `condition` (EAE or CFA) and
`replicate` are parsed from the sample name (`<condition>_brain_<replicate>`);
`condition` is the contrast tested in the differential expression section and
the library is the unit of replication, 3 EAE animals against 2 CFA animals.

```{r}
base_dir <- "/date/gcb/gcb_MZ/Humanized_mice_OD"

# knitr's root.dir points at the data directory, not the repository, so the
# config is addressed absolutely. Adjust repo_dir if the checkout moves.
repo_dir    <- "/home/mattia/Humanized_mice_EAE"
config_path <- file.path(repo_dir, "Preprocessing/config/config.yaml")
stopifnot(file.exists(config_path))

cfg           <- yaml::read_yaml(config_path)
brain_samples <- grep("brain", names(cfg$samples), value = TRUE)

sample_sheet <- data.frame(
  sample           = brain_samples,
  stringsAsFactors = FALSE
)
sample_sheet$condition <- sub("_.*$", "", sample_sheet$sample)
sample_sheet$replicate <- sub("^.*_", "", sample_sheet$sample)

sample_sheet$condition <- factor(sample_sheet$condition, levels = c("CFA", "EAE"))

# Sanity check: replication available per condition
table(sample_sheet$condition)
sample_sheet
```

##3. Create the Seurat object

Let's start with loading the actual data. We will load separately all the brain samples samples from the cell ranger output and then everything will be merged into a unique object thet will be called brain hMS.

1.load all the seurat object corrected by cell bender

DoubletFinder is run on each library independently, so that the artificial
doublets are built from the same cell composition as the library being scored.
The expected multiplet rate is not a single value for all samples: it is read
off the 10x Chromium table from each library's own recovered-cell count, then
reduced by the homotypic proportion, which is the fraction of doublets formed
by two cells of the same type and therefore not detectable. `pK` is likewise
estimated per sample with `paramSweep`/`find.pK` rather than fixed at 0.09.

```{r}
# Expected multiplet rate as a function of recovered cells, interpolated from
# the 10x Chromium 3' table (~0.8% per 1,000 cells, near-linear in this range).
# Each library is scored with the rate implied by its own cell count instead of
# a single 0.075 for every sample, which corresponds to ~10,000 recovered cells
# and over-filters the smaller libraries.
doublet_rate_10x <- function(n_cells) {
  ref_cells <- c(  500,  1000,  2000,  3000,  4000,  5000,
                  6000,  7000,  8000,  9000, 10000)
  ref_rate  <- c(0.004, 0.008, 0.016, 0.023, 0.031, 0.039,
                 0.046, 0.054, 0.061, 0.069, 0.076)
  as.numeric(approx(ref_cells, ref_rate, xout = n_cells, rule = 2)$y)
}

mt_cutoff <- 5

# Load each sample and create a Seurat object (with unique cell names)
seurat_list <- list()
for (i in seq_len(nrow(sample_sheet))) {
  sample  <- sample_sheet$sample[i]
  h5_path <- file.path(base_dir, sample, "RNA_AAAAGGGG", "cellbender_output_seurat.h5")
  if (!file.exists(h5_path)) {
    warning("Missing file: ", h5_path)
    next
  }
  message("Loading ", sample)

  data <- Read10X_h5(filename = h5_path, use.names = TRUE)

  # Normalize feature names: collapse underscores and multiple dashes into a single dash
  fn <- rownames(data)
  fn <- gsub("_+", "-", fn)      # underscores -> dash
  fn <- gsub("-{2,}", "-", fn)   # collapse multiple dashes
  rownames(data) <- fn

  # Prefix cell barcodes with sample name so they are unique across samples
  colnames(data) <- paste0(sample, "_", colnames(data))

  seurat_obj <- CreateSeuratObject(
    counts = data,
    project = sample,
    min.cells = 3,
    min.features = 200
  )

  # Per-sample QC before merging: mouse (mm10) and human (hg38) mitochondrial
  # fractions are summed, because the reference is the joint GRCh38 + mm10 one.
  seurat_obj[["percent.mt.mm10"]] <- PercentageFeatureSet(seurat_obj, pattern = "mt-")
  seurat_obj[["percent.mt.hg38"]] <- PercentageFeatureSet(seurat_obj, pattern = "MT-")
  seurat_obj[["percent.mt"]] <-
    seurat_obj[["percent.mt.mm10"]] + seurat_obj[["percent.mt.hg38"]]

  seurat_obj <- subset(seurat_obj, subset = percent.mt < mt_cutoff)

  # Carry the design metadata on every cell
  seurat_obj$sample    <- sample
  seurat_obj$condition <- as.character(sample_sheet$condition[i])
  seurat_obj$replicate <- sample_sheet$replicate[i]

  # Light preprocessing for doublet detection (on each sample independently using SCT)
  seurat_obj <- SCTransform(seurat_obj, verbose = FALSE)
  seurat_obj <- RunPCA(seurat_obj, verbose = FALSE)
  seurat_obj <- FindNeighbors(seurat_obj, dims = 1:10, verbose = FALSE)
  seurat_obj <- FindClusters(seurat_obj, resolution = 0.1, verbose = FALSE)
  seurat_obj <- RunUMAP(seurat_obj, dims = 1:10, verbose = FALSE)

  umap_before <- DimPlot(seurat_obj, reduction = "umap", group.by = "seurat_clusters") + ggtitle(paste(sample, "before doublet removal"))

  # Per-sample DoubletFinder (SCT input)
  # pK sets the neighbourhood size used to compute pANN and its optimum depends
  # on the number of cells and of distinct populations, so it is estimated on
  # each library rather than fixed at 0.09 for all of them.
  sweep_res   <- paramSweep(seurat_obj, PCs = 1:10, sct = TRUE)
  sweep_stats <- summarizeSweep(sweep_res, GT = FALSE)
  bcmvn       <- find.pK(sweep_stats)
  pK <- as.numeric(as.character(bcmvn$pK[which.max(bcmvn$BCmetric)]))
  if (length(pK) == 0 || is.na(pK)) pK <- 0.09   # fall back to the fixed value
  pN <- 0.25
  message("  pK = ", pK)

  # Expected number of doublets, from this library's own cell count and
  # corrected downward for the homotypic pairs that cannot be detected.
  rate <- doublet_rate_10x(ncol(seurat_obj))

  # modelHomotypic needs a clustering fine enough to separate cell types; the
  # res = 0.1 used for the QC UMAP is too coarse and inflates the homotypic
  # estimate. A finer clustering is computed here for this purpose only.
  seurat_obj <- FindClusters(seurat_obj, resolution = 0.8,
                             cluster.name = "clusters_for_homotypic",
                             verbose = FALSE)
  homotypic <- modelHomotypic(seurat_obj$clusters_for_homotypic)

  nExp_poi     <- round(rate * ncol(seurat_obj))
  nExp_poi.adj <- round(nExp_poi * (1 - homotypic))
  message("  cells = ", ncol(seurat_obj),
          ", rate = ", signif(rate, 3),
          ", homotypic = ", signif(homotypic, 3),
          ", nExp = ", nExp_poi, " -> ", nExp_poi.adj)

  seurat_obj <- doubletFinder(seurat_obj, PCs = 1:10, pN = pN, pK = pK, nExp = nExp_poi, reuse.pANN = NULL, sct = TRUE)

  df_col <- grep("^pANN_", colnames(seurat_obj@meta.data), value = TRUE)[1]
  if (is.na(df_col)) stop("DoubletFinder pANN metadata not found")

  seurat_obj <- doubletFinder(seurat_obj, PCs = 1:10, pN = pN, pK = pK, nExp = nExp_poi.adj, reuse.pANN = df_col, sct = TRUE)

  # The adjusted classification is the one keyed to nExp_poi.adj; select it by
  # name rather than by position so the choice does not depend on column order.
  df_class_col <- grep(paste0("^DF.classifications_.*_", nExp_poi.adj, "$"),
                       colnames(seurat_obj@meta.data), value = TRUE)
  if (length(df_class_col) == 0) {
    df_class_col <- tail(grep("^DF.classifications_",
                              colnames(seurat_obj@meta.data), value = TRUE), 1)
  }
  if (length(df_class_col) == 0) stop("DoubletFinder classification column not found")
  message("Using doublet classification column: ", df_class_col)

  seurat_obj$doublet_class <- seurat_obj@meta.data[[df_class_col]]
  seurat_obj$doublet_pANN  <- seurat_obj@meta.data[[df_col]]

  sel_singlets <- seurat_obj$doublet_class == "Singlet"
  if (sum(sel_singlets, na.rm = TRUE) == 0) {
    stop("No cells annotated as Singlet in ", df_class_col)
  }
  seurat_obj <- subset(seurat_obj, cells = colnames(seurat_obj)[which(sel_singlets)])

  # Recompute UMAP after doublet removal (SCT after singlet filtering)
  # Remove the old SCT assay 
  DefaultAssay(seurat_obj) <- "RNA"
  seurat_obj[["SCT"]] <- NULL
  seurat_obj <- SCTransform(seurat_obj, verbose = TRUE)
  seurat_obj <- RunPCA(seurat_obj, verbose = TRUE)
  seurat_obj <- FindNeighbors(seurat_obj, dims = 1:10, verbose = TRUE)
  seurat_obj <- FindClusters(seurat_obj, resolution = 0.1, verbose = TRUE)
  seurat_obj <- RunUMAP(seurat_obj, dims = 1:10, verbose = TRUE)

  umap_after <- DimPlot(seurat_obj, reduction = "umap", group.by = "seurat_clusters") + ggtitle(paste(sample, "after doublet removal"))

  # Print per-sample before/after UMAP side-by-side
  print(umap_before + umap_after)

  seurat_list[[sample]] <- seurat_obj
}
```



2.Now all the different samples are going to be merged in one unique object 

```{r}
humanized_mouse_brain <- Reduce(function(x, y) merge(x, y), seurat_list)

rm(seurat_list); gc()

# Remove metadata from previous single-sample doublet removal analysis. The
# pANN / DF.classifications columns are keyed to each sample's own pK and nExp,
# so they are matched by pattern rather than by their exact names.
drop_cols <- grep("^(pANN_|DF.classifications_|nCount_SCT|nFeature_SCT|SCT_snn_res|seurat_clusters|clusters_for_homotypic)",
                  colnames(humanized_mouse_brain@meta.data), value = TRUE)
for (cc in drop_cols) humanized_mouse_brain@meta.data[[cc]] <- NULL
```

filter away cells that have a very high mitocondrial content

```{r}
humanized_mouse_brain <- subset(humanized_mouse_brain, subset = percent.mt < mt_cutoff)
```


##4.Quality controls and cell selection

In this sections quality controls based selection of good will be performed on the merged object will be performed before proceeding with the rest of then analysis. 

1.Let's first plot some QCs metrics to have a look at the global dataset, this will give a general idea of the dataset quality

```{r}
# add the mitochondrial gene precentage 
# mouse mito (mm10)
humanized_mouse_brain[["percent.mt.mm10"]] <- PercentageFeatureSet(
  humanized_mouse_brain,
  pattern = "mt-"
)

# human mito (hg38)
humanized_mouse_brain[["percent.mt.hg38"]] <- PercentageFeatureSet(
  humanized_mouse_brain,
  pattern = "MT-"
)
humanized_mouse_brain[["percent.mt"]] <- humanized_mouse_brain[["percent.mt.mm10"]] + humanized_mouse_brain[["percent.mt.hg38"]]
```

check human vs mouse cells

```{r}
# Assume this is your merged object
obj <- humanized_mouse_brain
DefaultAssay(obj) <- "RNA"
obj <- JoinLayers(obj)
# Count UMIs per cell for each species (adjust prefixes as needed)
human_umis <- Matrix::colSums(
  GetAssayData(obj, assay = "RNA", layer = "counts")[grep("^GRCh38-", rownames(obj)), , drop = FALSE]
)
mouse_umis <- Matrix::colSums(
  GetAssayData(obj, assay = "RNA", layer = "counts")[grep("^mm10-", rownames(obj)), , drop = FALSE]
)

# Assign species label based on which count is higher
obj$species <- ifelse(human_umis > mouse_umis, "human", "mouse")

# Optionally mark ambiguous/mixed cells (e.g., 20–80% human)
ratio <- human_umis / (human_umis + mouse_umis)
obj$species[ratio > 0.2 & ratio < 0.8] <- "mixed"

# Write back
humanized_mouse_brain <- obj

#how many cells per sample
table(humanized_mouse_brain$orig.ident)

#how many cell per species
table(humanized_mouse_brain$species)

#how many cell per replicate
table(humanized_mouse_brain$sample)

```


```{r,fig.height=10, fig.width=10}
# Visualize QC metrics as a violin plot

QC_RNA <- data.frame(sample =  humanized_mouse_brain@meta.data[["sample"]],
                     Genes = humanized_mouse_brain@meta.data[["nFeature_RNA"]],
                     RNA = humanized_mouse_brain@meta.data[["nCount_RNA"]],
                     Mt = humanized_mouse_brain@meta.data[["percent.mt"]])
QC_RNA$sample <- as.factor(QC_RNA$sample)

Transcripts <- qc_violin(QC_RNA, "RNA",   "Number of UMI per cell")
Genes       <- qc_violin(QC_RNA, "Genes", "Number of Genes per cell")
Mt          <- qc_violin(QC_RNA, "Mt",    "Percentage of Mt genes")

Transcripts
Genes
Mt

```

let's also visualize how is the relationship between counts, genes and mito content

```{r}

plot1 <- FeatureScatter(humanized_mouse_brain, feature1 = "nCount_RNA", feature2 = "percent.mt")
plot2 <- FeatureScatter(humanized_mouse_brain, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
plot1 + plot2
```







##5.Data integration, dimensionlity reduction and clustering

In this section we will perform the data integration, dimensionality reduction and clustering of the

##5.1 Dimensionality reduction and clustering without data integration
Despite in this experiment there are not multiple data batches and replicate we will first perform a clustering with data ingtegration and without and look at the difference

1.Data normalization, variable features finding PCA

```{r}
options(future.globals.maxSize = 100e+09)

humanized_mouse_brain[["SCT"]] <- NULL

# Log-normalized analysis on RNA
humanized_mouse_brain <- NormalizeData(humanized_mouse_brain, assay = "RNA")
humanized_mouse_brain <- FindVariableFeatures(humanized_mouse_brain, assay = "RNA")
humanized_mouse_brain <- ScaleData(humanized_mouse_brain, assay = "RNA")
humanized_mouse_brain <- RunPCA(humanized_mouse_brain, assay = "RNA", reduction.name = "pca_lognorm")

# SCT analysis on SCT assay
humanized_mouse_brain <- SCTransform(humanized_mouse_brain, verbose = TRUE)
humanized_mouse_brain <- FindVariableFeatures(humanized_mouse_brain, assay = "SCT")
humanized_mouse_brain <- RunPCA(humanized_mouse_brain, assay = "SCT", reduction.name = "pca_SCT")
```

2.Having a look at the elbow plot to understand how many PCs are really informative, this value will be used in the subsequent analysis steps 
```{r}
p1 <- ElbowPlot(humanized_mouse_brain, reduction = "pca_SCT")
p1
```
After 15 SD the elbow plot becomes flat indicating that the majority of the information is explained by the first 15 dimension, this value will be used for the rest of the analysis.

3.Clustering without data integration

```{r}
humanized_mouse_brain <- FindNeighbors(humanized_mouse_brain, dims = 1:15, reduction = "pca_SCT")
humanized_mouse_brain <- FindClusters(humanized_mouse_brain, resolution = 0.1, cluster.name = "unintegrated_clusters")
```
4. Plot UMAP with samples and clustering
```{r,fig.height=10, fig.width=10}
humanized_mouse_brain <- RunUMAP(humanized_mouse_brain, dims = 1:15, reduction = "pca_SCT", reduction.name = "umap.unintegrated")
# visualize by batch and cell type annotation

my_colors <- c(
  "#1F77B4", "#FF7F0E", "#2CA02C", "#D62728", "#9467BD",
  "#8C564B", "#17BECF", "#BCBD22", "#7F7F7F", "#e377c2",
  "#aec7e8", "#ffbb78", "#98df8a"
)


p1 <- DimPlot(humanized_mouse_brain, reduction = "umap.unintegrated",
              group.by = "unintegrated_clusters", cols = my_colors) + umap_theme()
p2 <- DimPlot(humanized_mouse_brain, reduction = "umap.unintegrated",
              group.by = "condition") + umap_theme()
p3 <- DimPlot(humanized_mouse_brain, reduction = "umap.unintegrated",
              group.by = "sample") + umap_theme()

p1
p2
p3
saveRDS(humanized_mouse_brain,"/date/gcb/gcb_MZ/Humanized_mice_OD/humanized_mouse_brain_doublets.rds")
```

```{r,fig.height=10, fig.width=20}
colors_vector <- c(
  "CFA_brain_1" = "#E41A1C", 
  "CFA_brain_2" = "#377EB8", 
  "EAE_brain_1" = "#984EA3", 
  "EAE_brain_2" = "#FF7F00", 
  "EAE_brain_3" = "#FFFF33"
)

cluster_composition_plot(humanized_mouse_brain$unintegrated_clusters,
                         humanized_mouse_brain$sample,
                         colors_vector)
```


4.Now let's plot some markers genes to rename and homogenize clusters. After that the final steps of the analysis will be performed.The code use to generate the plots is used in the Supplementary Fig.S3C.


```{r,fig.height=80, fig.width=80}
markers <- c(
  # Cluster 0 - Cerebellar Granule Cells
  "mm10-Gabra6", "mm10-Cbln1", "mm10-Cbln3", "mm10-Neurod1", "mm10-Barhl2",
  # Cluster 1 - Excitatory Cortical Neurons (Glutamatergic)
  "mm10-Tbr1", "mm10-Satb2", "mm10-Emx1", "mm10-Neurod6",
  # Cluster 2 - GABAergic Interneurons (SST/VIP subtypes)
  "mm10-Lhx6", "mm10-Lhx8", "mm10-Sst", "mm10-Vip", "mm10-Tac2", "mm10-Slc5a7", "mm10-Nxph2",
  # Cluster 3 - GABAergic Inhibitory Neurons (Striatal/MGE-derived)
  "mm10-Dlx1", "mm10-Dlx2", "mm10-Gad1", "mm10-Sp9", "mm10-Meis2", "mm10-Pbx3",
  # Cluster 4 - Mature Oligodendrocytes
  "mm10-Plp1", "mm10-Mbp", "mm10-Apod", "mm10-Ptgds",
  # Cluster 5 - Astrocytes
  "mm10-Atp1a2", "mm10-Slc1a3", "mm10-Apoe", "mm10-Gja1", "mm10-Gli1", "mm10-Mfge8",
  # Cluster 6 - Medium Spiny Neurons (Striatal)
  "mm10-Drd1", "mm10-Drd2", "mm10-Adora2a", "mm10-Gpr88", "mm10-Rgs9", "mm10-Rarb",
  # Cluster 7 - Myelinating Oligodendrocytes
  "mm10-Klk6", "mm10-Gjc2", "mm10-Ugt8a", "mm10-Gjb1",
  # Cluster 8 - Cerebellar Inhibitory Interneurons (MLI/Pvalb+)
  "mm10-Pvalb", "mm10-Pax2", "mm10-Kit", "mm10-Tfap2a", "mm10-Tfap2b", "mm10-Skor1",
  # Cluster 9 - Microglia
  "mm10-Tmem119", "mm10-Hexb", "mm10-Trem2", "mm10-Tyrobp", "mm10-Inpp5d", "mm10-Itgam", "mm10-Cd14", "mm10-Adgre1",
  # Cluster 10 - Thalamic Neurons
  "mm10-Synpo2", "mm10-Zfhx3", "mm10-Shox2", "mm10-Lef1", "mm10-Slitrk6",
  # Cluster 11 - Oligodendrocyte Precursor Cells (OPCs)
  "mm10-Pdgfra", "mm10-Cspg4", "mm10-Vcan",
  # Cluster 12 - Purkinje Cells (Cerebellar)
  "mm10-Car8", "mm10-Pcp2", "mm10-Ppp1r17", "mm10-Col18a1", "mm10-Atp2a3"
)

# 1) create a list of plots (one per gene)
plots <- FeaturePlot(
  humanized_mouse_brain,
  features = markers,
  reduction = "umap.unintegrated",
  cols = c("lightgrey", "magenta"),
  combine = FALSE
)

# 2) apply consistent theme / title formatting to every plot
plots <- lapply(plots, function(p) {
  p +
    labs(x = "UMAP 1", y = "UMAP 2") +
    theme_minimal() +
    theme(
      plot.title = element_text(hjust = 0.5, size = 64, face = "bold"),
      axis.title.x = element_text(size = 50),
      axis.title.y = element_text(size = 50),
      legend.text = element_text(size = 50),
      legend.title = element_text(size = 52),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      axis.text.x = element_blank(),
      axis.text.y = element_blank()
    )
})

# 3) combine into a grid (e.g., 2 columns)
p <- wrap_plots(plots, ncol = 4)
p
```


5.Cluster renaming, based on the very broad classification OL, OPCs or microglia, this clustering to subset only MOLs to subsequently use them in the next analysis. The UMAP plot with the samples correponds to fig.3C and the UMAP plot with the broad cell types corresponds to Fig.S3B. 

```{r,fig.height=10, fig.width=10}
broad_clustering <- c("Cerebellar_Granule_Cells","Excitatory_Cortical_Neu","SST_Gaba_Neu","Stria_MGE_Inter_Neu","MOL",
                      "Astrocytes","Medium_Spiny_Neu_Stria","Myelin_MOL","Cerebellar_Inhib_Interneu_Pvalb+",
                      "Microglia","Thal_Neu","OPcs","Purkinje_Cells")

names(broad_clustering) <- levels(humanized_mouse_brain)
humanized_mouse_brain <- RenameIdents(humanized_mouse_brain, broad_clustering)
humanized_mouse_brain$broad_clustering <- Idents(humanized_mouse_brain) #create a new metadata colums with the broad clustering

# Check the new identities
table(Idents(humanized_mouse_brain))

my_colors <- c(
  "#1F77B4", "#FF7F0E", "#2CA02C", "#D62728", "#9467BD",
  "#8C564B", "#17BECF", "#BCBD22", "#7F7F7F", "#e377c2",
  "#aec7e8", "#ffbb78"
)

# UMAP with the condition (Fig.3C)
p1 <- DimPlot(humanized_mouse_brain, reduction = "umap.unintegrated",
              group.by = "condition") + umap_theme()

# UMAP with the broad cell types (Fig.S3B)
p2 <- DimPlot(humanized_mouse_brain, reduction = "umap.unintegrated",
              group.by = "broad_clustering", cols = my_colors) + umap_theme()

p1
p2

#update cluster calling myelin_MOL and MOL as unique cluster
humanized_mouse_brain$broad_clustering <- as.character(humanized_mouse_brain$broad_clustering)
humanized_mouse_brain$broad_clustering[grepl("^myelin[_ ]?mol$", humanized_mouse_brain$broad_clustering, ignore.case=TRUE)] <- "MOL"
humanized_mouse_brain$broad_clustering <- factor(humanized_mouse_brain$broad_clustering)
Idents(humanized_mouse_brain) <- humanized_mouse_brain$broad_clustering
table(humanized_mouse_brain$broad_clustering)
```

Dotplot for the broad clustering 


```{r,fig.height=20, fig.width=27}

markers <- c(
  # Cluster 0 - Cerebellar Granule Cells
  "mm10-Gabra6", "mm10-Cbln1", "mm10-Cbln3", "mm10-Neurod1", "mm10-Barhl2",
  # Cluster 1 - Excitatory Cortical Neurons (Glutamatergic)
  "mm10-Tbr1", "mm10-Satb2", "mm10-Emx1", "mm10-Neurod6",
  # Cluster 2 - GABAergic Interneurons (SST/VIP subtypes)
  "mm10-Lhx6", "mm10-Lhx8", "mm10-Sst", "mm10-Vip", "mm10-Tac2", "mm10-Slc5a7", "mm10-Nxph2",
  # Cluster 3 - GABAergic Inhibitory Neurons (Striatal/MGE-derived)
  "mm10-Dlx1", "mm10-Dlx2", "mm10-Gad1", "mm10-Sp9", "mm10-Meis2", "mm10-Pbx3",
  # Cluster 4 - Mature Oligodendrocytes
  "mm10-Plp1", "mm10-Mbp", "mm10-Apod", "mm10-Ptgds",
  # Cluster 5 - Astrocytes
  "mm10-Atp1a2", "mm10-Slc1a3", "mm10-Apoe", "mm10-Gja1", "mm10-Gli1", "mm10-Mfge8",
  # Cluster 6 - Medium Spiny Neurons (Striatal)
  "mm10-Drd1", "mm10-Drd2", "mm10-Adora2a", "mm10-Gpr88", "mm10-Rgs9", "mm10-Rarb",
  # Cluster 7 - Myelinating Oligodendrocytes
  #"mm10-Klk6", "mm10-Gjc2", "mm10-Ugt8a", "mm10-Gjb1",
  # Cluster 8 - Cerebellar Inhibitory Interneurons (MLI/Pvalb+)
  "mm10-Pvalb", "mm10-Pax2", "mm10-Kit", "mm10-Tfap2a", "mm10-Tfap2b", "mm10-Skor1",
  # Cluster 9 - Microglia
  "mm10-Tmem119", "mm10-Hexb", "mm10-Trem2", "mm10-Tyrobp", "mm10-Inpp5d", "mm10-Itgam", "mm10-Cd14", "mm10-Adgre1",
  # Cluster 10 - Thalamic Neurons
  "mm10-Synpo2", "mm10-Zfhx3", "mm10-Shox2", "mm10-Lef1", "mm10-Slitrk6",
  # Cluster 11 - Oligodendrocyte Precursor Cells (OPCs)
  "mm10-Pdgfra", "mm10-Cspg4", "mm10-Vcan",
  # Cluster 12 - Purkinje Cells (Cerebellar)
  "mm10-Car8", "mm10-Pcp2", "mm10-Ppp1r17", "mm10-Col18a1", "mm10-Atp2a3"
)

p1 <- DotPlot(humanized_mouse_brain, features = markers, group.by = "broad_clustering", cols = c("lightgrey", "firebrick")) +
  xlab("")+
  scale_x_discrete(labels = markers) + 
  dotplot_theme()
p1
```





##6. Reclustering of the glial subsets

Two broad populations are taken out of the whole-brain object and analysed on
their own: the oligodendroglia (MOL and OPCs) and the microglia. Each of them is
renormalised and reclustered with `recluster_subset()`, which names every
reduction, graph and cluster column after the subset, so that the two objects
can be carried side by side without overwriting each other.

##6.1 Oligodendroglia (MOL and OPCs)

1.Subset the whole dataset and recluster on the oligodendroglial cells only

```{r}
OL <- subset(humanized_mouse_brain, idents = c("MOL", "OPcs"))

OL <- recluster_subset(OL, "OL", dims = 1:15, resolution = 0.1)
```

2.Plot UMAP with samples and the new oligodendroglial clusters

```{r}
my_colors <- c(
  "#1F77B4",
  "#FF7F0E",
  "#2CA02C",
  "#D62728",
  "#9467BD",
  "#8C564B",
  "#17BECF"
)

p1 <- DimPlot(OL, reduction = "umap_OL", group.by = "OL_clusters",
              cols = my_colors) + umap_theme()          # UMAP with new clustering
p2 <- DimPlot(OL, reduction = "umap_OL", group.by = "condition") + umap_theme()
p3 <- DimPlot(OL, reduction = "umap_OL", group.by = "sample") + umap_theme()
p4 <- DimPlot(OL, reduction = "umap_OL", group.by = "broad_clustering") + umap_theme()

p1
p2
p3
p4
```

clusters proportions

```{r}
colors_vector <- c(
  "CFA_brain_1" = "#E41A1C",
  "CFA_brain_2" = "#377EB8",
  "EAE_brain_1" = "#984EA3",
  "EAE_brain_2" = "#FF7F00",
  "EAE_brain_3" = "#FFFF33"
)

cluster_composition_plot(OL$OL_clusters, OL$sample, colors_vector)
```

3.Plotting marker genes to review the clustering and rename the different populations of MOLs

```{r,fig.height=20, fig.width=20}

markers <- c(
  #MOL5/6–EAE-b
  "mm10-Serpina3n","mm10-Nlrc5",
  #MOL5/6–EAE-a
  "mm10-Plin4","mm10-Sult1a1","mm10-Zfand4",
  #MOL1/2 EAE
  "mm10-Klk8","mm10-Itga8", "mm10-Tlr3",
  #MHC-II
  "mm10-H2-Aa", "mm10-H2-Ab1", "mm10-H2-Eb1", "mm10-Cd74", "mm10-Ctss",
  #interferon
  "mm10-Ifih1", "mm10-Irf7", "mm10-Irgm1", "mm10-Irgm2", "mm10-Igtp","mm10-Zbp1",
   #cycling_DAO
  "mm10-Cdkn1a","mm10-Ccnd1","mm10-Nav3","mm10-Tmsb10","mm10-Lrrfip1",
  # OPC1
  "mm10-Rph3a", "mm10-Adora2b","mm10-Pdgfra","mm10-Cspg4",
  # OPC2
  "mm10-Phyhd1","mm10-Fcgr2b","mm10-Mylk",
  # OPC3
  "mm10-Slc14a1","mm10-Lgals1","mm10-Myrf",
  #MOL 5/6
  "mm10-Ptgds", "mm10-Opalin", "mm10-Qdpr", "mm10-Il33", "mm10-Apoe", "mm10-Cd63", "mm10-Jph4","mm10-Fgfr2"

)

p1 <- DotPlot(OL, features = markers, group.by = "OL_clusters", cols = c("lightgrey", "firebrick")) +
  xlab("")+
  scale_x_discrete(labels = markers) +
  dotplot_theme()
p1
```


```{r,fig.height=20, fig.width=20}

markers <- c(
  #MOL1
  "mm10-Fosb","mm10-Dusp1","mm10-Dnajb1","mm10-Btg2","mm10-Hspa1a","mm10-Hspa1b","mm10-Tob1","mm10-Mal",
  #MOL2
  "mm10-Anxa5","mm10-Klk6","mm10-Mgst3","mm10-S100b","mm10-Tppp3","mm10-Cyp51","mm10-Plekhb1",
  # MFOL 1/2
  "mm10-Ctps","mm10-Tmem141","mm10-Opalin","mm10-Ttyh1", "mm10-Plekha1", "mm10-Ddr1", "mm10-Cd9", "mm10-Sirt2","mm10-Arpc1b",
  # NFOL1/2
  "mm10-Sema6a","mm10-Mag","mm10-Mobp","mm10-Mbp","mm10-Rras2","mm10-Prom1"


)

p1 <- DotPlot(OL, features = markers, group.by = "OL_clusters", cols = c("lightgrey", "firebrick")) +
  xlab("")+
  scale_x_discrete(labels = markers) +
  dotplot_theme()
p1
```

4.Marker features calculations in OLs

Markers of every cluster against all the remaining cells, positive only, saved
with one worksheet per cluster.

```{r}
OL_markers <- subset_markers(
  OL,
  file.path(results_dir, "cluster_markers", "Oligodendroglia_markers.xlsx")
)
```

Plotting cluster Markers

```{r,fig.height=20, fig.width=30}

markers <- c(
  # Cluster 0 - Myelinating Oligodendrocytes (MOL)
  "mm10-Ugt8a","mm10-Car2","mm10-Cdh20","mm10-Rtn4","mm10-Rnf220","mm10-Sox2ot","mm10-St18","mm10-Ptprd","mm10-Plp1","mm10-Nkain2",
  # Cluster 1 - Mature Oligodendrocytes (MOL) - homeostatic/metabolic
  "mm10-Garnl3","mm10-Mt1","mm10-Zbtb16","mm10-Polr3e","mm10-Slc6a1","mm10-Slc25a13","mm10-Otud7a","mm10-Mast3","mm10-Elmo1","mm10-Plcl1",
  # Cluster 2 - Mature Oligodendrocytes (MOL) - Apod/Ptgds-high
  "mm10-Apod","mm10-Ptgds","mm10-Phyhd1","mm10-Rsrp1","mm10-Edil3","mm10-Acaca","mm10-Neat1","mm10-St6galnac3","mm10-Anln","mm10-Hif3a",
  # Cluster 3 - Neuronal contamination (Cerebellar Granule Cells)
  "mm10-Kcnip4","mm10-Meg3","mm10-Chn2","mm10-Car10","mm10-Fgf14","mm10-Kcnd2","mm10-Cntnap2","mm10-Rbfox1","mm10-Negr1","mm10-Nrxn1",
  # Cluster 4 - OPCs / Committed OL Precursors (COP)
  "mm10-Lhfpl3","mm10-Xylt1","mm10-Tnr","mm10-Sgcd","mm10-Dscam","mm10-Adarb2","mm10-Ptprz1","mm10-Lrrtm4","mm10-Csmd1","mm10-Kcnip1",
  # Cluster 5 - OPCs (early/proliferative)
  "mm10-Nxph1","mm10-Gpc5","mm10-Sox5","mm10-Sox6","mm10-Slc35f1","mm10-Pcdh7","mm10-Ptprt","mm10-Luzp2","mm10-Nav3","mm10-Pdgfra",
  # Cluster 6 - Neuronal contamination (GABAergic/Glutamatergic)
  "mm10-Grin2b","mm10-Kcnb2","mm10-Atp1b1","mm10-Dclk1","mm10-Tenm3","mm10-Epha6","mm10-Kalrn","mm10-Gria1","mm10-Dlgap2","mm10-Etl4"
)

p1 <- DotPlot(OL, features = markers, group.by = "OL_clusters", cols = c("lightgrey", "firebrick")) +
  xlab("")+
  scale_x_discrete(labels = markers) +
  dotplot_theme()
p1
```

5.Compute the GO terms significantly enriched

```{r}
# marker genes computed above
entrez_OL <- entrez_by_cluster(OL_markers)

ego_list <- go_bp_by_cluster(entrez_OL)   # list of enriched terms per cluster
```

6.Create an Excel file to save the results, to be attached as supplementary material

```{r}
write_enrichment_workbook(
  ego_list,
  file.path(base_dir, "GO_enrichment_by_cluster_OL.xlsx"),
  label = "GO terms"
)
```

7.Plot the functional enrichment results for GO terms

```{r,fig.height=20, fig.width=40}
ego_df <- enrichment_to_df(ego_list)

p1 <- plot_enrichment_dotplot(ego_df, "GO BP enrichment per cluster")
save_plots(list(Func_enrich_GO = p1), suffix = "_OL")
p1
```

8.Functional enrichment analysis using KEGG pathways

```{r}
ekegg_list <- kegg_by_cluster(entrez_OL)

kegg_df <- enrichment_to_df(ekegg_list)
```

9.Create an Excel file to save the KEGG results

```{r}
write_enrichment_workbook(
  ekegg_list,
  file.path(base_dir, "KEGG_enrichment_by_cluster_OL.xlsx"),
  label = "KEGG pathways"
)
```

10.Plot the KEGG results

```{r,fig.height=20, fig.width=40}
p1 <- plot_enrichment_dotplot(kegg_df, "KEGG enrichment per cluster",
                              y_lab = "Pathway")
save_plots(list(Func_enrich_KEGG = p1), suffix = "_OL")
p1
```

##6.2 Microglia

1.Subset the whole dataset and recluster on the microglial cells only

```{r}
Microglia <- subset(humanized_mouse_brain, idents = c("Microglia"))

Microglia <- recluster_subset(Microglia, "Microglia", dims = 1:15, resolution = 0.1)
```

2.Plot UMAP with samples and the new microglial clusters

```{r}
p1 <- DimPlot(Microglia, reduction = "umap_Microglia",
              group.by = "Microglia_clusters") + umap_theme()
p2 <- DimPlot(Microglia, reduction = "umap_Microglia", group.by = "condition") + umap_theme()
p3 <- DimPlot(Microglia, reduction = "umap_Microglia", group.by = "sample") + umap_theme()
p4 <- DimPlot(Microglia, reduction = "umap_Microglia", group.by = "broad_clustering") + umap_theme()

p1
p2
p3
p4
```

3.Marker features calculations in the microglia

```{r}
Microglia_markers <- subset_markers(
  Microglia,
  file.path(base_dir, "Seurat_clusters_Microglia.xlsx")
)
```

Plotting cluster Markers

```{r,fig.height=20, fig.width=30}

markers <- c(
  # Cluster 0 - Neuronal contamination (Cerebellar Granule Cells)
  "mm10-Kcnip4","mm10-Kcnd2","mm10-Fgf14","mm10-Ralyl","mm10-Nrxn1","mm10-Rims1","mm10-Rbfox1","mm10-Tenm1","mm10-Car10","mm10-Negr1",
  # Cluster 1 - Homeostatic Microglia
  "mm10-Plxdc2","mm10-Tanc2","mm10-Csmd3","mm10-Nav2","mm10-Srgap2","mm10-Zfhx3","mm10-Rapgef5","mm10-Tgfbr1","mm10-Elmo1","mm10-Frmd4a",
  # Cluster 2 - Neuronal contamination (GABAergic/Glutamatergic)
  "mm10-Lrrc7","mm10-Grin2b","mm10-Kcnh7","mm10-Tenm3","mm10-Grm5","mm10-Epha6","mm10-Meis2","mm10-Myo16","mm10-Sntg1","mm10-Cpne4",
  # Cluster 3 - Oligodendrocyte contamination
  "mm10-Prr5l","mm10-Sec14l5","mm10-Slc24a2","mm10-Edil3","mm10-Arap2","mm10-Efnb3","mm10-Mog","mm10-D7Ertd443e","mm10-Mag","mm10-Ppp1r16b",
  # Cluster 4 - Activated/Phagocytic Microglia (DAM-like)
  "mm10-Arhgap24","mm10-Fkbp5","mm10-Gcnt2","mm10-C1qa","mm10-Ssh2","mm10-Cd244a","mm10-C1qb","mm10-Man2b1","mm10-Cd300lf","mm10-Npl",
  # Cluster 5 - Microglia - metabolically active / stressed
  "mm10-Cst3","mm10-Psap","mm10-mt-Co3","mm10-mt-Atp6","mm10-mt-Co2","mm10-mt-Co1","mm10-Mef2a","mm10-Snx24","mm10-Inpp5d","mm10-Rps27a"
)

p1 <- DotPlot(Microglia, features = markers, group.by = "Microglia_clusters",
              cols = c("lightgrey", "firebrick")) +
  xlab("")+
  scale_x_discrete(labels = markers) +
  dotplot_theme()
p1
```

4.Compute the GO terms significantly enriched

```{r}
# marker genes computed above
entrez_Microglia <- entrez_by_cluster(Microglia_markers)

ego_list <- go_bp_by_cluster(entrez_Microglia)
```

5.Create an Excel file to save the results, to be attached as supplementary material

```{r}
write_enrichment_workbook(
  ego_list,
  file.path(results_dir, "functional_enrichment",
            "GO_enrichment_by_cluster_Microglia.xlsx"),
  label = "GO terms"
)
```

6.Plot the functional enrichment results for GO terms

```{r,fig.height=20, fig.width=40}
ego_df <- enrichment_to_df(ego_list)

p1 <- plot_enrichment_dotplot(ego_df, "GO BP enrichment per cluster")
save_plots(list(Func_enrich_GO = p1), suffix = "_Microglia")
p1
```

##6.3 Reclustering of every broad cluster

The same subset-and-recluster sequence is applied programmatically to the broad
clusters listed in `broad_clusters`, so that any population can be inspected on
its own cells without duplicating the code above.


```{r}
# 1) cluster set
#broad_clusters <- unique(humanized_mouse_brain$broad_clustering)
broad_clusters <- "MOL"

# storage
clusters_objects <- list()

for (bc in broad_clusters) {
  bc_clean <- make.names(bc)
  obj_name <- bc_clean
  message("Processing broad cluster: ", bc, " --> ", obj_name)

  # subset and recluster on the cells of this broad cluster only. The PCA keeps
  # its historical "pca<name>_SCT" name so that objects written by earlier runs
  # stay readable.
  OL <- subset(humanized_mouse_brain, subset = broad_clustering == bc)
  OL <- recluster_subset(OL, obj_name, dims = 1:15, resolution = 0.1,
                         pca_name = paste0("pca", obj_name, "_SCT"))

  # store into list by key and save object
  clusters_objects[[obj_name]] <- OL
  saveRDS(OL, file = file.path(results_dir, "objects", paste0(obj_name, "_SCT.rds")))

  # markers
  OL_markers <- subset_markers(
    OL,
    file.path(results_dir, "cluster_markers", paste0(bc, "_markers.xlsx"))
  )

  # GO
  cat("Calculating GO for cluster: ", bc)
  entrez_list <- entrez_by_cluster(OL_markers)
  ego <- go_bp_by_cluster(entrez_list)

  # one file per cluster ...
  for (cl in names(ego)) {
    if (!is.null(ego[[cl]]) && nrow(as.data.frame(ego[[cl]])) > 0) {
      write.xlsx(as.data.frame(ego[[cl]]),
                 file.path(base_dir, paste0(obj_name, "_GO_cluster", cl, ".xlsx")),
                 sheetName = paste0("c", cl), overwrite = TRUE)
    }
  }

  # ... and one workbook with every cluster of this broad cluster
  write_enrichment_workbook(
    ego,
    file.path(results_dir, "functional_enrichment",
              paste0("GO_enrichment_by_cluster_", bc, ".xlsx")),
    label = "GO terms"
  )

  ego_df <- enrichment_to_df(ego)
  p1 <- plot_enrichment_dotplot(ego_df, "GO BP enrichment per cluster")
  save_plots(list(Func_enrich_GO = p1), suffix = paste0("_", bc),
             width = 24, height = 26)
}

```

UMAP plot of reclustering

```{r}
for (bc in broad_clusters) {
  bc_clean  <- make.names(bc)
  obj       <- clusters_objects[[bc_clean]]
  umap_name <- paste0("umap_", bc_clean)
  clu_name  <- paste0(bc_clean, "_clusters")

  p1 <- DimPlot(obj, reduction = umap_name, group.by = clu_name) + umap_theme()
  p2 <- DimPlot(obj, reduction = umap_name, group.by = "condition") + umap_theme()
  p3 <- DimPlot(obj, reduction = umap_name, group.by = "sample") + umap_theme()

  print(p1)
  print(p2)
  print(p3)

  save_plots(list(umap_clusters = p1, umap_condition = p2, umap_sample = p3),
             suffix = paste0("_", bc_clean))
}
```






Save objects

`OL` here is the object left by the loop above, i.e. the last broad cluster that
was reclustered, not the MOL + OPC object of section 6.1.

```{r}
obj_dir <- file.path(results_dir, "objects")

saveRDS(Microglia, file.path(obj_dir, "Microglia_SCT.rds"))
saveRDS(OL,        file.path(obj_dir, "Oligodendroglia_SCT.rds"))
saveRDS(humanized_mouse_brain,
        file.path(obj_dir, "humanized_mouse_brain_whole_doublets.rds"))
```




##7. Differential Gene Expression Analysis: EAE vs CFA

Pseudobulk DESeq2 is used to compare EAE (n = 3 animals) vs CFA (n = 2 animals)
within each cluster. Raw counts are aggregated per sample per cluster before running
DESeq2, which avoids pseudo-replication and handles unequal group sizes naturally.
CFA is the reference level; positive log2FC values indicate higher expression in EAE.

##7.1 Setup: load libraries and check the condition metadata

`condition` is carried on every cell from the sample sheet of section 2, so it
only needs to be checked here. DESeq2 is attached at this point rather than in
the setup chunk, so that its S4 generics do not mask the dplyr verbs used above.

```{r}
library(DESeq2)
library(ggrepel)

table(humanized_mouse_brain$sample, humanized_mouse_brain$condition)
```

##7.2 Helper functions

```{r}
# Pseudobulk DESeq2 for one cluster ----------------------------------------
# Returns a data.frame of DESeq2 results (EAE vs CFA), or NULL if the cluster
# does not meet quality filters.
run_pseudobulk_dge <- function(seurat_obj,
                                cluster_col,
                                cluster_id,
                                sample_col    = "sample",
                                condition_col = "condition",
                                min_cells     = 10,
                                min_counts    = 10) {
  idx <- seurat_obj@meta.data[[cluster_col]] == cluster_id
  if (sum(idx) < min_cells) {
    message("Skipping cluster ", cluster_id, ": only ", sum(idx), " cells")
    return(NULL)
  }

  sub_obj <- seurat_obj[, idx]
  DefaultAssay(sub_obj) <- "RNA"
  sub_obj <- JoinLayers(sub_obj)
  counts_mat <- GetAssayData(sub_obj, assay = "RNA", layer = "counts")

  # Aggregate counts per sample (pseudobulk)
  samples <- unique(sub_obj@meta.data[[sample_col]])
  pb_mat <- vapply(samples, function(s) {
    s_cells <- sub_obj@meta.data[[sample_col]] == s
    as.numeric(Matrix::rowSums(counts_mat[, s_cells, drop = FALSE]))
  }, numeric(nrow(counts_mat)))
  rownames(pb_mat) <- rownames(counts_mat)

  # Sample-level metadata; CFA is the reference level
  col_meta <- data.frame(
    sample    = samples,
    condition = factor(
      sapply(samples, function(s)
        unique(sub_obj@meta.data[[condition_col]][sub_obj@meta.data[[sample_col]] == s])),
      levels = c("CFA", "EAE")
    ),
    row.names = samples
  )

  if (sum(col_meta$condition == "EAE") == 0 || sum(col_meta$condition == "CFA") == 0) {
    message("Skipping cluster ", cluster_id, ": only one condition represented")
    return(NULL)
  }

  # Filter lowly expressed genes
  pb_mat <- pb_mat[rowSums(pb_mat) >= min_counts, , drop = FALSE]
  if (nrow(pb_mat) < 10) {
    message("Skipping cluster ", cluster_id, ": too few expressed genes after filtering")
    return(NULL)
  }

  tryCatch({
    dds <- DESeqDataSetFromMatrix(countData = round(pb_mat),
                                   colData   = col_meta,
                                   design    = ~ condition)
    dds <- DESeq(dds, quiet = TRUE)
    res <- results(dds, contrast = c("condition", "EAE", "CFA"))
    res_df <- as.data.frame(res)
    res_df$gene    <- rownames(res_df)
    res_df$cluster <- as.character(cluster_id)
    res_df[order(res_df$padj, na.last = TRUE), ]
  }, error = function(e) {
    message("DESeq2 error – cluster ", cluster_id, ": ", conditionMessage(e))
    NULL
  })
}

# Volcano plot helper -------------------------------------------------------
plot_volcano <- function(res_df, cluster_label,
                          padj_thresh = 0.05, lfc_thresh = 0.5, top_n = 15) {
  df <- res_df %>%
    filter(!is.na(padj), !is.na(log2FoldChange)) %>%
    mutate(
      sig = case_when(
        padj < padj_thresh & log2FoldChange >  lfc_thresh ~ "Up in EAE",
        padj < padj_thresh & log2FoldChange < -lfc_thresh ~ "Up in CFA",
        TRUE ~ "NS"
      )
    ) %>%
    mutate(label = ifelse(sig != "NS" & rank(padj, ties.method = "first") <= top_n,
                          gene, NA_character_))

  n_up <- sum(df$sig == "Up in EAE")
  n_dn <- sum(df$sig == "Up in CFA")

  ggplot(df, aes(x = log2FoldChange, y = -log10(padj), color = sig)) +
    geom_point(alpha = 0.6, size = 1.5) +
    ggrepel::geom_text_repel(aes(label = label), size = 3, max.overlaps = 20, na.rm = TRUE) +
    scale_color_manual(
      values = c("Up in EAE" = "firebrick", "Up in CFA" = "steelblue", "NS" = "grey70")
    ) +
    geom_vline(xintercept = c(-lfc_thresh, lfc_thresh), linetype = "dashed", colour = "grey40") +
    geom_hline(yintercept = -log10(padj_thresh),         linetype = "dashed", colour = "grey40") +
    annotate("text", x = Inf,  y = Inf, hjust = 1.1, vjust = 1.0,
             label = paste0("Up in EAE: ", n_up), color = "firebrick",  size = 4.5, fontface = "bold") +
    annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 1.0,
             label = paste0("Up in CFA: ", n_dn), color = "steelblue", size = 4.5, fontface = "bold") +
    labs(
      title = paste("EAE vs CFA –", cluster_label),
      x     = "log2 Fold Change (EAE / CFA)",
      y     = "-log10(adjusted p-value)",
      color = NULL
    ) +
    theme_classic() +
    theme(
      plot.title  = element_text(hjust = 0.5, size = 16, face = "bold"),
      axis.text   = element_text(size = 12),
      axis.title  = element_text(size = 14),
      legend.text = element_text(size = 12)
    )
}
```

##7.3 DGE within each harmony cluster (whole-brain object)

```{r}
broad_clustering_ids <- levels(humanized_mouse_brain$broad_clustering)

dge_brain_list <- lapply(broad_clustering_ids, function(cl) {
  message("Running DGE for harmony cluster: ", cl)
  run_pseudobulk_dge(
    seurat_obj  = humanized_mouse_brain,
    cluster_col = "broad_clustering",
    cluster_id  = cl
  )
})
names(dge_brain_list) <- broad_clustering_ids

# Combine into one data.frame
dge_brain_all <- do.call(rbind, Filter(Negate(is.null), dge_brain_list))

# Save to Excel (one sheet per cluster)
wb_brain <- createWorkbook()
for (cl in names(dge_brain_list)) {
  if (!is.null(dge_brain_list[[cl]])) {
    sheet_name <- substr(paste0("Cluster_", cl), 1, 31)
    addWorksheet(wb_brain, sheet_name)
    writeData(wb_brain, sheet_name, dge_brain_list[[cl]])
  }
}
saveWorkbook(wb_brain,
             "/date/gcb/gcb_MZ/Humanized_mice_OD/Results/DGE_EAE_vs_CFA_harmony_clusters.xlsx",
             overwrite = TRUE)
cat("Saved whole-brain DGE results.\n")
```

Volcano plots – whole-brain harmony clusters

```{r, fig.height=6, fig.width=8}
for (cl in names(dge_brain_list)) {
  if (!is.null(dge_brain_list[[cl]]) && sum(!is.na(dge_brain_list[[cl]]$padj)) > 0) {
    p <- plot_volcano(dge_brain_list[[cl]], cl)
    ggplot2::ggsave(
      file.path("/date/gcb/gcb_MZ/Humanized_mice_OD/Results/Plots/",
                paste0("volcano_DGE_", cl, "_brain.png")),
      p, width = 8, height =6, dpi = 300
    )
    print(p)
  }
}
```

##7.4 Summary: number of significant DEGs per cluster (padj < 0.05, |log2FC| > 0.5)

```{r}
summarise_dge <- function(dge_list) {
  do.call(rbind, lapply(names(dge_list), function(cl) {
    df <- dge_list[[cl]]
    if (is.null(df)) return(data.frame(cluster = cl, n_up_EAE = NA, n_up_CFA = NA))
    df <- df[!is.na(df$padj), ]
    data.frame(
      cluster  = cl,
      n_up_EAE = sum(df$padj < 0.05 & df$log2FoldChange >  0.5),
      n_up_CFA = sum(df$padj < 0.05 & df$log2FoldChange < -0.5)
    )
  }))
}

dge_summary <- summarise_dge(dge_brain_list)
print(dge_summary)

wb_summary <- createWorkbook()
addWorksheet(wb_summary, "DGE_summary")
writeData(wb_summary, "DGE_summary", dge_summary)
saveWorkbook(wb_summary,
             "/date/gcb/gcb_MZ/Humanized_mice_OD/Results/DGE_EAE_vs_CFA_summary.xlsx",
             overwrite = TRUE)
```

##7.5 Functional enrichment analysis of DEGs using GO terms

1.Compute the GO terms significantly enriched

```{r}
# Build named gene lists: one entry per cluster and direction (EAE-up / CFA-up)
dge_gene_lists <- list()
for (cl in names(dge_brain_list)) {
  df <- dge_brain_list[[cl]]
  if (is.null(df)) next
  df <- df[!is.na(df$padj), ]
  up_eae <- df$gene[df$padj < 0.05 & df$log2FoldChange >  0.5]
  up_cfa <- df$gene[df$padj < 0.05 & df$log2FoldChange < -0.5]
  if (length(up_eae) > 0) dge_gene_lists[[paste0("Cluster_", cl, "_EAE_up")]] <- up_eae
  if (length(up_cfa) > 0) dge_gene_lists[[paste0("Cluster_", cl, "_CFA_up")]] <- up_cfa
}

# convert symbols to ENTREZID
all_dge_genes_clean <- sub("^mm10-", "", unique(unlist(dge_gene_lists)))
gene_map_dge <- bitr(all_dge_genes_clean, fromType = "SYMBOL", toType = "ENTREZID", OrgDb = org.Mm.eg.db)
sym2ent_dge <- setNames(gene_map_dge$ENTREZID, gene_map_dge$SYMBOL)

entrez_dge_lists <- lapply(dge_gene_lists, function(x) na.omit(unname(sym2ent_dge[sub("^mm10-", "", x)])))

ego_dge_list <- lapply(names(entrez_dge_lists), function(cl){
  enrichGO(gene          = entrez_dge_lists[[cl]],
           OrgDb         = org.Mm.eg.db,
           keyType       = "ENTREZID",
           ont           = "BP",
           pAdjustMethod = "BH",
           pvalueCutoff  = 0.05,
           qvalueCutoff  = 0.2,
           readable      = TRUE)
}) #list of enriched terms per cluster and direction
names(ego_dge_list) <- names(entrez_dge_lists)
```

2.Create an Excel file to save the results, to be attached as supplementary material

```{r}

# Create a new workbook
wb <- createWorkbook()

# Loop through each cluster and add enrichment results as a sheet
for (cluster_name in names(ego_dge_list)) {

  # Extract results from enrichResult object
  if (!is.null(ego_dge_list[[cluster_name]]) && nrow(ego_dge_list[[cluster_name]]@result) > 0) {

    # Get the results as a data frame
    go_results <- as.data.frame(ego_dge_list[[cluster_name]])

    # Create sheet name (Excel has 31 char limit for sheet names)
    # Preserve the direction suffix (_EAE_up / _CFA_up) when truncating
    suffix <- ifelse(grepl("_EAE_up$", cluster_name), "_EAE_up", "_CFA_up")
    prefix <- sub("_(EAE|CFA)_up$", "", cluster_name)
    sheet_name <- paste0(substr(prefix, 1, 31 - nchar(suffix)), suffix)

    # Add worksheet
    addWorksheet(wb, sheet_name)

    # Write data
    writeData(wb, sheet_name, go_results)

    cat(sprintf("Added sheet for cluster: %s (%d GO terms)\n",
                cluster_name, nrow(go_results)))

  } else {
    cat(sprintf("No significant GO terms for cluster: %s\n", cluster_name))
  }
}

# Save the workbook
saveWorkbook(wb,
             "/date/gcb/gcb_MZ/Humanized_mice_OD/Results/DGE_EAE_vs_CFA_GO_enrichment.xlsx",
             overwrite = TRUE)

cat("\n✓ GO enrichment results saved to Excel file!\n")

```

3.Plot the functional enrichment results for GO terms – one plot per cluster

```{r,fig.height=20, fig.width=20}
# Collect GO BP results into one data.frame

ego_dge_df <- map2_df(names(ego_dge_list), ego_dge_list, ~{
  if (is.null(.y) || nrow(as.data.frame(.y)) == 0) return(NULL)
  df <- as.data.frame(.y)
  df$cluster <- .x
  df$mLog10Padj <- -log10(df$p.adjust)
  df
})

# One plot per harmony cluster (combining EAE_up and CFA_up directions)
cluster_ids <- unique(sub("_(EAE|CFA)_up$", "", ego_dge_df$cluster))

for (cl in cluster_ids) {

  plot_df <- ego_dge_df %>%
    filter(grepl(paste0("^", cl, "_(EAE|CFA)_up$"), cluster)) %>%
    group_by(cluster) %>%
    slice_max(order_by = mLog10Padj, n = 10, with_ties = FALSE) %>%
    ungroup() %>%
    mutate(Description = fct_reorder(Description, mLog10Padj))

  if (nrow(plot_df) == 0) next

  p1 <- ggplot(plot_df, aes(x = cluster, y = Description, size = Count, color = mLog10Padj)) +
    geom_point() +
    scale_color_gradient(low = "gold", high = "red") +
    labs(title = paste("GO BP enrichment –", cl), x = "Direction", y = "Term", color = "-log10(FDR)") +
    theme_bw(base_size = 12) + theme_classic()+
    theme(
      axis.text.x = element_text(hjust = 1, vjust = 0.5, family = "Arial", size = 24, angle = 45),
      axis.text.y = element_text(family = "Arial", size = 24),
      axis.title.y = element_text(family = "Arial", size =28, face = "bold"),
      axis.title.x = element_text(family = "Arial", size =28, face = "bold"),
      legend.text = element_text(size = 24, family = "Arial"),
      legend.title = element_text(size = 32, family = "Arial", face = "bold")
    )
  ggplot2::ggsave(file.path("/date/gcb/gcb_MZ/Humanized_mice_OD/Results/Plots/",
                             paste0("Func_enrich_GO_DGE_", cl, ".png")),
                  p1, width = 14, height = 16, dpi = 300)
  print(p1)
}

```

##7.6 Functional enrichment analysis of DEGs using KEGG pathways

1.Kegg pathways analysis

```{r}
ekegg_dge_list <- lapply(names(entrez_dge_lists), function(cl){
  enrichKEGG(gene          = entrez_dge_lists[[cl]],
             organism      = "mmu",
             pvalueCutoff  = 0.05)
})
names(ekegg_dge_list) <- names(entrez_dge_lists)


kegg_dge_df <- map2_df(names(ekegg_dge_list), ekegg_dge_list, ~{
  if (is.null(.y) || nrow(as.data.frame(.y)) == 0) return(NULL)
  df <- as.data.frame(.y)
  df$cluster <- .x
  df$mLog10Padj <- -log10(df$p.adjust)
  df
})
```

2.Create an Excel file to save the results, to be attached as supplementary material

```{r}

# Save to Excel - one sheet per cluster and direction
wb_kegg <- createWorkbook()

for (cluster_name in names(ekegg_dge_list)) {

  if (!is.null(ekegg_dge_list[[cluster_name]]) && nrow(ekegg_dge_list[[cluster_name]]@result) > 0) {

    # Get results as data frame
    kegg_results <- as.data.frame(ekegg_dge_list[[cluster_name]])

    # Add mLog10Padj column
    kegg_results$mLog10Padj <- -log10(kegg_results$p.adjust)

    # Create sheet name (Excel has 31 char limit for sheet names)
    # Preserve the direction suffix (_EAE_up / _CFA_up) when truncating
    suffix <- ifelse(grepl("_EAE_up$", cluster_name), "_EAE_up", "_CFA_up")
    prefix <- sub("_(EAE|CFA)_up$", "", cluster_name)
    sheet_name <- paste0(substr(prefix, 1, 31 - nchar(suffix)), suffix)

    # Add worksheet and write data
    addWorksheet(wb_kegg, sheet_name)
    writeData(wb_kegg, sheet_name, kegg_results)

    cat(sprintf("Added sheet for cluster: %s (%d KEGG pathways)\n",
                cluster_name, nrow(kegg_results)))

  } else {
    cat(sprintf("No significant KEGG pathways for cluster: %s\n", cluster_name))
  }
}

# Save workbook with one sheet per cluster
saveWorkbook(wb_kegg,
             "/date/gcb/gcb_MZ/Humanized_mice_OD/Results/DGE_EAE_vs_CFA_KEGG_enrichment.xlsx",
             overwrite = TRUE)

cat("\n✓ KEGG enrichment results saved (one sheet per cluster)!\n")

```

3.Plot the results – one plot per cluster

```{r,fig.height=20, fig.width=20}
cluster_ids_kegg <- unique(sub("_(EAE|CFA)_up$", "", kegg_dge_df$cluster))

for (cl in cluster_ids_kegg) {

  plot_df_kegg <- kegg_dge_df %>%
    filter(grepl(paste0("^", cl, "_(EAE|CFA)_up$"), cluster)) %>%
    group_by(cluster) %>%
    slice_max(order_by = mLog10Padj, n = 10, with_ties = FALSE) %>%
    ungroup() %>%
    mutate(Description = fct_reorder(Description, mLog10Padj))

  if (nrow(plot_df_kegg) == 0) next

  p1 <- ggplot(plot_df_kegg, aes(x = cluster, y = Description, size = Count, color = mLog10Padj)) +
    geom_point() +
    scale_color_gradient(low = "gold", high = "red") +
    labs(title = paste("KEGG enrichment –", cl), x = "Direction", y = "Pathway", color = "-log10(FDR)") +
    theme_bw(base_size = 12)+theme_classic()+
    theme(
      axis.text.x = element_text(hjust = 1, vjust = 0.5, family = "Arial", size = 24, angle = 45),
      axis.text.y = element_text(family = "Arial", size = 24),
      axis.title.y = element_text(family = "Arial", size =28, face = "bold"),
      axis.title.x = element_text(family = "Arial", size =28, face = "bold"),
      legend.text = element_text(size = 24, family = "Arial"),
      legend.title = element_text(size = 32, family = "Arial", face = "bold")
    )
  ggplot2::ggsave(file.path("/date/gcb/gcb_MZ/Humanized_mice_OD/Results/Plots/",
                             paste0("Func_enrich_KEGG_DGE_", cl, ".png")),
                  p1, width = 14, height = 16, dpi = 300)
  print(p1)
}

```




##8. Label Transfer for MOL subset (mouse cells only)

Label transfer is performed using three independent mouse oligodendrocyte reference
datasets. Only mouse cells from the MOL subset are used as query so that gene names
(mm10- prefix space) match the reference objects.

```{r}
MOL <- readRDS("/date/gcb/gcb_MZ/Humanized_mice_OD/Results/objects/MOL_SCT.rds")
DefaultAssay(OL) <- "RNA"
OL[["RNA"]] <- JoinLayers(OL[["RNA"]])


# Strip mm10- prefix so gene names match standard mouse references (Pandey, Zuchero)
counts_lt <- GetAssayData(MOL, assay = "RNA", layer = "counts")
rownames(counts_lt) <- gsub("^mm10-", "", rownames(counts_lt))
MOL_lt <- CreateSeuratObject(counts = counts_lt, meta.data = MOL@meta.data)
MOL_lt <- NormalizeData(MOL_lt)
rm(counts_lt)
```

##8.1 Label transfer – Pandey et al. panDAO (Cell Reports, 2022)

1. Perform label transfer

```{r}
# Load and capture the original object name
loaded_name <- load("/date/gcb/gcb_MZ/MOL_FGF_10x/oligos_final_pub.RObj") #Pandey Dataset can be donwloaded from Zenodo

# Assign it to DA_oligo using the captured name
DA_oligo <- get(loaded_name[1])

DA_oligo_v5 <- UpdateSeuratObject(object =DA_oligo)
DA_oligo_v5@version

DefaultAssay(DA_oligo_v5) <- "RNA"
DefaultAssay(OL)          <- "RNA"

# CCA needs scale.data on both
DA_oligo_v5 <- NormalizeData(DA_oligo_v5) |>
               FindVariableFeatures() |>
               ScaleData()

# 1. keep only the mouse-tagged genes
mm_mask  <- grepl("^mm10-", rownames(OL))
OL_mouse <- OL[mm_mask, ]

# 2. pull raw counts and rename rows
cts <- GetAssayData(OL_mouse, assay = "RNA", layer = "counts")
rownames(cts) <- sub("^mm10-", "", rownames(cts))

# 3. guard against duplicate names produced by stripping (rare but possible)
dup <- duplicated(rownames(cts))
if (any(dup)) {
    message(sum(dup), " duplicate gene symbols after prefix stripping; collapsing by sum")
    cts <- rowsum(as.matrix(cts), group = rownames(cts))
}

# 4. rebuild the Seurat object, carrying over cell metadata
OL_clean <- CreateSeuratObject(
    counts    = cts,
    meta.data = OL_mouse@meta.data
)
DefaultAssay(OL_clean) <- "RNA"
OL_clean <- NormalizeData(OL_clean) |>
            FindVariableFeatures() |>
            ScaleData()
feats <- intersect(VariableFeatures(OL_clean), rownames(DA_oligo_v5))

length(feats) 

# 5. (optional) carry over any reductions / cell embeddings you want to keep
 OL_clean[["umap"]] <- OL_mouse[["umap_OL"]]
  
OL_clean <- transfer_labels(
    reference = DA_oligo_v5,
    query     = OL_clean,
    features  = feats,
    refdata   = "finalInterps",
    suffix    = "DA_oligo"
)
```

2. UMAP colored by predicted label

```{r, fig.height=8, fig.width=12}
colors <- c(
  "#0072B2", "#E69F00", "green", "#D55E00", "#CC79A7",
  "red", "#F0E442", "#000000", "#332288", "#88CCEE",
  "#44AA99", "#117733", "#999933", "#DDCC77", "#661100",
  "#AA4499", "#882255"
)

p1 <- DimPlot(OL_clean, reduction = "umap", group.by = "predicted.id_DA_oligo",
              cols = colors) +
  ggtitle("Pandey panDAO – predicted labels") +
  umap_theme(legend_title_face = "plain")

my_colors <- c(
  "#1F77B4",
  "#FF7F0E",
  "#2CA02C",
  "#D62728",
  "#9467BD",
  "#8C564B",
  "#17BECF"
)

p2 <- DimPlot(OL_clean, reduction = "umap", group.by = "OL_clusters",
              cols = my_colors) + umap_theme()   # UMAP with new clustering

p1
p2
```

3. Violin plots of similarity scores per OL cluster

```{r, fig.height=12, fig.width=24}
prediction_score_violins(OL_clean, "OL_clusters")
```

##8.2 Label transfer – Zuchero EAE dataset (Nature Medicine, 2018)

1. Perform label transfer

```{r}
EAE_Oligo <- readRDS("/date/gcb/gcb_MZ/MOL_FGF_10x/EAE_SS2_NatMed2018_seuratv5.rds")

DefaultAssay(EAE_Oligo) <- "RNA"


OL_clean <- transfer_labels(
    reference = EAE_Oligo,
    query     = OL_clean,
    features  = feats,
    refdata   = "Renamed_clusternames",
    suffix    = "EAE_oligo"
)

rm(EAE_Oligo)
gc()
```

2. UMAP colored by predicted label

```{r, fig.height=8, fig.width=14}
colors <- c(
  "#A6CEE3", "#1F78B4", "#B2DF8A", "#33A02C", "black", "#E31A1C",
  "#FDBF6F", "#FF7F00", "#CAB2D6", "#6A3D9A", "#FFFF99", "#B15928",
  "#8DD3C7", "#FFFFB3", "#BEBADA", "#FB8072", "#80B1D3", "gold",
  "#B3DE69", "#FCCDE5", "#D9D9D9", "#BC80BD", "#CCEBC5"
)

p1 <- DimPlot(OL_clean, reduction = "umap", group.by = "predicted.id_EAE_oligo",
              cols = colors) +
  ggtitle("Humanized mice vs Spinal cord EAE – predicted labels") +
  umap_theme(legend_title_face = "plain")

my_colors <- c(
  "#1F77B4",
  "#FF7F0E",
  "#2CA02C",
  "#D62728",
  "#9467BD",
  "#8C564B",
  "#17BECF"
)


p2 <- DimPlot(OL_clean, reduction = "umap", group.by = "OL_clusters",
              cols = my_colors) + umap_theme()   # UMAP with new clustering


p1
p2
```

3. Violin plots of similarity scores per OL cluster

```{r, fig.height=12, fig.width=24}
prediction_score_violins(OL_clean, "OL_clusters")
```


##8.3 Label transfer – Absinta 

1. Perform label transfer

```{r}
Human_MS <- readRDS("/date/gcb/gcb_MZ/MOL_FGF_10x/ms_lesions_MM10orthologs_snRNAseq_seuratV5_WMEA_updated_WMOLGdown3K.rds")

DefaultAssay(Human_MS) <- "RNA"


OL_clean <- transfer_labels(
    reference = Human_MS,
    query     = OL_clean,
    features  = feats,
    refdata   = "type_fine",
    suffix    = "Human_MS"
)
```

2. UMAP colored by predicted label

```{r, fig.height=8, fig.width=14}
colors <- c(
  "#A6CEE3", "#1F78B4", "#B2DF8A", "#33A02C", "black", "#E31A1C",
  "#FDBF6F", "#FF7F00", "#CAB2D6", "#6A3D9A", "#FFFF99", "#B15928",
  "#8DD3C7", "#FFFFB3", "#BEBADA", "#FB8072", "#80B1D3", "gold",
  "#B3DE69", "#FCCDE5", "#D9D9D9", "#BC80BD", "#CCEBC5"
)

p1 <- DimPlot(OL_clean, reduction = "umap", group.by = "predicted.id_Human_MS",
              cols = colors) +
  ggtitle("Humanized mice vs Human MS – predicted labels") +
  umap_theme(legend_title_face = "plain")

my_colors <- c(
  "#1F77B4",
  "#FF7F0E",
  "#2CA02C",
  "#D62728",
  "#9467BD",
  "#8C564B",
  "#17BECF"
)


p2 <- DimPlot(OL_clean, reduction = "umap", group.by = "OL_clusters",
              cols = my_colors) + umap_theme()   # UMAP with new clustering


p1
p2
```


3. Violin plots of similarity scores per OL cluster

```{r, fig.height=12, fig.width=24}
prediction_score_violins(OL_clean, "OL_clusters")
```



##9. Gene module scores from Falcao, Pandey, McNair, Absinta

Published signatures are scored on the oligodendroglial object with
`AddModuleScore` and compared between conditions within each cluster. The
condition is read from the `condition` metadata column, not from `orig.ident`,
which carries the full sample name.

first we load the helper function

```{r}

# --- Prefix-aware feature matching ----------------------------------------
# Genes in the Seurat object are prefixed with "mm10-" (mouse) or "GRCh38-" (human).
# Given a vector of bare symbols, return the matching features in the object.
match_features <- function(seurat_obj, genes, species = "mouse", assay = "RNA") {
  feats <- rownames(seurat_obj[[assay]])
  out <- c()
  if (species %in% c("mouse", "both")) {
    # Capitalise first letter only -> mouse-style symbol
    g_mm <- paste0(toupper(substr(genes, 1, 1)),
                   tolower(substr(genes, 2, nchar(genes))))
    out <- c(out, intersect(paste0("mm10-", g_mm), feats))
  }
  if (species %in% c("human", "both")) {
    out <- c(out, intersect(paste0("GRCh38-", toupper(genes)), feats))
  }
  unique(out)
}

# --- AddModuleScore wrapper -----------------------------------------------
# Scores a signature on the object and stores it as a clean metadata column.
score_signature <- function(seurat_obj, genes, name,
                            species = "mouse", assay = "RNA", ctrl = 50) {
  feats <- match_features(seurat_obj, genes, species, assay)
  message(sprintf("'%s' (%s): %d / %d genes matched",
                  name, species, length(feats), length(genes)))
  if (length(feats) < 3) {
    warning(sprintf("Skipping '%s' - <3 genes matched.", name))
    seurat_obj[[name]] <- NA_real_
    return(seurat_obj)
  }
  seurat_obj <- AddModuleScore(seurat_obj, features = list(feats),
                               name = name, assay = assay, ctrl = ctrl)
  # AddModuleScore appends "1" to the name - rename to a clean column
  seurat_obj@meta.data[[name]] <- seurat_obj@meta.data[[paste0(name, "1")]]
  seurat_obj@meta.data[[paste0(name, "1")]] <- NULL
  seurat_obj
}

# =============================================================================
# 2. FILE-FORMAT-SPECIFIC LOADERS
# =============================================================================

# Loader 1: LONG format Seurat FindAllMarkers output (Absinta 2021)
# Columns: p_val, avg_logFC, pct.1, pct.2, p_val_adj, cluster, gene
# Row 1 is a title row that must be skipped (skip_rows = 1).
load_markers_long <- function(path, sheet, cluster_map = NULL,
                              padj_max = 0.05, logfc_min = 0.25,
                              top_n = NULL, skip_rows = 1) {
  df <- read_excel(path, sheet = sheet, skip = skip_rows)
  colnames(df) <- tolower(colnames(df))
  fc_col <- intersect(c("avg_logfc", "avg_log2fc"), colnames(df))[1]
  stopifnot(!is.na(fc_col),
            "gene" %in% colnames(df),
            "cluster" %in% colnames(df))
  if (!"p_val_adj" %in% colnames(df)) df$p_val_adj <- NA_real_

  # Rename via base R to avoid !! on character strings
  colnames(df)[colnames(df) == fc_col] <- "logfc"

  df <- df %>%
    filter(!is.na(gene), !is.na(cluster)) %>%
    filter(is.na(p_val_adj) | p_val_adj <= padj_max) %>%
    filter(logfc >= logfc_min)

  df$cluster <- as.character(df$cluster)
  if (!is.null(cluster_map)) {
    df <- df %>% filter(cluster %in% names(cluster_map))
    df$cluster <- unname(cluster_map[df$cluster])
  }
  if (!is.null(top_n)) {
    df <- df %>%
      group_by(cluster) %>%
      slice_max(logfc, n = top_n, with_ties = FALSE) %>%
      ungroup()
  }
  split(df$gene, df$cluster) %>%
    lapply(function(x) unique(na.omit(as.character(x))))
}

# Loader 2: WIDE format with one FC column per cluster (Falcão 2018 DE sheets)
load_markers_wide <- function(path, sheet, fc_min = 5, top_n = 100,
                              cluster_map = NULL, skip_rows = 0) {
  df <- read_excel(path, sheet = sheet, skip = skip_rows)
  colnames(df)[1] <- "gene"
  df <- df %>% filter(!is.na(gene))
  num_cols <- names(df)[-1][sapply(df[-1], is.numeric)]

  out <- lapply(num_cols, function(cl) {
    df %>% select(gene, value = all_of(cl)) %>%
      filter(!is.na(value), value >= fc_min) %>%
      arrange(desc(value)) %>%
      slice_head(n = top_n) %>%
      pull(gene) %>% unique()
  })
  names(out) <- num_cols

  if (!is.null(cluster_map)) {
    keep <- intersect(names(cluster_map), names(out))
    out <- out[keep]
    names(out) <- cluster_map[keep]
  }
  out[sapply(out, length) > 0]
}

# Loader 3: NNMF / module-loading matrices (Falcão 2018 GenemodulesNNMF)
load_modules <- function(path, sheet, modules = NULL, top_n = 100,
                         min_loading = NULL, skip_rows = 0) {
  df <- read_excel(path, sheet = sheet, skip = skip_rows)
  colnames(df)[1] <- "gene"
  df <- df %>% filter(!is.na(gene))
  mod_cols <- names(df)[-1][sapply(df[-1], is.numeric)]
  if (!is.null(modules)) mod_cols <- intersect(as.character(modules), mod_cols)

  out <- lapply(mod_cols, function(m) {
    sub <- df %>% select(gene, value = all_of(m)) %>%
      filter(!is.na(value)) %>% arrange(desc(value))
    if (!is.null(min_loading)) sub <- sub %>% filter(value >= min_loading)
    sub %>% slice_head(n = top_n) %>% pull(gene) %>% unique()
  })
  names(out) <- paste0("module_", mod_cols)
  out[sapply(out, length) > 0]
}

# Loader 4: PANDEY format - single wide sheet with PAIRED LogFC/FDR columns.
# signature_defs is a named list of list(logfc=..., fdr=..., direction=...).
load_markers_paired <- function(path, sheet, signature_defs,
                                gene_col = "symbol",
                                logfc_min = 1, fdr_max = 0.05,
                                top_n = 100, skip_rows = 1) {
  df <- read_excel(path, sheet = sheet, skip = skip_rows)
  stopifnot(gene_col %in% colnames(df))
  out <- lapply(signature_defs, function(s) {
    direction <- if (is.null(s$direction)) "up" else s$direction
    sub <- df %>% select(gene = all_of(gene_col),
                         lfc  = all_of(s$logfc),
                         fdr  = all_of(s$fdr)) %>%
      filter(!is.na(gene), !is.na(lfc), !is.na(fdr), fdr <= fdr_max)
    sub <- if (direction == "up")
      sub %>% filter(lfc >=  logfc_min) %>% arrange(desc(lfc))
    else
      sub %>% filter(lfc <= -logfc_min) %>% arrange(lfc)
    sub %>% slice_head(n = top_n) %>% pull(gene) %>% unique()
  })
  out[sapply(out, length) > 0]
}

# Loader 5: McNAIR format - one sheet per cluster, sheet name = cluster name.
# Per-sheet columns: symbol, is_marker, log2fc, FDR, ...
load_markers_per_sheet <- function(path, sheets = NULL,
                                   gene_col = "symbol",
                                   marker_col = "is_marker",
                                   logfc_col = "log2fc",
                                   fdr_col = "FDR",
                                   logfc_min = 1, fdr_max = 0.05,
                                   require_marker = TRUE,
                                   top_n = 100) {
  all_sheets <- excel_sheets(path)
  if (is.null(sheets)) sheets <- all_sheets
  sheets <- intersect(sheets, all_sheets)

  out <- lapply(sheets, function(s) {
    df <- suppressMessages(read_excel(path, sheet = s))
    needed <- c(gene_col, logfc_col, fdr_col)
    if (!all(needed %in% colnames(df))) return(NULL)

    if (require_marker && marker_col %in% colnames(df)) {
      df <- df[df[[marker_col]] %in% c(TRUE, "TRUE", "True", 1), , drop = FALSE]
    }

    df <- df %>% as_tibble() %>%
      filter(
        !is.na(.data[[gene_col]]),
        !is.na(.data[[logfc_col]]),
        !is.na(.data[[fdr_col]]),
        .data[[fdr_col]]  <= fdr_max,
        .data[[logfc_col]] >= logfc_min
      ) %>%
      arrange(desc(.data[[logfc_col]])) %>%
      slice_head(n = top_n)

    unique(df[[gene_col]])
  })

  names(out) <- sheets
  out[!sapply(out, is.null) & sapply(out, length) > 0]
}
```



Load all signature files

```{r}

# =============================================================================
# 3. LOAD ALL FOUR PAPERS - OL LINEAGE ONLY
# =============================================================================
sig_dir <- "/date/gcb/gcb_MZ/Humanized_mice_OD/signatures/"   # files sit in the project root

# ---- PANDEY 2022 (all signatures here are MOL-related) -------------------
pandey_defs <- list(
  Pandey_DA2_mol56  = list(logfc = "LogFC_mol56DA2_vs_mol56",
                           fdr   = "FDR_mol56DA2_vs_mol56",  direction = "up"),
  Pandey_DA1_mol56  = list(logfc = "LogFC_mol56DA1_vs_mol56",
                           fdr   = "FDR_mol56DA1_vs_mol56",  direction = "up"),
  Pandey_IFN_OL     = list(logfc = "LogFC_mol56IFN_m56",
                           fdr   = "FDR_mol56IFN_m56",       direction = "up"),
  Pandey_DA1_mol2   = list(logfc = "LogFC_mol2da1_vs_mol2",
                           fdr   = "FDR_mol2da1_vs_mol2",    direction = "up"),
  Pandey_DAO_coarse = list(logfc = "LogFC", fdr = "FDR",     direction = "up")
)
pandey <- load_markers_paired(
  file.path(sig_dir, "pandey_2022.xlsx"),
  sheet = "TableS2",
  signature_defs = pandey_defs,
  gene_col = "symbol",
  logfc_min = 1, fdr_max = 0.05, top_n = 100,
  skip_rows = 1
)

# ---- McNAIR 2024/25 (OPC + Oligo sheets only) ----------------------------
mcnair_ol_sheets <- c(
  "OPC_1", "OPC_2", "COP",
  "Oligo_A", "Oligo_B", "Oligo_C", "Oligo_D",
  "Oligo_E", "Oligo_F", "Oligo_G"
)
mcnair_ol <- load_markers_per_sheet(
  file.path(sig_dir, "mcnair_2024.xlsx"),
  sheets = mcnair_ol_sheets,
  logfc_min = 1, fdr_max = 0.05, top_n = 100
)
names(mcnair_ol) <- paste0("McNair_", names(mcnair_ol))

# ---- ABSINTA 2021 (OL = S5, OPC = S6) ------------------------------------
# These cluster_maps are PLACEHOLDERS. Check Fig. 3-4 of Absinta et al. 2021
# and rename to the labels the authors actually use, especially the
# disease-associated OL/OPC clusters.
absinta_ol_map <- c(
  "0" = "Oligo_A",
  "1" = "Oligo_B",
  "2" = "Oligo_C",
  "3" = "Oligo_Stress",   # HSPH1, HSPD1, FOS — clear UPR/heat-shock signature
  "4" = "Oligo_D",
  "5" = "Oligo_Imm",      # OPCML, DSCAM, NRXN1 — immature/remyelinating-like
  "6" = "Oligo_Mito",     # MT-ND3, MT-CO2, MT-CO3 — mitochondria-dominated
  "7" = "DA_Oligo"        # LINGO1, HSP90AA1, NEAT1 — LINGO1 marks failed remyelination
)

absinta_opc_map <- c(
  "0" = "OPC_A",
  "1" = "Stressed_OPC",   # HSP90AA1, HMGCS1 — stress + cholesterol biosynthesis (matches paper text)
  "2" = "OPC_B",          # XIST-dominant — likely female-enriched subcluster
  "3" = "OPC_C",
  "4" = "OPC_D",
  "5" = "OPC_Prolif",     # RPL32, RPS24, EEF1A1 — ribosomal/actively translating
  "6" = "PreOL",          # ST18, MAP7 — premyelinating (matches paper's ~1% premyelinating OPC)
  "7" = "OPC_E",          # BCAS1 — possibly early newly-formed OL
  "8" = "Immune_OPC"      # FCGR2A, ACSL1, APBB1IP — immunological OPC (paper: 23% at lesion edge)
)
absinta_ol <- load_markers_long(
  file.path(sig_dir, "absinta_2021.xlsx"),   # corrected filename
  sheet = "Table S5_OL", cluster_map = absinta_ol_map,
  padj_max = 0.05, logfc_min = 0.25, top_n = 100
)

absinta_opc <- load_markers_long(
  file.path(sig_dir, "absinta_2021.xlsx"),
  sheet = "Table S6_OPC", cluster_map = absinta_opc_map,
  padj_max = 0.05, logfc_min = 0.25, top_n = 100
)
names(absinta_ol)  <- paste0("Absinta_", names(absinta_ol))
names(absinta_opc) <- paste0("Absinta_", names(absinta_opc))

# ---- FALCÃO 2018 (only runs if the file has been uploaded) ---------------
falcao_file <- file.path(sig_dir, "falcao_2018.xlsx")
falcao_ol <- list(); falcao_modules <- list()
if (file.exists(falcao_file)) {
  falcao_ol_map <- c(
    "diffclustersOPC1"          = "Falcao_OPC1",
    "diffclustersOPC2"          = "Falcao_OPC2",
    "diffclustersOPC3"          = "Falcao_OPC3",
    "diffclustersOPC_Cycling"   = "Falcao_OPC_Cycling",
    "diffclustersCOP"           = "Falcao_COP",
    "diffclustersNFOL"          = "Falcao_NFOL",
    "diffclustersMOL1_2_EAE"    = "Falcao_MOL1_2_EAE",
    "diffclustersMOL2_Cntrl_A"  = "Falcao_MOL2_Cntrl",
    "diffclustersMOL3_EAE"      = "Falcao_MOL3_EAE",
    "diffclustersMOL5_6_Cntrl_A"= "Falcao_MOL5_6_Cntrl_A",
    "diffclustersMOL5_6_Cntrl_B"= "Falcao_MOL5_6_Cntrl_B",
    "diffclustersMOL5_6_EAE_A"  = "Falcao_MOL5_6_EAE_A",
    "diffclustersMOL5_6_EAE_B"  = "Falcao_MOL5_6_EAE_B",
    "diffclustersMOL5_6_EAE_C"  = "Falcao_MOL5_6_EAE_C"
  )
  falcao_ol <- load_markers_wide(
    falcao_file, sheet = "DifferentialExpressionOLs",
    fc_min = 5, top_n = 100, cluster_map = falcao_ol_map
  )
  falcao_modules <- load_modules(
    falcao_file, sheet = "GenemodulesNNMF",
    modules = c(1, 2, 3, 4, 6), top_n = 100
  )
  names(falcao_modules)[names(falcao_modules) == "module_1"] <- "Falcao_immuneOL_NNMF1"
} else {
  message("Falcão file not found at ", falcao_file,
          " - skipping Falcão signatures.")
}

# ---- Combine + record species per signature ------------------------------
signatures <- c(pandey, mcnair_ol,
                absinta_ol, absinta_opc,
                falcao_ol, falcao_modules)

sig_species <- c(
  setNames(rep("mouse", length(pandey)),         names(pandey)),
  setNames(rep("human", length(mcnair_ol)),      names(mcnair_ol)),
  setNames(rep("human", length(absinta_ol)),     names(absinta_ol)),
  setNames(rep("human", length(absinta_opc)),    names(absinta_opc)),
  setNames(rep("mouse", length(falcao_ol)),      names(falcao_ol)),
  setNames(rep("mouse", length(falcao_modules)), names(falcao_modules))
)

tibble(signature = names(signatures),
       n_genes   = lengths(signatures),
       species   = sig_species[names(signatures)]) %>% print(n = 100)
```

```{r score_signatures}
# =============================================================================
# 4. SCORE EVERY SIGNATURE ON THE OL OBJECT
# =============================================================================
DefaultAssay(OL) <- "RNA"

for (nm in names(signatures)) {
  OL <- score_signature(OL,
                        genes   = signatures[[nm]],
                        name    = nm,
                        species = sig_species[[nm]])
}

ol_features <- intersect(names(signatures), colnames(OL@meta.data))
ol_features
```

```{r plot_OL_signatures, fig.width=16, fig.height=14}
# =============================================================================
# 5. UMAP FEATUREPLOTS + SPLIT VIOLIN PLOTS
# =============================================================================
p <- FeaturePlot(OL,
            features = ol_features,
            reduction = "umap_OL",
            order = TRUE,
            ncol = 4,
            min.cutoff = "q5",
            max.cutoff = "q95") &
  scale_color_gradientn(colors = c("lightgrey", "#fdae61", "#d7191c"))
print(p)

p1<-VlnPlot(OL,
        features = ol_features,
        group.by = "OL_clusters",
        split.by = "condition",
        pt.size = 0,
        ncol = 2) &
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
print(p1)
```


```{r stats_EAE_vs_CFA}
# =============================================================================
# 7. STATISTICAL COMPARISON EAE vs CFA, PER CLUSTER, PER SIGNATURE
# =============================================================================
compare_scores <- function(seurat_obj, score_cols, cluster_col,
                           condition_col = "condition") {
  form <- as.formula(paste("score ~", condition_col))

  seurat_obj@meta.data %>%
    select(all_of(c(cluster_col, condition_col, score_cols))) %>%
    pivot_longer(all_of(score_cols),
                 names_to = "signature", values_to = "score") %>%
    drop_na(score) %>%
    filter(is.finite(score)) %>%                              # catches Inf / NaN
    group_by(signature, .data[[cluster_col]], .data[[condition_col]]) %>%
    filter(n() >= 2) %>%                                      # >= 2 obs per condition
    group_by(signature, .data[[cluster_col]]) %>%
    filter(n_distinct(.data[[condition_col]]) >= 2) %>%       # both conditions present
    group_modify(~ tryCatch(                                  # skip groups that still fail
      rstatix::wilcox_test(.x, form),
      error = function(e) tibble()
    )) %>%
    adjust_pvalue(method = "BH") %>%
    add_significance("p.adj")
}
ol_stats <- compare_scores(OL, ol_features, "OL_clusters")
ol_stats
```

```{r boxplots_with_pvals, fig.width=26, fig.height=40}
# =============================================================================
# 8. BOXPLOTS WITH WILCOXON SIGNIFICANCE BRACKETS
# =============================================================================
plot_score_box <- function(seurat_obj, score, cluster_col,
                           condition_col = "condition") {
  df <- FetchData(seurat_obj, vars = c(score, cluster_col, condition_col))
  colnames(df)[colnames(df) == score]         <- "score"
  colnames(df)[colnames(df) == cluster_col]   <- "cluster"
  colnames(df)[colnames(df) == condition_col] <- "condition"
  df <- drop_na(df, score)

  ggplot(df, aes(x = cluster, y = score, fill = condition)) +
    geom_boxplot(outlier.size = 0.2, alpha = 0.85) +
    stat_compare_means(aes(group = condition),
                       method = "wilcox.test",
                       label = "p.signif",
                       hide.ns = TRUE) +
    scale_fill_manual(values = c("CFA" = "#4575b4", "EAE" = "#d73027")) +
    theme_classic(base_size = 10) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    labs(title = score, x = NULL, y = "Module score")
}

ol_boxes <- lapply(ol_features, plot_score_box,
                   seurat_obj = OL, cluster_col = "OL_clusters")
wrap_plots(ol_boxes, ncol = 2)
```

```{r heatmap_mean_scores, fig.width=12, fig.height=8}
# =============================================================================
# 9. HEATMAP - MEAN SIGNATURE SCORE PER (CLUSTER × CONDITION)
# =============================================================================
make_score_matrix <- function(seurat_obj, scores, cluster_col,
                              condition_col = "condition") {
  seurat_obj@meta.data %>%
    select(all_of(c(cluster_col, condition_col, scores))) %>%
    mutate(group = paste(.[[cluster_col]], .[[condition_col]], sep = "_")) %>%
    select(-all_of(c(cluster_col, condition_col))) %>%
    group_by(group) %>%
    summarise(across(everything(), mean, na.rm = TRUE)) %>%
    column_to_rownames("group") %>%
    as.matrix() %>% t()
}

ol_mat <- make_score_matrix(OL, ol_features, "OL_clusters")

ol_mat_clean <- ol_mat

# replace NA / NaN / Inf with 0
ol_mat_clean[!is.finite(ol_mat_clean)] <- 0

# drop rows with zero variance — they collapse to NaN after row-scaling
ol_mat_clean <- ol_mat_clean[apply(ol_mat_clean, 1, var) > 0, , drop = FALSE]

pheatmap(ol_mat_clean,
         scale = "row",
         cluster_cols = TRUE,
         color = colorRampPalette(c("navy", "white", "firebrick"))(50),
         main = "OL-lineage signatures - mean score by cluster × condition")
```

##10. Session info

```{r}
sessionInfo()
```
