---
title: "Humanized mice EAE analysis - CD45+ sorted cells (Brain / Meninges / Spleen)"
author: "Mattia Zaghi"
date: "2026-09-02"
output:
  html_document: default
  word_document: default
  pdf_document: default
---

## What is this vignette for?

This vignette processes the **CD45+ sorted** samples defined in
`Preprocessing/config/config_CD45+_sorted.yaml`. The design differs from the
other datasets in this repository in two ways that shape every downstream
choice made below:

* Three **tissues** were sorted from each animal: `Brain`, `Meninges`, `Spleen`.
  Cells from the same animal therefore appear in three libraries, so libraries
  are *not* independent replicates of the condition effect.
* Three **conditions** were collected: `CFA` (n = 3), `EAE-NonSymp` (n = 3) and
  `EAE-Symp` (**n = 1**).

The single symptomatic animal is the critical constraint. With one biological
replicate there is no within-group variance to estimate, so no statistical
test of `EAE-Symp` vs anything is valid - a nominal p-value from such a
comparison reflects only within-animal cell-to-cell noise and will be
anticonservative by a wide margin. This vignette therefore reserves formal
testing for `EAE-NonSymp` vs `CFA` (3 vs 3, per tissue) and treats
`EAE-Symp` descriptively: it is embedded, clustered, annotated and plotted
alongside the rest, and reported as effect sizes without p-values.

Everything else mirrors `analysis.Rmd`: CellBender-corrected input, per-sample
QC, per-sample DoubletFinder, merge, SCTransform, clustering, and pseudobulk
DESeq2 within each cluster.

##1. Load packages and configure directories

```{r setup, include=FALSE}
library(Seurat)
library(future)
library(tidyverse)
library(dplyr)
library(ggthemes)
library(scales)
library(cowplot)
library(RColorBrewer)
library(openxlsx)
library(purrr)
library(forcats)
library(hdf5r)
library(patchwork)
library(pals)
library(knitr)
library(DoubletFinder)
library(ggpubr)
library(pheatmap)
library(DESeq2)
library(ggrepel)
library(yaml)
opts_knit$set(root.dir = "/ws/gcb/MZ/Humanized_mice_OD/CD45/") # change accordingly
```

##2. Sample sheet

The sample names are read from the preprocessing config rather than retyped, so
this vignette cannot drift out of sync with what Cell Ranger actually
processed. Condition and tissue are parsed from the sample name
(`<condition>_<tissue>_<replicate>`), and `animal` identifies the mouse that a
library came from - the unit of replication for every condition contrast.

```{r}
base_dir <- "/ws/gcb/MZ/Humanized_mice_OD/CD45/"

# knitr's root.dir points at the data directory, not the repository, so the
# config is addressed absolutely. Adjust repo_dir if the checkout moves.
repo_dir    <- "/home/mattia/Humanized_mice_EAE"
config_path <- file.path(repo_dir, "Preprocessing/config/config_CD45+_sorted.yaml")
stopifnot(file.exists(config_path))

cfg          <- yaml::read_yaml(config_path)
cd45_samples <- names(cfg$samples)

sample_sheet <- data.frame(
  sample = cd45_samples,
  stringsAsFactors = FALSE
)
sample_sheet$condition <- sub("_[^_]+_[^_]+$", "", sample_sheet$sample)
sample_sheet$tissue    <- sub("^.*_([^_]+)_[^_]+$", "\\1", sample_sheet$sample)
sample_sheet$replicate <- sub("^.*_", "", sample_sheet$sample)
# One animal contributed Brain + Meninges + Spleen, so the animal id is the
# condition and the replicate number, without the tissue.
sample_sheet$animal <- paste(sample_sheet$condition, sample_sheet$replicate,
                             sep = "_")

sample_sheet$condition <- factor(sample_sheet$condition,
                                 levels = c("CFA", "EAE-NonSymp", "EAE-Symp"))
sample_sheet$tissue    <- factor(sample_sheet$tissue,
                                 levels = c("Brain", "Meninges", "Spleen"))

# Sanity check: replication available for each condition x tissue cell
with(sample_sheet, table(condition, tissue))
sample_sheet
```

##3. Create the Seurat object

Loading is as in `analysis.Rmd`: the CellBender-corrected h5 per sample,
feature names normalised, barcodes prefixed with the sample name, and the
combined mouse + human mitochondrial fraction computed against the joint
GRCh38-and-mm10 reference.

```{r}
# Expected multiplet rate as a function of recovered cells, interpolated from
# the 10x Chromium 3' table (~0.8% per 1,000 cells, near-linear in this range).
# These libraries recovered ~5,000 cells each, so the expected rate is ~3.9% -
# not the 0.075 inherited from analysis.Rmd, which corresponds to ~10,000 cells
# and would discard roughly 270 singlets per library as doublets.
doublet_rate_10x <- function(n_cells) {
  ref_cells <- c(  500,  1000,  2000,  3000,  4000,  5000,
                  6000,  7000,  8000,  9000, 10000)
  ref_rate  <- c(0.004, 0.008, 0.016, 0.023, 0.031, 0.039,
                 0.046, 0.054, 0.061, 0.069, 0.076)
  as.numeric(approx(ref_cells, ref_rate, xout = n_cells, rule = 2)$y)
}

mt_cutoff <- 5

seurat_list <- list()

for (i in seq_len(nrow(sample_sheet))) {
  sample  <- sample_sheet$sample[i]
  h5_path <- file.path(base_dir, sample, "RNA_AAAAGGGG/epochs/",
                       "cellbender_output_seurat.h5")
  if (!file.exists(h5_path)) {
    warning("Missing file: ", h5_path)
    next
  }
  message("Loading ", sample)

  data <- Read10X_h5(filename = h5_path, use.names = TRUE)
  if (is.list(data)) data <- data[["Gene Expression"]]

  # Normalize feature names: collapse underscores and multiple dashes
  fn <- rownames(data)
  fn <- gsub("_+", "-", fn)
  fn <- gsub("-{2,}", "-", fn)
  rownames(data) <- fn

  colnames(data) <- paste0(sample, "_", colnames(data))

  seurat_obj <- CreateSeuratObject(
    counts       = data,
    project      = sample,
    min.cells    = 3,
    min.features = 200
  )

  seurat_obj[["percent.mt.mm10"]] <- PercentageFeatureSet(seurat_obj, pattern = "mt-")
  seurat_obj[["percent.mt.hg38"]] <- PercentageFeatureSet(seurat_obj, pattern = "MT-")
  seurat_obj[["percent.mt"]] <-
    seurat_obj[["percent.mt.mm10"]] + seurat_obj[["percent.mt.hg38"]]

  seurat_obj <- subset(seurat_obj, subset = percent.mt < mt_cutoff)

  # Carry the design metadata on every cell
  seurat_obj$sample    <- sample
  seurat_obj$condition <- as.character(sample_sheet$condition[i])
  seurat_obj$tissue    <- as.character(sample_sheet$tissue[i])
  seurat_obj$animal    <- sample_sheet$animal[i]

  seurat_list[[sample]] <- seurat_obj
}
```

##4. Doublet removal

Per-sample DoubletFinder, run on each library independently so that the
artificial-doublet population is built from the same cell composition as the
library it is scoring.

Two departures from `analysis.Rmd`, both motivated in the evaluation section at
the end of this vignette:

1. **`pK` is estimated per sample** with `paramSweep`/`find.pK` instead of being
   fixed at 0.09. `pK` sets the neighbourhood size used to compute pANN, and its
   optimum depends on the number of cells and the number of distinct
   populations. A CD45+ Meninges library and a CD45+ Spleen library differ in
   both, so one hard-coded value cannot suit them all.
2. **The homotypic-proportion adjustment is applied downward, not upward.**
   `nExp` should be the expected doublet count *reduced* by the fraction that
   are homotypic and therefore undetectable, i.e. `nExp * (1 - homotypic)`.

```{r}
for (sample in names(seurat_list)) {
  message("Doublet detection: ", sample)
  seurat_obj <- seurat_list[[sample]]

  # Light preprocessing for doublet detection (per sample, SCT)
  seurat_obj <- SCTransform(seurat_obj, verbose = FALSE)
  seurat_obj <- RunPCA(seurat_obj, verbose = FALSE)
  seurat_obj <- FindNeighbors(seurat_obj, dims = 1:10, verbose = FALSE)
  seurat_obj <- FindClusters(seurat_obj, resolution = 0.1, verbose = FALSE)
  seurat_obj <- RunUMAP(seurat_obj, dims = 1:10, verbose = FALSE)

  umap_before <- DimPlot(seurat_obj, reduction = "umap",
                         group.by = "seurat_clusters") +
    ggtitle(paste(sample, "before doublet removal"))

  # ---- pK estimation (per sample) ----
  sweep_res   <- paramSweep(seurat_obj, PCs = 1:10, sct = TRUE)
  sweep_stats <- summarizeSweep(sweep_res, GT = FALSE)
  bcmvn       <- find.pK(sweep_stats)
  pK <- as.numeric(as.character(
    bcmvn$pK[which.max(bcmvn$BCmetric)]
  ))
  if (is.na(pK) || length(pK) == 0) pK <- 0.09  # fall back to the fixed value
  message("  pK = ", pK)
  pN <- 0.25

  # ---- expected number of doublets, corrected for homotypic pairs ----
  # The rate is taken from this library's own cell count, so a library that
  # recovered fewer cells is not over-filtered.
  rate <- doublet_rate_10x(ncol(seurat_obj))

  # modelHomotypic needs a clustering fine enough to separate cell types; the
  # res = 0.1 used for the QC UMAP is too coarse and inflates the homotypic
  # estimate. A finer clustering is computed here for this purpose only.
  seurat_obj <- FindClusters(seurat_obj, resolution = 0.8,
                             cluster.name = "clusters_for_homotypic",
                             verbose = FALSE)
  homotypic <- modelHomotypic(seurat_obj$clusters_for_homotypic)

  nExp_poi <- round(rate * ncol(seurat_obj))
  nExp_adj <- round(nExp_poi * (1 - homotypic))
  message("  cells = ", ncol(seurat_obj),
          ", rate = ", signif(rate, 3),
          ", homotypic = ", signif(homotypic, 3),
          ", nExp = ", nExp_poi, " -> ", nExp_adj)

  seurat_obj <- doubletFinder(seurat_obj, PCs = 1:10, pN = pN, pK = pK,
                              nExp = nExp_poi, reuse.pANN = NULL, sct = TRUE)

  df_col <- grep("^pANN_", colnames(seurat_obj@meta.data), value = TRUE)[1]
  if (is.na(df_col)) stop("DoubletFinder pANN metadata not found")

  seurat_obj <- doubletFinder(seurat_obj, PCs = 1:10, pN = pN, pK = pK,
                              nExp = nExp_adj, reuse.pANN = df_col, sct = TRUE)

  # The adjusted classification is the one keyed to nExp_adj; select it by name
  # rather than by position so the choice does not depend on column ordering.
  df_class_col <- grep(paste0("^DF.classifications_.*_", nExp_adj, "$"),
                       colnames(seurat_obj@meta.data), value = TRUE)
  if (length(df_class_col) == 0) {
    df_class_col <- tail(grep("^DF.classifications_",
                              colnames(seurat_obj@meta.data), value = TRUE), 1)
  }
  if (length(df_class_col) == 0) stop("DoubletFinder classification not found")
  message("  using ", df_class_col)

  seurat_obj$doublet_class <- seurat_obj@meta.data[[df_class_col]]
  seurat_obj$doublet_pANN  <- seurat_obj@meta.data[[df_col]]

  # Diagnostic: where do the called doublets sit? A cluster that is almost
  # entirely doublet-called is a real doublet cluster; doublets scattered
  # evenly across all clusters suggest the rate is set too high.
  print(DimPlot(seurat_obj, reduction = "umap", group.by = "doublet_class") +
          ggtitle(paste(sample, "doublet calls")))
  print(table(cluster = seurat_obj$seurat_clusters,
              call    = seurat_obj$doublet_class))
  print(VlnPlot(seurat_obj, features = c("nFeature_RNA", "nCount_RNA"),
                group.by = "doublet_class", pt.size = 0) +
          patchwork::plot_annotation(title = paste(sample, "complexity by call")))

  sel_singlets <- seurat_obj$doublet_class == "Singlet"
  if (sum(sel_singlets, na.rm = TRUE) == 0) {
    stop("No cells annotated as Singlet in ", sample)
  }
  seurat_obj <- subset(seurat_obj,
                       cells = colnames(seurat_obj)[which(sel_singlets)])

  # Recompute after doublet removal
  DefaultAssay(seurat_obj) <- "RNA"
  seurat_obj[["SCT"]] <- NULL
  seurat_obj <- SCTransform(seurat_obj, verbose = FALSE)
  seurat_obj <- RunPCA(seurat_obj, verbose = FALSE)
  seurat_obj <- FindNeighbors(seurat_obj, dims = 1:10, verbose = FALSE)
  seurat_obj <- FindClusters(seurat_obj, resolution = 0.1, verbose = FALSE)
  seurat_obj <- RunUMAP(seurat_obj, dims = 1:10, verbose = FALSE)

  umap_after <- DimPlot(seurat_obj, reduction = "umap",
                        group.by = "seurat_clusters") +
    ggtitle(paste(sample, "after doublet removal"))

  print(umap_before + umap_after)

  seurat_list[[sample]] <- seurat_obj
}
```

Record how many cells each library lost, so an outlying sample is visible
before it influences the merged object.

```{r}
doublet_summary <- do.call(rbind, lapply(names(seurat_list), function(s) {
  data.frame(sample = s, cells_retained = ncol(seurat_list[[s]]))
}))
doublet_summary <- left_join(doublet_summary, sample_sheet, by = "sample")
doublet_summary

ggplot(doublet_summary, aes(x = sample, y = cells_retained, fill = tissue)) +
  geom_col() +
  coord_flip() +
  labs(x = NULL, y = "Cells retained after doublet removal") +
  theme_classic(base_size = 12)
```

##5. Merge and cluster

```{r}
humanized_mouse_cd45 <- Reduce(function(x, y) merge(x, y), seurat_list)
rm(seurat_list); gc()

# Remove per-sample doublet-detection metadata, which is meaningless once the
# libraries are merged (the columns are keyed to each sample's own nExp)
md <- humanized_mouse_cd45@meta.data
drop_cols <- grep("^(pANN_|DF.classifications_|nCount_SCT|nFeature_SCT|SCT_snn_res|seurat_clusters|clusters_for_homotypic)",
                  colnames(md), value = TRUE)
for (cc in drop_cols) humanized_mouse_cd45@meta.data[[cc]] <- NULL

humanized_mouse_cd45 <- subset(humanized_mouse_cd45,
                               subset = percent.mt < mt_cutoff)
humanized_mouse_cd45
```

Normalisation follows `analysis.Rmd`: a log-normalised RNA branch (kept for
marker plots and for the pseudobulk step, which reads raw counts) and an SCT
branch used for dimensionality reduction and clustering.

```{r}
options(future.globals.maxSize = 100e+09)

DefaultAssay(humanized_mouse_cd45)<-"RNA"

humanized_mouse_cd45[["SCT"]] <- NULL

# Log-normalized analysis on RNA
humanized_mouse_cd45 <- NormalizeData(humanized_mouse_cd45, assay = "RNA")
humanized_mouse_cd45 <- FindVariableFeatures(humanized_mouse_cd45, assay = "RNA")
humanized_mouse_cd45 <- ScaleData(humanized_mouse_cd45, assay = "RNA")
humanized_mouse_cd45 <- RunPCA(humanized_mouse_cd45, assay = "RNA",
                               reduction.name = "pca_lognorm")

# SCT analysis on SCT assay
humanized_mouse_cd45 <- SCTransform(humanized_mouse_cd45, verbose = TRUE)
humanized_mouse_cd45 <- FindVariableFeatures(humanized_mouse_cd45, assay = "SCT")
humanized_mouse_cd45 <- RunPCA(humanized_mouse_cd45, assay = "SCT",
                               reduction.name = "pca_SCT")
```

Read the elbow before fixing `n_dims`. The value below is the one used for the
unsorted brain; a CD45+ sorted population is less heterogeneous, so confirm it
rather than inheriting it.

```{r}
ElbowPlot(humanized_mouse_cd45, reduction = "pca_SCT", ndims = 50)
```

```{r}
n_dims <- 20  # <- set from the elbow plot above

humanized_mouse_cd45 <- FindNeighbors(humanized_mouse_cd45, dims = 1:n_dims,
                                      reduction = "pca_SCT")
humanized_mouse_cd45 <- FindClusters(humanized_mouse_cd45, resolution = 0.5,
                                     cluster.name = "unintegrated_clusters")
humanized_mouse_cd45 <- RunUMAP(humanized_mouse_cd45, dims = 1:n_dims,
                                reduction = "pca_SCT",
                                reduction.name = "umap.unintegrated")
```

```{r, fig.height=6, fig.width=18}
my_colors <- as.vector(pals::polychrome(36))

p1 <- DimPlot(humanized_mouse_cd45, reduction = "umap.unintegrated",
              group.by = "unintegrated_clusters", cols = my_colors,
              label = TRUE) + labs(x = "UMAP 1", y = "UMAP 2")
p2 <- DimPlot(humanized_mouse_cd45, reduction = "umap.unintegrated",
              group.by = "tissue") + labs(x = "UMAP 1", y = "UMAP 2")
p3 <- DimPlot(humanized_mouse_cd45, reduction = "umap.unintegrated",
              group.by = "condition") + labs(x = "UMAP 1", y = "UMAP 2")
p1 + p2 + p3


```

Cluster composition by tissue and by animal. The second panel is the one to
read sceptically: a cluster carried by a single animal is a batch effect
candidate, not a biological population.

```{r, fig.height=8, fig.width=14}
comp_tissue <- as.data.frame(table(cluster = humanized_mouse_cd45$unintegrated_clusters,
                                   tissue  = humanized_mouse_cd45$tissue))
comp_animal <- as.data.frame(table(cluster = humanized_mouse_cd45$unintegrated_clusters,
                                   animal  = humanized_mouse_cd45$animal))

q1 <- ggplot(comp_tissue, aes(x = cluster, y = Freq, fill = tissue)) +
  geom_col(position = "fill") + scale_y_continuous(labels = percent) +
  labs(x = "Cluster", y = "Fraction of cells", fill = "Tissue") +
  theme_classic(base_size = 14)
q2 <- ggplot(comp_animal, aes(x = cluster, y = Freq, fill = animal)) +
  geom_col(position = "fill") + scale_y_continuous(labels = percent) +
  labs(x = "Cluster", y = "Fraction of cells", fill = "Animal") +
  theme_classic(base_size = 14)
q1 / q2
```

Cluster markers.

```{r}
DefaultAssay(humanized_mouse_cd45) <- "SCT"
humanized_mouse_cd45 <- PrepSCTFindMarkers(humanized_mouse_cd45)
cd45_markers <- FindAllMarkers(humanized_mouse_cd45, only.pos = TRUE)

write.xlsx(cd45_markers,
           "/ws/gcb/MZ/Humanized_mice_OD/Results/CD45_cluster_markers.xlsx",
           overwrite = TRUE)
cd45_markers %>% group_by(cluster) %>% slice_max(avg_log2FC, n = 10)
```

```{r}
#remove cluster 13 which is from ambient RNA contamination

humanized_mouse <- subset(humanized_mouse, idents = c("0", "1","2","3","4","5","6","7","8","9","10","11","12",
                                               "14","15","16"))
#rename fine and broad celltype annotation

fine_global <- c(
  "0"  = "CD4 T (memory/Th17)",
  "1"  = "Pre-B",
  "2"  = "B naive",
  "3"  = "CD4 T naive/CM",
  "4"  = "CD8 T (GZMK+ EM)",
  "5"  = "Monocytes classical",
  "6"  = "Monocytes non-class.",
  "7"  = "B cells (IGLC2-hi)",
  "8"  = "cDC2 / mo-DC",
  "9"  = "B cells (IGHM/IGLC3)",
  "10" = "Proliferating",
  "11" = "Neutrophils mature",
  "12" = "Macrophages",
  "13" = "Pre-B proliferating",
  "14" = "CD8 T activated",
  "15" = "NK",
  "16" = "B transitional",
  "17" = "pDC",
  "18" = "Pro-B / early lymph.",
  "19" = "Neutrophils immature",
  "20" = "B cells (IGLC1-hi)",
  "21" = "MAIT / innate-like T",
  "22" = "Plasma cells",
  "23" = "cDC1",
  "24" = "Mast cells",
  "25" = "Plasma cells (IGLV+)",
  "26" = "B germinal-centre",
  "27" = "Plasmablasts",
  "28" = "Plasma cells (BCMA+)"
)
humanized_mouse_cd45<-RenameIdents(humanized_mouse_cd45,fine_global)
humanized_mouse_cd45@meta.data$fine_clusters<-humanized_mouse_cd45@active.ident

lineage_global <- c(
  "CD4 T (memory/Th17)"  = "CD4 T",
  "Pre-B"                = "B progenitor",
  "B naive"              = "B",
  "CD4 T naive/CM"       = "CD4 T",
  "CD8 T (GZMK+ EM)"     = "CD8 T",
  "Monocytes classical"  = "Monocyte",
  "Monocytes non-class." = "Monocyte",
  "B cells (IGLC2-hi)"   = "B",
  "cDC2 / mo-DC"         = "DC",
  "B cells (IGHM/IGLC3)" = "B",
  "Proliferating"        = "Proliferating",
  "Neutrophils mature"   = "Neutrophil",
  "Macrophages"          = "Macrophage",
  "Pre-B proliferating"  = "B progenitor",
  "CD8 T activated"      = "CD8 T",
  "NK"                   = "NK",
  "B transitional"       = "B",
  "pDC"                  = "DC",
  "Pro-B / early lymph." = "B progenitor",
  "Neutrophils immature" = "Neutrophil",
  "B cells (IGLC1-hi)"   = "B",
  "MAIT / innate-like T" = "Innate-like T",
  "Plasma cells"         = "Plasma",
  "cDC1"                 = "DC",
  "Mast cells"           = "Mast",
  "Plasma cells (IGLV+)" = "Plasma",
  "B germinal-centre"    = "B",
  "Plasmablasts"         = "Plasma",
  "Plasma cells (BCMA+)" = "Plasma"
)
humanized_mouse_cd45<-RenameIdents(humanized_mouse_cd45,lineage_global)
humanized_mouse_cd45@meta.data$broad_clusters<-humanized_mouse_cd45@active.ident

saveRDS(humanized_mouse_cd45,
        "/date/gcb/gcb_MZ/Humanized_mice_OD/Results/objects/humanized_mouse_cd45_SCT.rds")
```

Export umap coordinates and Clusters

```{r}
suppressPackageStartupMessages({ library(Seurat); library(openxlsx) })

obj_dir  <- "/ws/gcb/MZ/Humanized_mice_OD/Results"
out_xlsx <- "/ws/gcb/MZ/Humanized_mice_OD/Results/CD45_metadata.xlsx"

wb <- createWorkbook()
  mk<-"CD45"
  md <- humanized_mouse_cd45@meta.data
  md$barcode<-colnames(humanized_mouse_cd45)
  df <- data.frame(cell = rownames(md), md, row.names = NULL,
                   check.names = FALSE, stringsAsFactors = FALSE)
  addWorksheet(wb, mk)
  writeData(wb, mk, df)
  cat(sprintf("  %-9s : %d cells x %d metadata columns\n", mk, nrow(df), ncol(df) - 1))

saveWorkbook(wb, out_xlsx, overwrite = TRUE)


obj_dir  <- "/ws/gcb/MZ/Humanized_mice_OD/Results"
out_xlsx <- "/ws/gcb/MZ/Humanized_mice_OD/Results/CD45_UMAP_coordinates_harmony.xlsx"

wb <- createWorkbook()
  md <- humanized_mouse_cd45@reductions$umap.unintegrated@cell.embeddings
  df <- data.frame(cell = rownames(md), md, row.names = NULL,
                   check.names = FALSE, stringsAsFactors = FALSE)
  addWorksheet(wb, mk)
  writeData(wb, mk, df)

  saveWorkbook(wb, out_xlsx, overwrite = TRUE)
  
```

save full count matrix for future

```{r}
library("zellkonverter")
library(SingleCellExperiment)
sce <- as.SingleCellExperiment(humanized_mouse_cd45, assay = "RNA")
writeH5AD(sce, "/ws/gcb/MZ/Humanized_mice_OD/Results/humanized_mouse_cd45.h5ad")

library(Matrix); library(data.table)

obj <- humanized_mouse_cd45

DefaultAssay(obj) <- "RNA"
obj <- JoinLayers(obj)
mat <- GetAssayData(obj, assay = "RNA", layer = "data")   # or layer = "counts" for raw

outfile   <- "/ws/gcb/MZ/Humanized_mice_OD/Results/Count_Matrix.csv.gz"
dir.create(dirname(outfile), recursive = TRUE, showWarnings = FALSE)

chunk <- 2000
idx   <- split(seq_len(ncol(mat)), ceiling(seq_len(ncol(mat)) / chunk))

for (i in seq_along(idx)) {
  blk <- as.matrix(t(mat[, idx[[i]], drop = FALSE]))    # cells x genes
  dt  <- data.table(cell = rownames(blk), blk)
  fwrite(dt, outfile, append = i > 1, col.names = i == 1, compress = "gzip")
  message("chunk ", i, "/", length(idx))
}
```


##6. Reclustering based on brain regions 

```{r}
# 1) cluster set
tissue <- unique(humanized_mouse_cd45$tissue)

# storage
clusters_objects <- list()

for (bc in tissue) {
  bc_clean <- make.names(bc)
  obj_name <- paste0(bc_clean)
  message("Processing tissue: ", bc, " --> ", obj_name)

  # subset
  TS <- subset(humanized_mouse_cd45, subset = tissue == bc)

  # prep
  DefaultAssay(TS) <- "RNA"
  TS[["SCT"]] <- NULL
  TS <- SCTransform(TS, vst.flavor = "v2", verbose = FALSE)
  TS <- RunPCA(TS, verbose = FALSE, reduction.name = paste0("pca",obj_name,"_SCT"))
  TS <- FindNeighbors(TS, reduction = paste0("pca",obj_name,"_SCT"), dims = 1:20,
                      graph.name = c(paste0(obj_name,"_nn"), paste0(obj_name,"_snn")))
  TS <- FindClusters(TS, resolution = 0.5,
                     graph.name = paste0(obj_name,"_snn"), cluster.name = paste0(obj_name,"_clusters"))
  TS <- RunUMAP(TS, reduction = paste0("pca",obj_name,"_SCT"), dims = 1:20,
                reduction.name = paste0("umap_",obj_name))
  # store into list by key
  clusters_objects[[obj_name]] <- TS
  # save object
  saveRDS(TS, file = file.path("/date/gcb/gcb_MZ/Humanized_mice_OD/Results/objects/",
                               paste0(obj_name, "_CD45_SCT.rds")))
  # markers
  DefaultAssay(TS) <- "SCT"
  TS <- PrepSCTFindMarkers(TS, assay = "SCT", verbose = FALSE)
  TS_markers <- FindAllMarkers(TS, only.pos = TRUE)

  # Create a new workbook
  wb <- createWorkbook()

  # Loop through each cluster and add data to a new sheet
  for (cluster_value in unique(TS_markers$cluster)) {
    cluster_data <- subset(TS_markers, cluster == cluster_value)
    sheet_name <- paste0("Cluster_", cluster_value)
    addWorksheet(wb, sheet_name)
    writeData(wb, sheet_name, cluster_data)
}

# Save the workbook to a file
saveWorkbook(wb, paste0("/date/gcb/gcb_MZ/Humanized_mice_OD/Results/cluster_markers/",bc,"_CD45_markers.xlsx"), overwrite = TRUE)
}
```

Save Umap and Metadata coordinates


```{r}
suppressPackageStartupMessages({ library(Seurat); library(openxlsx) })

obj_dir  <- "/date/gcb/gcb_MZ/Humanized_mice_OD/Results/UMAP_Metadata_count_matrices/"
out_xlsx <- "/date/gcb/gcb_MZ/Humanized_mice_OD/Results/UMAP_Metadata_count_matrices/Reclustering_CD45_metadata.xlsx"


cat("marks:", paste(names(clusters_objects), collapse = ", "), "\n")

wb <- createWorkbook()
for (mk in names(clusters_objects)) {
  md <- clusters_objects[[mk]]@meta.data
  md$barcode<-colnames(clusters_objects[[mk]])
  df <- data.frame(cell = rownames(md), md, row.names = NULL,
                   check.names = FALSE, stringsAsFactors = FALSE)
  addWorksheet(wb, mk)
  writeData(wb, mk, df)
  cat(sprintf("  %-9s : %d cells x %d metadata columns\n", mk, nrow(df), ncol(df) - 1))
}
saveWorkbook(wb, out_xlsx, overwrite = TRUE)
cat("Wrote:", out_xlsx, "\n")

obj_dir  <- "/date/gcb/gcb_MZ/Humanized_mice_OD/Results/UMAP_Metadata_count_matrices"
out_xlsx <- "/date/gcb/gcb_MZ/Humanized_mice_OD/Results/UMAP_Metadata_count_matrices/Reclustering_CD45_UMAP.xlsx"

cat("marks:", paste(names(clusters_objects), collapse = ", "), "\n")

wb <- createWorkbook()
for (mk in names(clusters_objects)) {
  md <- clusters_objects[[mk]]@reductions[[paste0("umap_",mk)]]@cell.embeddings
  df <- data.frame(cell = rownames(md), md, row.names = NULL,
                   check.names = FALSE, stringsAsFactors = FALSE)
  addWorksheet(wb, mk)
  writeData(wb, mk, df)
  cat(sprintf("  %-9s : %d cells x %d metadata columns\n", mk, nrow(df), ncol(df) - 1))
}
saveWorkbook(wb, out_xlsx, overwrite = TRUE)
cat("Wrote:", out_xlsx, "\n")


```



##6. Differential gene expression: EAE-NonSymp vs CFA

Pseudobulk DESeq2 within each cluster, with the animal as the unit of
replication. Two features of this design need handling that the unsorted brain
dataset did not:

* **Each animal contributed three tissues.** Aggregating per animal x tissue
  and fitting `~ tissue + condition` blocks the tissue effect while keeping the
  condition term estimated across animals. Tissue is dropped automatically when
  a cluster is confined to one tissue, which would otherwise leave the model
  rank-deficient.
* **`EAE-Symp` has n = 1** and is excluded from testing. It is summarised
  separately below without p-values.

CFA is the reference level, so a positive log2FC means higher in EAE.

```{r}

run_pseudobulk_dge_cd45 <- function(seurat_obj,
                                    cluster_col,
                                    cluster_id,
                                    test_conditions  = c("CFA", "EAE"),
                                    animal_col       = "sample",
                                    tissue_col       = "tissue",
                                    condition_col    = "disease",
                                    min_cells        = 10,
                                    min_cells_per_pb = 10,
                                    min_counts       = 10) {

  idx <- seurat_obj@meta.data[[cluster_col]] == cluster_id &
    seurat_obj@meta.data[[condition_col]] %in% test_conditions
  if (sum(idx) < min_cells) {
    message("Skipping cluster ", cluster_id, ": only ", sum(idx), " testable cells")
    return(NULL)
  }
  

  sub_obj <- seurat_obj[, idx]
  DefaultAssay(sub_obj) <- "RNA"
  sub_obj <- JoinLayers(sub_obj)
  counts_mat <- GetAssayData(sub_obj, assay = "RNA", layer = "counts")
  meta <- sub_obj@meta.data

  # One pseudobulk profile per animal x tissue
  pb_group <- paste(meta[[animal_col]], meta[[tissue_col]], sep = "__")

  keep_groups <- names(which(table(pb_group) >= min_cells_per_pb))
  if (length(keep_groups) < 4) {
    message("Skipping cluster ", cluster_id,
            ": fewer than 4 usable pseudobulk groups")
    return(NULL)
  }

  pb_mat <- vapply(keep_groups, function(g) {
    as.numeric(Matrix::rowSums(counts_mat[, pb_group == g, drop = FALSE]))
  }, numeric(nrow(counts_mat)))
  rownames(pb_mat) <- rownames(counts_mat)

  col_meta <- data.frame(
    group     = keep_groups,
    animal    = sub("__.*$", "", keep_groups),
    tissue    = factor(sub("^.*__", "", keep_groups)),
    row.names = keep_groups
  )
  col_meta$condition <- factor(
    vapply(col_meta$animal, function(a)
      unique(as.character(meta[[condition_col]][meta[[animal_col]] == a]))[1],
      character(1)),
    levels = test_conditions
  )

  # Require at least two animals per condition, or the dispersion estimate is
  # not identifiable and the test is not interpretable.
  n_animals <- tapply(col_meta$animal, col_meta$condition,
                      function(x) length(unique(x)))
  if (any(is.na(n_animals)) || any(n_animals < 2)) {
    message("Skipping cluster ", cluster_id,
            ": fewer than 2 animals in one condition")
    return(NULL)
  }

  col_meta$tissue <- droplevels(col_meta$tissue)
  use_tissue <- nlevels(col_meta$tissue) > 1 &&
    !any(rowSums(table(col_meta$tissue, col_meta$condition) > 0) < 2)
  design <- if (use_tissue) ~ tissue + condition else ~ condition
  message("Cluster ", cluster_id, ": ", nrow(col_meta),
          " pseudobulk groups, design ", deparse(design))

  pb_mat <- pb_mat[rowSums(pb_mat) >= min_counts, , drop = FALSE]
  if (nrow(pb_mat) < 10) {
    message("Skipping cluster ", cluster_id, ": too few expressed genes")
    return(NULL)
  }

  tryCatch({
    dds <- DESeqDataSetFromMatrix(countData = round(pb_mat),
                                  colData   = col_meta,
                                  design    = design)
    dds <- DESeq(dds, quiet = TRUE)
    res <- results(dds, contrast = c("condition", test_conditions[2],
                                     test_conditions[1]))
    res_df <- as.data.frame(res)
    res_df$gene    <- rownames(res_df)
    res_df$cluster <- as.character(cluster_id)
    res_df$design  <- deparse(design)
    res_df$n_pb    <- nrow(col_meta)
    res_df[order(res_df$padj, na.last = TRUE), ]
  }, error = function(e) {
    message("DESeq2 error - cluster ", cluster_id, ": ", conditionMessage(e))
    NULL
  })
}

```

DGE within each cluster, pooling tissues and blocking on tissue.

```{r}
tissue <- names(clusters_objects)

for (bc in tissue) {
  cluster_ids <- levels(clusters_objects[[bc]]@meta.data[[paste0(bc, "_clusters")]])

  dge_cd45_list <- lapply(cluster_ids, function(cl) {
    message("[", bc, "] Running DGE for cluster: ", cl)
    tryCatch(
      run_pseudobulk_dge_cd45(
        seurat_obj  = clusters_objects[[bc]],
        cluster_col = paste0(bc, "_clusters"),
        cluster_id  = cl
      ),
      error = function(e) { message("  skipped: ", conditionMessage(e)); NULL }
    )
  })
  names(dge_cd45_list) <- cluster_ids

  wb_cd45 <- createWorkbook()                     # inside the loop
  for (cl in names(dge_cd45_list)) {
    res <- dge_cd45_list[[cl]]
    if (is.null(res) || nrow(res) == 0) next
    res <- data.frame(gene = rownames(res), res, row.names = NULL)
    sheet_name <- substr(gsub("[\\[\\]:*?/\\\\]", "_", paste0("Cluster_", cl)), 1, 31)
    addWorksheet(wb_cd45, sheet_name)
    writeData(wb_cd45, sheet_name, res)
  }

  if (length(names(wb_cd45)) == 0) { message("No results for ", bc, " - nothing saved."); next }

  saveWorkbook(wb_cd45,
    file.path("/ws/gcb/MZ/Humanized_mice_OD/Results",
              paste0("DGE_EAE-NonSymp_vs_CFA_CD45_", bc, "_clusters.xlsx")),
    overwrite = TRUE)
  message("Saved ", bc, ".")
}
```

PCA per tissue

```{r}
library(DESeq2); library(ggplot2); library(ggrepel); library(Matrix)
for (bc in tissue) {
obj <- clusters_objects[[bc]]
DefaultAssay(obj) <- "RNA"
obj <- JoinLayers(obj)
cts <- GetAssayData(obj, assay = "RNA", layer = "counts")

# sparse aggregation: genes x samples (avoids densifying 28k x 22k)
g  <- factor(obj$sample)
mm <- sparse.model.matrix(~ 0 + g); colnames(mm) <- levels(g)
pb <- as.matrix(cts %*% mm)

cd <- data.frame(
  sample  = colnames(pb),
  detail  = factor(sub("_.*$", "", colnames(pb))),   # CFA / EAE-NonSymp / EAE-Symp
  row.names = colnames(pb)
)
cd$disease <- factor(ifelse(cd$detail == "CFA", "CFA", "EAE"), levels = c("CFA", "EAE"))

dds <- DESeqDataSetFromMatrix(round(pb), cd, ~ disease)
dds <- dds[rowSums(counts(dds)) >= 10, ]
vsd <- vst(dds, blind = TRUE)          # blind = TRUE: ignore the design, honest QC

p <- plotPCA(vsd, intgroup = "detail", returnData = TRUE)
pv <- round(100 * attr(p, "percentVar"))

p1<-ggplot(p, aes(PC1, PC2, colour = detail, label = name)) +
  geom_point(size = 4) + geom_text_repel(size = 3, show.legend = FALSE) +
  labs(x = paste0("PC1: ", pv[1], "% variance"),
       y = paste0("PC2: ", pv[2], "% variance"), colour = NULL,
       title = paste0(bc," pseudobulk, 7 animals")) +
  coord_fixed() + theme_bw()
print(p1)
qc  <- data.frame(
  depth   = colSums(pb)[p$name],
  nGene   = colSums(pb > 0)[p$name],
  PC1     = p$PC1
)
p2<-cor(qc$PC1, qc$depth); cor(qc$PC1, qc$nGene)
print(p2)
}
```



Per-tissue DGE within each cluster - the stratified version, 3 vs 3 animals in
each tissue. This is the comparison to report when a cluster behaves
differently in Brain, Meninges and Spleen.

```{r}
tissues <- levels(droplevels(factor(humanized_mouse_cd45$tissue)))

dge_by_tissue <- list()
for (tis in tissues) {
  obj_tis <- subset(humanized_mouse_cd45, subset = tissue == tis)
  for (cl in levels(droplevels(obj_tis$unintegrated_clusters))) {
    key <- paste(tis, cl, sep = "_cluster")
    message("Running DGE for ", key)
    dge_by_tissue[[key]] <- run_pseudobulk_dge_cd45(
      seurat_obj  = obj_tis,
      cluster_col = "unintegrated_clusters",
      cluster_id  = cl
    )
  }
  rm(obj_tis); gc()
}

wb_tissue <- createWorkbook()
for (key in names(dge_by_tissue)) {
  if (!is.null(dge_by_tissue[[key]])) {
    sheet_name <- substr(key, 1, 31)
    addWorksheet(wb_tissue, sheet_name)
    writeData(wb_tissue, sheet_name, dge_by_tissue[[key]])
  }
}
saveWorkbook(wb_tissue,
             "/date/gcb/gcb_MZ/Humanized_mice_OD/Results/DGE_EAE-NonSymp_vs_CFA_CD45_by_tissue.xlsx",
             overwrite = TRUE)
cat("Saved CD45 per-tissue DGE results.\n")
```


##7. The symptomatic animal, described rather than tested

With n = 1 there is no valid test, but the animal is still informative. Two
things can honestly be reported: how its cluster composition compares to the
groups that do have replication, and the size of its expression shift for genes
that were already identified in the replicated `EAE-NonSymp` vs `CFA` contrast.
No p-values are produced here by design.

```{r, fig.height=6, fig.width=12}
prop <- humanized_mouse_cd45@meta.data %>%
  count(animal, condition, tissue, unintegrated_clusters) %>%
  group_by(animal, tissue) %>%
  mutate(frac = n / sum(n)) %>%
  ungroup()

ggplot(prop, aes(x = unintegrated_clusters, y = frac, colour = condition)) +
  geom_point(position = position_dodge(width = 0.6), size = 2) +
  facet_wrap(~ tissue, ncol = 1) +
  scale_y_continuous(labels = percent) +
  labs(x = "Cluster", y = "Fraction of cells in library", colour = "Condition") +
  theme_classic(base_size = 13)
```

Effect size for the symptomatic animal on the replicated hits: pseudobulk CPM
per animal x tissue, so the symptomatic value can be read against the spread of
the CFA and non-symptomatic animals.

```{r}
pseudobulk_cpm <- function(seurat_obj, cluster_col, cluster_id,
                           min_cells_per_pb = 10) {
  idx <- seurat_obj@meta.data[[cluster_col]] == cluster_id
  if (sum(idx) < min_cells_per_pb) return(NULL)
  sub_obj <- seurat_obj[, idx]
  DefaultAssay(sub_obj) <- "RNA"
  sub_obj <- JoinLayers(sub_obj)
  cm   <- GetAssayData(sub_obj, assay = "RNA", layer = "counts")
  meta <- sub_obj@meta.data
  grp  <- paste(meta$animal, meta$tissue, sep = "__")
  keep <- names(which(table(grp) >= min_cells_per_pb))
  if (length(keep) == 0) return(NULL)
  mat <- vapply(keep, function(g)
    as.numeric(Matrix::rowSums(cm[, grp == g, drop = FALSE])),
    numeric(nrow(cm)))
  rownames(mat) <- rownames(cm)
  cpm <- sweep(mat, 2, colSums(mat), "/") * 1e6
  as.data.frame(cpm) %>%
    mutate(gene = rownames(cpm)) %>%
    pivot_longer(-gene, names_to = "group", values_to = "cpm") %>%
    mutate(animal  = sub("__.*$", "", group),
           tissue  = sub("^.*__", "", group),
           cluster = as.character(cluster_id))
}

# Example: the top replicated hits of one cluster, with the symptomatic animal
# shown alongside. Change `cl` to inspect other clusters.
cl <- cluster_ids[1]
if (!is.null(dge_cd45_list[[cl]])) {
  top_genes <- head(dge_cd45_list[[cl]]$gene[
    which(dge_cd45_list[[cl]]$padj < 0.05)], 12)

  if (length(top_genes) > 0) {
    cpm_long <- pseudobulk_cpm(humanized_mouse_cd45,
                               "unintegrated_clusters", cl) %>%
      filter(gene %in% top_genes) %>%
      left_join(distinct(sample_sheet, animal, condition), by = "animal")

    print(
      ggplot(cpm_long, aes(x = condition, y = cpm, colour = condition)) +
        geom_point(position = position_jitter(width = 0.15), size = 2) +
        facet_wrap(~ gene, scales = "free_y") +
        labs(x = NULL, y = "Pseudobulk CPM",
             title = paste("Cluster", cl,
                           "- replicated hits, all animals shown")) +
        theme_classic(base_size = 12) +
        theme(axis.text.x = element_text(angle = 45, hjust = 1),
              legend.position = "none")
    )
  }
}
```

##8. Session info

```{r}
sessionInfo()
```
