---
title: "Humanized mice EAE analysis Brain"
author: "Mattia Zaghi"
date: "2026-03-06"
output:
  word_document: default
  html_document: default
  pdf_document: default
---

## What is this vignette for?
This Vignette contains all the necessary data analysis steps followed to analyzed the single cell RNA-seq of Mature humanized_mouse_braindendrocytes treated with FGF and Tomato controls including the code to generate the different figures.


##1.Load packages, configure directories and create the Seurat object

This section of the code is necessary to load all the necessary packages for the analysis, configure the directories and create the Seurat object that is going to be used for the rest of the analysis.

##1.1 Load packages, configure directories

1.First load all the necessary packages in order to run all the subsequent steps & set the working directory where all the output of this analysis will be saved and stored

```{r setup, include=FALSE}
library(Seurat)
library(future)
library(tidyverse)
library(dplyr)
library(tidyverse)
library(ggthemes)
library(scales)
library(cowplot)
library(RColorBrewer)
library(openxlsx)
library(homologene)
library(openxlsx)
library(clusterProfiler)
library(org.Mm.eg.db)
library(org.Hs.eg.db)
library(purrr)
library(forcats)
library(readxl)
library(hdf5r)
library(patchwork)
library(pals)
require("knitr")


library(DoubletFinder)
opts_knit$set(root.dir = "/date/gcb/gcb_MZ/Humanized_mice_OD/")# change accordingly

#source("/home/mattia/FGF_signaling_induces_disease-associated_humanized_mouse_braindendrocytes/functions.R")
```

##1.2 Create the seurat object 

Let's start with loading the actual data. We will load separately all the brain samples samples from the cell ranger output and then everything will be merged into a unique object thet will be called brain hMS. 

1.load all the seurat object corrected by cell bender 
```{r}

# Base folder containing all samples
base_dir <- "/date/gcb/gcb_MZ/Humanized_mice_OD"

# Find all sample folders that contain "brain" in the name
brain_samples <- list.dirs(base_dir, recursive = FALSE, full.names = FALSE)
brain_samples <- brain_samples[grepl("brain", brain_samples)]

# Load each sample and create a Seurat object (with unique cell names)
seurat_list <- list()
for (sample in brain_samples) {
  h5_path <- file.path(base_dir, sample, "RNA_AAAAGGGG", "cellbender_output_seurat.h5")
  if (!file.exists(h5_path)) {
    warning("Missing file: ", h5_path)
    next
  }
  message("Loading ", sample)

  data <- Read10X_h5(filename = h5_path, use.names = TRUE)

  # Normalize feature names: collapse underscores and multiple dashes into a single dash
  fn <- rownames(data)
  fn <- gsub("_+", "-", fn)      # underscores -> dash
  fn <- gsub("-{2,}", "-", fn)   # collapse multiple dashes
  rownames(data) <- fn

  # Prefix cell barcodes with sample name so they are unique across samples
  colnames(data) <- paste0(sample, "_", colnames(data))

  seurat_obj <- CreateSeuratObject(
    counts = data,
    project = sample,
    min.cells = 3,
    min.features = 200
  )

  # Per-sample QC before merging
  seurat_obj[["percent.mt.mm10"]] <- PercentageFeatureSet(
  seurat_obj,
  pattern = "mt-"
)

# human mito (hg38)
seurat_obj[["percent.mt.hg38"]] <- PercentageFeatureSet(
  seurat_obj,
  pattern = "MT-"
)
seurat_obj[["percent.mt"]]<-seurat_obj[["percent.mt.mm10"]]+seurat_obj[["percent.mt.hg38"]]

  seurat_obj <- subset(seurat_obj, subset = percent.mt < 5)

  # Light preprocessing for doublet detection (on each sample independently using SCT)
  seurat_obj <- SCTransform(seurat_obj, verbose = FALSE)
  seurat_obj <- RunPCA(seurat_obj, verbose = FALSE)
  seurat_obj <- FindNeighbors(seurat_obj, dims = 1:10, verbose = FALSE)
  seurat_obj <- FindClusters(seurat_obj, resolution = 0.1, verbose = FALSE)
  seurat_obj <- RunUMAP(seurat_obj, dims = 1:10, verbose = FALSE)

  umap_before <- DimPlot(seurat_obj, reduction = "umap", group.by = "seurat_clusters") + ggtitle(paste(sample, "before doublet removal"))

  # Per-sample DoubletFinder (SCT input)
  nExp_poi <- round(0.075 * ncol(seurat_obj))
  nExp_poi.adj <- round(nExp_poi * 1.25)
  pK <- 0.09
  pN <- 0.25

  seurat_obj <- doubletFinder(seurat_obj, PCs = 1:10, pN = pN, pK = pK, nExp = nExp_poi, reuse.pANN = NULL, sct = TRUE)

  df_col <- grep("pANN_", colnames(seurat_obj@meta.data), value = TRUE)[1]
  if (is.na(df_col)) stop("DoubletFinder pANN metadata not found")

  seurat_obj <- doubletFinder(seurat_obj, PCs = 1:10, pN = pN, pK = pK, nExp = nExp_poi.adj, reuse.pANN = df_col, sct = TRUE)

  df_class_cols <- grep("DF.classifications_", colnames(seurat_obj@meta.data), value = TRUE)
  if (length(df_class_cols) == 0) stop("DoubletFinder classification column not found")

  # Use the most recent classification column (should be the second pass)
  df_class_col <- tail(df_class_cols, 1)
  message("Using doublet classification column: ", df_class_col)

  sel_singlets <- seurat_obj@meta.data[[df_class_col]] == "Singlet"
  if (sum(sel_singlets, na.rm = TRUE) == 0) {
    stop("No cells annotated as Singlet in ", df_class_col)
  }
  seurat_obj <- subset(seurat_obj, cells = rownames(seurat_obj@meta.data)[sel_singlets])

  # Recompute UMAP after doublet removal (SCT after singlet filtering)
  # Remove the old SCT assay 
  DefaultAssay(seurat_obj)<-"RNA"
  seurat_obj[["SCT"]] <- NULL
  seurat_obj <- SCTransform(seurat_obj, verbose = TRUE)
  seurat_obj <- RunPCA(seurat_obj, verbose = TRUE)
  seurat_obj <- FindNeighbors(seurat_obj, dims = 1:10, verbose = TRUE)
  seurat_obj <- FindClusters(seurat_obj, resolution = 0.1, verbose = TRUE)
  seurat_obj <- RunUMAP(seurat_obj, dims = 1:10, verbose = TRUE)

  umap_after <- DimPlot(seurat_obj, reduction = "umap", group.by = "seurat_clusters") + ggtitle(paste(sample, "after doublet removal"))

  # Print per-sample before/after UMAP side-by-side
  print(umap_before + umap_after)

  seurat_list[[sample]] <- seurat_obj
}
```



2.Now all the different samples are going to be merged in one unique object 

```{r}
seurat_list <- lapply(names(seurat_list), function(s) {
  obj <- seurat_list[[s]]
  obj$sample <- s
  obj
})
names(seurat_list) <- names(seurat_list)

humanized_mouse_brain <- Reduce(function(x, y) merge(x, y), seurat_list)

rm(seurat_list)
#remove metadata from previous single sample doublet removal analysis
humanized_mouse_brain@meta.data[["nCount_SCT"]]<-NULL
humanized_mouse_brain@meta.data[["nFeature_SCT"]]<-NULL
humanized_mouse_brain@meta.data[["SCT_snn_res.0.1"]]<-NULL
humanized_mouse_brain@meta.data[["seurat_clusters"]]<-NULL
obj@meta.data[["pANN_0.25_0.09_879"]]<-NULL
humanized_mouse_brain@meta.data[["pANN_0.25_0.09_867"]]<-NULL
```

filter away cells that have a very high mitocondrial content 

```{r}
humanized_mouse_brain <- subset(humanized_mouse_brain, subset = percent.mt < 5)
```


##2.Quality controls and cell selection

In this sections quality controls based selection of good will be performed on the merged object will be performed before proceeding with the rest of then analysis. 

1.Let's first plot some QCs metrics to have a look at the global dataset, this will give a general idea of the dataset quality

```{r}
# add the mitochondrial gene precentage 
# mouse mito (mm10)
humanized_mouse_brain[["percent.mt.mm10"]] <- PercentageFeatureSet(
  humanized_mouse_brain,
  pattern = "mt-"
)

# human mito (hg38)
humanized_mouse_brain[["percent.mt.hg38"]] <- PercentageFeatureSet(
  humanized_mouse_brain,
  pattern = "MT-"
)
humanized_mouse_brain[["percent.mt"]]<-humanized_mouse_brain[["percent.mt.mm10"]]+humanized_mouse_brain[["percent.mt.hg38"]]
```

check human vs mouse cells

```{r}
# Assume this is your merged object
obj <- humanized_mouse_brain
DefaultAssay(obj)<-"RNA"
obj<-JoinLayers(obj)
# Count UMIs per cell for each species (adjust prefixes as needed)
human_umis <- Matrix::colSums(
  GetAssayData(obj, assay = "RNA", layer = "counts")[grep("^GRCh38-", rownames(obj)), , drop = FALSE]
)
mouse_umis <-Matrix::colSums(
  GetAssayData(obj, assay = "RNA", layer = "counts")[grep("^mm10-", rownames(obj)), , drop = FALSE]
)

# Assign species label based on which count is higher
obj$species <- ifelse(human_umis > mouse_umis, "human", "mouse")

# Optionally mark ambiguous/mixed cells (e.g., 20–80% human)
ratio <- human_umis / (human_umis + mouse_umis)
obj$species[ratio > 0.2 & ratio < 0.8] <- "mixed"

# Write back
humanized_mouse_brain <- obj

#how many cells per sample
table(humanized_mouse_brain$orig.ident)

#how many cell per species
table(humanized_mouse_brain$species)

#how many cell per replicate
table(humanized_mouse_brain$sample)

```


```{r,fig.height=10, fig.width=10}
# Visualize QC metrics as a violin plot

QC_RNA <- data.frame(sample =  humanized_mouse_brain@meta.data[["sample"]],
                     Genes = humanized_mouse_brain@meta.data[["nFeature_RNA"]],
                     RNA = humanized_mouse_brain@meta.data[["nCount_RNA"]],
                     Mt = humanized_mouse_brain@meta.data[["percent.mt"]])
QC_RNA$sample <- as.factor(QC_RNA$sample)

# Function to add mean and median annotations for each sample above the violin plots
add_mean_median_annotation <- function(plot, data, x_var, y_var) {
  samples <- unique(data[[x_var]])
  for (sample in samples) {
    sample_data <- data[data[[x_var]] == sample, ]
    mean_value <- mean(sample_data[[y_var]], na.rm = TRUE)
    median_value <- median(sample_data[[y_var]], na.rm = TRUE)
    plot <- plot + 
      annotate("text", x = sample, y = max(sample_data[[y_var]], na.rm = TRUE) * 1.05, 
               label = paste("Mean:", round(mean_value, 2)), color = "blue", size = 5, fontface = "bold") +
      annotate("text", x = sample, y = max(sample_data[[y_var]], na.rm = TRUE) * 1.50, 
               label = paste("Median:", round(median_value, 2)), color = "red", size = 5, fontface = "bold")
  }
  return(plot)
}

# Plot for RNA
Transcripts <- ggplot(QC_RNA) +
  aes(x = sample, y = RNA, fill = sample, color = sample) +
  geom_violin(trim = TRUE, color = "#000000") +
  geom_boxplot(width = 0.1, color = "#000000", fill = "#ffffff", outlier.shape = NA) +
  scale_fill_manual(values = c(
"#1f77b4",  # blue
            "#ff7f0e",  # orange
            "#2ca02c",  # green
            "#d62728",  # red
            "#9467bd"
))+
  scale_color_manual(values = c(
"#1f77b4",  # blue
            "#ff7f0e",  # orange
            "#2ca02c",  # green
            "#d62728",  # red
            "#9467bd"
))+
  ggthemes::theme_base() +
  xlab("") +
  ylab("Number of UMI per cell") +
  theme(panel.border = element_rect()) +
  theme_classic() + theme(legend.position = "none") +
  theme(axis.text.x = element_text(size = 18, family = "Arial", angle = 45, hjust = 1),
        axis.text.y = element_text(size = 18, family = "Arial"),
        axis.title.y = element_text(size = 18, family = "Arial"),
        axis.line = element_line(size = 1))

Transcripts <- add_mean_median_annotation(Transcripts, QC_RNA, "sample", "RNA")

# Plot for Genes
Genes <- ggplot(QC_RNA) +
  aes(x = sample, y = Genes, fill = sample, color = sample) +
  geom_violin(trim = TRUE, color = "#000000") +
  geom_boxplot(width = 0.1, color = "#000000", fill = "#ffffff", outlier.shape = NA) +
  scale_fill_manual(values = c(
"#1f77b4",  # blue
            "#ff7f0e",  # orange
            "#2ca02c",  # green
            "#d62728",  # red
            "#9467bd"
))+
  scale_color_manual(values = c(
"#1f77b4",  # blue
            "#ff7f0e",  # orange
            "#2ca02c",  # green
            "#d62728",  # red
            "#9467bd"
))+
  ggthemes::theme_base() +
  xlab("") +
  ylab("Number of Genes per cell") +
  theme(panel.border = element_rect()) +
  theme_classic() + theme(legend.position = "none") +
  theme(axis.text.x = element_text(size = 18, family = "Arial", angle = 45, hjust = 1),
        axis.text.y = element_text(size = 18, family = "Arial"),
        axis.title.y = element_text(size = 18, family = "Arial"),
        axis.line = element_line(size = 1))

Genes <- add_mean_median_annotation(Genes, QC_RNA, "sample", "Genes")

# Plot for Mt
Mt <- ggplot(QC_RNA) +
  aes(x = sample, y =   Mt, fill = sample, color = sample) +
  geom_violin(trim = TRUE, color = "#000000") +
  geom_boxplot(width = 0.1, color = "#000000", fill = "#ffffff", outlier.shape = NA) +
  scale_fill_manual(values = c(
"#1f77b4",  # blue
            "#ff7f0e",  # orange
            "#2ca02c",  # green
            "#d62728",  # red
            "#9467bd"
))+
  scale_color_manual(values = c(
"#1f77b4",  # blue
            "#ff7f0e",  # orange
            "#2ca02c",  # green
            "#d62728",  # red
            "#9467bd"
))+
  ggthemes::theme_base() +
  xlab("") +
  ylab("Percentage of Mt genes") +
  theme(panel.border = element_rect()) +
  theme_classic() + theme(legend.position = "none") +
  theme(axis.text.x = element_text(size = 18, family = "Arial", angle = 45, hjust = 1),
        axis.text.y = element_text(size = 18, family = "Arial"),
        axis.title.y = element_text(size = 18, family = "Arial"),
        axis.line = element_line(size = 1))

Mt <- add_mean_median_annotation(Mt, QC_RNA, "sample", "Mt")

Transcripts
Genes
Mt

```

let's also visualize how is the relationship between counts, genes and mito content

```{r}

plot1 <- FeatureScatter(humanized_mouse_brain, feature1 = "nCount_RNA", feature2 = "percent.mt")
plot2 <- FeatureScatter(humanized_mouse_brain, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
plot1 + plot2
```







##3.Data integration, dimensionlity reduction and clustering

In this section we will perform the data integration, dimensionality reduction and clustering of the

##3.1 Dimensionality reduction and clustering without data integration
Despite in this experiment there are not multiple data batches and replicate we will first perform a clustering with data ingtegration and without and look at the difference

1.Data normalization, variable features finding PCA

```{r}
options(future.globals.maxSize = 100e+09)

humanized_mouse_brain[["SCT"]]<-NULL

# Log-normalized analysis on RNA
humanized_mouse_brain <- NormalizeData(humanized_mouse_brain, assay = "RNA")
humanized_mouse_brain <- FindVariableFeatures(humanized_mouse_brain, assay = "RNA")
humanized_mouse_brain <- ScaleData(humanized_mouse_brain, assay = "RNA")
humanized_mouse_brain <- RunPCA(humanized_mouse_brain, assay = "RNA", reduction.name = "pca_lognorm")

# SCT analysis on SCT assay
humanized_mouse_brain <- SCTransform(humanized_mouse_brain, verbose = TRUE)
humanized_mouse_brain <- FindVariableFeatures(humanized_mouse_brain, assay = "SCT")
humanized_mouse_brain <- RunPCA(humanized_mouse_brain, assay = "SCT", reduction.name = "pca_SCT")
```

2.Having a look at the elbow plot to understand how many PCs are really informative, this value will be used in the subsequent analysis steps 
```{r}
p1<-ElbowPlot(humanized_mouse_brain,reduction = "pca_SCT")
p1
```
After 15 SD the elbow plot becomes flat indicating that the majority of the information is explained by the first 15 dimension, this value will be used for the rest of the analysis.

3.Clustering without data integration

```{r}
humanized_mouse_brain <- FindNeighbors(humanized_mouse_brain, dims = 1:15, reduction = "pca_SCT")
humanized_mouse_brain <- FindClusters(humanized_mouse_brain, resolution = 0.1, cluster.name = "unintegrated_clusters")
```
4. Plot UMAP with samples and clustering
```{r,fig.height=10, fig.width=10}
humanized_mouse_brain <- RunUMAP(humanized_mouse_brain, dims = 1:15, reduction = "pca_SCT", reduction.name = "umap.unintegrated")
# visualize by batch and cell type annotation

my_colors <- c(
  "#1F77B4", "#FF7F0E", "#2CA02C", "#D62728", "#9467BD",
  "#8C564B", "#17BECF", "#BCBD22", "#7F7F7F", "#e377c2",
  "#aec7e8", "#ffbb78", "#98df8a"
)


p1<-DimPlot(humanized_mouse_brain, reduction = "umap.unintegrated", group.by = c("unintegrated_clusters"),cols = my_colors)+labs(x = "UMAP 1", y = "UMAP 2")+ #UMAP with samples
 theme_minimal() +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      size = 24,
      face = "bold"
    ),
    axis.title.x = element_text(size = 20),
    axis.title.y = element_text(size = 20),
    legend.text = element_text(size = 16),         # Legend labels
    legend.title = element_text(size = 24, face = "bold"),  # Legend title
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank()
  )

p2<-DimPlot(humanized_mouse_brain, reduction = "umap.unintegrated", group.by = c("orig.ident"))+labs(x = "UMAP 1", y = "UMAP 2")+ #UMAP with samples
 theme_minimal() +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      size = 24,
      face = "bold"
    ),
    axis.title.x = element_text(size = 20),
    axis.title.y = element_text(size = 20),
    legend.text = element_text(size = 16),         # Legend labels
    legend.title = element_text(size = 24, face = "bold"),  # Legend title
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank()
  )

p3<-DimPlot(humanized_mouse_brain, reduction = "umap.unintegrated", group.by = c("sample"))+labs(x = "UMAP 1", y = "UMAP 2")+ #UMAP with samples
 theme_minimal() +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      size = 24,
      face = "bold"
    ),
    axis.title.x = element_text(size = 20),
    axis.title.y = element_text(size = 20),
    legend.text = element_text(size = 16),         # Legend labels
    legend.title = element_text(size = 24, face = "bold"),  # Legend title
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank()
  )


p1 
p2
p3
saveRDS(humanized_mouse_brain,"/date/gcb/gcb_MZ/Humanized_mice_OD/humanized_mouse_brain_doublets.rds")
```

```{r,fig.height=10, fig.width=20}
colors_vector <- c(
  "CFA_brain_1" = "#E41A1C", 
  "CFA_brain_2" = "#377EB8", 
  "EAE_brain_1" = "#984EA3", 
  "EAE_brain_2" = "#FF7F00", 
  "EAE_brain_3" = "#FFFF33"
)

# Create the contingency table
t <- table(Cluster=humanized_mouse_brain$unintegrated_clusters, Batch=humanized_mouse_brain@meta.data[["sample"]])
t <- t[, rev(names(colors_vector))]

# Convert the table to a data frame
df <- as.data.frame(t)

# Calculate the percentage of each plate within each cluster
df <- df %>%
  group_by(Cluster) %>%mutate(Percent = Freq / sum(Freq) * 100)

# Plot the data
ggplot(df, aes(x = Cluster, y = Percent, fill = Batch)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_y_continuous(labels = percent_format(), expand = c(0, 0)) +
  scale_fill_manual(values = rev(colors_vector)) +
  theme_classic() +
  theme(
    axis.text.x = element_text( hjust = 1, vjust = 0.5, family = "Arial", size = 16),
    axis.text.y = element_text(family = "Arial", size = 16),
    axis.title.x = element_text(family = "Arial", size = 18),
    axis.title.y = element_text(family = "Arial", size = 18),
    legend.text = element_text(size = 16,family = "Arial"),
    legend.title = element_text(size = 18,family = "Arial")
  ) +
  labs(y = "Percentage of cells", x = "Cluster", fill = "Animal") +
  guides(fill = guide_legend(ncol = 3, title.position = "top", title.hjust = 0.5))
```







##3.2 Dimensionality reduction and clustering with data integration
In this section the dimensionality reduction will be performed including data integration and SCT transformation which should make the dataset more homogeneus and prevent any differences that are not due to biological variation.

1. SCT transformation, PCA running, clustering and dimensionality reduction again using 15 PCs as shown by the elbowplot above

```{r}

humanized_mouse_brain <- IntegrateLayers(
  object = humanized_mouse_brain,
  method = HarmonyIntegration,
  orig.reduction = "pca_SCT", new.reduction = "harmony",
  normalization.method = "SCT",
  verbose = F
)
humanized_mouse_brain <- FindNeighbors(humanized_mouse_brain, dims = 1:15, reduction = "harmony")
humanized_mouse_brain <- FindClusters(humanized_mouse_brain, resolution = 0.1,cluster.name = "harmony_clusters")
```

2. Perform Dimensionality reduction
```{r}
humanized_mouse_brain <- RunUMAP(humanized_mouse_brain, dims = 1:15, reduction = "harmony",reduction.name = "umap.harmony")
```

3.Plot UMAP with clusters and sample to check that the data integration effectively results in a better homogeneus dataset 

```{r}


pal13 <- pals::glasbey(13)         # 13 very distinct colors
pal13

p1<-DimPlot(humanized_mouse_brain, reduction = "umap.harmony", group.by = c("harmony_clusters"), cols = pal13)+labs(x = "UMAP 1", y = "UMAP 2")+theme_minimal()+
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
     plot.title = element_blank()

  )


p2<-DimPlot(humanized_mouse_brain, reduction = "umap.harmony", group.by = c("sample"))+labs(x = "UMAP 1", y = "UMAP 2")+
  ggtitle("") +  # Add your title here
 theme_minimal() +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      size = 24,
      face = "bold"
    ),
    axis.title.x = element_text(size = 20),
    axis.title.y = element_text(size = 20),
    legend.text = element_text(size = 16),         # Legend labels
    legend.title = element_text(size = 24, face = "bold"),  # Legend title
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank()
  )

col<- c(
  mouse = "grey80",
  human = "firebrick3"
)


p3<-DimPlot(humanized_mouse_brain, reduction = "umap.SCT", group.by = c("species"), cols = col)+labs(x = "UMAP 1", y = "UMAP 2")+
  ggtitle("") +  # Add your title here
 theme_minimal() +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      size = 24,
      face = "bold"
    ),
    axis.title.x = element_text(size = 20),
    axis.title.y = element_text(size = 20),
    legend.text = element_text(size = 16),         # Legend labels
    legend.title = element_text(size = 24, face = "bold"),  # Legend title
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank()
  )

p1
p2
p3
```


clusters proportions

```{r}
colors_vector <- c(
  "CFA_brain_1" = "#E41A1C", 
  "CFA_brain_2" = "#377EB8", 
  "EAE_brain_1" = "#984EA3", 
  "EAE_brain_2" = "#FF7F00", 
  "EAE_brain_3" = "#FFFF33"
)

# Create the contingency table
t <- table(Cluster=humanized_mouse_brain$harmony_clusters, Batch=humanized_mouse_brain@meta.data[["sample"]])
t <- t[, rev(names(colors_vector))]

# Convert the table to a data frame
df <- as.data.frame(t)

# Calculate the percentage of each plate within each cluster
df <- df %>%
  group_by(Cluster) %>%mutate(Percent = Freq / sum(Freq) * 100)

# Plot the data
ggplot(df, aes(x = Cluster, y = Percent, fill = Batch)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_y_continuous(labels = percent_format(), expand = c(0, 0)) +
  scale_fill_manual(values = rev(colors_vector)) +
  theme_classic() +
  theme(
    axis.text.x = element_text( hjust = 1, vjust = 0.5, family = "Arial", size = 16),
    axis.text.y = element_text(family = "Arial", size = 16),
    axis.title.x = element_text(family = "Arial", size = 18),
    axis.title.y = element_text(family = "Arial", size = 18),
    legend.text = element_text(size = 16,family = "Arial"),
    legend.title = element_text(size = 18,family = "Arial")
  ) +
  labs(y = "Percentage of cells", x = "Cluster", fill = "Animal") +
  guides(fill = guide_legend(ncol = 3, title.position = "top", title.hjust = 0.5))
```


7.Compute and save cluster markers of the different population in the MOL subclustering. This will be one of the supplementary tables to be attached to be included. 

```{r}

# find markers for every cluster compared to all remaining cells, report only the positive
# ones

DefaultAssay(humanized_mouse_brain)<-"SCT"
humanized_mouse_brain <- PrepSCTFindMarkers(humanized_mouse_brain, assay = "SCT",verbose = T)
humanized_mouse_brain_markers <- FindAllMarkers(humanized_mouse_brain, only.pos = TRUE)


 # Create a new workbook
wb <- createWorkbook()

# Loop through each cluster and add data to a new sheet
for (cluster_value in unique(humanized_mouse_brain_markers$cluster)) {
  cluster_data <- subset(humanized_mouse_brain_markers, cluster == cluster_value)
  sheet_name <- paste0("Cluster_", cluster_value)
  addWorksheet(wb, sheet_name)
  writeData(wb, sheet_name, cluster_data)
}

# Save the workbook to a file
saveWorkbook(wb, "/date/gcb/gcb_MZ/Humanized_mice_OD/Seurat_clusters_broad_brain.xlsx", overwrite = TRUE)
```

4.Now let's plot some markers genes to rename and homogenize clusters. After that the final steps of the analysis will be performed.The code use to generate the plots is used in the Supplementary Fig.S3C.


```{r,fig.height=80, fig.width=80}
markers <- c(
  # Cluster 0 - Cerebellar Granule Cells
  "mm10-Gabra6", "mm10-Cbln1", "mm10-Cbln3", "mm10-Neurod1", "mm10-Barhl2",
  # Cluster 1 - Excitatory Cortical Neurons (Glutamatergic)
  "mm10-Tbr1", "mm10-Satb2", "mm10-Emx1", "mm10-Neurod6",
  # Cluster 2 - GABAergic Interneurons (SST/VIP subtypes)
  "mm10-Lhx6", "mm10-Lhx8", "mm10-Sst", "mm10-Vip", "mm10-Tac2", "mm10-Slc5a7", "mm10-Nxph2",
  # Cluster 3 - GABAergic Inhibitory Neurons (Striatal/MGE-derived)
  "mm10-Dlx1", "mm10-Dlx2", "mm10-Gad1", "mm10-Sp9", "mm10-Meis2", "mm10-Pbx3",
  # Cluster 4 - Mature Oligodendrocytes
  "mm10-Plp1", "mm10-Mbp", "mm10-Apod", "mm10-Ptgds",
  # Cluster 5 - Astrocytes
  "mm10-Atp1a2", "mm10-Slc1a3", "mm10-Apoe", "mm10-Gja1", "mm10-Gli1", "mm10-Mfge8",
  # Cluster 6 - Medium Spiny Neurons (Striatal)
  "mm10-Drd1", "mm10-Drd2", "mm10-Adora2a", "mm10-Gpr88", "mm10-Rgs9", "mm10-Rarb",
  # Cluster 7 - Myelinating Oligodendrocytes
  "mm10-Klk6", "mm10-Gjc2", "mm10-Ugt8a", "mm10-Gjb1",
  # Cluster 8 - Cerebellar Inhibitory Interneurons (MLI/Pvalb+)
  "mm10-Pvalb", "mm10-Pax2", "mm10-Kit", "mm10-Tfap2a", "mm10-Tfap2b", "mm10-Skor1",
  # Cluster 9 - Microglia
  "mm10-Tmem119", "mm10-Hexb", "mm10-Trem2", "mm10-Tyrobp", "mm10-Inpp5d", "mm10-Itgam", "mm10-Cd14", "mm10-Adgre1",
  # Cluster 10 - Thalamic Neurons
  "mm10-Synpo2", "mm10-Zfhx3", "mm10-Shox2", "mm10-Lef1", "mm10-Slitrk6",
  # Cluster 11 - Oligodendrocyte Precursor Cells (OPCs)
  "mm10-Pdgfra", "mm10-Cspg4", "mm10-Vcan",
  # Cluster 12 - Purkinje Cells (Cerebellar)
  "mm10-Car8", "mm10-Pcp2", "mm10-Ppp1r17", "mm10-Col18a1", "mm10-Atp2a3"
)

# 1) create a list of plots (one per gene)
plots <- FeaturePlot(
  humanized_mouse_brain,
  features = markers,
  reduction = "umap.unintegrated",
  cols = c("lightgrey", "magenta"),
  combine = FALSE
)

# 2) apply consistent theme / title formatting to every plot
plots <- lapply(plots, function(p) {
  p +
    labs(x = "UMAP 1", y = "UMAP 2") +
    theme_minimal() +
    theme(
      plot.title = element_text(hjust = 0.5, size = 64, face = "bold"),
      axis.title.x = element_text(size = 50),
      axis.title.y = element_text(size = 50),
      legend.text = element_text(size = 50),
      legend.title = element_text(size = 52),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      axis.text.x = element_blank(),
      axis.text.y = element_blank()
    )
})

# 3) combine into a grid (e.g., 2 columns)
p <- wrap_plots(plots, ncol = 4)
p
```


5.Cluster renaming, based on the very broad classification OL, OPCs or microglia, this clustering to subset only MOLs to subsequently use them in the next analysis. The UMAP plot with the samples correponds to fig.3C and the UMAP plot with the broad cell types corresponds to Fig.S3B. 

```{r,fig.height=10, fig.width=10}
broad_clustering <- c("Cerebellar_Granule_Cells","Excitatory_Cortical_Neu","SST_Gaba_Neu","Stria_MGE_Inter_Neu","MOL",
                      "Astrocytes","Medium_Spiny_Neu_Stria","Myelin_MOL","Cerebellar_Inhib_Interneu_Pvalb+",
                      "Microglia","Thal_Neu","OPcs","Purkinje_Cells")

names(broad_clustering) <- levels(humanized_mouse_brain)
humanized_mouse_brain <- RenameIdents(humanized_mouse_brain, broad_clustering)
humanized_mouse_brain$broad_clustering <- Idents(humanized_mouse_brain) #create a new metadata colums with the broad clustering

# Check the new identities
table(Idents(humanized_mouse_brain))

my_colors <- c(
  "#1F77B4", "#FF7F0E", "#2CA02C", "#D62728", "#9467BD",
  "#8C564B", "#17BECF", "#BCBD22", "#7F7F7F", "#e377c2",
  "#aec7e8", "#ffbb78", "#98df8a"
)

p1<-DimPlot(humanized_mouse_brain, reduction = "umap.rpca_SCT",group.by = "orig.ident")+labs(x = "UMAP 1", y = "UMAP 2")+ #UMAP with samples
 theme_minimal() +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      size = 24,
      face = "bold"
    ),
    axis.title.x = element_text(size = 20),
    axis.title.y = element_text(size = 20),
    legend.text = element_text(size = 16),         # Legend labels
    legend.title = element_text(size = 24, face = "bold"),  # Legend title
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank()
  )

colors_3 <- c("red", "#56B4E9", "#009E73")  # select the colors you want for the UMAP

p2<-DimPlot(humanized_mouse_brain, reduction = "umap.unintegrated",group.by = "broad_clustering", cols = my_colors)+labs(x = "UMAP 1", y = "UMAP 2")+ #UMAP with samples
 theme_minimal() +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      size = 24,
      face = "bold"
    ),
    axis.title.x = element_text(size = 20),
    axis.title.y = element_text(size = 20),
    legend.text = element_text(size = 16),         # Legend labels
    legend.title = element_text(size = 24, face = "bold"),  # Legend title
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank()
  )


p2
```

Dotplot for the broad clustering 


```{r,fig.height=20, fig.width=27}

markers <- c(
  # Cluster 0 - Cerebellar Granule Cells
  "mm10-Gabra6", "mm10-Cbln1", "mm10-Cbln3", "mm10-Neurod1", "mm10-Barhl2",
  # Cluster 1 - Excitatory Cortical Neurons (Glutamatergic)
  "mm10-Tbr1", "mm10-Satb2", "mm10-Emx1", "mm10-Neurod6",
  # Cluster 2 - GABAergic Interneurons (SST/VIP subtypes)
  "mm10-Lhx6", "mm10-Lhx8", "mm10-Sst", "mm10-Vip", "mm10-Tac2", "mm10-Slc5a7", "mm10-Nxph2",
  # Cluster 3 - GABAergic Inhibitory Neurons (Striatal/MGE-derived)
  "mm10-Dlx1", "mm10-Dlx2", "mm10-Gad1", "mm10-Sp9", "mm10-Meis2", "mm10-Pbx3",
  # Cluster 4 - Mature Oligodendrocytes
  "mm10-Plp1", "mm10-Mbp", "mm10-Apod", "mm10-Ptgds",
  # Cluster 5 - Astrocytes
  "mm10-Atp1a2", "mm10-Slc1a3", "mm10-Apoe", "mm10-Gja1", "mm10-Gli1", "mm10-Mfge8",
  # Cluster 6 - Medium Spiny Neurons (Striatal)
  "mm10-Drd1", "mm10-Drd2", "mm10-Adora2a", "mm10-Gpr88", "mm10-Rgs9", "mm10-Rarb",
  # Cluster 7 - Myelinating Oligodendrocytes
  "mm10-Klk6", "mm10-Gjc2", "mm10-Ugt8a", "mm10-Gjb1",
  # Cluster 8 - Cerebellar Inhibitory Interneurons (MLI/Pvalb+)
  "mm10-Pvalb", "mm10-Pax2", "mm10-Kit", "mm10-Tfap2a", "mm10-Tfap2b", "mm10-Skor1",
  # Cluster 9 - Microglia
  "mm10-Tmem119", "mm10-Hexb", "mm10-Trem2", "mm10-Tyrobp", "mm10-Inpp5d", "mm10-Itgam", "mm10-Cd14", "mm10-Adgre1",
  # Cluster 10 - Thalamic Neurons
  "mm10-Synpo2", "mm10-Zfhx3", "mm10-Shox2", "mm10-Lef1", "mm10-Slitrk6",
  # Cluster 11 - Oligodendrocyte Precursor Cells (OPCs)
  "mm10-Pdgfra", "mm10-Cspg4", "mm10-Vcan",
  # Cluster 12 - Purkinje Cells (Cerebellar)
  "mm10-Car8", "mm10-Pcp2", "mm10-Ppp1r17", "mm10-Col18a1", "mm10-Atp2a3"
)

p1<-DotPlot(humanized_mouse_brain, features = markers,group.by = "broad_clustering",cols = c("lightgrey", "firebrick"))+
  xlab("")+
  scale_x_discrete(labels = markers) + 
  theme(
    axis.text.x = element_text( hjust = 1, vjust = 0.5, family = "Arial", size = 24,angle = 90),
    axis.text.y = element_text(family = "Arial", size = 20),
    axis.title.y = element_text(family = "Arial", size =24),
    legend.text = element_text(size = 20,family = "Arial"),
    legend.title = element_text(size = 24,family = "Arial")
  ) 
p1
```





##4.MOL subsetting, and reclustering of MOLs only 

##4.1 Select only MOLs from the whole dataset

With this line of code we will create a new Seurat object that will contain only MOLs that we will use for the subsequents analysis steps

```{r}



OL <- subset(humanized_mouse_brain, idents = c("Myelin_MOL","MOL","OPcs"))            


```

##4.2 Reclustering of MOLs

1.Renormalize the datasets and recluster using only MOL cells   

```{r}
DefaultAssay(OL)<-"RNA"

# Remove the old SCT assay 
OL[["SCT"]] <- NULL

# Now run SCTransform renormalizing with only the OL_OPC cells 
OL <- SCTransform(OL, vst.flavor = "v2", verbose = FALSE)

# Continue with custom-named reductions
OL <- RunPCA(OL, verbose = T, reduction.name = "pca_OL_SCT")


OL<- FindNeighbors(OL, reduction = "pca_OL_SCT", dims = 1:15, 
                      graph.name = c("OL_nn", "OL_snn"))

OL <- FindClusters(OL, resolution = 0.1, 
                     graph.name = "OL_snn",
                     cluster.name = "OL_clusters")

OL <- RunUMAP(OL, reduction = "pca_OL_SCT", dims = 1:15, 
                reduction.name = "umap_OL")


```

2.Plot UMAP with samples and the New MOLs clusters

```{r}

my_colors <- c(
  "#1F77B4",
  "#FF7F0E",
  "#2CA02C",
  "#D62728",
  "#9467BD",
  "#8C564B",
  "#17BECF"
)

p1<-DimPlot(OL, reduction = "umap_OL", group.by = c("OL_clusters"), cols=my_colors)+labs(x = "UMAP 1", y = "UMAP 2")+ #UMAP with new clustering
 theme_minimal() +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      size = 24,
      face = "bold"
    ),
    axis.title.x = element_text(size = 20),
    axis.title.y = element_text(size = 20),
    legend.text = element_text(size = 16),         # Legend labels
    legend.title = element_text(size = 24, face = "bold"),  # Legend title
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank()
  )

p2<-DimPlot(OL, reduction = "umap_OL", group.by = c("orig.ident"))+labs(x = "UMAP 1", y = "UMAP 2")+ #UMAP with samples
 theme_minimal() +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      size = 24,
      face = "bold"
    ),
    axis.title.x = element_text(size = 20),
    axis.title.y = element_text(size = 20),
    legend.text = element_text(size = 16),         # Legend labels
    legend.title = element_text(size = 24, face = "bold"),  # Legend title
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank()
  )

p3<-DimPlot(OL, reduction = "umap_OL", group.by = c("sample"))+labs(x = "UMAP 1", y = "UMAP 2")+ #UMAP with samples
 theme_minimal() +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      size = 24,
      face = "bold"
    ),
    axis.title.x = element_text(size = 20),
    axis.title.y = element_text(size = 20),
    legend.text = element_text(size = 16),         # Legend labels
    legend.title = element_text(size = 24, face = "bold"),  # Legend title
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank()
  )

p4<-DimPlot(OL, reduction = "umap_OL", group.by = c("broad_clustering"))+labs(x = "UMAP 1", y = "UMAP 2")+ #UMAP with samples
 theme_minimal() +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      size = 24,
      face = "bold"
    ),
    axis.title.x = element_text(size = 20),
    axis.title.y = element_text(size = 20),
    legend.text = element_text(size = 16),         # Legend labels
    legend.title = element_text(size = 24, face = "bold"),  # Legend title
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank()
  )

p1
p2
p3
p4
```







clusters proportions

```{r}
colors_vector <- c(
  "CFA_brain_1" = "#E41A1C", 
  "CFA_brain_2" = "#377EB8", 
  "EAE_brain_1" = "#984EA3", 
  "EAE_brain_2" = "#FF7F00", 
  "EAE_brain_3" = "#FFFF33"
)

# Create the contingency table
t <- table(Cluster=OL$OL_clusters, Batch=OL@meta.data[["sample"]])
t <- t[, rev(names(colors_vector))]

# Convert the table to a data frame
df <- as.data.frame(t)

# Calculate the percentage of each plate within each cluster
df <- df %>%
  group_by(Cluster) %>%mutate(Percent = Freq / sum(Freq) * 100)

# Plot the data
ggplot(df, aes(x = Cluster, y = Percent, fill = Batch)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_y_continuous(labels = percent_format(), expand = c(0, 0)) +
  scale_fill_manual(values = rev(colors_vector)) +
  theme_classic() +
  theme(
    axis.text.x = element_text( hjust = 1, vjust = 0.5, family = "Arial", size = 16),
    axis.text.y = element_text(family = "Arial", size = 16),
    axis.title.x = element_text(family = "Arial", size = 18),
    axis.title.y = element_text(family = "Arial", size = 18),
    legend.text = element_text(size = 16,family = "Arial"),
    legend.title = element_text(size = 18,family = "Arial")
  ) +
  labs(y = "Percentage of cells", x = "Cluster", fill = "Animal") +
  guides(fill = guide_legend(ncol = 3, title.position = "top", title.hjust = 0.5))
```

3.Plotting marker genes to review the clustering and rename the different populations of MOLs

```{r,fig.height=20, fig.width=20}

markers <- c(
  #MOL5/6–EAE-b
  "mm10-Serpina3n","mm10-Nlrc5",
  #MOL5/6–EAE-a
  "mm10-Plin4","mm10-Sult1a1","mm10-Zfand4",
  #MOL1/2 EAE
  "mm10-Klk8","mm10-Itga8", "mm10-Tlr3",
  #MHC-II
  "mm10-H2-Aa", "mm10-H2-Ab1", "mm10-H2-Eb1", "mm10-Cd74", "mm10-Ctss",
  #interferon
  "mm10-Ifih1", "mm10-Irf7", "mm10-Irgm1", "mm10-Irgm2", "mm10-Igtp","mm10-Zbp1",
   #cycling_DAO
  "mm10-Cdkn1a","mm10-Ccnd1","mm10-Nav3","mm10-Tmsb10","mm10-Lrrfip1",
  # OPC1
  "mm10-Rph3a", "mm10-Adora2b","mm10-Pdgfra","mm10-Cspg4",
  # OPC2
  "mm10-Phyhd1","mm10-Fcgr2b","mm10-Mylk",
  # OPC3
  "mm10-Slc14a1","mm10-Lgals1","mm10-Myrf",
  #MOL 5/6
  "mm10-Ptgds", "mm10-Opalin", "mm10-Qdpr", "mm10-Il33", "mm10-Apoe", "mm10-Cd63", "mm10-Jph4","mm10-Fgfr2"
  
)

p1<-DotPlot(OL, features = markers,group.by = "OL_clusters",cols = c("lightgrey", "firebrick"))+
  xlab("")+
  scale_x_discrete(labels = markers) + 
  theme(
    axis.text.x = element_text( hjust = 1, vjust = 0.5, family = "Arial", size = 24,angle = 90),
    axis.text.y = element_text(family = "Arial", size = 20),
    axis.title.y = element_text(family = "Arial", size =24),
    legend.text = element_text(size = 20,family = "Arial"),
    legend.title = element_text(size = 24,family = "Arial")
  ) 
p1
```


```{r,fig.height=20, fig.width=20}

markers <- c(
  #MOL1
  "mm10-Fosb","mm10-Dusp1","mm10-Dnajb1","mm10-Btg2","mm10-Hspa1a","mm10-Hspa1b","mm10-Tob1","mm10-Mal",
  #MOL2
  "mm10-Anxa5","mm10-Klk6","mm10-Mgst3","mm10-S100b","mm10-Tppp3","mm10-Cyp51","mm10-Plekhb1",
  # MFOL 1/2
  "mm10-Ctps","mm10-Tmem141","mm10-Opalin","mm10-Ttyh1", "mm10-Plekha1", "mm10-Ddr1", "mm10-Cd9", "mm10-Sirt2","mm10-Arpc1b",
  # NFOL1/2
  "mm10-Sema6a","mm10-Mag","mm10-Mobp","mm10-Mbp","mm10-Rras2","mm10-Prom1"

  
)

p1<-DotPlot(OL, features = markers,group.by = "OL_clusters",cols = c("lightgrey", "firebrick"))+
  xlab("")+
  scale_x_discrete(labels = markers) + 
  theme(
    axis.text.x = element_text( hjust = 1, vjust = 0.5, family = "Arial", size = 24,angle = 90),
    axis.text.y = element_text(family = "Arial", size = 20),
    axis.title.y = element_text(family = "Arial", size =24),
    legend.text = element_text(size = 20,family = "Arial"),
    legend.title = element_text(size = 24,family = "Arial")
  ) 
p1
```

Marker features calculations in OLs

```{r}

# find markers for every cluster compared to all remaining cells, report only the positive
# ones

DefaultAssay(OL)<-"SCT"
OL<- PrepSCTFindMarkers(OL, assay = "SCT",verbose = T)
OL_markers<- FindAllMarkers(OL, only.pos = TRUE)


 # Create a new workbook
wb <- createWorkbook()

# Loop through each cluster and add data to a new sheet
for (cluster_value in unique(OL_markers$cluster)) {
  cluster_data <- subset(OL_markers, cluster == cluster_value)
  sheet_name <- paste0("Cluster_", cluster_value)
  addWorksheet(wb, sheet_name)
  writeData(wb, sheet_name, cluster_data)
}

# Save the workbook to a file
saveWorkbook(wb, "/date/gcb/gcb_MZ/Humanized_mice_OD/Seurat_clusters_OLs.xlsx", overwrite = TRUE)
```

Plotting cluster Markers



```{r,fig.height=20, fig.width=30}



markers <- c(
  # Cluster 0 - Myelinating Oligodendrocytes (MOL)
  "mm10-Ugt8a","mm10-Car2","mm10-Cdh20","mm10-Rtn4","mm10-Rnf220","mm10-Sox2ot","mm10-St18","mm10-Ptprd","mm10-Plp1","mm10-Nkain2",
  # Cluster 1 - Mature Oligodendrocytes (MOL) - homeostatic/metabolic
  "mm10-Garnl3","mm10-Mt1","mm10-Zbtb16","mm10-Polr3e","mm10-Slc6a1","mm10-Slc25a13","mm10-Otud7a","mm10-Mast3","mm10-Elmo1","mm10-Plcl1",
  # Cluster 2 - Mature Oligodendrocytes (MOL) - Apod/Ptgds-high
  "mm10-Apod","mm10-Ptgds","mm10-Phyhd1","mm10-Rsrp1","mm10-Edil3","mm10-Acaca","mm10-Neat1","mm10-St6galnac3","mm10-Anln","mm10-Hif3a",
  # Cluster 3 - Neuronal contamination (Cerebellar Granule Cells)
  "mm10-Kcnip4","mm10-Meg3","mm10-Chn2","mm10-Car10","mm10-Fgf14","mm10-Kcnd2","mm10-Cntnap2","mm10-Rbfox1","mm10-Negr1","mm10-Nrxn1",
  # Cluster 4 - OPCs / Committed OL Precursors (COP)
  "mm10-Lhfpl3","mm10-Xylt1","mm10-Tnr","mm10-Sgcd","mm10-Dscam","mm10-Adarb2","mm10-Ptprz1","mm10-Lrrtm4","mm10-Csmd1","mm10-Kcnip1",
  # Cluster 5 - OPCs (early/proliferative)
  "mm10-Nxph1","mm10-Gpc5","mm10-Sox5","mm10-Sox6","mm10-Slc35f1","mm10-Pcdh7","mm10-Ptprt","mm10-Luzp2","mm10-Nav3","mm10-Pdgfra",
  # Cluster 6 - Neuronal contamination (GABAergic/Glutamatergic)
  "mm10-Grin2b","mm10-Kcnb2","mm10-Atp1b1","mm10-Dclk1","mm10-Tenm3","mm10-Epha6","mm10-Kalrn","mm10-Gria1","mm10-Dlgap2","mm10-Etl4"
)

  


p1<-DotPlot(OL, features = markers,group.by = "OL_clusters",cols = c("lightgrey", "firebrick"))+
  xlab("")+
  scale_x_discrete(labels = markers) + 
  theme(
    axis.text.x = element_text( hjust = 1, vjust = 0.5, family = "Arial", size = 24,angle = 90),
    axis.text.y = element_text(family = "Arial", size = 20),
    axis.title.y = element_text(family = "Arial", size =24),
    legend.text = element_text(size = 20,family = "Arial"),
    legend.title = element_text(size = 24,family = "Arial")
  ) 
p1
```









1.Compute the GO terms significantly enriched

```{r}
# marker genes computed above

pos_markers <- OL_markers%>%
  dplyr::select(cluster, gene)

# convert symbols to ENTREZID per cluster
marker_list <- split(pos_markers$gene, pos_markers$cluster)

# convert once (union) and map back
all_genes <- unique(unlist(marker_list))

all_genes_clean <- sub("^mm10-", "", all_genes)

gene_map <- bitr(all_genes_clean, fromType = "SYMBOL", toType = "ENTREZID", OrgDb = org.Mm.eg.db)
sym2ent <- setNames(gene_map$ENTREZID, gene_map$SYMBOL)

marker_list_clean <- lapply(marker_list, function(x) sub("^mm10-", "", x))

entrez_by_cluster <- lapply(marker_list_clean, function(x) na.omit(unname(sym2ent[x])))

ego_list <- lapply(names(entrez_by_cluster), function(cl){
  enrichGO(gene          = entrez_by_cluster[[cl]],
           OrgDb         = org.Mm.eg.db,
           keyType       = "ENTREZID",
           ont           = "BP",
           pAdjustMethod = "BH",
           pvalueCutoff  = 0.05,
           qvalueCutoff  = 0.2,
           readable      = TRUE)
}) #list of enriched terms per cluster
names(ego_list) <- names(entrez_by_cluster)

```


2.Create an Excel file to save the results, to be attached as supplementary material

```{r}

# Create a new workbook
wb <- createWorkbook()

# Loop through each cluster and add enrichment results as a sheet
for (cluster_name in names(ego_list)) {
  
  # Extract results from enrichResult object
  if (!is.null(ego_list[[cluster_name]]) && nrow(ego_list[[cluster_name]]@result) > 0) {
    
    # Get the results as a data frame
    go_results <- as.data.frame(ego_list[[cluster_name]])
    
    # Create sheet name (Excel has 31 char limit for sheet names)
    sheet_name <- substr(cluster_name, 1, 31)
    
    # Add worksheet
    addWorksheet(wb, sheet_name)
    
    # Write data
    writeData(wb, sheet_name, go_results)
    
    cat(sprintf("Added sheet for cluster: %s (%d GO terms)\n", 
                cluster_name, nrow(go_results)))
    
  } else {
    cat(sprintf("No significant GO terms for cluster: %s\n", cluster_name))
  }
}

# Save the workbook
saveWorkbook(wb, 
             "/date/gcb/gcb_MZ/Humanized_mice_OD/GO_enrichment_by_cluster_OL.xlsx", 
             overwrite = TRUE)

cat("\n✓ GO enrichment results saved to Excel file!\n")

```

3.Plot the functional enrichment results results for GO terms


```{r,fig.height=20, fig.width=40}
# Collect GO BP results into one data.frame

ego_df <- map2_df(names(ego_list), ego_list, ~{
  if (is.null(.y) || nrow(as.data.frame(.y)) == 0) return(NULL)
  df <- as.data.frame(.y)
  df$cluster <- .x
  df$mLog10Padj <- -log10(df$p.adjust)
  df
})

# Choose top terms per cluster and plot

plot_df <- ego_df %>%
  group_by(cluster) %>%
  slice_max(order_by = mLog10Padj, n = 10, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(Description = fct_reorder(Description, mLog10Padj))

p1<-ggplot(plot_df, aes(x = cluster, y = Description, size = Count, color = mLog10Padj)) +
  geom_point() +
  scale_color_gradient(low = "gold", high = "red") +
  labs(title = "GO BP enrichment per cluster", x = "Cluster", y = "Term", color = "-log10(FDR)") +
  theme_bw(base_size = 12) + theme_classic()+
  theme(
    axis.text.x = element_text( hjust = 1, vjust = 0.5, family = "Arial", size = 24,angle = 45),
    axis.text.y = element_text(family = "Arial", size = 24),
    axis.title.y = element_text(family = "Arial", size =28, face = "bold"),
    axis.title.x = element_text(family = "Arial", size =28, face = "bold"),
    legend.text = element_text(size = 24,family = "Arial"),
    legend.title = element_text(size = 32,family = "Arial",face = "bold")
  ) 
plots <- list(Func_enrich_GO=p1)
for (nm in names(plots)) {
ggplot2::ggsave(file.path("/date/gcb/gcb_MZ/MOL_FGF_10x/Plots/", paste0(nm, ".png")), plots[[nm]], width =14, height = 16, dpi = 300)
}
p1

```
##8.2 Functional enrichment analysis using KEGG pathways
Functional enrichment using Kegg pathways

1.Kegg pathways analysis

```{r}
ekegg_list <- lapply(names(entrez_by_cluster), function(cl){
  enrichKEGG(gene          = entrez_by_cluster[[cl]],
             organism      = "mmu",
             pvalueCutoff  = 0.05)
})
names(ekegg_list) <- names(entrez_by_cluster)


kegg_df <- map2_df(names(ekegg_list), ekegg_list, ~{
  if (is.null(.y) || nrow(as.data.frame(.y)) == 0) return(NULL)
  df <- as.data.frame(.y)
  df$cluster <- .x
  df$mLog10Padj <- -log10(df$p.adjust)
  df
})
```

2.Create an Excel file to save the results, to be attached as supplementary material

```{r}

# Save to Excel - Option 1: One sheet per cluster
wb_kegg <- createWorkbook()

for (cluster_name in names(ekegg_list)) {
  
  if (!is.null(ekegg_list[[cluster_name]]) && nrow(ekegg_list[[cluster_name]]@result) > 0) {
    
    # Get results as data frame
    kegg_results <- as.data.frame(ekegg_list[[cluster_name]])
    
    # Add mLog10Padj column
    kegg_results$mLog10Padj <- -log10(kegg_results$p.adjust)
    
    # Create sheet name
    sheet_name <- substr(cluster_name, 1, 31)
    
    # Add worksheet and write data
    addWorksheet(wb_kegg, sheet_name)
    writeData(wb_kegg, sheet_name, kegg_results)
    
    cat(sprintf("Added sheet for cluster: %s (%d KEGG pathways)\n", 
                cluster_name, nrow(kegg_results)))
    
  } else {
    cat(sprintf("No significant KEGG pathways for cluster: %s\n", cluster_name))
  }
}

# Save workbook with one sheet per cluster
saveWorkbook(wb_kegg, 
             "/date/gcb/gcb_MZ/Humanized_mice_OD/KEGG_enrichment_by_cluster_OL.xlsx", 
             overwrite = TRUE)

cat("\n✓ KEGG enrichment results saved (one sheet per cluster)!\n")

```

3.Plot the results

```{r,fig.height=20, fig.width=40}
plot_df_kegg <- kegg_df %>%
  group_by(cluster) %>%
  slice_max(order_by = mLog10Padj, n = 10, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(Description = fct_reorder(Description, mLog10Padj))

p1<-ggplot(plot_df_kegg, aes(x = cluster, y = Description, size = Count, color = mLog10Padj)) +
  geom_point() +
  scale_color_gradient(low = "gold", high = "red") +
  labs(title = "KEGG enrichment per cluster", x = "Cluster", y = "Pathway", color = "-log10(FDR)") +
  theme_bw(base_size = 12)+theme_classic()+
  theme(
    axis.text.x = element_text( hjust = 1, vjust = 0.5, family = "Arial", size = 24,angle = 45),
    axis.text.y = element_text(family = "Arial", size = 24),
    axis.title.y = element_text(family = "Arial", size =28, face = "bold"),
    axis.title.x = element_text(family = "Arial", size =28, face = "bold"),
    legend.text = element_text(size = 24,family = "Arial"),
    legend.title = element_text(size = 32,family = "Arial",face = "bold")
  ) 
plots <- list(Func_enrich_KEGG=p1)
for (nm in names(plots)) {
ggplot2::ggsave(file.path("/date/gcb/gcb_MZ/MOL_FGF_10x/Plots/", paste0(nm, ".png")), plots[[nm]], width =14, height = 16, dpi = 300)
}
p1

```

1.Perform label transfer between the MOL dataset and the Pandey_dataset

```{r}
DefaultAssay(OL)<-"RNA"
OL[["RNA"]] <- JoinLayers(OL[["RNA"]])
```

```{r}
# Load and capture the original object name
loaded_name <- load("/date/gcb/gcb_MZ/MOL_FGF_10x/oligos_final_pub.RObj") #Pandey Dataset can be donwloaded from Zenodo

# Assign it to DA_humanized_mouse_brain using the captured name
DA_humanized_mouse_brain <- get(loaded_name[1])

DA_humanized_mouse_brain_v5 <- UpdateSeuratObject(object =DA_humanized_mouse_brain)
DA_humanized_mouse_brain_v5@version

OL[["RNA"]] <- as(OL[["RNA"]], "Assay5")
OL[["dummy"]] <- CreateAssayObject(counts = GetAssayData(OL, assay = "RNA", layer = "counts")[1:10, ])

transfer_anchors <- FindTransferAnchors(
  reference = DA_humanized_mouse_brain_v5,
  query = OL,
  reference.assay = "RNA",
  query.assay = "RNA",
  reduction = "cca",
  npcs = 50,
  dims = 1:50,
  features = VariableFeatures(OL@assays[["RNA"]])
)

OL[["dummy"]] <- NULL
  

OL<- MapQuery(
  anchorset =transfer_anchors, 
  reference = DA_humanized_mouse_brain_v5,
  query =  OL,
  refdata = "finalInterps")
#rename predictions columns 
cn <- colnames(OL@meta.data)
cn[cn == "predicted.id"] <- "predicted.id_DA_humanized_mouse_brain"
cn[cn == "predicted.id.score"] <- "predicted.id.score_DA_humanized_mouse_brain"
colnames(OL@meta.data) <- cn
```
2.Plot the UMAP with clusters renamed based on the label transfer results and the similarity score of each cluster of MOLs with the cluster from Pandey et al. both to be inserted in the main figure of the scRNA

```{r}
# Combine palettes to get 23 distinct colors

colors<-  c(
  "#0072B2", "#E69F00", "green", "#D55E00", "#CC79A7",
  "red", "#F0E442", "#000000", "#332288", "#88CCEE",
  "#44AA99", "#117733", "#999933", "#DDCC77", "#661100",
  "#AA4499", "#882255"
)

p1<-DimPlot(MOL ,reduction = "umap_MOL",group.by = "predicted.id_DA_humanized_mouse_brain",cols = colors)+
  labs(x = "UMAP 1", y = "UMAP 2") +
  ggtitle("DA humanized_mouse_brains Clusters") +
  theme_minimal() +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      size = 24,
      face = "bold"
    ),
    axis.title.x = element_text(size = 20),
    axis.title.y = element_text(size = 20),
    legend.text = element_text(size = 16),         # Legend labels
    legend.title = element_text(size = 24),  # Legend title
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank()
  )


plots <- list(umap_MOL_predicted.id_DA_humanized_mouse_brain=p1)
for (nm in names(plots)) {
ggplot2::ggsave(file.path("/date/gcb/gcb_MZ/MOL_FGF_10x/Plots/", paste0(nm, ".png")), plots[[nm]], width =12, height = 8, dpi = 300)
}
p1
rm(DA_humanized_mouse_brain)
rm(DA_humanized_mouse_brain_v5)
gc()
```

3.Violin plot of each cluster similarity score. This code takes the prediction score associated to each cell in the MOL cluster and plot the predicted value 

```{r}
# Get cluster identities (your 5 populations)
clusters <- unique(MOL$MOL_fine_clustering)
print(paste("Creating plots for", length(clusters), "clusters"))

# Loop through each cluster
for (cluster_id in clusters) {
  
  # Create empty data frame to store all scores
  all_scores <- data.frame()
  
  # Get the prediction.score.id assay
  assay_data <- GetAssayData(MOL, assay = "prediction.score.id", slot = "data")
  
  # Get cells in this cluster
  cluster_cells <- colnames(MOL)[MOL$MOL_fine_clustering == cluster_id]
  
  # Extract scores for each of the 17 reference cell types
  for (ref_celltype in rownames(assay_data)) {
    scores <- assay_data[ref_celltype, cluster_cells]
    
    temp_df <- data.frame(
      score = as.numeric(scores),
      reference_celltype = ref_celltype,
      stringsAsFactors = FALSE
    )
    
    all_scores <- rbind(all_scores, temp_df)
  }
  
  # Create violin plot with all 17 reference cell types on x-axis
  p <- ggplot(all_scores, aes(x = reference_celltype, y = score, fill = reference_celltype)) +
    geom_violin(trim = TRUE, scale = "width") +
    geom_boxplot(width = 0.1, fill = "white", outlier.shape = NA) +
    ggtitle(paste("Cluster:", cluster_id)) +
    xlab("Reference Cell Type") +
    ylab("Similarity Score") +
    ylim(0, 1) +
    theme_classic() +
    theme(
      plot.title = element_text(hjust = 0.5, size = 36, face = "bold"),
      axis.text.x = element_text(hjust = 1, vjust = 0.5, family = "Arial", 
                                 size = 14, angle = 90),  # Smaller for 17 labels
      axis.text.y = element_text(family = "Arial", size = 20),
      axis.title.x = element_text(family = "Arial", size = 24, face = "bold"),
      axis.title.y = element_text(family = "Arial", size = 24, face = "bold"),
      legend.position = "none"
    )
  
   # Display plot in the R Markdown output
  print(p)
  
  # Save plot
  filename <- file.path("/date/gcb/gcb_MZ/MOL_FGF_10x/Plots/", 
                       paste0("Similarity_cluster_DAO_humanized_mouse_brain", cluster_id, ".png"))
  ggplot2::ggsave(filename, p, width = 24, height = 12, dpi = 300)
  
  print(paste("✓ Saved plot for cluster:", cluster_id))
}

```

Feature plot MOL

```{r,fig.height=80, fig.width=60}
markers <- c("mm10-Klk8","mm10-Itga8", "mm10-Tlr3", "mm10-Trim34a", #MOL1/2 EAE
             "mm10-Plin4","mm10-Sult1a1","mm10-Zfand4", #MOL5/6–EAE-a
             "mm10-S100a10","mm10-Rnase4","mm10-Tnfrsf1a",#MOL5/6–EAE-b
             "mm10-Serpina3n","mm10-Nlrc5","mm10-Igtp","mm10-Fgfr2","mm10-B2m", #DA OL
             "mm10-Piezo2","mm10-Masp1","mm10-C4b","mm10-Klk8","mm10-Il33", #DAO_1
             "mm10-Psmb8","mm10-Stat1","mm10-H2-T23","mm10-Psmb9", #IFN_DAO
             "mm10-Cdkn1a","mm10-Ccnd1","mm10-Nav3","mm10-Tmsb10","mm10-Lrrfip1" #cycling_DAO
             )
#Code to plot them and save them 
for (gene in markers) {
  p <- FeaturePlot(
    OL, 
    features = gene, 
    reduction = "umap_OL",
    cols = c("lightgrey", "magenta")
  )+
  labs(x = "UMAP 1", y = "UMAP 2") +
  ggtitle( paste0(gene)) +
  theme_minimal() +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      size =48,
      face = "bold"
    ),
    axis.title.x = element_text(size = 26),
    axis.title.y = element_text(size = 26),
    legend.text = element_text(size = 24),         # Legend labels
    legend.title = element_text(size = 28),  # Legend title
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank())
 print(p)
  # Save each plot
  #filename <- file.path("/date/gcb/gcb_MZ/MOL_FGF_10x/Plots/", paste0("Feature_plot_humanized_mouse_brain_", gene, ".png"))
  #ggplot2::ggsave(filename, p, width = 20, height = 12, dpi = 300)
}
#code to plot the markers here in the vignette

# 1) create a list of plots (one per gene)
plots <- FeaturePlot(
  OL,
  features = markers,
  reduction = "umap_OL",
  cols = c("lightgrey", "magenta"),
  combine = FALSE
)

# 2) apply consistent theme / title formatting to every plot
plots <- lapply(plots, function(p) {
  p +
    labs(x = "UMAP 1", y = "UMAP 2") +
    theme_minimal() +
    theme(
      aspect.ratio = 1,
      plot.title = element_text(hjust = 0.5, size = 64, face = "bold"),
      axis.title.x = element_text(size = 50),
      axis.title.y = element_text(size = 50),
      legend.text = element_text(size = 50),
      legend.title = element_text(size = 52),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      axis.text.x = element_blank(),
      axis.text.y = element_blank()
    )
})

# 3) combine into a grid (e.g., 2 columns)
p <- wrap_plots(plots, ncol = 4)
p
```

##4.1 Select only MOLs from the whole dataset

With this line of code we will create a new Seurat object that will contain only MOLs that we will use for the subsequents analysis steps

```{r}


Microglia <- subset(humanized_mouse_brain, idents = c("Microglia"))            


```

##4.2 Reclustering of MOLs

1.Renormalize the datasets and recluster using only MOL cells   

```{r}
DefaultAssay(Microglia)<-"RNA"

# Remove the Microgliad SCT assay 
Microglia[["SCT"]] <- NULL

# Now run SCTransform renormalizing with only the Microglia_OPC cells 
Microglia <- SCTransform(Microglia, vst.flavor = "v2", verbose = FALSE)

# Continue with custom-named reductions
Microglia <- RunPCA(Microglia, verbose = T, reduction.name = "pca_Microglia_SCT")


Microglia<- FindNeighbors(Microglia, reduction = "pca_Microglia_SCT", dims = 1:15, 
                      graph.name = c("Microglia_nn", "Microglia_snn"))

Microglia <- FindClusters(Microglia, resolution = 0.1, 
                     graph.name = "Microglia_snn",
                     cluster.name = "Microglia_clusters")

Microglia <- RunUMAP(Microglia, reduction = "pca_Microglia_SCT", dims = 1:15, 
                reduction.name = "umap_Microglia")


```

```{r}
p1<-DimPlot(Microglia, reduction = "umap_Microglia", group.by = c("Microglia_clusters"))+labs(x = "UMAP 1", y = "UMAP 2")+ #UMAP with new clustering
 theme_minimal() +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      size = 24,
      face = "bold"
    ),
    axis.title.x = element_text(size = 20),
    axis.title.y = element_text(size = 20),
    legend.text = element_text(size = 16),         # Legend labels
    legend.title = element_text(size = 24, face = "bold"),  # Legend title
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank()
  )

p2<-DimPlot(Microglia, reduction = "umap_Microglia", group.by = c("orig.ident"))+labs(x = "UMAP 1", y = "UMAP 2")+ #UMAP with samples
 theme_minimal() +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      size = 24,
      face = "bold"
    ),
    axis.title.x = element_text(size = 20),
    axis.title.y = element_text(size = 20),
    legend.text = element_text(size = 16),         # Legend labels
    legend.title = element_text(size = 24, face = "bold"),  # Legend title
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank()
  )

p3<-DimPlot(Microglia, reduction = "umap_Microglia", group.by = c("sample"))+labs(x = "UMAP 1", y = "UMAP 2")+ #UMAP with samples
 theme_minimal() +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      size = 24,
      face = "bold"
    ),
    axis.title.x = element_text(size = 20),
    axis.title.y = element_text(size = 20),
    legend.text = element_text(size = 16),         # Legend labels
    legend.title = element_text(size = 24, face = "bold"),  # Legend title
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank()
  )

p4<-DimPlot(Microglia, reduction = "umap_Microglia", group.by = c("broad_clustering"))+labs(x = "UMAP 1", y = "UMAP 2")+ #UMAP with samples
 theme_minimal() +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      size = 24,
      face = "bold"
    ),
    axis.title.x = element_text(size = 20),
    axis.title.y = element_text(size = 20),
    legend.text = element_text(size = 16),         # Legend labels
    legend.title = element_text(size = 24, face = "bold"),  # Legend title
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank()
  )

p1
p2
p3
p4
```

Marker features calculations in OLs

```{r}

# find markers for every cluster compared to all remaining cells, report only the positive
# ones

DefaultAssay(Microglia)<-"SCT"
Microglia<- PrepSCTFindMarkers(Microglia, assay = "SCT",verbose = T)
Microglia_markers<- FindAllMarkers(Microglia, only.pos = TRUE)


 # Create a new workbook
wb <- createWorkbook()

# Loop through each cluster and add data to a new sheet
for (cluster_value in unique(Microglia_markers$cluster)) {
  cluster_data <- subset(Microglia_markers, cluster == cluster_value)
  sheet_name <- paste0("Cluster_", cluster_value)
  addWorksheet(wb, sheet_name)
  writeData(wb, sheet_name, cluster_data)
}

# Save the workbook to a file
saveWorkbook(wb, "/date/gcb/gcb_MZ/Humanized_mice_OD/Seurat_clusters_Microglia.xlsx", overwrite = TRUE)
```

Plotting cluster Markers



```{r,fig.height=20, fig.width=30}



markers <- c(
  # Cluster 0 - Neuronal contamination (Cerebellar Granule Cells)
  "mm10-Kcnip4","mm10-Kcnd2","mm10-Fgf14","mm10-Ralyl","mm10-Nrxn1","mm10-Rims1","mm10-Rbfox1","mm10-Tenm1","mm10-Car10","mm10-Negr1",
  # Cluster 1 - Homeostatic Microglia
  "mm10-Plxdc2","mm10-Tanc2","mm10-Csmd3","mm10-Nav2","mm10-Srgap2","mm10-Zfhx3","mm10-Rapgef5","mm10-Tgfbr1","mm10-Elmo1","mm10-Frmd4a",
  # Cluster 2 - Neuronal contamination (GABAergic/Glutamatergic)
  "mm10-Lrrc7","mm10-Grin2b","mm10-Kcnh7","mm10-Tenm3","mm10-Grm5","mm10-Epha6","mm10-Meis2","mm10-Myo16","mm10-Sntg1","mm10-Cpne4",
  # Cluster 3 - Oligodendrocyte contamination
  "mm10-Prr5l","mm10-Sec14l5","mm10-Slc24a2","mm10-Edil3","mm10-Arap2","mm10-Efnb3","mm10-Mog","mm10-D7Ertd443e","mm10-Mag","mm10-Ppp1r16b",
  # Cluster 4 - Activated/Phagocytic Microglia (DAM-like)
  "mm10-Arhgap24","mm10-Fkbp5","mm10-Gcnt2","mm10-C1qa","mm10-Ssh2","mm10-Cd244a","mm10-C1qb","mm10-Man2b1","mm10-Cd300lf","mm10-Npl",
  # Cluster 5 - Microglia - metabolically active / stressed
  "mm10-Cst3","mm10-Psap","mm10-mt-Co3","mm10-mt-Atp6","mm10-mt-Co2","mm10-mt-Co1","mm10-Mef2a","mm10-Snx24","mm10-Inpp5d","mm10-Rps27a"
)
  


p1<-DotPlot(OL, features = markers,group.by = "OL_clusters",cols = c("lightgrey", "firebrick"))+
  xlab("")+
  scale_x_discrete(labels = markers) + 
  theme(
    axis.text.x = element_text( hjust = 1, vjust = 0.5, family = "Arial", size = 24,angle = 90),
    axis.text.y = element_text(family = "Arial", size = 20),
    axis.title.y = element_text(family = "Arial", size =24),
    legend.text = element_text(size = 20,family = "Arial"),
    legend.title = element_text(size = 24,family = "Arial")
  ) 
p1
```
1.Compute the GO terms significantly enriched

```{r}
# marker genes computed above

pos_markers <- Microglia_markers%>%
  dplyr::select(cluster, gene)

# convert symbMicroglias to ENTREZID per cluster
marker_list <- split(pos_markers$gene, pos_markers$cluster)

# convert once (union) and map back
all_genes <- unique(unlist(marker_list))

all_genes_clean <- sub("^mm10-", "", all_genes)

gene_map <- bitr(all_genes_clean, fromType = "SYMBOL", toType = "ENTREZID", OrgDb = org.Mm.eg.db)
sym2ent <- setNames(gene_map$ENTREZID, gene_map$SYMBOL)

marker_list_clean <- lapply(marker_list, function(x) sub("^mm10-", "", x))

entrez_by_cluster <- lapply(marker_list_clean, function(x) na.omit(unname(sym2ent[x])))

ego_list <- lapply(names(entrez_by_cluster), function(cl){
  enrichGO(gene          = entrez_by_cluster[[cl]],
           OrgDb         = org.Mm.eg.db,
           keyType       = "ENTREZID",
           ont           = "BP",
           pAdjustMethod = "BH",
           pvalueCutoff  = 0.05,
           qvalueCutoff  = 0.2,
           readable      = TRUE)
}) #list of enriched terms per cluster
names(ego_list) <- names(entrez_by_cluster)

```


2.Create an Excel file to save the results, to be attached as supplementary material

```{r}

# Create a new workbook
wb <- createWorkbook()

# Loop through each cluster and add enrichment results as a sheet
for (cluster_name in names(ego_list)) {
  
  # Extract results from enrichResult object
  if (!is.null(ego_list[[cluster_name]]) && nrow(ego_list[[cluster_name]]@result) > 0) {
    
    # Get the results as a data frame
    go_results <- as.data.frame(ego_list[[cluster_name]])
    
    # Create sheet name (Excel has 31 char limit for sheet names)
    sheet_name <- substr(cluster_name, 1, 31)
    
    # Add worksheet
    addWorksheet(wb, sheet_name)
    
    # Write data
    writeData(wb, sheet_name, go_results)
    
    cat(sprintf("Added sheet for cluster: %s (%d GO terms)\n", 
                cluster_name, nrow(go_results)))
    
  } else {
    cat(sprintf("No significant GO terms for cluster: %s\n", cluster_name))
  }
}

# Save the workbook
saveWorkbook(wb, 
             "/date/gcb/gcb_MZ/Humanized_mice_OD/GO_enrichment_by_cluster_Microglia.xlsx", 
             overwrite = TRUE)

cat("\n✓ GO enrichment results saved to Excel file!\n")

```


3.Plot the functional enrichment results results for GO terms


```{r,fig.height=20, fig.width=40}
# Collect GO BP results into one data.frame

ego_df <- map2_df(names(ego_list), ego_list, ~{
  if (is.null(.y) || nrow(as.data.frame(.y)) == 0) return(NULL)
  df <- as.data.frame(.y)
  df$cluster <- .x
  df$mLog10Padj <- -log10(df$p.adjust)
  df
})

# Choose top terms per cluster and plot

plot_df <- ego_df %>%
  group_by(cluster) %>%
  slice_max(order_by = mLog10Padj, n = 10, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(Description = fct_reorder(Description, mLog10Padj))

p1<-ggplot(plot_df, aes(x = cluster, y = Description, size = Count, color = mLog10Padj)) +
  geom_point() +
  scale_color_gradient(low = "gold", high = "red") +
  labs(title = "GO BP enrichment per cluster", x = "Cluster", y = "Term", color = "-log10(FDR)") +
  theme_bw(base_size = 12) + theme_classic()+
  theme(
    axis.text.x = element_text( hjust = 1, vjust = 0.5, family = "Arial", size = 24,angle = 45),
    axis.text.y = element_text(family = "Arial", size = 24),
    axis.title.y = element_text(family = "Arial", size =28, face = "bold"),
    axis.title.x = element_text(family = "Arial", size =28, face = "bold"),
    legend.text = element_text(size = 24,family = "Arial"),
    legend.title = element_text(size = 32,family = "Arial",face = "bold")
  ) 
plots <- list(Func_enrich_GO=p1)
for (nm in names(plots)) {
ggplot2::ggsave(file.path("/date/gcb/gcb_MZ/MOL_FGF_10x/Plots/", paste0(nm, ".png")), plots[[nm]], width =14, height = 16, dpi = 300)
}
p1

```






7.Compute and save cluster markers of the different population in the MOL subclustering. This will be one of the supplementary tables to be attached to be included. 

```{r}

# find markers for every cluster compared to all remaining cells, report only the positive
# ones

DefaultAssay(MOL)<-"SCT"
MOL <- PrepSCTFindMarkers(MOL, assay = "SCT",verbose = T)
MOL_markers <- FindAllMarkers(MOL, only.pos = TRUE)


 # Create a new workbook
wb <- createWorkbook()

# Loop through each cluster and add data to a new sheet
for (cluster_value in unique(MOL_markers$cluster)) {
  cluster_data <- subset(MOL_markers, cluster == cluster_value)
  sheet_name <- paste0("Cluster_", cluster_value)
  addWorksheet(wb, sheet_name)
  writeData(wb, sheet_name, cluster_data)
}

# Save the workbook to a file
saveWorkbook(wb, "/date/gcb/gcb_MZ/MOL_FGF_10x/Tables/Seurat_clusters_MOL_scRNA_10x.xlsx", overwrite = TRUE)
```

##5.Comparison with other datasets of DAOs and EAE
In this section the MOL dataset will be compared with other datasets in which DAOs have been described to define which type are induced by FGF. 

##5.1 Label transfer with panDAO paper Pandey et al., Cell Reports, 2022 
The first dataset that will be compared to our MOLs is a panDAO datasets cross studies which includes DAO from different type of disease model. 

1.Perform label transfer between the MOL dataset and the Pandey_dataset

```{r}
DefaultAssay(MOL)<-"RNA"
MOL[["RNA"]] <- JoinLayers(MOL[["RNA"]])
```

```{r}
# Load and capture the original object name
loaded_name <- load("/proj/user/mattia/humanized_mouse_brains_final_pub.RObj") #Pandey Dataset can be donwloaded from Zenodo

# Assign it to DA_humanized_mouse_brain using the captured name
DA_humanized_mouse_brain <- get(loaded_name[1])

DA_humanized_mouse_brain_v5 <- UpdateSeuratObject(object =DA_humanized_mouse_brain)
DA_humanized_mouse_brain_v5@version

transfer_anchors <- FindTransferAnchors(reference =DA_humanized_mouse_brain_v5, query = MOL,
                                            reference.assay = "RNA",
                                        reduction = "cca",
                                        npcs = 50,
                                        dims = 1:50,
  query.assay = "RNA",
  features = VariableFeatures(MOL@assays[["RNA"]]),)
  

MOL<- MapQuery(
  anchorset =transfer_anchors, 
  reference = DA_humanized_mouse_brain_v5,
  query =  MOL,
  refdata = "finalInterps")
#rename predictions columns 
cn <- colnames(MOL@meta.data)
cn[cn == "predicted.id"] <- "predicted.id_DA_humanized_mouse_brain"
cn[cn == "predicted.id.score"] <- "predicted.id.score_DA_humanized_mouse_brain"
colnames(MOL@meta.data) <- cn
```
2.Plot the UMAP with clusters renamed based on the label transfer results and the similarity score of each cluster of MOLs with the cluster from Pandey et al. both to be inserted in the main figure of the scRNA

```{r}
# Combine palettes to get 23 distinct colors

colors<-  c(
  "#0072B2", "#E69F00", "green", "#D55E00", "#CC79A7",
  "red", "#F0E442", "#000000", "#332288", "#88CCEE",
  "#44AA99", "#117733", "#999933", "#DDCC77", "#661100",
  "#AA4499", "#882255"
)

p1<-DimPlot(MOL ,reduction = "umap_MOL",group.by = "predicted.id_DA_humanized_mouse_brain",cols = colors)+
  labs(x = "UMAP 1", y = "UMAP 2") +
  ggtitle("DA humanized_mouse_brains Clusters") +
  theme_minimal() +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      size = 24,
      face = "bold"
    ),
    axis.title.x = element_text(size = 20),
    axis.title.y = element_text(size = 20),
    legend.text = element_text(size = 16),         # Legend labels
    legend.title = element_text(size = 24),  # Legend title
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank()
  )


plots <- list(umap_MOL_predicted.id_DA_humanized_mouse_brain=p1)
for (nm in names(plots)) {
ggplot2::ggsave(file.path("/date/gcb/gcb_MZ/MOL_FGF_10x/Plots/", paste0(nm, ".png")), plots[[nm]], width =12, height = 8, dpi = 300)
}
p1
rm(DA_humanized_mouse_brain)
rm(DA_humanized_mouse_brain_v5)
gc()
```

3.Violin plot of each cluster similarity score. This code takes the prediction score associated to each cell in the MOL cluster and plot the predicted value 

```{r}
# Get cluster identities (your 5 populations)
clusters <- unique(MOL$MOL_fine_clustering)
print(paste("Creating plots for", length(clusters), "clusters"))

# Loop through each cluster
for (cluster_id in clusters) {
  
  # Create empty data frame to store all scores
  all_scores <- data.frame()
  
  # Get the prediction.score.id assay
  assay_data <- GetAssayData(MOL, assay = "prediction.score.id", slot = "data")
  
  # Get cells in this cluster
  cluster_cells <- colnames(MOL)[MOL$MOL_fine_clustering == cluster_id]
  
  # Extract scores for each of the 17 reference cell types
  for (ref_celltype in rownames(assay_data)) {
    scores <- assay_data[ref_celltype, cluster_cells]
    
    temp_df <- data.frame(
      score = as.numeric(scores),
      reference_celltype = ref_celltype,
      stringsAsFactors = FALSE
    )
    
    all_scores <- rbind(all_scores, temp_df)
  }
  
  # Create violin plot with all 17 reference cell types on x-axis
  p <- ggplot(all_scores, aes(x = reference_celltype, y = score, fill = reference_celltype)) +
    geom_violin(trim = TRUE, scale = "width") +
    geom_boxplot(width = 0.1, fill = "white", outlier.shape = NA) +
    ggtitle(paste("Cluster:", cluster_id)) +
    xlab("Reference Cell Type") +
    ylab("Similarity Score") +
    ylim(0, 1) +
    theme_classic() +
    theme(
      plot.title = element_text(hjust = 0.5, size = 36, face = "bold"),
      axis.text.x = element_text(hjust = 1, vjust = 0.5, family = "Arial", 
                                 size = 14, angle = 90),  # Smaller for 17 labels
      axis.text.y = element_text(family = "Arial", size = 20),
      axis.title.x = element_text(family = "Arial", size = 24, face = "bold"),
      axis.title.y = element_text(family = "Arial", size = 24, face = "bold"),
      legend.position = "none"
    )
  
   # Display plot in the R Markdown output
  print(p)
  
  # Save plot
  filename <- file.path("/date/gcb/gcb_MZ/MOL_FGF_10x/Plots/", 
                       paste0("Similarity_cluster_DAO_humanized_mouse_brain", cluster_id, ".png"))
  ggplot2::ggsave(filename, p, width = 24, height = 12, dpi = 300)
  
  print(paste("✓ Saved plot for cluster:", cluster_id))
}

```

##5.2 Module score calculation using the same gene module used in Pandey et al.
In this section using the same gene set from Pandey et al., their enrichment into the MOL clusters will be performed. 

1.Extract the module from the supplementary material of Pandey et al., this code is structured in a way that uniforms the gene names since in the dataset from the paper some gene names are different than the ones used in the MOLs data

```{r}

# Read markers
DA_markers <- read.csv2("/proj/user/mattia/Markers_Cell_rep_2022.csv")

# Extract UPREGULATED genes for each signature
genes_DA1_up <- DA_markers %>% 
  dplyr::filter(DAO_fine_upregulatedGenes == "DA1") %>% 
  dplyr::select(symbol)

genes_DA2_up <- DA_markers %>% 
  dplyr::filter(DAO_fine_upregulatedGenes == "DA2") %>% 
  dplyr::select(symbol)

genes_IFN_up <- DA_markers %>% 
  dplyr::filter(DAO_fine_upregulatedGenes == "DA1_IFN") %>% 
  dplyr::select(symbol)

# Extract DOWNREGULATED genes for each signature
genes_DA1_down <- DA_markers %>% 
  dplyr::filter(DAO_fine_downregulatedGenes == "DA1") %>% 
  dplyr::select(symbol)

genes_DA2_down <- DA_markers %>% 
  dplyr::filter(DAO_fine_downregulatedGenes == "DA2") %>% 
  dplyr::select(symbol)

genes_IFN_down <- DA_markers %>% 
  dplyr::filter(DAO_fine_downregulatedGenes == "DA1_IFN") %>% 
  dplyr::select(symbol)

# Create lists with both up and down genes
sets <- list(
  DA1 = list(up = genes_DA1_up$symbol, down = genes_DA1_down$symbol),
  DA2 = list(up = genes_DA2_up$symbol, down = genes_DA2_down$symbol),
  IFN = list(up = genes_IFN_up$symbol, down = genes_IFN_down$symbol)
)

# Replacement dictionary for gene name corrections
repl <- c(
  "Cac1h"   = "Cacna1h",
  "Kcb1"    = "Kcnb1",
  "Plx1"    = "Plxna1",
  "Serpi3n" = "Serpina3n",
  "Rse4"    = "Rgs4",
  "prt"     = "Prt",
  "Sp91"    = "Spp1",
  "v3"      = "Vcan"
)

# Function to apply replacements to a character vector
apply_replacements <- function(x, map) {
  i <- match(x, names(map), nomatch = 0L)
  x[which(i > 0L)] <- map[i[i > 0L]]
  unique(x)
}

# Apply replacements to NESTED lists (up and down separately)
sets_fixed <- lapply(sets, function(gene_set) {
  list(
    up = apply_replacements(gene_set$up, repl),
    down = apply_replacements(gene_set$down, repl)
  )
})

# Filter to keep only genes present in your Seurat object
present_only <- function(v, obj) intersect(v, rownames(obj))

sets_fixed_present <- lapply(sets_fixed, function(gene_set) {
  list(
    up = present_only(gene_set$up, MOL),
    down = present_only(gene_set$down, MOL)
  )
})

# Inspect what was changed - FIXED VERSION
changed <- lapply(names(sets), function(nm) {
  # Process up-regulated genes
  old_up <- sets[[nm]]$up
  new_up <- sets_fixed[[nm]]$up
  changed_up <- old_up[old_up %in% names(repl)]
  
  # Process down-regulated genes
  old_down <- sets[[nm]]$down
  new_down <- sets_fixed[[nm]]$down
  changed_down <- old_down[old_down %in% names(repl)]
  
  # Combine both
  df_up <- if(length(changed_up) > 0) {
    data.frame(
      set = nm,
      direction = "up",
      from = changed_up,
      to = repl[changed_up],
      stringsAsFactors = FALSE
    )
  } else NULL
  
  df_down <- if(length(changed_down) > 0) {
    data.frame(
      set = nm,
      direction = "down",
      from = changed_down,
      to = repl[changed_down],
      stringsAsFactors = FALSE
    )
  } else NULL
  
  rbind(df_up, df_down)
})

changed <- do.call(rbind, changed)
print("Gene name changes:")
print(changed)

# Check missing genes
missing_in_obj <- lapply(names(sets_fixed), function(nm) {
  list(
    up = setdiff(sets_fixed[[nm]]$up, rownames(MOL)),
    down = setdiff(sets_fixed[[nm]]$down, rownames(MOL))
  )
})
names(missing_in_obj) <- names(sets_fixed)

print("\nGenes missing in object:")
print(missing_in_obj)

```


2. To calculate the gene set score the same way as in Pandey et al. an ad hoc function need to be created, to apply the same methodology to perform the calculation. 


3.Apply the funciton and calculate the score
```{r}
DefaultAssay(MOL) <- "RNA"
MOL[["RNA"]] <- JoinLayers(MOL[["RNA"]])
# Calculate signed scores for each gene set
cat("Calculating DA1 signed score:\n")
MOL$DA1_score <- calculate_signed_score(
  MOL, 
  geneset_up = sets$DA1$up,
  geneset_down = sets$DA1$down,
  assay = "RNA"
)

cat("\nCalculating DA2 signed score:\n")
MOL$DA2_score <- calculate_signed_score(
  MOL, 
  geneset_up = sets$DA2$up,
  geneset_down = sets$DA2$down,
  assay = "RNA"
)

cat("\nCalculating IFN signed score:\n")
MOL$IFN_score <- calculate_signed_score(
  MOL, 
  geneset_up = sets$IFN$up,
  geneset_down = sets$IFN$down,
  assay = "RNA"
)


```

4.Plot the score that should go in figure 3. In this code the enrichment for each set is plotted separately


```{r}

# Select the grouping variable used in the plot (e.g., cluster_renamed)
grp <- "MOL_fine_clustering"  # or "seurat_clusters" or your custom label

# Panel C: DA1-enriched gene set score across OL subtypes
pC <- ggplot(MOL@meta.data, aes_string(x = grp, y = "DA1_score", color = grp)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6) +
  geom_jitter(width = 0.2, size = 0.7, alpha = 0.5) +
  theme_classic(base_family = "Arial") +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 24),
    axis.text.y = element_text(size = 24),
    axis.title.y = element_text(size = 28, face = "bold"),
    legend.position = "none"
  ) +
  labs(x = NULL, y = "MOL5/6_DA1 enriched gene set score")

# Panel D: DA2-enriched
pD <- ggplot(MOL@meta.data, aes_string(x = grp, y = "DA2_score", color = grp)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6) +
  geom_jitter(width = 0.2, size = 0.7, alpha = 0.5) +
  theme_classic(base_family = "Arial") +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 24),
    axis.text.y = element_text(size = 24),
    axis.title.y = element_text(size = 28, face = "bold"),
    legend.position = "none"
  ) +
  labs(x = NULL, y = "MOL5/6_DA2 enriched gene set score")

# Panel E: IFN-enriched
pE <- ggplot(MOL@meta.data, aes_string(x = grp, y = "IFN_score", color = grp)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6) +
  geom_jitter(width = 0.2, size = 0.7, alpha = 0.5) +
  theme_classic(base_family = "Arial") +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 24),
    axis.text.y = element_text(size = 24),
    axis.title.y = element_text(size = 28, face = "bold"),
    legend.position = "none"
  ) +
  labs(x = NULL, y = "MOL5/6_IFN enriched gene set score")



plots <- list(MOL5_6_DA1=pC,MOL5_6_DA2=pD,MOL5_6_DAIFN=pE)
for (nm in names(plots)) {
ggplot2::ggsave(file.path("/date/gcb/gcb_MZ/MOL_FGF_10x/Plots/", paste0(nm, ".png")), plots[[nm]], width =20, height = 12, dpi = 300)
}
pC
pD
pE
```
##6 Comparison with EAE DAO from Nat. Med., 2018
This section is for the comparison between the simlarity of the MOLs with the EAE study.

1.Label transfer between the EAE dataset and the MOL

1.1 in case a subset of the EAE_humanized_mouse_brain before doing the label transfer is disired it can be done like this

```{r}
EAE_humanized_mouse_brain<-readRDS("/proj/tmp/tmp_EA/MZ_Zuchero/EAE_SS2_NatMed2018_seuratv5.rds")

EAE_humanized_mouse_brain_subset <- subset(humanized_mouse_brain, idents = c("MOL_1_2_EAE","MOL_2_Cntrl_A","MOL_3_EAE","MOL_6_Cntrl_A","MOL_6_Cntrl_B",
                                             "MOL_5_6_EAE_A","MOL_5_6_EAE_B","MOL_5_6_EAE_C"))  
```

```{r}

#EAE_humanized_mouse_brain<-readRDS("/proj/tmp/tmp_EA/MZ_Zuchero/EAE_SS2_NatMed2018_seuratv5.rds")

transfer_anchors <- FindTransferAnchors(reference =EAE_humanized_mouse_brain, query = MOL,
                                            reference.assay = "RNA",
                                        reduction = "cca",
                                        npcs = 50,
                                        dims = 1:50,
  query.assay = "RNA",
  features = VariableFeatures(MOL@assays[["RNA"]]),)
  

MOL<- MapQuery(
  anchorset =transfer_anchors, 
  reference = EAE_humanized_mouse_brain,
  query =  MOL,
  refdata = "Renamed_clusternames")

#rename predictions columns 
cn <- colnames(MOL@meta.data)
cn[cn == "predicted.id"] <- "predicted.id_EAE_humanized_mouse_brain"
cn[cn == "predicted.id.score"] <- "predicted.id.score_EAE_humanized_mouse_brain"
colnames(MOL@meta.data) <- cn

```
2. UMAP plot of predicted clusters from EAE dataset

```{r,fig.height=8, fig.width=14}


# Combine palettes to get 23 distinct colors
colors <- c(brewer.pal(12, "Paired"), brewer.pal(11, "Set3"))
colors<- c("#A6CEE3","#1F78B4","#B2DF8A","#33A02C","black","#E31A1C","#FDBF6F","#FF7F00","#CAB2D6","#6A3D9A","#FFFF99","#B15928","#8DD3C7","#FFFFB3",
           "#BEBADA","#FB8072","#80B1D3","gold","#B3DE69","#FCCDE5","#D9D9D9","#BC80BD","#CCEBC5")
p1<-DimPlot(MOL ,reduction = "umap_MOL",group.by = "predicted.id_EAE_humanized_mouse_brain",cols = colors)+
  labs(x = "UMAP 1", y = "UMAP 2") +
  ggtitle("DA MOLs Clusters") +
  theme_minimal() +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      size = 24,
      face = "bold"
    ),
    axis.title.x = element_text(size = 20),
    axis.title.y = element_text(size = 20),
    legend.text = element_text(size = 16),         # Legend labels
    legend.title = element_text(size = 24),  # Legend title
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank()
  )

plots <- list(EAE_MOL_label_transfer=p1)
for (nm in names(plots)) {
ggplot2::ggsave(file.path("/date/gcb/gcb_MZ/MOL_FGF_10x/Plots/", paste0(nm, ".png")), plots[[nm]], width =14, height = 8, dpi = 300)
}
p1

```

3.Violin plot of each cluster similarity score. This code takes the prediction score associated to each cell in the MOL cluster and plot the predicted value. Same as the code used for Pandey et al. dataset

```{r}
# Get cluster identities (your 5 populations)
clusters <- unique(MOL$MOL_fine_clustering)
print(paste("Creating plots for", length(clusters), "clusters"))

# Loop through each cluster
for (cluster_id in clusters) {
  
  # Create empty data frame to store all scores
  all_scores <- data.frame()
  
  # Get the prediction.score.id assay
  assay_data <- GetAssayData(MOL, assay = "prediction.score.id", slot = "data")
  
  # Get cells in this cluster
  cluster_cells <- colnames(MOL)[MOL$MOL_fine_clustering == cluster_id]
  
  # Extract scores for each of the 17 reference cell types
  for (ref_celltype in rownames(assay_data)) {
    scores <- assay_data[ref_celltype, cluster_cells]
    
    temp_df <- data.frame(
      score = as.numeric(scores),
      reference_celltype = ref_celltype,
      stringsAsFactors = FALSE
    )
    
    all_scores <- rbind(all_scores, temp_df)
  }
  
  # Create violin plot with all 17 reference cell types on x-axis
  p <- ggplot(all_scores, aes(x = reference_celltype, y = score, fill = reference_celltype)) +
    geom_violin(trim = TRUE, scale = "width") +
    geom_boxplot(width = 0.1, fill = "white", outlier.shape = NA) +
    ggtitle(paste("Cluster:", cluster_id)) +
    xlab("Reference Cell Type") +
    ylab("Similarity Score") +
    ylim(0, 1) +
    theme_classic() +
    theme(
      plot.title = element_text(hjust = 0.5, size = 36, face = "bold"),
      axis.text.x = element_text(hjust = 1, vjust = 0.5, family = "Arial", 
                                 size = 14, angle = 90),  # Smaller for 17 labels
      axis.text.y = element_text(family = "Arial", size = 20),
      axis.title.x = element_text(family = "Arial", size = 24, face = "bold"),
      axis.title.y = element_text(family = "Arial", size = 24, face = "bold"),
      legend.position = "none"
    )
  
   # Display plot in the R Markdown output
  print(p)
  
  # Save plot
  filename <- file.path("/date/gcb/gcb_MZ/MOL_FGF_10x/Plots/", 
                       paste0("Similarity_cluster_EAE_dataset", cluster_id, ".png"))
  ggplot2::ggsave(filename, p, width = 24, height = 12, dpi = 300)
  
  print(paste("✓ Saved plot for cluster:", cluster_id))
}

```

3.Calculating a gene module score for MOL_5_6_EAE_B in the MOL FGF dataset

3.1 First we upload the 

```{r}

# By sheet name
EAE_humanized_mouse_brain_diff_gene <- read_excel("/date/gcb/gcb_MZ/MOL_FGF_10x//EAE_humanized_mouse_brain_Modules.xlsx", sheet = "DifferentialExpressionOLs") %>% 
  dplyr::rename(Gene=1)

EAE_humanized_mouse_brain_diff_gene

```

3.2 Let's select the Top100 gene differentially expressed MOL_5_6_EAE_B and use them as module to calculate the score, the top 100 genes up and downregulated will be selected since we need both to perform the module score evaluation like it has been done above


```{r}


# Extract top 100 upregulated gene names
MOL_5_6_EAE_up <- EAE_humanized_mouse_brain_diff_gene %>%
  filter(!is.na(diffclustersMOL5_6_EAE_B)) %>%
  slice_max(diffclustersMOL5_6_EAE_B, n = 100) %>%
  pull(Gene)  # or pull(1) if ...1 is the first column

# Extract top 100 downregulated gene names
MOL_5_6_EAE_down <- EAE_humanized_mouse_brain_diff_gene %>%
  filter(!is.na(diffclustersMOL5_6_EAE_B)) %>%
  slice_min(diffclustersMOL5_6_EAE_B, n = 100) %>%
  pull(Gene)

# Create the set of up and downregulated genes to submit to the function 

# Create lists with both up and down genes
sets <- list(
  MOL_5_6_EAE_B = list(up = MOL_5_6_EAE_up, down = MOL_5_6_EAE_down)
)

```

3.3 Now let's use the same method use on the Pandey et al. dataset to calculate the module score


```{r}
DefaultAssay(MOL) <- "RNA"
MOL[["RNA"]] <- JoinLayers(MOL[["RNA"]])
# Calculate signed scores for each gene set
cat("Calculating MOL_5_6_EAE_B signed score:\n")
MOL$MOL_5_6_EAE_B_score <- calculate_signed_score(
  MOL, 
  geneset_up = sets$MOL_5_6_EAE_B$up,
  geneset_down = sets$MOL_5_6_EAE_B$down,
  assay = "RNA"
)


```
3.4 Plot the signature score 

```{r,fig.height=8, fig.width=8}
# Select the grouping variable used in the plot (e.g., cluster_renamed)
grp <- "MOL_fine_clustering"  # or "seurat_clusters" or your custom label

# Panel C: DA1-enriched gene set score across OL subtypes
pC <- ggplot(MOL@meta.data, aes_string(x = grp, y = "MOL_5_6_EAE_B_score", color = grp)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6) +
  geom_jitter(width = 0.2, size = 0.7, alpha = 0.5) +
  theme_classic(base_family = "Arial") +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 24),
    axis.text.y = element_text(size = 24),
    axis.title.y = element_text(size = 28, face = "bold"),
    legend.position = "none"
  ) +
  labs(x = NULL, y = "MOL_5_6_EAE_B enriched gene set score")


plots <- list(MOL_5_6_EAE_B_score=pC)
for (nm in names(plots)) {
ggplot2::ggsave(file.path("/date/gcb/gcb_MZ/MOL_FGF_10x//Plots/", paste0(nm, ".png")), plots[[nm]], width =20, height = 12, dpi = 300)
}
pC

```



##7 Plotting the gene expression profile of interesting genes

This code chunk can be used to plot the gene expression profile of genes of interest both as feature plot and violin plot

```{r}
markers <- c("B2m","C4b","Serpina3n","Cdkn1a","Mki67")  # marker genes

for (gene in markers) {
  p <- FeaturePlot(
    MOL, 
    features = gene, 
    reduction = "umap_MOL",
    cols = c("lightgrey", "magenta")
  )+
  labs(x = "UMAP 1", y = "UMAP 2") +
  ggtitle( paste0(gene)) +
  theme_minimal() +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      size =48,
      face = "bold"
    ),
    axis.title.x = element_text(size = 26),
    axis.title.y = element_text(size = 26),
    legend.text = element_text(size = 24),         # Legend labels
    legend.title = element_text(size = 28),  # Legend title
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank())
  # Display plot in the R Markdown output
  print(p)
  
  # Save each plot
  filename <- file.path("/date/gcb/gcb_MZ/MOL_FGF_10x/Plots/", paste0("Feature_plot_", gene, ".png"))
  ggplot2::ggsave(filename, p, width = 20, height = 12, dpi = 300)
}

for (gene in markers) {
  p <- VlnPlot(
    MOL, 
    features = gene, 
    group.by = "MOL_fine_clustering", 
    pt.size = 0  # This removes the dots
  ) +
    xlab("") +
    theme(
      plot.title = element_text(
      hjust = 0.5,
      size =48,
      face = "bold"
    ),
      axis.text.x = element_text(hjust = 1, vjust = 0.5, family = "Arial", size = 20, angle = 90),
      axis.text.y = element_text(family = "Arial", size = 20),
      axis.title.y = element_text(family = "Arial", size = 24),
      legend.text = element_text(size = 20, family = "Arial"),
      legend.title = element_text(size = 24, family = "Arial")
    )
  # Display plot in the R Markdown output
  print(p)
  
  # Save each plot
  filename <- file.path("/date/gcb/gcb_MZ/MOL_FGF_10x/Plots/", paste0("Violin_plot_", gene, ".png"))
  ggplot2::ggsave(filename, p, width = 20, height = 12, dpi = 300)
}

```


##8 Functional enrichement analysis

In this section the functional enrichment analysis using the marker genes associated will be executed. In one part of the analysis the functional enrichment will be performed using Gene Ontology terms and in the second part using KEGG pathways. 

##8.1 Functional enrichment analysis using Gene Ontology terms
Functional enrichment using Gene Ontology terms

1.Compute the GO terms significantly enriched

```{r}
# marker genes computed above

pos_markers <- MOL_markers%>%
  dplyr::select(cluster, gene)

# convert symbols to ENTREZID per cluster
marker_list <- split(pos_markers$gene, pos_markers$cluster)

# convert once (union) and map back
all_genes <- unique(unlist(marker_list))
gene_map <- bitr(all_genes, fromType = "SYMBOL", toType = "ENTREZID", OrgDb = org.Mm.eg.db)
sym2ent <- setNames(gene_map$ENTREZID, gene_map$SYMBOL)

entrez_by_cluster <- lapply(marker_list, function(x) na.omit(unname(sym2ent[x])))

ego_list <- lapply(names(entrez_by_cluster), function(cl){
  enrichGO(gene          = entrez_by_cluster[[cl]],
           OrgDb         = org.Mm.eg.db,
           keyType       = "ENTREZID",
           ont           = "BP",
           pAdjustMethod = "BH",
           pvalueCutoff  = 0.05,
           qvalueCutoff  = 0.2,
           readable      = TRUE)
}) #list of enriched terms per cluster
names(ego_list) <- names(entrez_by_cluster)

```


2.Create an Excel file to save the results, to be attached as supplementary material

```{r}

# Create a new workbook
wb <- createWorkbook()

# Loop through each cluster and add enrichment results as a sheet
for (cluster_name in names(ego_list)) {
  
  # Extract results from enrichResult object
  if (!is.null(ego_list[[cluster_name]]) && nrow(ego_list[[cluster_name]]@result) > 0) {
    
    # Get the results as a data frame
    go_results <- as.data.frame(ego_list[[cluster_name]])
    
    # Create sheet name (Excel has 31 char limit for sheet names)
    sheet_name <- substr(cluster_name, 1, 31)
    
    # Add worksheet
    addWorksheet(wb, sheet_name)
    
    # Write data
    writeData(wb, sheet_name, go_results)
    
    cat(sprintf("Added sheet for cluster: %s (%d GO terms)\n", 
                cluster_name, nrow(go_results)))
    
  } else {
    cat(sprintf("No significant GO terms for cluster: %s\n", cluster_name))
  }
}

# Save the workbook
saveWorkbook(wb, 
             "/date/gcb/gcb_MZ/MOL_FGF_10x/Tables/GO_enrichment_by_cluster.xlsx", 
             overwrite = TRUE)

cat("\n✓ GO enrichment results saved to Excel file!\n")

```

3.Plot the functional enrichment results results for GO terms


```{r,fig.height=25, fig.width=20}
# Collect GO BP results into one data.frame

ego_df <- map2_df(names(ego_list), ego_list, ~{
  if (is.null(.y) || nrow(as.data.frame(.y)) == 0) return(NULL)
  df <- as.data.frame(.y)
  df$cluster <- .x
  df$mLog10Padj <- -log10(df$p.adjust)
  df
})

# Choose top terms per cluster and plot

plot_df <- ego_df %>%
  group_by(cluster) %>%
  slice_max(order_by = mLog10Padj, n = 10, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(Description = fct_reorder(Description, mLog10Padj))

p1<-ggplot(plot_df, aes(x = cluster, y = Description, size = Count, color = mLog10Padj)) +
  geom_point() +
  scale_color_gradient(low = "gold", high = "red") +
  labs(title = "GO BP enrichment per cluster", x = "Cluster", y = "Term", color = "-log10(FDR)") +
  theme_bw(base_size = 12) + theme_classic()+
  theme(
    axis.text.x = element_text( hjust = 1, vjust = 0.5, family = "Arial", size = 20,angle = 90),
    axis.text.y = element_text(family = "Arial", size = 24),
    axis.title.y = element_text(family = "Arial", size =28, face = "bold"),
    legend.text = element_text(size = 24,family = "Arial"),
    legend.title = element_text(size = 32,family = "Arial",face = "bold")
  ) 
plots <- list(Func_enrich_GO=p1)
for (nm in names(plots)) {
ggplot2::ggsave(file.path("/date/gcb/gcb_MZ/MOL_FGF_10x/Plots/", paste0(nm, ".png")), plots[[nm]], width =14, height = 16, dpi = 300)
}
p1

```
##8.2 Functional enrichment analysis using KEGG pathways
Functional enrichment using Kegg pathways

1.Kegg pathways analysis

```{r}
ekegg_list <- lapply(names(entrez_by_cluster), function(cl){
  enrichKEGG(gene          = entrez_by_cluster[[cl]],
             organism      = "mmu",
             pvalueCutoff  = 0.05)
})
names(ekegg_list) <- names(entrez_by_cluster)


kegg_df <- map2_df(names(ekegg_list), ekegg_list, ~{
  if (is.null(.y) || nrow(as.data.frame(.y)) == 0) return(NULL)
  df <- as.data.frame(.y)
  df$cluster <- .x
  df$mLog10Padj <- -log10(df$p.adjust)
  df
})
```

2.Create an Excel file to save the results, to be attached as supplementary material

```{r}

# Save to Excel - Option 1: One sheet per cluster
wb_kegg <- createWorkbook()

for (cluster_name in names(ekegg_list)) {
  
  if (!is.null(ekegg_list[[cluster_name]]) && nrow(ekegg_list[[cluster_name]]@result) > 0) {
    
    # Get results as data frame
    kegg_results <- as.data.frame(ekegg_list[[cluster_name]])
    
    # Add mLog10Padj column
    kegg_results$mLog10Padj <- -log10(kegg_results$p.adjust)
    
    # Create sheet name
    sheet_name <- substr(cluster_name, 1, 31)
    
    # Add worksheet and write data
    addWorksheet(wb_kegg, sheet_name)
    writeData(wb_kegg, sheet_name, kegg_results)
    
    cat(sprintf("Added sheet for cluster: %s (%d KEGG pathways)\n", 
                cluster_name, nrow(kegg_results)))
    
  } else {
    cat(sprintf("No significant KEGG pathways for cluster: %s\n", cluster_name))
  }
}

# Save workbook with one sheet per cluster
saveWorkbook(wb_kegg, 
             "/date/gcb/gcb_MZ/MOL_FGF_10x/Tables/KEGG_enrichment_by_cluster.xlsx", 
             overwrite = TRUE)

cat("\n✓ KEGG enrichment results saved (one sheet per cluster)!\n")

```

3.Plot the results

```{r,fig.height=25, fig.width=20}
plot_df_kegg <- kegg_df %>%
  group_by(cluster) %>%
  slice_max(order_by = mLog10Padj, n = 10, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(Description = fct_reorder(Description, mLog10Padj))

p1<-ggplot(plot_df_kegg, aes(x = cluster, y = Description, size = Count, color = mLog10Padj)) +
  geom_point() +
  scale_color_gradient(low = "gold", high = "red") +
  labs(title = "KEGG enrichment per cluster", x = "Cluster", y = "Pathway", color = "-log10(FDR)") +
  theme_bw(base_size = 12)+theme_classic()+
  theme(
    axis.text.x = element_text( hjust = 1, vjust = 0.5, family = "Arial", size = 20,angle = 90),
    axis.text.y = element_text(family = "Arial", size = 16),
    axis.title.y = element_text(family = "Arial", size =16),
    legend.text = element_text(size = 20,family = "Arial"),
    legend.title = element_text(size = 24,family = "Arial")
  ) 

plots <- list(Func_enrich_KEGG=p1)
for (nm in names(plots)) {
ggplot2::ggsave(file.path("/date/gcb/gcb_MZ/MOL_FGF_10x/Plots/", paste0(nm, ".png")), plots[[nm]], width =14, height = 16, dpi = 300)
}
p1

```
##9 Human MS data integration   
In this final section a comparison between the MOL dataset and the human MS dataset transferred to the mouse genome is performed.

##9.1 Pathology annotation
The first analysis is the pathology annotation that will be again performed as label transfer (previosly done)

1.Transfer the pathology label to the MOL dataset

```{r}

#and disease stage or control labels
predicted_pathology<-readRDS("/proj/tmp/tmp_EA/MZ_Zuchero/Zuchero_MOL_FGF_Absinta_MM10_orthologs_predicted_pathology.rds")

# 2. Subset to cells actually in your Seurat object (ensure rownames match Seurat cell names)
pred_df <-predicted_pathology[rownames(predicted_pathology) %in% colnames(MOL), ]

# 3. Reorder rows to match Seurat cell order
pred_df <- pred_df[colnames(MOL), ]

# 4. Add all prediction columns as metadata
MOL <- AddMetaData(object = MOL, metadata = pred_df)


```


2.Plot the pathology label into the UMAP plot of MOLs


```{r}

p<-DimPlot(MOL ,reduction = "umap_MOL",group.by = "Absinta_2021_pathology",cols=colors)+
  labs(x = "UMAP 1", y = "UMAP 2") +
  ggtitle("MS pathology Absinta, 2021") +
  theme_minimal() +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      size = 24,
      face = "bold"
    ),
    axis.title.x = element_text(size = 20),
    axis.title.y = element_text(size = 20),
    legend.text = element_text(size = 16),         # Legend labels
    legend.title = element_text(size = 24),  # Legend title
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank()
  )

plots <- list(Absinta_pathology_label_transfer=p)
for (nm in names(plots)) {
ggplot2::ggsave(file.path("/date/gcb/gcb_MZ/MOL_FGF_10x/Plots/", paste0(nm, ".png")), plots[[nm]], width =14, height = 8, dpi = 300)
}
p
```

3.Plot the similarity of each cluster to each pathological annotation as violin plot

```{r,fig.height=10, fig.width=20}
# Define pathology annotation columns
pathology_columns <- c(
  "prediction.score.chronic_active_MS_lesion_edge",
  "prediction.score.chronic_inactive_MS_lesion_edge",
  "prediction.score.MS_lesion_core",
  "prediction.score.MS_periplaque_white_matter",
  "prediction.score.control_white_matter"
)

# Create cleaner labels for plotting
pathology_labels <- c(
  "Chronic Active Lesion Edge",
  "Chronic Inactive Lesion Edge",
  "MS Lesion Core",
  "Periplaque White Matter",
  "Control White Matter"
)

# Get cluster identities
clusters <- unique(MOL$MOL_fine_clustering)
print(paste("Creating pathology plots for", length(clusters), "clusters"))

# Loop through each cluster
for (cluster_id in clusters) {
  
  # Create empty data frame to store all scores
  all_scores <- data.frame()
  
  # Get cells in this cluster
  cluster_cells <- colnames(MOL)[MOL$MOL_fine_clustering == cluster_id]
  
  # Extract scores for each pathology type from metadata
  for (i in seq_along(pathology_columns)) {
    col_name <- pathology_columns[i]
    label_name <- pathology_labels[i]
    
    # Get scores for this pathology type and this cluster
    scores <- MOL@meta.data[[col_name]][MOL$MOL_fine_clustering == cluster_id]
    
    temp_df <- data.frame(
      score = as.numeric(scores),
      pathology_type = label_name,
      stringsAsFactors = FALSE
    )
    
    all_scores <- rbind(all_scores, temp_df)
  }
  
  # Create violin plot with all pathology types on x-axis
  p <- ggplot(all_scores, aes(x = pathology_type, y = score, fill = pathology_type)) +
    geom_violin(trim = TRUE, scale = "width") +
    geom_boxplot(width = 0.1, fill = "white", outlier.shape = NA) +
    ggtitle(paste("Pathology Similarity - Cluster:", cluster_id)) +
    xlab("Pathology Type") +
    ylab("Similarity Score") +
    ylim(0, 1) +
    theme_classic() +
    theme(
      plot.title = element_text(hjust = 0.5, size = 36, face = "bold"),
      axis.text.x = element_text(hjust = 1, vjust = 0.5, family = "Arial", 
                                 size = 14, angle = 90),
      axis.text.y = element_text(family = "Arial", size = 20),
      axis.title.x = element_text(family = "Arial", size = 24, face = "bold"),
      axis.title.y = element_text(family = "Arial", size = 24, face = "bold"),
      legend.position = "none"
    )
  
  # Display plot in R Markdown output
  print(p)
  
  # Save plot
  filename <- file.path("/date/gcb/gcb_MZ/MOL_FGF_10x/Plots/", 
                       paste0("Pathology_similarity_cluster_", cluster_id, ".png"))
  ggplot2::ggsave(filename, p, width = 20, height = 12, dpi = 300)
  
  print(paste("✓ Saved pathology plot for cluster:", cluster_id))
}

```

##9.2 Cluster annotation
In this section using the ortholog dataset a label transfer between the human data will be performed with the FGF data to understand to which cluster the FGF dataset is more similar 

1.Transfer the cell label to the MOL dataset, we will first select only the predicted score for OLs, since in our dataset we only have the mature OLs

```{r}

#and disease stage or control labels
predicted_celltype_OL<-readRDS("/date/gcb/gcb_MZ/MOL_FGF_10x//Zuchero_MOL_FGF_Absinta_MM10_orthologs_predicted_celltypes.rds") %>% 
  dplyr::filter(Absinta_2021_celltypes %in% c("OL_0", "OL_1", "OL_2","OL_3","OL_4","OL_4","OL_5","OL_6","OL_7")) %>% 
  dplyr::select(prediction.score.OL_0,prediction.score.OL_1,prediction.score.OL_2,prediction.score.OL_3,
               prediction.score.OL_4,prediction.score.OL_5,prediction.score.OL_6,prediction.score.OL_7,Absinta_2021_celltypes)

# 2. Subset to cells actually in your Seurat object (ensure rownames match Seurat cell names)
pred_df <-predicted_celltype_OL[rownames(predicted_celltype_OL) %in% colnames(MOL), ]

# 3. Reorder rows to match Seurat cell order
pred_df <- pred_df[colnames(MOL), ]

# 4. Add all prediction columns as metadata
MOL <- AddMetaData(object = MOL, metadata = pred_df)

```

2.Plot the pathology label into the UMAP plot of MOLs


```{r}

p<-DimPlot(MOL ,reduction = "umap_MOL",group.by = "Absinta_2021_celltypes")+
  labs(x = "UMAP 1", y = "UMAP 2") +
  ggtitle("MS pathology Absinta, 2021") +
  theme_minimal() +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      size = 24,
      face = "bold"
    ),
    axis.title.x = element_text(size = 20),
    axis.title.y = element_text(size = 20),
    legend.text = element_text(size = 16),         # Legend labels
    legend.title = element_text(size = 24),  # Legend title
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank()
  )

plots <- list(Absinta_celltype_label_transfer=p)
for (nm in names(plots)) {
ggplot2::ggsave(file.path("/date/gcb/gcb_MZ/MOL_FGF_10x//Plots/", paste0(nm, ".png")), plots[[nm]], width =14, height = 8, dpi = 300)
}
p
```

3.Plot the similarity of each cluster to each cell type as violin plot

```{r,fig.height=10, fig.width=20}
# Define celltypes annotation columns
celltypes_columns <- c(
  "prediction.score.OL_0",
  "prediction.score.OL_1",
  "prediction.score.OL_2",
  "prediction.score.OL_3",
  "prediction.score.OL_4",
  "prediction.score.OL_5",
  "prediction.score.OL_6",
  "prediction.score.OL_7"
)

# Create cleaner labels for plotting
celltypes_labels <- c(
  "OL_0",
  "OL_1",
  "OL_2",
  "OL_3",
  "OL_4",
  "OL_5",
  "OL_6",
  "OL_7"
)

# Get cluster identities
clusters <- unique(MOL$MOL_fine_clustering)
print(paste("Creating celltypes plots for", length(clusters), "clusters"))

# Loop through each cluster
for (cluster_id in clusters) {
  
  # Create empty data frame to store all scores
  all_scores <- data.frame()
  
  # Get cells in this cluster
  cluster_cells <- colnames(MOL)[MOL$MOL_fine_clustering == cluster_id]
  
  # Extract scores for each celltypes type from metadata
  for (i in seq_along(celltypes_columns)) {
    col_name <- celltypes_columns[i]
    label_name <- celltypes_labels[i]
    
    # Get scores for this celltypes type and this cluster
    scores <- MOL@meta.data[[col_name]][MOL$MOL_fine_clustering == cluster_id]
    
    temp_df <- data.frame(
      score = as.numeric(scores),
      celltypes_type = label_name,
      stringsAsFactors = FALSE
    )
    
    all_scores <- rbind(all_scores, temp_df)
  }
  
  # Create violin plot with all celltypes types on x-axis
  p <- ggplot(all_scores, aes(x = celltypes_type, y = score, fill = celltypes_type)) +
    geom_violin(trim = TRUE, scale = "width") +
    geom_boxplot(width = 0.1, fill = "white", outlier.shape = NA) +
    ggtitle(paste("celltypes Similarity - Cluster:", cluster_id)) +
    xlab("celltypes Type") +
    ylab("Similarity Score") +
    ylim(0, 1) +
    theme_classic() +
    theme(
      plot.title = element_text(hjust = 0.5, size = 36, face = "bold"),
      axis.text.x = element_text(hjust = 1, vjust = 0.5, family = "Arial", 
                                 size = 14, angle = 90),
      axis.text.y = element_text(family = "Arial", size = 20),
      axis.title.x = element_text(family = "Arial", size = 24, face = "bold"),
      axis.title.y = element_text(family = "Arial", size = 24, face = "bold"),
      legend.position = "none"
    )
  
  # Display plot in R Markdown output
  print(p)
  
  # Save plot
  filename <- file.path("/date/gcb/gcb_MZ/MOL_FGF_10x//Plots/", 
                       paste0("celltypes_similarity_cluster_", cluster_id, ".png"))
  ggplot2::ggsave(filename, p, width = 20, height = 12, dpi = 300)
  
  print(paste("✓ Saved celltypes plot for cluster:", cluster_id))
}

```




#9.3 FGF specific module for Human integration

To check for the presence of a gene expression module that is typical of FGF also in human an FGF specific gene module needs to be created to do so the top 100 upregulated and downregulated maker genes for each FGF specifc cell cluster will be selected.The dataset in which the modules will be computed into are the humanized_mouse_brains from Macnair et al. Neuron, 2025 (Roche) and Absinta et al., Nature, 2021. 

1. Load the cluster markers from the mouse FGF dataset and select the top 100 dowregulated and upregulated makers based on padj value and use them to build a gene module of the different clusters of the dataset

```{r}

# Path to your Excel file
excel_file <- "/date/gcb/gcb_MZ/MOL_FGF_10x//clustersALL_MOL_scRNA_10x.xlsx"

# Get all sheet names (each sheet = one cluster)
sheet_names <- excel_sheets(excel_file)

# Read data from each sheet and create nested list
sets <- lapply(sheet_names, function(sheet) {
  # Read the sheet
  cluster_df <- read_excel(excel_file, sheet = sheet)
  
  # Split into upregulated (positive log2FC) and downregulated (negative log2FC)
  # Then rank by most significant (lowest p_val_adj)
  
  # Top 100 UPREGULATED genes (positive avg_log2FC, sorted by p_val_adj)
  up_genes <- cluster_df %>%
    filter(avg_log2FC > 0) %>%                  # Only upregulated
    arrange(p_val_adj) %>%                       # Most significant first
    slice_head(n = 100) %>%                      # Top 100
    pull(gene)
  
  # Top 100 DOWNREGULATED genes (negative avg_log2FC, sorted by p_val_adj)
  down_genes <- cluster_df %>%
    filter(avg_log2FC < 0) %>%                  # Only downregulated
    arrange(p_val_adj) %>%                       # Most significant first
    slice_head(n = 100) %>%                      # Top 100
    pull(gene)
  
  # Return as list with up and down
  list(up = up_genes, down = down_genes)
})

# Name the list elements by cluster (sheet names)
names(sets) <- sheet_names

# Verify
for (cluster_name in names(sets)) {
  cat(sprintf("%s: %d up genes, %d down genes\n", 
              cluster_name, 
              length(sets[[cluster_name]]$up),
              length(sets[[cluster_name]]$down)))
}
```

2. Since these are mouse genes before being used to compute the gene modules they need to be converted into human genes


```{r}
library(homologene)

# Your existing sets structure:
# sets$cluster_name$up    <- mouse genes (upregulated)
# sets$cluster_name$down  <- mouse genes (downregulated)

# Convert to human genes
sets_human <- lapply(sets, function(cluster) {
  
  # Convert up genes
  if (length(cluster$up) > 0) {
    up_conversion <- mouse2human(cluster$up, db = homologeneData2)
    up_human <- unique(up_conversion$humanGene)
  } else {
    up_human <- character(0)
  }
  
  # Convert down genes  
  if (length(cluster$down) > 0) {
    down_conversion <- mouse2human(cluster$down, db = homologeneData2)
    down_human <- unique(down_conversion$humanGene)
  } else {
    down_human <- character(0)
  }
  
  list(up = up_human, down = down_human)
})

# Check conversion efficiency
cat("\nConversion Summary:\n")
cat(sprintf("%-20s %10s %10s\n", "Cluster", "Up Genes", "Down Genes"))
cat(strrep("-", 45), "\n")

for (cluster_name in names(sets_human)) {
  cat(sprintf("%-20s %3d -> %3d  %3d -> %3d\n",
              cluster_name,
              length(sets[[cluster_name]]$up),
              length(sets_human[[cluster_name]]$up),
              length(sets[[cluster_name]]$down),
              length(sets_human[[cluster_name]]$down)))
}


```


3. load the Roche dataset


```{r}


# Read original object
MS_Lesions_Roche <- readRDS("/date/gcb/gcb_MZ/MOL_FGF_10x/ms_lesions_snRNAseq_seuratV5_WMEA_updated_WMOLGdown3K.rds")


```

4. Now let's use the same method use on the Pandey et al. dataset to calculate the module score. Different function to handle the ENSEMBL annotation in the Roche dataset. 


```{r}
# Assuming sets_human and your Seurat object (MS_Lesions_Roche) are loaded
DefaultAssay(MS_Lesions_Roche) <- "RNA"
# Set idents to a metadata column
Idents(MS_Lesions_Roche) <- "type_fine"

# Loop over each cluster in sets_human
for (cluster_name in names(sets_human)) {
  cat("Calculating", cluster_name, "signed score:\n")
  
  up_genes <- sets_human[[cluster_name]]$up
  down_genes <- sets_human[[cluster_name]]$down
  
  # Calculate enrichment score using your custom function for each cluster
  score_name <- paste0(cluster_name, "_score")
  MS_Lesions_Roche[[score_name]] <- calculate_signed_score_with_ensembl(
    MS_Lesions_Roche, 
    geneset = up_genes,
    geneset_down = down_genes,
    control_cluster = "humanized_mouse_brain_D",
    cluster_col = "type_fine",
    assay = "RNA"
  )
}


```


5. Plot the signature score 

```{r,fig.height=10, fig.width=20}

# Define the grouping variable
grp <- "type_fine"

# Define which clusters to include in the plots
clusters_to_plot <- c("humanized_mouse_brain_A", "humanized_mouse_brain_B", "humanized_mouse_brain_C", "humanized_mouse_brain_D", "humanized_mouse_brain_E","humanized_mouse_brain_F","humanized_mouse_brain_G")


# Define all score names to plot
score_names <- c("Cluster_OL_1_score", "Cluster_OL_2_score", "Cluster_DAO_IFN_score", 
                 "Cluster_DAO_FGF_immune_score", "Cluster_DAO_FGF_cycling_score")

# Create output directory if it doesn't exist
output_dir <- "/date/gcb/gcb_MZ/MOL_FGF_10x//Plots/"
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

# Filter metadata to selected clusters
metadata_filtered <- MS_Lesions_Roche@meta.data %>%
  filter(type_fine %in% clusters_to_plot)
# Or to exclude: filter(!type_fine %in% clusters_to_exclude)

# Loop through each score and create plots
for (score_name in score_names) {
  cat("Creating plot for:", score_name, "\n")
  
  # Create the plot with filtered data
  p <- ggplot(metadata_filtered, 
              aes_string(x = grp, y = score_name, color = grp)) +
    geom_boxplot(outlier.shape = NA, alpha = 0.6) +
    geom_jitter(width = 0.2, size = 0.7, alpha = 0.5) +
    theme_classic(base_family = "Arial") +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, size = 24),
      axis.text.y = element_text(size = 24),
      axis.title.y = element_text(size = 28, face = "bold"),
      legend.position = "none"
    ) +
    labs(x = NULL, 
         y = paste(gsub("_score", "", score_name), "enriched gene set score"))
  
  # Print the plot
  print(p)
  
  # Save the plot
  output_file <- file.path(output_dir, paste0(score_name, ".png"))
  ggsave(output_file, plot = p, width = 20, height = 12, dpi = 300)
  
  cat("Saved:", output_file, "\n\n")
}


```


6. Plot some specific genes into the roche dataset form the FGF humanized_mouse_brain

```{r}
# Define markers
markers <- c("CDKN1A", "BTG2", "GDF15", "C4B", "PIEZO2", "B2M", "SERPINA3","STAT3")

# Define which clusters to plot
clusters_to_plot <- c("humanized_mouse_brain_A", "humanized_mouse_brain_B", "humanized_mouse_brain_C", "humanized_mouse_brain_D", "humanized_mouse_brain_E","humanized_mouse_brain_F","humanized_mouse_brain_G")
# Or use: clusters_to_plot <- grep("humanized_mouse_brain", unique(MS_Lesions_Roche$type_fine), value = TRUE)

# Subset Seurat object to selected clusters
MS_Lesions_subset  <- subset(MS_Lesions_Roche, subset = type_fine %in% clusters_to_plot)

# Create output directory
output_dir <- "/date/gcb/gcb_MZ/MOL_FGF_10x//Plots/"
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

# Plot each marker
for (gene in markers) {
  cat("Plotting:", gene, "\n")
  
  p <- VlnPlot(
    MS_Lesions_subset,  # Use subset instead
    features = gene, 
    group.by = "type_fine"
  ) +
  ggtitle(paste0(gene)) +
  theme_minimal() +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      size = 48,
      face = "bold"
    ),
    axis.title.x = element_text(size = 26),
    axis.title.y = element_text(size = 26),
    legend.text = element_text(size = 24),
    legend.title = element_text(size = 28),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 20),  # Show cluster names
    axis.text.y = element_text(size = 20)
  )
  
  print(p)
  
  # Save each plot
  filename <- file.path(output_dir, paste0("VlnPlot_", gene, ".png"))
  ggsave(filename, p, width = 20, height = 12, dpi = 300)
  
  cat("Saved:", filename, "\n\n")
}


```


7.Dotplot of mouse markers into Roche dataset

```{r,fig.height=10, fig.width=20}
markers <- c("PTGDS","OPALIN","KIRREL3","PRICKLE1", #OL1
             "S100B","PLIN3","ENPP6","ROBO1", #OL2
             "SERPINA3","PIEZO2","MASP1","C4B", #DAO_1
             "IRGM","PSMB8","STAT1","HLA-E","PSMB9", #IFN_DAO
             "CDKN1A","CCND1","NAV3","TMSB10","LRRFIP1", #cycling_DAO
             "IL33"
)

ensembl_markers <- get_ensembl_features(markers, MS_Lesions_Roche)

# Define which clusters to plot
clusters_to_plot <- c("humanized_mouse_brain_A", "humanized_mouse_brain_B", "humanized_mouse_brain_C", "humanized_mouse_brain_D", "humanized_mouse_brain_E","humanized_mouse_brain_F","humanized_mouse_brain_G")
# Or use: clusters_to_plot <- grep("humanized_mouse_brain", unique(MS_Lesions_Roche$type_fine), value = TRUE)

# Subset Seurat object to selected clusters
MS_Lesions_subset  <- subset(MS_Lesions_Roche, subset = type_fine %in% clusters_to_plot)

# Create a named vector: names are ENSEMBL features, values are symbol only
feature_labels <- setNames(markers, ensembl_markers)

p1<-DotPlot(MS_Lesions_subset, features = ensembl_markers,group.by = "type_fine",cols = c("lightgrey", "firebrick"))+
  xlab("")+
  scale_x_discrete(labels = feature_labels) + 
  theme(
    axis.text.x = element_text( hjust = 1, vjust = 0.5, family = "Arial", size = 20,angle = 90),
    axis.text.y = element_text(family = "Arial", size = 20),
    axis.title.y = element_text(family = "Arial", size =24),
    legend.text = element_text(size = 20,family = "Arial"),
    legend.title = element_text(size = 24,family = "Arial")
  ) 
p1
```


8.Load the Absinta Human dataset

```{r}

Absinta_hg38<- readRDS("/date/gcb/gcb_MZ/MOL_FGF_10x//GSE180759_Absinta_seuratV5.rds")

```

9. Enrichment modules into the Absinta human

```{r}
# Assuming sets_human and your Seurat object (MS_Lesions_Roche) are loaded
DefaultAssay(Absinta_hg38) <- "RNA"
# Set idents to a metadata column
Idents(Absinta_hg38) <- "final_annot"

# Loop over each cluster in sets_human
for (cluster_name in names(sets_human)) {
  cat("Calculating", cluster_name, "signed score:\n")
  
  up_genes <- sets_human[[cluster_name]]$up
  down_genes <- sets_human[[cluster_name]]$down
  
  # Calculate enrichment score using your custom function for each cluster
  score_name <- paste0(cluster_name, "_score")
  Absinta_hg38[[score_name]] <- calculate_signed_score(
    Absinta_hg38, geneset_up = up_genes,
    geneset_down = down_genes,
    control_cluster = "OL_0",
    cluster_col = "final_annot",
    assay = "RNA"
  )
}
```

10.Plot the signature score from the Absinta dataset

```{r,fig.height=10, fig.width=20}

# Define the grouping variable
grp <- "final_annot"

# Define which clusters to include in the plots
clusters_to_plot <- c("OL_0","OL_1","OL_2","OL_3","OL_4","OL_5","OL_5","OL_6","OL_7")


# Define all score names to plot
score_names <- c("Cluster_OL_1_score", "Cluster_OL_2_score", "Cluster_DAO_IFN_score", 
                 "Cluster_DAO_FGF_immune_score", "Cluster_DAO_FGF_cycling_score")

# Create output directory if it doesn't exist
output_dir <- "/date/gcb/gcb_MZ/MOL_FGF_10x//Plots/"
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

# Filter metadata to selected clusters
metadata_filtered <- Absinta_hg38@meta.data %>%
  filter(final_annot %in% clusters_to_plot)
# Or to exclude: filter(!type_fine %in% clusters_to_exclude)

# Loop through each score and create plots
for (score_name in score_names) {
  cat("Creating plot for:", score_name, "\n")
  
  # Create the plot with filtered data
  p <- ggplot(metadata_filtered, 
              aes_string(x = grp, y = score_name, color = grp)) +
    geom_boxplot(outlier.shape = NA, alpha = 0.6) +
    geom_jitter(width = 0.2, size = 0.7, alpha = 0.5) +
    theme_classic(base_family = "Arial") +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, size = 24),
      axis.text.y = element_text(size = 24),
      axis.title.y = element_text(size = 28, face = "bold"),
      legend.position = "none"
    ) +
    labs(x = NULL, 
         y = paste(gsub("_score", "", score_name), "enriched gene set score"))
  
  # Print the plot
  print(p)
  
  # Save the plot
  output_file <- file.path(output_dir, paste0(score_name, ".png"))
  ggsave(output_file, plot = p, width = 20, height = 12, dpi = 300)
  
  cat("Saved:", output_file, "\n\n")
}


```
11.Plot some relevant genes in the Absinta datasets from the FGF datasets

```{r}
# Define markers
markers <- c("CDKN1A", "BTG2", "GDF15", "C4B", "PIEZO2", "B2M","STAT3")

# Define which clusters to plot
clusters_to_plot <- c("OL_0", "OL_1", "OL_2", "OL_3", "OL_4","OL_5","OL_6","OL_7")
# Or use: clusters_to_plot <- grep("humanized_mouse_brain", unique(MS_Lesions_Roche$type_fine), value = TRUE)

# Subset Seurat object to selected clusters
MS_Lesions_subset <- subset(Absinta_hg38, subset = final_annot %in% clusters_to_plot)

# Create output directory
output_dir <- "/date/gcb/gcb_MZ/MOL_FGF_10x//Plots/"
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

# Plot each marker
for (gene in markers) {
  cat("Plotting:", gene, "\n")
  
  p <- VlnPlot(
    MS_Lesions_subset,  # Use subset instead
    features = gene, 
    group.by = "final_annot"
  ) +
  ggtitle(paste0(gene)) +
  theme_minimal() +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      size = 48,
      face = "bold"
    ),
    axis.title.x = element_text(size = 26),
    axis.title.y = element_text(size = 26),
    legend.text = element_text(size = 24),
    legend.title = element_text(size = 28),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 20),  # Show cluster names
    axis.text.y = element_text(size = 20)
  )
  
  print(p)
  
  # Save each plot
  filename <- file.path(output_dir, paste0("VlnPlot_", gene, ".png"))
  ggsave(filename, p, width = 20, height = 12, dpi = 300)
  
  cat("Saved:", filename, "\n\n")
}

```
12. Dotplot of some interesting markers in the Absinta dataset 

```{r}

markers <- c("PTGDS","OPALIN","KIRREL3","PRICKLE1", #OL1
             "S100B","PLIN3","ENPP6","ROBO1", #OL2
             "SERPINA3","PIEZO2","MASP1","C4B", #DAO_1
             "IRGM","PSMB8","STAT1","HLA-E","PSMB9", #IFN_DAO
             "CDKN1A","CCND1","NAV3","TMSB10","LRRFIP1", #cycling_DAO
             "IL33"
)

# Define which clusters to plot
clusters_to_plot <- c("OL_0", "OL_1", "OL_2", "OL_3", "OL_4","OL_5","OL_6","OL_7")
# Or use: clusters_to_plot <- grep("humanized_mouse_brain", unique(MS_Lesions_Roche$type_fine), value = TRUE)

# Subset Seurat object to selected clusters
Absinta_hg38_subset  <- subset(Absinta_hg38, subset = final_annot %in% clusters_to_plot)

p1<-DotPlot(MS_Lesions_subset, features = markers,group.by = "final_annot",cols = c("lightgrey", "firebrick"))+
  xlab("")+
  theme(
    axis.text.x = element_text( hjust = 1, vjust = 0.5, family = "Arial", size = 20,angle = 90),
    axis.text.y = element_text(family = "Arial", size = 20),
    axis.title.y = element_text(family = "Arial", size =24),
    legend.text = element_text(size = 20,family = "Arial"),
    legend.title = element_text(size = 24,family = "Arial")
  ) 
p1
```

#9.4 Roche dataset lable transfer human to mouse

In this section a label transfer from the annotation of the Roche Dataset to the MOLs will be performed. Also in this case the Roche dataset has been first annotated to the mouse genome find the closest ortolog for each gene. 

1.Transfer the diagnosis label to the MOL dataset

```{r}

#and disease stage or control labels
predicted_diagnosis<-readRDS("/date/gcb/gcb_MZ/MOL_FGF_10x/Zuchero_MOL_FGF_McNair_WMOLG_MM10_orthologs_predicted_diagnosis_vargenes2K.rds")

# 2. Subset to cells actually in your Seurat object (ensure rownames match Seurat cell names)
pred_df <-predicted_diagnosis[rownames(predicted_diagnosis) %in% colnames(MOL), ]

# 3. Reorder rows to match Seurat cell order
pred_df <- pred_df[colnames(MOL), ]

# 4. Add all prediction columns as metadata
MOL <- AddMetaData(object = MOL, metadata = pred_df)


```

2.Plot the pathology label into the UMAP plot of MOLs


```{r}

p<-DimPlot(MOL ,reduction = "umap_MOL",group.by = "McNair_2025_diagnosis")+
  labs(x = "UMAP 1", y = "UMAP 2") +
  ggtitle("McNair diagnosis") +
  theme_minimal() +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      size = 24,
      face = "bold"
    ),
    axis.title.x = element_text(size = 20),
    axis.title.y = element_text(size = 20),
    legend.text = element_text(size = 16),         # Legend labels
    legend.title = element_text(size = 24),  # Legend title
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank()
  )

plots <- list(Macnair_diagnosislabel_transfer=p)
for (nm in names(plots)) {
ggplot2::ggsave(file.path("/date/gcb/gcb_MZ/MOL_FGF_10x//Plots/", paste0(nm, ".png")), plots[[nm]], width =14, height = 8, dpi = 300)
}
p
```

3.Plot the similarity of each cluster to each pathological annotation as violin plot

```{r,fig.height=10, fig.width=20}
# Define pathology annotation columns
pathology_columns <- c(
  "prediction.score.CTR",
  "prediction.score.SPMS",
  "prediction.score.RRMS",
  "prediction.score.PPMS"
)

# Create cleaner labels for plotting
pathology_labels <- c(
  "Controls",
  "Secondary Progressive",
  "Relapsing Remitting",
  "Primary Progressive"
)

# Get cluster identities
clusters <- unique(MOL$MOL_fine_clustering)
print(paste("Creating pathology plots for", length(clusters), "clusters"))

# Loop through each cluster
for (cluster_id in clusters) {
  
  # Create empty data frame to store all scores
  all_scores <- data.frame()
  
  # Get cells in this cluster
  cluster_cells <- colnames(MOL)[MOL$MOL_fine_clustering == cluster_id]
  
  # Extract scores for each pathology type from metadata
  for (i in seq_along(pathology_columns)) {
    col_name <- pathology_columns[i]
    label_name <- pathology_labels[i]
    
    # Get scores for this pathology type and this cluster
    scores <- MOL@meta.data[[col_name]][MOL$MOL_fine_clustering == cluster_id]
    
    temp_df <- data.frame(
      score = as.numeric(scores),
      pathology_type = label_name,
      stringsAsFactors = FALSE
    )
    
    all_scores <- rbind(all_scores, temp_df)
  }
  
  # Create violin plot with all pathology types on x-axis
  p <- ggplot(all_scores, aes(x = pathology_type, y = score, fill = pathology_type)) +
    geom_violin(trim = TRUE, scale = "width") +
    geom_boxplot(width = 0.1, fill = "white", outlier.shape = NA) +
    ggtitle(paste("Diagnosis Similarity - Cluster:", cluster_id)) +
    xlab("Diagnosis Type") +
    ylab("Similarity Score") +
    ylim(0, 1) +
    theme_classic() +
    theme(
      plot.title = element_text(hjust = 0.5, size = 36, face = "bold"),
      axis.text.x = element_text(hjust = 1, vjust = 0.5, family = "Arial", 
                                 size = 14, angle = 90),
      axis.text.y = element_text(family = "Arial", size = 20),
      axis.title.x = element_text(family = "Arial", size = 24, face = "bold"),
      axis.title.y = element_text(family = "Arial", size = 24, face = "bold"),
      legend.position = "none"
    )
  
  # Display plot in R Markdown output
  print(p)
  
  # Save plot
  filename <- file.path("/date/gcb/gcb_MZ/MOL_FGF_10x//Plots/", 
                       paste0("Pathology_similarity_cluster_", cluster_id, ".png"))
  ggplot2::ggsave(filename, p, width = 20, height = 12, dpi = 300)
  
  print(paste("✓ Saved diagnosis plot for cluster:", cluster_id))
}

```

4.Transfer the lesion type label to the MOL dataset

```{r}

#and disease stage or control labels
predicted_lesions<-readRDS("/date/gcb/gcb_MZ/MOL_FGF_10x/Zuchero_MOL_FGF_McNair_WMOLG_MM10_orthologs_predicted_lesiontype_vargenes2K.rds")

# 2. Subset to cells actually in your Seurat object (ensure rownames match Seurat cell names)
pred_df <-predicted_lesions[rownames(predicted_diagnosis) %in% colnames(MOL), ]

# 3. Reorder rows to match Seurat cell order
pred_df <- pred_df[colnames(MOL), ]

# 4. Add all prediction columns as metadata
MOL <- AddMetaData(object = MOL, metadata = pred_df)


```

5.Plot the pathology label into the UMAP plot of MOLs


```{r}

p<-DimPlot(MOL ,reduction = "umap_MOL",group.by = "McNair_2025_lesiontype3")+
  labs(x = "UMAP 1", y = "UMAP 2") +
  ggtitle("McNair Lesion type") +
  theme_minimal() +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      size = 24,
      face = "bold"
    ),
    axis.title.x = element_text(size = 20),
    axis.title.y = element_text(size = 20),
    legend.text = element_text(size = 16),         # Legend labels
    legend.title = element_text(size = 24),  # Legend title
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank()
  )

plots <- list(Macnair_lesion_label_transfer=p)
for (nm in names(plots)) {
ggplot2::ggsave(file.path("/date/gcb/gcb_MZ/MOL_FGF_10x//Plots/", paste0(nm, ".png")), plots[[nm]], width =14, height = 8, dpi = 300)
}
p
```
6.Plot the similarity of each cluster to each pathological annotation as violin plot

```{r,fig.height=10, fig.width=20}
# Define pathology annotation columns
pathology_columns <- c(
  "prediction.score.WM",
  "prediction.score.CIL",
  "prediction.score.RL",
  "prediction.score.NAWM",
  "prediction.score.AL",
  "prediction.score.CAL"
)

# Create cleaner labels for plotting
pathology_labels <- c(
  "White Matter",
  "Chronic Inactive Lesion",
  "Repairing Lesions",
  "Normal Appearing White Matter",
  "Acute Lesions",
  "Chronic Acute Lesions"
)

# Get cluster identities
clusters <- unique(MOL$MOL_fine_clustering)
print(paste("Creating lesion plots for", length(clusters), "clusters"))

# Loop through each cluster
for (cluster_id in clusters) {
  
  # Create empty data frame to store all scores
  all_scores <- data.frame()
  
  # Get cells in this cluster
  cluster_cells <- colnames(MOL)[MOL$MOL_fine_clustering == cluster_id]
  
  # Extract scores for each pathology type from metadata
  for (i in seq_along(pathology_columns)) {
    col_name <- pathology_columns[i]
    label_name <- pathology_labels[i]
    
    # Get scores for this pathology type and this cluster
    scores <- MOL@meta.data[[col_name]][MOL$MOL_fine_clustering == cluster_id]
    
    temp_df <- data.frame(
      score = as.numeric(scores),
      pathology_type = label_name,
      stringsAsFactors = FALSE
    )
    
    all_scores <- rbind(all_scores, temp_df)
  }
  
  # Create violin plot with all pathology types on x-axis
  p <- ggplot(all_scores, aes(x = pathology_type, y = score, fill = pathology_type)) +
    geom_violin(trim = TRUE, scale = "width") +
    geom_boxplot(width = 0.1, fill = "white", outlier.shape = NA) +
    ggtitle(paste("Lesion Similarity - Cluster:", cluster_id)) +
    xlab("Lesion Type") +
    ylab("Similarity Score") +
    ylim(0, 1) +
    theme_classic() +
    theme(
      plot.title = element_text(hjust = 0.5, size = 36, face = "bold"),
      axis.text.x = element_text(hjust = 1, vjust = 0.5, family = "Arial", 
                                 size = 14, angle = 90),
      axis.text.y = element_text(family = "Arial", size = 20),
      axis.title.x = element_text(family = "Arial", size = 24, face = "bold"),
      axis.title.y = element_text(family = "Arial", size = 24, face = "bold"),
      legend.position = "none"
    )
  
  # Display plot in R Markdown output
  print(p)
  
  # Save plot
  filename <- file.path("/date/gcb/gcb_MZ/MOL_FGF_10x//Plots/", 
                       paste0("Pathology_similarity_cluster_", cluster_id, ".png"))
  ggplot2::ggsave(filename, p, width = 20, height = 12, dpi = 300)
  
  print(paste("✓ Saved lesion plot for cluster:", cluster_id))
}

```

8.Transfer the cell type label to the MOL dataset

```{r}

#and disease stage or control labels
predicted_celltype<-readRDS("/date/gcb/gcb_MZ/MOL_FGF_10x/Zuchero_MOL_FGF_McNair_WMOLG_MM10_orthologs_predicted_celltype_vargenes2K.rds")

# 2. Subset to cells actually in your Seurat object (ensure rownames match Seurat cell names)
pred_df <-predicted_celltype[rownames(predicted_celltype) %in% colnames(MOL), ]

# 3. Reorder rows to match Seurat cell order
pred_df <- pred_df[colnames(MOL), ]

# 4. Add all prediction columns as metadata
MOL <- AddMetaData(object = MOL, metadata = pred_df)


```

9.Plot the pathology label into the UMAP plot of MOLs


```{r}

p<-DimPlot(MOL ,reduction = "umap_MOL",group.by = "McNair_2025_celltype")+
  labs(x = "UMAP 1", y = "UMAP 2") +
  ggtitle("McNair Lesion type") +
  theme_minimal() +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      size = 24,
      face = "bold"
    ),
    axis.title.x = element_text(size = 20),
    axis.title.y = element_text(size = 20),
    legend.text = element_text(size = 16),         # Legend labels
    legend.title = element_text(size = 24),  # Legend title
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank()
  )

plots <- list(Macnair_celltype_label_transfer=p)
for (nm in names(plots)) {
ggplot2::ggsave(file.path("/date/gcb/gcb_MZ/MOL_FGF_10x/Plots/", paste0(nm, ".png")), plots[[nm]], width =14, height = 8, dpi = 300)
}
p
```


10.Plot the similarity of each cluster to each pathological annotation as violin plot

```{r,fig.height=10, fig.width=20}
# Define pathology annotation columns
pathology_columns <- c(
  "prediction.score.humanized_mouse_brain_A",
  "prediction.score.humanized_mouse_brain_Astro",
  "prediction.score.humanized_mouse_brain_B",
  "prediction.score.humanized_mouse_brain_C",
  "prediction.score.humanized_mouse_brain_D",
  "prediction.score.humanized_mouse_brain_E",
  "prediction.score.humanized_mouse_brain_F",
  "prediction.score.humanized_mouse_brain_G",
  "prediction.score.OPC_1",
  "prediction.score.OPC_2"
)

# Create cleaner labels for plotting
pathology_labels <- c(
  "humanized_mouse_brain_A",
  "humanized_mouse_brain_Astro",
  "humanized_mouse_brain_B",
  "humanized_mouse_brain_C",
  "humanized_mouse_brain_D",
  "humanized_mouse_brain_E",
  "humanized_mouse_brain_F",
  "humanized_mouse_brain_G",
  "OPC_1",
  "OPC_2"
)

# Get cluster identities
clusters <- unique(MOL$MOL_fine_clustering)
print(paste("Creating cell type plots for", length(clusters), "clusters"))

# Loop through each cluster
for (cluster_id in clusters) {
  
  # Create empty data frame to store all scores
  all_scores <- data.frame()
  
  # Get cells in this cluster
  cluster_cells <- colnames(MOL)[MOL$MOL_fine_clustering == cluster_id]
  
  # Extract scores for each pathology type from metadata
  for (i in seq_along(pathology_columns)) {
    col_name <- pathology_columns[i]
    label_name <- pathology_labels[i]
    
    # Get scores for this pathology type and this cluster
    scores <- MOL@meta.data[[col_name]][MOL$MOL_fine_clustering == cluster_id]
    
    temp_df <- data.frame(
      score = as.numeric(scores),
      pathology_type = label_name,
      stringsAsFactors = FALSE
    )
    
    all_scores <- rbind(all_scores, temp_df)
  }
  
  # Create violin plot with all pathology types on x-axis
  p <- ggplot(all_scores, aes(x = pathology_type, y = score, fill = pathology_type)) +
    geom_violin(trim = TRUE, scale = "width") +
    geom_boxplot(width = 0.1, fill = "white", outlier.shape = NA) +
    ggtitle(paste("Cell type Similarity - Cluster:", cluster_id)) +
    xlab("Cell Type") +
    ylab("Similarity Score") +
    ylim(0, 1) +
    theme_classic() +
    theme(
      plot.title = element_text(hjust = 0.5, size = 36, face = "bold"),
      axis.text.x = element_text(hjust = 1, vjust = 0.5, family = "Arial", 
                                 size = 14, angle = 90),
      axis.text.y = element_text(family = "Arial", size = 20),
      axis.title.x = element_text(family = "Arial", size = 24, face = "bold"),
      axis.title.y = element_text(family = "Arial", size = 24, face = "bold"),
      legend.position = "none"
    )
  
  # Display plot in R Markdown output
  print(p)
  
  # Save plot
  filename <- file.path("/date/gcb/gcb_MZ/MOL_FGF_10x//Plots/", 
                       paste0("Pathology_similarity_cluster_", cluster_id, ".png"))
  ggplot2::ggsave(filename, p, width = 20, height = 12, dpi = 300)
  
  print(paste("✓ Saved cell type plot for cluster:", cluster_id))
}

```


#9.5 Label transfer to the Macnair dataset converted to mouse using top 2000 varibale features and subsetting only Mature humanized_mouse_brains

1.To do this analysis first we need to perform the label transfer from the Macnair dataset to the MOLs only using the 2000 varibale genes intersection and subsetting only the Mature humanized_mouse_brain population in the Macnair data

```{r}
Macnair_mouse<-readRDS("/date/gcb/gcb_EA/DataRefs2025/MS_MacNair_OLGWM_seurat/Predicted_McnairMM10_Zuchero/ms_lesions_MM10orthologs_snRNAseq_seuratV5_WMEA_updated_WMOLGdown3K.rds")

Idents(Macnair_mouse)<-"type_fine"

#select only the MOLs

Macnair_mouse_subset <- subset(Macnair_mouse, idents = c("humanized_mouse_brain_A",
  "humanized_mouse_brain_B",
  "humanized_mouse_brain_C",
  "humanized_mouse_brain_D",
  "humanized_mouse_brain_E",
  "humanized_mouse_brain_F",
  "humanized_mouse_brain_G"))  
```

```{r}
# Step 1: Get variable features from both datasets
var_features_reference <- VariableFeatures(Macnair_mouse_subset[["RNA"]])
var_features_query <- VariableFeatures(MOL[["RNA"]])

# Step 2: Find intersection of variable features
shared_features <- intersect(var_features_reference, var_features_query)

cat(sprintf("Reference variable features: %d\n", length(var_features_reference)))
cat(sprintf("Query variable features: %d\n", length(var_features_query)))
cat(sprintf("Shared variable features: %d\n", length(shared_features)))

# Step 3: Use shared features for anchor finding
transfer_anchors <- FindTransferAnchors(
  reference = Macnair_mouse_subset, 
  query = MOL,
  reference.assay = "RNA",
  query.assay = "RNA",
  reduction = "cca",
  npcs = 50,
  dims = 1:50,
  features = shared_features  # Use intersected features
)

# Step 4: Map query cell type
MOL <- MapQuery(
  anchorset = transfer_anchors, 
  reference = Macnair_mouse_subset,
  query = MOL,
  refdata = "type_fine"
)

# Step 5: Rename prediction columns cell type
cn <- colnames(MOL@meta.data)
cn[cn == "predicted.id"] <- "predicted.id_Macnair_mouse_type_fine"
cn[cn == "predicted.id.score"] <- "predicted.id.score_Macnair_mouse_type_fine"
colnames(MOL@meta.data) <- cn

```

2. Umap plot of the predicted cell types

```{r}

p<-DimPlot(MOL ,reduction = "umap_MOL",group.by = "predicted.id_Macnair_mouse_type_fine")+
  labs(x = "UMAP 1", y = "UMAP 2") +
  ggtitle("McNair Lesion type") +
  theme_minimal() +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      size = 24,
      face = "bold"
    ),
    axis.title.x = element_text(size = 20),
    axis.title.y = element_text(size = 20),
    legend.text = element_text(size = 16),         # Legend labels
    legend.title = element_text(size = 24),  # Legend title
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank()
  )

plots <- list(Macnair_celltype_label_transfer=p)
for (nm in names(plots)) {
ggplot2::ggsave(file.path("/date/gcb/gcb_MZ/MOL_FGF_10x//Plots/", paste0(nm, ".png")), plots[[nm]], width =14, height = 8, dpi = 300)
}
p
```


3. Violin plot of the cell type score similarity in each MOL cluster

```{r}
# Get cluster identities (your 5 populations)
clusters <- unique(MOL$MOL_fine_clustering)
print(paste("Creating plots for", length(clusters), "clusters"))

# Loop through each cluster
for (cluster_id in clusters) {
  
  # Create empty data frame to store all scores
  all_scores <- data.frame()
  
  # Get the prediction.score.id assay
  assay_data <- GetAssayData(MOL, assay = "prediction.score.id", slot = "data")
  
  # Get cells in this cluster
  cluster_cells <- colnames(MOL)[MOL$MOL_fine_clustering == cluster_id]
  
  # Extract scores for each of the 17 reference cell types
  for (ref_celltype in rownames(assay_data)) {
    scores <- assay_data[ref_celltype, cluster_cells]
    
    temp_df <- data.frame(
      score = as.numeric(scores),
      reference_celltype = ref_celltype,
      stringsAsFactors = FALSE
    )
    
    all_scores <- rbind(all_scores, temp_df)
  }
  
  # Create violin plot with all 17 reference cell types on x-axis
  p <- ggplot(all_scores, aes(x = reference_celltype, y = score, fill = reference_celltype)) +
    geom_violin(trim = TRUE, scale = "width") +
    geom_boxplot(width = 0.1, fill = "white", outlier.shape = NA) +
    ggtitle(paste("Cluster:", cluster_id)) +
    xlab("Reference Cell Type") +
    ylab("Similarity Score") +
    ylim(0, 1) +
    theme_classic() +
    theme(
      plot.title = element_text(hjust = 0.5, size = 36, face = "bold"),
      axis.text.x = element_text(hjust = 1, vjust = 0.5, family = "Arial", 
                                 size = 14, angle = 90),  # Smaller for 17 labels
      axis.text.y = element_text(family = "Arial", size = 20),
      axis.title.x = element_text(family = "Arial", size = 24, face = "bold"),
      axis.title.y = element_text(family = "Arial", size = 24, face = "bold"),
      legend.position = "none"
    )
  
   # Display plot in the R Markdown output
  print(p)
  
  # Save plot
  filename <- file.path("/date/gcb/gcb_MZ/MOL_FGF_10x//Plots/", 
                       paste0("Similarity_cluster_Macnair_dataset", cluster_id, ".png"))
  ggplot2::ggsave(filename, p, width = 24, height = 12, dpi = 300)
  
  print(paste("✓ Saved plot for cluster:", cluster_id))
}

```




4. Map query diagnosis 


```{r}
# Step 1: Map query diagnosis
MOL <- MapQuery(
  anchorset = transfer_anchors, 
  reference = Macnair_mouse_subset,
  query = MOL,
  refdata = "diagnosis"
)

# Step 2: Rename prediction columns cell type
cn <- colnames(MOL@meta.data)
cn[cn == "predicted.id"] <- "predicted.id_Macnair_mouse_diagnosis"
cn[cn == "predicted.id.score"] <- "predicted.id.score_Macnair_mouse_diagnosis"
colnames(MOL@meta.data) <- cn
```

5. Umap plot of the predicted diagnosis 

```{r}

p<-DimPlot(MOL ,reduction = "umap_MOL",group.by = "predicted.id_Macnair_mouse_diagnosis")+
  labs(x = "UMAP 1", y = "UMAP 2") +
  ggtitle("McNair diagnosis") +
  theme_minimal() +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      size = 24,
      face = "bold"
    ),
    axis.title.x = element_text(size = 20),
    axis.title.y = element_text(size = 20),
    legend.text = element_text(size = 16),         # Legend labels
    legend.title = element_text(size = 24),  # Legend title
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank()
  )

plots <- list(Macnair_diagnosislabel_transfer=p)
for (nm in names(plots)) {
ggplot2::ggsave(file.path("/date/gcb/gcb_MZ/MOL_FGF_10x//Plots/", paste0(nm, ".png")), plots[[nm]], width =14, height = 8, dpi = 300)
}
p
```


6. Violin plot of the diagnosis similarity in each MOL cluster

```{r}
# Get cluster identities (your 5 populations)
clusters <- unique(MOL$MOL_fine_clustering)
print(paste("Creating plots for", length(clusters), "clusters"))

# Loop through each cluster
for (cluster_id in clusters) {
  
  # Create empty data frame to store all scores
  all_scores <- data.frame()
  
  # Get the prediction.score.id assay
  assay_data <- GetAssayData(MOL, assay = "prediction.score.id", slot = "data")
  
  # Get cells in this cluster
  cluster_cells <- colnames(MOL)[MOL$MOL_fine_clustering == cluster_id]
  
  # Extract scores for each of the 17 reference cell types
  for (ref_celltype in rownames(assay_data)) {
    scores <- assay_data[ref_celltype, cluster_cells]
    
    temp_df <- data.frame(
      score = as.numeric(scores),
      reference_celltype = ref_celltype,
      stringsAsFactors = FALSE
    )
    
    all_scores <- rbind(all_scores, temp_df)
  }
  
  # Create violin plot with all 17 reference cell types on x-axis
  p <- ggplot(all_scores, aes(x = reference_celltype, y = score, fill = reference_celltype)) +
    geom_violin(trim = TRUE, scale = "width") +
    geom_boxplot(width = 0.1, fill = "white", outlier.shape = NA) +
    ggtitle(paste("Cluster:", cluster_id)) +
    xlab("Reference Cell Type") +
    ylab("Similarity Score") +
    ylim(0, 1) +
    theme_classic() +
    theme(
      plot.title = element_text(hjust = 0.5, size = 36, face = "bold"),
      axis.text.x = element_text(hjust = 1, vjust = 0.5, family = "Arial", 
                                 size = 14, angle = 90),  # Smaller for 17 labels
      axis.text.y = element_text(family = "Arial", size = 20),
      axis.title.x = element_text(family = "Arial", size = 24, face = "bold"),
      axis.title.y = element_text(family = "Arial", size = 24, face = "bold"),
      legend.position = "none"
    )
  
   # Display plot in the R Markdown output
  print(p)
  
  # Save plot
  filename <- file.path("/date/gcb/gcb_MZ/MOL_FGF_10x//Plots/", 
                       paste0("Similarity_cluster_Macnair_dataset", cluster_id, ".png"))
  ggplot2::ggsave(filename, p, width = 24, height = 12, dpi = 300)
  
  print(paste("✓ Saved plot for cluster:", cluster_id))
}

```

7. Map query lesion type


```{r}
# Step 1: Map query diagnosis
MOL <- MapQuery(
  anchorset = transfer_anchors, 
  reference = Macnair_mouse_subset,
  query = MOL,
  refdata = "lesion_type"
)

# Step 2: Rename prediction columns cell type
cn <- colnames(MOL@meta.data)
cn[cn == "predicted.id"] <- "predicted.id_Macnair_mouse_lesion_type"
cn[cn == "predicted.id.score"] <- "predicted.id.score_Macnair_mouse_lesion_type"
colnames(MOL@meta.data) <- cn
```

8. Umap plot of the predicted lesion type 

```{r}

p<-DimPlot(MOL ,reduction = "umap_MOL",group.by = "predicted.id_Macnair_mouse_lesion_type")+
  labs(x = "UMAP 1", y = "UMAP 2") +
  ggtitle("McNair lesion type") +
  theme_minimal() +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      size = 24,
      face = "bold"
    ),
    axis.title.x = element_text(size = 20),
    axis.title.y = element_text(size = 20),
    legend.text = element_text(size = 16),         # Legend labels
    legend.title = element_text(size = 24),  # Legend title
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank()
  )

plots <- list(Macnair_diagnosislabel_transfer=p)
for (nm in names(plots)) {
ggplot2::ggsave(file.path("/date/gcb/gcb_MZ/MOL_FGF_10x//Plots/", paste0(nm, ".png")), plots[[nm]], width =14, height = 8, dpi = 300)
}
p
```

9. Violin plot of the lesion type similarity in each MOL cluster

```{r}
# Get cluster identities (your 5 populations)
clusters <- unique(MOL$MOL_fine_clustering)
print(paste("Creating plots for", length(clusters), "clusters"))

# Loop through each cluster
for (cluster_id in clusters) {
  
  # Create empty data frame to store all scores
  all_scores <- data.frame()
  
  # Get the prediction.score.id assay
  assay_data <- GetAssayData(MOL, assay = "prediction.score.id", slot = "data")
  
  # Get cells in this cluster
  cluster_cells <- colnames(MOL)[MOL$MOL_fine_clustering == cluster_id]
  
  # Extract scores for each of the 17 reference cell types
  for (ref_celltype in rownames(assay_data)) {
    scores <- assay_data[ref_celltype, cluster_cells]
    
    temp_df <- data.frame(
      score = as.numeric(scores),
      reference_celltype = ref_celltype,
      stringsAsFactors = FALSE
    )
    
    all_scores <- rbind(all_scores, temp_df)
  }
  
  # Create violin plot with all 17 reference cell types on x-axis
  p <- ggplot(all_scores, aes(x = reference_celltype, y = score, fill = reference_celltype)) +
    geom_violin(trim = TRUE, scale = "width") +
    geom_boxplot(width = 0.1, fill = "white", outlier.shape = NA) +
    ggtitle(paste("Cluster:", cluster_id)) +
    xlab("Reference Cell Type") +
    ylab("Similarity Score") +
    ylim(0, 1) +
    theme_classic() +
    theme(
      plot.title = element_text(hjust = 0.5, size = 36, face = "bold"),
      axis.text.x = element_text(hjust = 1, vjust = 0.5, family = "Arial", 
                                 size = 14, angle = 90),  # Smaller for 17 labels
      axis.text.y = element_text(family = "Arial", size = 20),
      axis.title.x = element_text(family = "Arial", size = 24, face = "bold"),
      axis.title.y = element_text(family = "Arial", size = 24, face = "bold"),
      legend.position = "none"
    )
  
   # Display plot in the R Markdown output
  print(p)
  
  # Save plot
  filename <- file.path("/date/gcb/gcb_MZ/MOL_FGF_10x//Plots/", 
                       paste0("Similarity_cluster_Macnair_dataset", cluster_id, ".png"))
  ggplot2::ggsave(filename, p, width = 24, height = 12, dpi = 300)
  
  print(paste("✓ Saved plot for cluster:", cluster_id))
}

```



Save dataset

```{r}

saveRDS(MOL,"/date/gcb/gcb_MZ/MOL_FGF_10x/MOL.rds")
```



