---
title: "Humanized mice EAE analysis Brain"
author: "Mattia Zaghi"
date: "2026-03-06"
output:
  word_document: default
  html_document: default
  pdf_document: default
---

## What is this vignette for?
This Vignette contains all the necessary data analysis steps followed to analyzed the single cell RNA-seq of Mature humanized_mouse_braindendrocytes treated with FGF and Tomato controls including the code to generate the different figures.


##1.Load packages, configure directories and create the Seurat object

This section of the code is necessary to load all the necessary packages for the analysis, configure the directories and create the Seurat object that is going to be used for the rest of the analysis.

##1.1 Load packages, configure directories

1.First load all the necessary packages in order to run all the subsequent steps & set the working directory where all the output of this analysis will be saved and stored

```{r setup, include=FALSE}
library(Seurat)
library(future)
library(tidyverse)
library(dplyr)
library(ggthemes)
library(scales)
library(cowplot)
library(RColorBrewer)
library(openxlsx)
library(homologene)
library(clusterProfiler)
library(org.Mm.eg.db)
library(org.Hs.eg.db)
library(purrr)
library(forcats)
library(readxl)
library(hdf5r)
library(patchwork)
library(pals)
library(knitr)
library(DoubletFinder)
opts_knit$set(root.dir = "/date/gcb/gcb_MZ/Humanized_mice_OD/")# change accordingly

#source("/home/mattia/FGF_signaling_induces_disease-associated_humanized_mouse_braindendrocytes/functions.R")
```

##1.2 Create the seurat object 

Let's start with loading the actual data. We will load separately all the brain samples samples from the cell ranger output and then everything will be merged into a unique object thet will be called brain hMS. 

1.load all the seurat object corrected by cell bender 
```{r}

# Base folder containing all samples
base_dir <- "/date/gcb/gcb_MZ/Humanized_mice_OD"

# Find all sample folders that contain "brain" in the name
brain_samples <- list.dirs(base_dir, recursive = FALSE, full.names = FALSE)
brain_samples <- brain_samples[grepl("brain", brain_samples)]

# Load each sample and create a Seurat object (with unique cell names)
seurat_list <- list()
for (sample in brain_samples) {
  h5_path <- file.path(base_dir, sample, "RNA_AAAAGGGG", "cellbender_output_seurat.h5")
  if (!file.exists(h5_path)) {
    warning("Missing file: ", h5_path)
    next
  }
  message("Loading ", sample)

  data <- Read10X_h5(filename = h5_path, use.names = TRUE)

  # Normalize feature names: collapse underscores and multiple dashes into a single dash
  fn <- rownames(data)
  fn <- gsub("_+", "-", fn)      # underscores -> dash
  fn <- gsub("-{2,}", "-", fn)   # collapse multiple dashes
  rownames(data) <- fn

  # Prefix cell barcodes with sample name so they are unique across samples
  colnames(data) <- paste0(sample, "_", colnames(data))

  seurat_obj <- CreateSeuratObject(
    counts = data,
    project = sample,
    min.cells = 3,
    min.features = 200
  )

  # Per-sample QC before merging
  seurat_obj[["percent.mt.mm10"]] <- PercentageFeatureSet(
  seurat_obj,
  pattern = "mt-"
)

# human mito (hg38)
seurat_obj[["percent.mt.hg38"]] <- PercentageFeatureSet(
  seurat_obj,
  pattern = "MT-"
)
seurat_obj[["percent.mt"]] <- seurat_obj[["percent.mt.mm10"]] + seurat_obj[["percent.mt.hg38"]]

  seurat_obj <- subset(seurat_obj, subset = percent.mt < 5)

  # Light preprocessing for doublet detection (on each sample independently using SCT)
  seurat_obj <- SCTransform(seurat_obj, verbose = FALSE)
  seurat_obj <- RunPCA(seurat_obj, verbose = FALSE)
  seurat_obj <- FindNeighbors(seurat_obj, dims = 1:10, verbose = FALSE)
  seurat_obj <- FindClusters(seurat_obj, resolution = 0.1, verbose = FALSE)
  seurat_obj <- RunUMAP(seurat_obj, dims = 1:10, verbose = FALSE)

  umap_before <- DimPlot(seurat_obj, reduction = "umap", group.by = "seurat_clusters") + ggtitle(paste(sample, "before doublet removal"))

  # Per-sample DoubletFinder (SCT input)
  nExp_poi <- round(0.075 * ncol(seurat_obj))
  nExp_poi.adj <- round(nExp_poi * 1.25)
  pK <- 0.09
  pN <- 0.25

  seurat_obj <- doubletFinder(seurat_obj, PCs = 1:10, pN = pN, pK = pK, nExp = nExp_poi, reuse.pANN = NULL, sct = TRUE)

  df_col <- grep("pANN_", colnames(seurat_obj@meta.data), value = TRUE)[1]
  if (is.na(df_col)) stop("DoubletFinder pANN metadata not found")

  seurat_obj <- doubletFinder(seurat_obj, PCs = 1:10, pN = pN, pK = pK, nExp = nExp_poi.adj, reuse.pANN = df_col, sct = TRUE)

  df_class_cols <- grep("DF.classifications_", colnames(seurat_obj@meta.data), value = TRUE)
  if (length(df_class_cols) == 0) stop("DoubletFinder classification column not found")

  # Use the most recent classification column (should be the second pass)
  df_class_col <- tail(df_class_cols, 1)
  message("Using doublet classification column: ", df_class_col)

  sel_singlets <- seurat_obj@meta.data[[df_class_col]] == "Singlet"
  if (sum(sel_singlets, na.rm = TRUE) == 0) {
    stop("No cells annotated as Singlet in ", df_class_col)
  }
  seurat_obj <- subset(seurat_obj, cells = rownames(seurat_obj@meta.data)[sel_singlets])

  # Recompute UMAP after doublet removal (SCT after singlet filtering)
  # Remove the old SCT assay 
  DefaultAssay(seurat_obj) <- "RNA"
  seurat_obj[["SCT"]] <- NULL
  seurat_obj <- SCTransform(seurat_obj, verbose = TRUE)
  seurat_obj <- RunPCA(seurat_obj, verbose = TRUE)
  seurat_obj <- FindNeighbors(seurat_obj, dims = 1:10, verbose = TRUE)
  seurat_obj <- FindClusters(seurat_obj, resolution = 0.1, verbose = TRUE)
  seurat_obj <- RunUMAP(seurat_obj, dims = 1:10, verbose = TRUE)

  umap_after <- DimPlot(seurat_obj, reduction = "umap", group.by = "seurat_clusters") + ggtitle(paste(sample, "after doublet removal"))

  # Print per-sample before/after UMAP side-by-side
  print(umap_before + umap_after)

  seurat_list[[sample]] <- seurat_obj
}
```



2.Now all the different samples are going to be merged in one unique object 

```{r}
seurat_list <- lapply(names(seurat_list), function(s) {
  obj <- seurat_list[[s]]
  obj$sample <- s
  obj
})

humanized_mouse_brain <- Reduce(function(x, y) merge(x, y), seurat_list)

rm(seurat_list)
# Remove metadata from previous single-sample doublet removal analysis
humanized_mouse_brain@meta.data[["nCount_SCT"]]      <- NULL
humanized_mouse_brain@meta.data[["nFeature_SCT"]]    <- NULL
humanized_mouse_brain@meta.data[["SCT_snn_res.0.1"]] <- NULL
humanized_mouse_brain@meta.data[["seurat_clusters"]] <- NULL
humanized_mouse_brain@meta.data[["pANN_0.25_0.09_879"]] <- NULL
humanized_mouse_brain@meta.data[["pANN_0.25_0.09_867"]] <- NULL
```

filter away cells that have a very high mitocondrial content 

```{r}
humanized_mouse_brain <- subset(humanized_mouse_brain, subset = percent.mt < 5)
```


##2.Quality controls and cell selection

In this sections quality controls based selection of good will be performed on the merged object will be performed before proceeding with the rest of then analysis. 

1.Let's first plot some QCs metrics to have a look at the global dataset, this will give a general idea of the dataset quality

```{r}
# add the mitochondrial gene precentage 
# mouse mito (mm10)
humanized_mouse_brain[["percent.mt.mm10"]] <- PercentageFeatureSet(
  humanized_mouse_brain,
  pattern = "mt-"
)

# human mito (hg38)
humanized_mouse_brain[["percent.mt.hg38"]] <- PercentageFeatureSet(
  humanized_mouse_brain,
  pattern = "MT-"
)
humanized_mouse_brain[["percent.mt"]] <- humanized_mouse_brain[["percent.mt.mm10"]] + humanized_mouse_brain[["percent.mt.hg38"]]
```

check human vs mouse cells

```{r}
# Assume this is your merged object
obj <- humanized_mouse_brain
DefaultAssay(obj) <- "RNA"
obj <- JoinLayers(obj)
# Count UMIs per cell for each species (adjust prefixes as needed)
human_umis <- Matrix::colSums(
  GetAssayData(obj, assay = "RNA", layer = "counts")[grep("^GRCh38-", rownames(obj)), , drop = FALSE]
)
mouse_umis <- Matrix::colSums(
  GetAssayData(obj, assay = "RNA", layer = "counts")[grep("^mm10-", rownames(obj)), , drop = FALSE]
)

# Assign species label based on which count is higher
obj$species <- ifelse(human_umis > mouse_umis, "human", "mouse")

# Optionally mark ambiguous/mixed cells (e.g., 20–80% human)
ratio <- human_umis / (human_umis + mouse_umis)
obj$species[ratio > 0.2 & ratio < 0.8] <- "mixed"

# Write back
humanized_mouse_brain <- obj

#how many cells per sample
table(humanized_mouse_brain$orig.ident)

#how many cell per species
table(humanized_mouse_brain$species)

#how many cell per replicate
table(humanized_mouse_brain$sample)

```


```{r,fig.height=10, fig.width=10}
# Visualize QC metrics as a violin plot

QC_RNA <- data.frame(sample =  humanized_mouse_brain@meta.data[["sample"]],
                     Genes = humanized_mouse_brain@meta.data[["nFeature_RNA"]],
                     RNA = humanized_mouse_brain@meta.data[["nCount_RNA"]],
                     Mt = humanized_mouse_brain@meta.data[["percent.mt"]])
QC_RNA$sample <- as.factor(QC_RNA$sample)

# Function to add mean and median annotations for each sample above the violin plots
add_mean_median_annotation <- function(plot, data, x_var, y_var) {
  samples <- unique(data[[x_var]])
  for (sample in samples) {
    sample_data <- data[data[[x_var]] == sample, ]
    mean_value <- mean(sample_data[[y_var]], na.rm = TRUE)
    median_value <- median(sample_data[[y_var]], na.rm = TRUE)
    plot <- plot + 
      annotate("text", x = sample, y = max(sample_data[[y_var]], na.rm = TRUE) * 1.05, 
               label = paste("Mean:", round(mean_value, 2)), color = "blue", size = 5, fontface = "bold") +
      annotate("text", x = sample, y = max(sample_data[[y_var]], na.rm = TRUE) * 1.50, 
               label = paste("Median:", round(median_value, 2)), color = "red", size = 5, fontface = "bold")
  }
  return(plot)
}

# Plot for RNA
Transcripts <- ggplot(QC_RNA) +
  aes(x = sample, y = RNA, fill = sample, color = sample) +
  geom_violin(trim = TRUE, color = "#000000") +
  geom_boxplot(width = 0.1, color = "#000000", fill = "#ffffff", outlier.shape = NA) +
  scale_fill_manual(values = c(
"#1f77b4",  # blue
            "#ff7f0e",  # orange
            "#2ca02c",  # green
            "#d62728",  # red
            "#9467bd"
))+
  scale_color_manual(values = c(
"#1f77b4",  # blue
            "#ff7f0e",  # orange
            "#2ca02c",  # green
            "#d62728",  # red
            "#9467bd"
))+
  ggthemes::theme_base() +
  xlab("") +
  ylab("Number of UMI per cell") +
  theme(panel.border = element_rect()) +
  theme_classic() + theme(legend.position = "none") +
  theme(axis.text.x = element_text(size = 18, family = "Arial", angle = 45, hjust = 1),
        axis.text.y = element_text(size = 18, family = "Arial"),
        axis.title.y = element_text(size = 18, family = "Arial"),
        axis.line = element_line(size = 1))

Transcripts <- add_mean_median_annotation(Transcripts, QC_RNA, "sample", "RNA")

# Plot for Genes
Genes <- ggplot(QC_RNA) +
  aes(x = sample, y = Genes, fill = sample, color = sample) +
  geom_violin(trim = TRUE, color = "#000000") +
  geom_boxplot(width = 0.1, color = "#000000", fill = "#ffffff", outlier.shape = NA) +
  scale_fill_manual(values = c(
"#1f77b4",  # blue
            "#ff7f0e",  # orange
            "#2ca02c",  # green
            "#d62728",  # red
            "#9467bd"
))+
  scale_color_manual(values = c(
"#1f77b4",  # blue
            "#ff7f0e",  # orange
            "#2ca02c",  # green
            "#d62728",  # red
            "#9467bd"
))+
  ggthemes::theme_base() +
  xlab("") +
  ylab("Number of Genes per cell") +
  theme(panel.border = element_rect()) +
  theme_classic() + theme(legend.position = "none") +
  theme(axis.text.x = element_text(size = 18, family = "Arial", angle = 45, hjust = 1),
        axis.text.y = element_text(size = 18, family = "Arial"),
        axis.title.y = element_text(size = 18, family = "Arial"),
        axis.line = element_line(size = 1))

Genes <- add_mean_median_annotation(Genes, QC_RNA, "sample", "Genes")

# Plot for Mt
Mt <- ggplot(QC_RNA) +
  aes(x = sample, y =   Mt, fill = sample, color = sample) +
  geom_violin(trim = TRUE, color = "#000000") +
  geom_boxplot(width = 0.1, color = "#000000", fill = "#ffffff", outlier.shape = NA) +
  scale_fill_manual(values = c(
"#1f77b4",  # blue
            "#ff7f0e",  # orange
            "#2ca02c",  # green
            "#d62728",  # red
            "#9467bd"
))+
  scale_color_manual(values = c(
"#1f77b4",  # blue
            "#ff7f0e",  # orange
            "#2ca02c",  # green
            "#d62728",  # red
            "#9467bd"
))+
  ggthemes::theme_base() +
  xlab("") +
  ylab("Percentage of Mt genes") +
  theme(panel.border = element_rect()) +
  theme_classic() + theme(legend.position = "none") +
  theme(axis.text.x = element_text(size = 18, family = "Arial", angle = 45, hjust = 1),
        axis.text.y = element_text(size = 18, family = "Arial"),
        axis.title.y = element_text(size = 18, family = "Arial"),
        axis.line = element_line(size = 1))

Mt <- add_mean_median_annotation(Mt, QC_RNA, "sample", "Mt")

Transcripts
Genes
Mt

```

let's also visualize how is the relationship between counts, genes and mito content

```{r}

plot1 <- FeatureScatter(humanized_mouse_brain, feature1 = "nCount_RNA", feature2 = "percent.mt")
plot2 <- FeatureScatter(humanized_mouse_brain, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
plot1 + plot2
```







##3.Data integration, dimensionlity reduction and clustering

In this section we will perform the data integration, dimensionality reduction and clustering of the

##3.1 Dimensionality reduction and clustering without data integration
Despite in this experiment there are not multiple data batches and replicate we will first perform a clustering with data ingtegration and without and look at the difference

1.Data normalization, variable features finding PCA

```{r}
options(future.globals.maxSize = 100e+09)

humanized_mouse_brain[["SCT"]] <- NULL

# Log-normalized analysis on RNA
humanized_mouse_brain <- NormalizeData(humanized_mouse_brain, assay = "RNA")
humanized_mouse_brain <- FindVariableFeatures(humanized_mouse_brain, assay = "RNA")
humanized_mouse_brain <- ScaleData(humanized_mouse_brain, assay = "RNA")
humanized_mouse_brain <- RunPCA(humanized_mouse_brain, assay = "RNA", reduction.name = "pca_lognorm")

# SCT analysis on SCT assay
humanized_mouse_brain <- SCTransform(humanized_mouse_brain, verbose = TRUE)
humanized_mouse_brain <- FindVariableFeatures(humanized_mouse_brain, assay = "SCT")
humanized_mouse_brain <- RunPCA(humanized_mouse_brain, assay = "SCT", reduction.name = "pca_SCT")
```

2.Having a look at the elbow plot to understand how many PCs are really informative, this value will be used in the subsequent analysis steps 
```{r}
p1 <- ElbowPlot(humanized_mouse_brain, reduction = "pca_SCT")
p1
```
After 15 SD the elbow plot becomes flat indicating that the majority of the information is explained by the first 15 dimension, this value will be used for the rest of the analysis.

3.Clustering without data integration

```{r}
humanized_mouse_brain <- FindNeighbors(humanized_mouse_brain, dims = 1:15, reduction = "pca_SCT")
humanized_mouse_brain <- FindClusters(humanized_mouse_brain, resolution = 0.1, cluster.name = "unintegrated_clusters")
```
4. Plot UMAP with samples and clustering
```{r,fig.height=10, fig.width=10}
humanized_mouse_brain <- RunUMAP(humanized_mouse_brain, dims = 1:15, reduction = "pca_SCT", reduction.name = "umap.unintegrated")
# visualize by batch and cell type annotation

my_colors <- c(
  "#1F77B4", "#FF7F0E", "#2CA02C", "#D62728", "#9467BD",
  "#8C564B", "#17BECF", "#BCBD22", "#7F7F7F", "#e377c2",
  "#aec7e8", "#ffbb78", "#98df8a"
)


p1 <- DimPlot(humanized_mouse_brain, reduction = "umap.unintegrated", group.by = c("unintegrated_clusters"), cols = my_colors) + labs(x = "UMAP 1", y = "UMAP 2") + #UMAP with samples
 theme_minimal() +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      size = 24,
      face = "bold"
    ),
    axis.title.x = element_text(size = 20),
    axis.title.y = element_text(size = 20),
    legend.text = element_text(size = 16),         # Legend labels
    legend.title = element_text(size = 24, face = "bold"),  # Legend title
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank()
  )

p2 <- DimPlot(humanized_mouse_brain, reduction = "umap.unintegrated", group.by = c("orig.ident")) + labs(x = "UMAP 1", y = "UMAP 2") + #UMAP with samples
 theme_minimal() +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      size = 24,
      face = "bold"
    ),
    axis.title.x = element_text(size = 20),
    axis.title.y = element_text(size = 20),
    legend.text = element_text(size = 16),         # Legend labels
    legend.title = element_text(size = 24, face = "bold"),  # Legend title
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank()
  )

p3 <- DimPlot(humanized_mouse_brain, reduction = "umap.unintegrated", group.by = c("sample")) + labs(x = "UMAP 1", y = "UMAP 2") + #UMAP with samples
 theme_minimal() +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      size = 24,
      face = "bold"
    ),
    axis.title.x = element_text(size = 20),
    axis.title.y = element_text(size = 20),
    legend.text = element_text(size = 16),         # Legend labels
    legend.title = element_text(size = 24, face = "bold"),  # Legend title
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank()
  )


p1 
p2
p3
saveRDS(humanized_mouse_brain,"/date/gcb/gcb_MZ/Humanized_mice_OD/humanized_mouse_brain_doublets.rds")
```

```{r,fig.height=10, fig.width=20}
colors_vector <- c(
  "CFA_brain_1" = "#E41A1C", 
  "CFA_brain_2" = "#377EB8", 
  "EAE_brain_1" = "#984EA3", 
  "EAE_brain_2" = "#FF7F00", 
  "EAE_brain_3" = "#FFFF33"
)

# Create the contingency table
t <- table(Cluster=humanized_mouse_brain$unintegrated_clusters, Batch=humanized_mouse_brain@meta.data[["sample"]])
t <- t[, rev(names(colors_vector))]

# Convert the table to a data frame
df <- as.data.frame(t)

# Calculate the percentage of each plate within each cluster
df <- df %>%
  group_by(Cluster) %>% mutate(Percent = Freq / sum(Freq) * 100)

# Plot the data
ggplot(df, aes(x = Cluster, y = Percent, fill = Batch)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_y_continuous(labels = percent_format(), expand = c(0, 0)) +
  scale_fill_manual(values = rev(colors_vector)) +
  theme_classic() +
  theme(
    axis.text.x = element_text(hjust = 1, vjust = 0.5, family = "Arial", size = 16),
    axis.text.y = element_text(family = "Arial", size = 16),
    axis.title.x = element_text(family = "Arial", size = 18),
    axis.title.y = element_text(family = "Arial", size = 18),
    legend.text = element_text(size = 16, family = "Arial"),
    legend.title = element_text(size = 18, family = "Arial")
  ) +
  labs(y = "Percentage of cells", x = "Cluster", fill = "Animal") +
  guides(fill = guide_legend(ncol = 3, title.position = "top", title.hjust = 0.5))
```







##3.2 Dimensionality reduction and clustering with data integration
In this section the dimensionality reduction will be performed including data integration and SCT transformation which should make the dataset more homogeneus and prevent any differences that are not due to biological variation.

1. SCT transformation, PCA running, clustering and dimensionality reduction again using 15 PCs as shown by the elbowplot above

```{r}

humanized_mouse_brain <- IntegrateLayers(
  object = humanized_mouse_brain,
  method = HarmonyIntegration,
  orig.reduction = "pca_SCT", new.reduction = "harmony",
  normalization.method = "SCT",
  verbose = FALSE
)
humanized_mouse_brain <- FindNeighbors(humanized_mouse_brain, dims = 1:15, reduction = "harmony")
humanized_mouse_brain <- FindClusters(humanized_mouse_brain, resolution = 0.1, cluster.name = "harmony_clusters")
```

2. Perform Dimensionality reduction
```{r}
humanized_mouse_brain <- RunUMAP(humanized_mouse_brain, dims = 1:15, reduction = "harmony", reduction.name = "umap.harmony")
```

3.Plot UMAP with clusters and sample to check that the data integration effectively results in a better homogeneus dataset 

```{r}


pal13 <- pals::glasbey(13)         # 13 very distinct colors
pal13

p1 <- DimPlot(humanized_mouse_brain, reduction = "umap.harmony", group.by = c("harmony_clusters"), cols = pal13) + labs(x = "UMAP 1", y = "UMAP 2") + theme_minimal() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
     plot.title = element_blank()

  )


p2 <- DimPlot(humanized_mouse_brain, reduction = "umap.harmony", group.by = c("sample")) + labs(x = "UMAP 1", y = "UMAP 2") +
  ggtitle("") +  # Add your title here
 theme_minimal() +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      size = 24,
      face = "bold"
    ),
    axis.title.x = element_text(size = 20),
    axis.title.y = element_text(size = 20),
    legend.text = element_text(size = 16),         # Legend labels
    legend.title = element_text(size = 24, face = "bold"),  # Legend title
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank()
  )

col <- c(
  mouse = "grey80",
  human = "firebrick3"
)


p3 <- DimPlot(humanized_mouse_brain, reduction = "umap.SCT", group.by = c("species"), cols = col) + labs(x = "UMAP 1", y = "UMAP 2") +
  ggtitle("") +  # Add your title here
 theme_minimal() +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      size = 24,
      face = "bold"
    ),
    axis.title.x = element_text(size = 20),
    axis.title.y = element_text(size = 20),
    legend.text = element_text(size = 16),         # Legend labels
    legend.title = element_text(size = 24, face = "bold"),  # Legend title
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank()
  )

p1
p2
p3
```


clusters proportions

```{r}
colors_vector <- c(
  "CFA_brain_1" = "#E41A1C", 
  "CFA_brain_2" = "#377EB8", 
  "EAE_brain_1" = "#984EA3", 
  "EAE_brain_2" = "#FF7F00", 
  "EAE_brain_3" = "#FFFF33"
)

# Create the contingency table
t <- table(Cluster=humanized_mouse_brain$harmony_clusters, Batch=humanized_mouse_brain@meta.data[["sample"]])
t <- t[, rev(names(colors_vector))]

# Convert the table to a data frame
df <- as.data.frame(t)

# Calculate the percentage of each plate within each cluster
df <- df %>%
  group_by(Cluster) %>% mutate(Percent = Freq / sum(Freq) * 100)

# Plot the data
ggplot(df, aes(x = Cluster, y = Percent, fill = Batch)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_y_continuous(labels = percent_format(), expand = c(0, 0)) +
  scale_fill_manual(values = rev(colors_vector)) +
  theme_classic() +
  theme(
    axis.text.x = element_text(hjust = 1, vjust = 0.5, family = "Arial", size = 16),
    axis.text.y = element_text(family = "Arial", size = 16),
    axis.title.x = element_text(family = "Arial", size = 18),
    axis.title.y = element_text(family = "Arial", size = 18),
    legend.text = element_text(size = 16, family = "Arial"),
    legend.title = element_text(size = 18, family = "Arial")
  ) +
  labs(y = "Percentage of cells", x = "Cluster", fill = "Animal") +
  guides(fill = guide_legend(ncol = 3, title.position = "top", title.hjust = 0.5))
```


7.Compute and save cluster markers of the different population in the MOL subclustering. This will be one of the supplementary tables to be attached to be included. 

```{r}

# find markers for every cluster compared to all remaining cells, report only the positive
# ones

DefaultAssay(humanized_mouse_brain) <- "SCT"
humanized_mouse_brain <- PrepSCTFindMarkers(humanized_mouse_brain, assay = "SCT", verbose = TRUE)
humanized_mouse_brain_markers <- FindAllMarkers(humanized_mouse_brain, only.pos = TRUE)


 # Create a new workbook
wb <- createWorkbook()

# Loop through each cluster and add data to a new sheet
for (cluster_value in unique(humanized_mouse_brain_markers$cluster)) {
  cluster_data <- subset(humanized_mouse_brain_markers, cluster == cluster_value)
  sheet_name <- paste0("Cluster_", cluster_value)
  addWorksheet(wb, sheet_name)
  writeData(wb, sheet_name, cluster_data)
}

# Save the workbook to a file
saveWorkbook(wb, "/date/gcb/gcb_MZ/Humanized_mice_OD/Seurat_clusters_broad_brain.xlsx", overwrite = TRUE)
```

4.Now let's plot some markers genes to rename and homogenize clusters. After that the final steps of the analysis will be performed.The code use to generate the plots is used in the Supplementary Fig.S3C.


```{r,fig.height=80, fig.width=80}
markers <- c(
  # Cluster 0 - Cerebellar Granule Cells
  "mm10-Gabra6", "mm10-Cbln1", "mm10-Cbln3", "mm10-Neurod1", "mm10-Barhl2",
  # Cluster 1 - Excitatory Cortical Neurons (Glutamatergic)
  "mm10-Tbr1", "mm10-Satb2", "mm10-Emx1", "mm10-Neurod6",
  # Cluster 2 - GABAergic Interneurons (SST/VIP subtypes)
  "mm10-Lhx6", "mm10-Lhx8", "mm10-Sst", "mm10-Vip", "mm10-Tac2", "mm10-Slc5a7", "mm10-Nxph2",
  # Cluster 3 - GABAergic Inhibitory Neurons (Striatal/MGE-derived)
  "mm10-Dlx1", "mm10-Dlx2", "mm10-Gad1", "mm10-Sp9", "mm10-Meis2", "mm10-Pbx3",
  # Cluster 4 - Mature Oligodendrocytes
  "mm10-Plp1", "mm10-Mbp", "mm10-Apod", "mm10-Ptgds",
  # Cluster 5 - Astrocytes
  "mm10-Atp1a2", "mm10-Slc1a3", "mm10-Apoe", "mm10-Gja1", "mm10-Gli1", "mm10-Mfge8",
  # Cluster 6 - Medium Spiny Neurons (Striatal)
  "mm10-Drd1", "mm10-Drd2", "mm10-Adora2a", "mm10-Gpr88", "mm10-Rgs9", "mm10-Rarb",
  # Cluster 7 - Myelinating Oligodendrocytes
  "mm10-Klk6", "mm10-Gjc2", "mm10-Ugt8a", "mm10-Gjb1",
  # Cluster 8 - Cerebellar Inhibitory Interneurons (MLI/Pvalb+)
  "mm10-Pvalb", "mm10-Pax2", "mm10-Kit", "mm10-Tfap2a", "mm10-Tfap2b", "mm10-Skor1",
  # Cluster 9 - Microglia
  "mm10-Tmem119", "mm10-Hexb", "mm10-Trem2", "mm10-Tyrobp", "mm10-Inpp5d", "mm10-Itgam", "mm10-Cd14", "mm10-Adgre1",
  # Cluster 10 - Thalamic Neurons
  "mm10-Synpo2", "mm10-Zfhx3", "mm10-Shox2", "mm10-Lef1", "mm10-Slitrk6",
  # Cluster 11 - Oligodendrocyte Precursor Cells (OPCs)
  "mm10-Pdgfra", "mm10-Cspg4", "mm10-Vcan",
  # Cluster 12 - Purkinje Cells (Cerebellar)
  "mm10-Car8", "mm10-Pcp2", "mm10-Ppp1r17", "mm10-Col18a1", "mm10-Atp2a3"
)

# 1) create a list of plots (one per gene)
plots <- FeaturePlot(
  humanized_mouse_brain,
  features = markers,
  reduction = "umap.unintegrated",
  cols = c("lightgrey", "magenta"),
  combine = FALSE
)

# 2) apply consistent theme / title formatting to every plot
plots <- lapply(plots, function(p) {
  p +
    labs(x = "UMAP 1", y = "UMAP 2") +
    theme_minimal() +
    theme(
      plot.title = element_text(hjust = 0.5, size = 64, face = "bold"),
      axis.title.x = element_text(size = 50),
      axis.title.y = element_text(size = 50),
      legend.text = element_text(size = 50),
      legend.title = element_text(size = 52),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      axis.text.x = element_blank(),
      axis.text.y = element_blank()
    )
})

# 3) combine into a grid (e.g., 2 columns)
p <- wrap_plots(plots, ncol = 4)
p
```


5.Cluster renaming, based on the very broad classification OL, OPCs or microglia, this clustering to subset only MOLs to subsequently use them in the next analysis. The UMAP plot with the samples correponds to fig.3C and the UMAP plot with the broad cell types corresponds to Fig.S3B. 

```{r,fig.height=10, fig.width=10}
broad_clustering <- c("Cerebellar_Granule_Cells","Excitatory_Cortical_Neu","SST_Gaba_Neu","Stria_MGE_Inter_Neu","MOL",
                      "Astrocytes","Medium_Spiny_Neu_Stria","Myelin_MOL","Cerebellar_Inhib_Interneu_Pvalb+",
                      "Microglia","Thal_Neu","OPcs","Purkinje_Cells")

names(broad_clustering) <- levels(humanized_mouse_brain)
humanized_mouse_brain <- RenameIdents(humanized_mouse_brain, broad_clustering)
humanized_mouse_brain$broad_clustering <- Idents(humanized_mouse_brain) #create a new metadata colums with the broad clustering

# Check the new identities
table(Idents(humanized_mouse_brain))

my_colors <- c(
  "#1F77B4", "#FF7F0E", "#2CA02C", "#D62728", "#9467BD",
  "#8C564B", "#17BECF", "#BCBD22", "#7F7F7F", "#e377c2",
  "#aec7e8", "#ffbb78"
)

p1 <- DimPlot(humanized_mouse_brain, reduction = "umap.rpca_SCT", group.by = "orig.ident") + labs(x = "UMAP 1", y = "UMAP 2") + #UMAP with samples
 theme_minimal() +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      size = 24,
      face = "bold"
    ),
    axis.title.x = element_text(size = 20),
    axis.title.y = element_text(size = 20),
    legend.text = element_text(size = 16),         # Legend labels
    legend.title = element_text(size = 24, face = "bold"),  # Legend title
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank()
  )

colors_3 <- c("red", "#56B4E9", "#009E73")  # select the colors you want for the UMAP

p2 <- DimPlot(humanized_mouse_brain, reduction = "umap.unintegrated", group.by = "broad_clustering", cols = my_colors) + labs(x = "UMAP 1", y = "UMAP 2") + #UMAP with samples
 theme_minimal() +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      size = 24,
      face = "bold"
    ),
    axis.title.x = element_text(size = 20),
    axis.title.y = element_text(size = 20),
    legend.text = element_text(size = 16),         # Legend labels
    legend.title = element_text(size = 24, face = "bold"),  # Legend title
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank()
  )


p2

#update cluster calling myelin_MOL and MOL as unique cluster
humanized_mouse_brain$broad_clustering <- as.character(humanized_mouse_brain$broad_clustering)
humanized_mouse_brain$broad_clustering[grepl("^myelin[_ ]?mol$", humanized_mouse_brain$broad_clustering, ignore.case=TRUE)] <- "MOL"
humanized_mouse_brain$broad_clustering <- factor(humanized_mouse_brain$broad_clustering)
Idents(humanized_mouse_brain) <- humanized_mouse_brain$broad_clustering
table(humanized_mouse_brain$broad_clustering)
```

Dotplot for the broad clustering 


```{r,fig.height=20, fig.width=27}

markers <- c(
  # Cluster 0 - Cerebellar Granule Cells
  "mm10-Gabra6", "mm10-Cbln1", "mm10-Cbln3", "mm10-Neurod1", "mm10-Barhl2",
  # Cluster 1 - Excitatory Cortical Neurons (Glutamatergic)
  "mm10-Tbr1", "mm10-Satb2", "mm10-Emx1", "mm10-Neurod6",
  # Cluster 2 - GABAergic Interneurons (SST/VIP subtypes)
  "mm10-Lhx6", "mm10-Lhx8", "mm10-Sst", "mm10-Vip", "mm10-Tac2", "mm10-Slc5a7", "mm10-Nxph2",
  # Cluster 3 - GABAergic Inhibitory Neurons (Striatal/MGE-derived)
  "mm10-Dlx1", "mm10-Dlx2", "mm10-Gad1", "mm10-Sp9", "mm10-Meis2", "mm10-Pbx3",
  # Cluster 4 - Mature Oligodendrocytes
  "mm10-Plp1", "mm10-Mbp", "mm10-Apod", "mm10-Ptgds",
  # Cluster 5 - Astrocytes
  "mm10-Atp1a2", "mm10-Slc1a3", "mm10-Apoe", "mm10-Gja1", "mm10-Gli1", "mm10-Mfge8",
  # Cluster 6 - Medium Spiny Neurons (Striatal)
  "mm10-Drd1", "mm10-Drd2", "mm10-Adora2a", "mm10-Gpr88", "mm10-Rgs9", "mm10-Rarb",
  # Cluster 7 - Myelinating Oligodendrocytes
  "mm10-Klk6", "mm10-Gjc2", "mm10-Ugt8a", "mm10-Gjb1",
  # Cluster 8 - Cerebellar Inhibitory Interneurons (MLI/Pvalb+)
  "mm10-Pvalb", "mm10-Pax2", "mm10-Kit", "mm10-Tfap2a", "mm10-Tfap2b", "mm10-Skor1",
  # Cluster 9 - Microglia
  "mm10-Tmem119", "mm10-Hexb", "mm10-Trem2", "mm10-Tyrobp", "mm10-Inpp5d", "mm10-Itgam", "mm10-Cd14", "mm10-Adgre1",
  # Cluster 10 - Thalamic Neurons
  "mm10-Synpo2", "mm10-Zfhx3", "mm10-Shox2", "mm10-Lef1", "mm10-Slitrk6",
  # Cluster 11 - Oligodendrocyte Precursor Cells (OPCs)
  "mm10-Pdgfra", "mm10-Cspg4", "mm10-Vcan",
  # Cluster 12 - Purkinje Cells (Cerebellar)
  "mm10-Car8", "mm10-Pcp2", "mm10-Ppp1r17", "mm10-Col18a1", "mm10-Atp2a3"
)

p1 <- DotPlot(humanized_mouse_brain, features = markers, group.by = "broad_clustering", cols = c("lightgrey", "firebrick")) +
  xlab("")+
  scale_x_discrete(labels = markers) + 
  theme(
    axis.text.x = element_text(hjust = 1, vjust = 0.5, family = "Arial", size = 24, angle = 90),
    axis.text.y = element_text(family = "Arial", size = 20),
    axis.title.y = element_text(family = "Arial", size =24),
    legend.text = element_text(size = 20, family = "Arial"),
    legend.title = element_text(size = 24, family = "Arial")
  ) 
p1
```





##4.MOL subsetting, and reclustering of MOLs only 

##4.1 Select only MOLs from the whole dataset

With this line of code we will create a new Seurat object that will contain only MOLs that we will use for the subsequents analysis steps

```{r}



OL <- subset(humanized_mouse_brain, idents = c("Myelin_MOL", "MOL", "OPcs"))


```

##4.2 Reclustering of MOLs

1.Renormalize the datasets and recluster using only MOL cells   

```{r}
DefaultAssay(OL) <- "RNA"

# Remove the old SCT assay
OL[["SCT"]] <- NULL

# Now run SCTransform renormalizing with only the OL_OPC cells
OL <- SCTransform(OL, vst.flavor = "v2", verbose = FALSE)

# Continue with custom-named reductions
OL <- RunPCA(OL, verbose = TRUE, reduction.name = "pca_OL_SCT")

OL <- FindNeighbors(OL, reduction = "pca_OL_SCT", dims = 1:15,
                      graph.name = c("OL_nn", "OL_snn"))

OL <- FindClusters(OL, resolution = 0.1, 
                     graph.name = "OL_snn",
                     cluster.name = "OL_clusters")

OL <- RunUMAP(OL, reduction = "pca_OL_SCT", dims = 1:15, 
                reduction.name = "umap_OL")


```

2.Plot UMAP with samples and the New MOLs clusters

```{r}

my_colors <- c(
  "#1F77B4",
  "#FF7F0E",
  "#2CA02C",
  "#D62728",
  "#9467BD",
  "#8C564B",
  "#17BECF"
)

p1 <- DimPlot(OL, reduction = "umap_OL", group.by = c("OL_clusters"), cols = my_colors) + labs(x = "UMAP 1", y = "UMAP 2") + #UMAP with new clustering
 theme_minimal() +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      size = 24,
      face = "bold"
    ),
    axis.title.x = element_text(size = 20),
    axis.title.y = element_text(size = 20),
    legend.text = element_text(size = 16),         # Legend labels
    legend.title = element_text(size = 24, face = "bold"),  # Legend title
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank()
  )

p2 <- DimPlot(OL, reduction = "umap_OL", group.by = c("orig.ident")) + labs(x = "UMAP 1", y = "UMAP 2") + #UMAP with samples
 theme_minimal() +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      size = 24,
      face = "bold"
    ),
    axis.title.x = element_text(size = 20),
    axis.title.y = element_text(size = 20),
    legend.text = element_text(size = 16),         # Legend labels
    legend.title = element_text(size = 24, face = "bold"),  # Legend title
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank()
  )

p3 <- DimPlot(OL, reduction = "umap_OL", group.by = c("sample")) + labs(x = "UMAP 1", y = "UMAP 2") + #UMAP with samples
 theme_minimal() +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      size = 24,
      face = "bold"
    ),
    axis.title.x = element_text(size = 20),
    axis.title.y = element_text(size = 20),
    legend.text = element_text(size = 16),         # Legend labels
    legend.title = element_text(size = 24, face = "bold"),  # Legend title
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank()
  )

p4 <- DimPlot(OL, reduction = "umap_OL", group.by = c("broad_clustering")) + labs(x = "UMAP 1", y = "UMAP 2") + #UMAP with samples
 theme_minimal() +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      size = 24,
      face = "bold"
    ),
    axis.title.x = element_text(size = 20),
    axis.title.y = element_text(size = 20),
    legend.text = element_text(size = 16),         # Legend labels
    legend.title = element_text(size = 24, face = "bold"),  # Legend title
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank()
  )

p1
p2
p3
p4
```







clusters proportions

```{r}
colors_vector <- c(
  "CFA_brain_1" = "#E41A1C", 
  "CFA_brain_2" = "#377EB8", 
  "EAE_brain_1" = "#984EA3", 
  "EAE_brain_2" = "#FF7F00", 
  "EAE_brain_3" = "#FFFF33"
)

# Create the contingency table
t <- table(Cluster=OL$OL_clusters, Batch=OL@meta.data[["sample"]])
t <- t[, rev(names(colors_vector))]

# Convert the table to a data frame
df <- as.data.frame(t)

# Calculate the percentage of each plate within each cluster
df <- df %>%
  group_by(Cluster) %>% mutate(Percent = Freq / sum(Freq) * 100)

# Plot the data
ggplot(df, aes(x = Cluster, y = Percent, fill = Batch)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_y_continuous(labels = percent_format(), expand = c(0, 0)) +
  scale_fill_manual(values = rev(colors_vector)) +
  theme_classic() +
  theme(
    axis.text.x = element_text(hjust = 1, vjust = 0.5, family = "Arial", size = 16),
    axis.text.y = element_text(family = "Arial", size = 16),
    axis.title.x = element_text(family = "Arial", size = 18),
    axis.title.y = element_text(family = "Arial", size = 18),
    legend.text = element_text(size = 16, family = "Arial"),
    legend.title = element_text(size = 18, family = "Arial")
  ) +
  labs(y = "Percentage of cells", x = "Cluster", fill = "Animal") +
  guides(fill = guide_legend(ncol = 3, title.position = "top", title.hjust = 0.5))
```

3.Plotting marker genes to review the clustering and rename the different populations of MOLs

```{r,fig.height=20, fig.width=20}

markers <- c(
  #MOL5/6–EAE-b
  "mm10-Serpina3n","mm10-Nlrc5",
  #MOL5/6–EAE-a
  "mm10-Plin4","mm10-Sult1a1","mm10-Zfand4",
  #MOL1/2 EAE
  "mm10-Klk8","mm10-Itga8", "mm10-Tlr3",
  #MHC-II
  "mm10-H2-Aa", "mm10-H2-Ab1", "mm10-H2-Eb1", "mm10-Cd74", "mm10-Ctss",
  #interferon
  "mm10-Ifih1", "mm10-Irf7", "mm10-Irgm1", "mm10-Irgm2", "mm10-Igtp","mm10-Zbp1",
   #cycling_DAO
  "mm10-Cdkn1a","mm10-Ccnd1","mm10-Nav3","mm10-Tmsb10","mm10-Lrrfip1",
  # OPC1
  "mm10-Rph3a", "mm10-Adora2b","mm10-Pdgfra","mm10-Cspg4",
  # OPC2
  "mm10-Phyhd1","mm10-Fcgr2b","mm10-Mylk",
  # OPC3
  "mm10-Slc14a1","mm10-Lgals1","mm10-Myrf",
  #MOL 5/6
  "mm10-Ptgds", "mm10-Opalin", "mm10-Qdpr", "mm10-Il33", "mm10-Apoe", "mm10-Cd63", "mm10-Jph4","mm10-Fgfr2"
  
)

p1 <- DotPlot(OL, features = markers, group.by = "OL_clusters", cols = c("lightgrey", "firebrick")) +
  xlab("")+
  scale_x_discrete(labels = markers) + 
  theme(
    axis.text.x = element_text(hjust = 1, vjust = 0.5, family = "Arial", size = 24, angle = 90),
    axis.text.y = element_text(family = "Arial", size = 20),
    axis.title.y = element_text(family = "Arial", size =24),
    legend.text = element_text(size = 20, family = "Arial"),
    legend.title = element_text(size = 24, family = "Arial")
  ) 
p1
```


```{r,fig.height=20, fig.width=20}

markers <- c(
  #MOL1
  "mm10-Fosb","mm10-Dusp1","mm10-Dnajb1","mm10-Btg2","mm10-Hspa1a","mm10-Hspa1b","mm10-Tob1","mm10-Mal",
  #MOL2
  "mm10-Anxa5","mm10-Klk6","mm10-Mgst3","mm10-S100b","mm10-Tppp3","mm10-Cyp51","mm10-Plekhb1",
  # MFOL 1/2
  "mm10-Ctps","mm10-Tmem141","mm10-Opalin","mm10-Ttyh1", "mm10-Plekha1", "mm10-Ddr1", "mm10-Cd9", "mm10-Sirt2","mm10-Arpc1b",
  # NFOL1/2
  "mm10-Sema6a","mm10-Mag","mm10-Mobp","mm10-Mbp","mm10-Rras2","mm10-Prom1"

  
)

p1 <- DotPlot(OL, features = markers, group.by = "OL_clusters", cols = c("lightgrey", "firebrick")) +
  xlab("")+
  scale_x_discrete(labels = markers) + 
  theme(
    axis.text.x = element_text(hjust = 1, vjust = 0.5, family = "Arial", size = 24, angle = 90),
    axis.text.y = element_text(family = "Arial", size = 20),
    axis.title.y = element_text(family = "Arial", size =24),
    legend.text = element_text(size = 20, family = "Arial"),
    legend.title = element_text(size = 24, family = "Arial")
  ) 
p1
```

Marker features calculations in OLs

```{r}

# find markers for every cluster compared to all remaining cells, report only the positive
# ones

  DefaultAssay(OL) <- "SCT"
  OL <- PrepSCTFindMarkers(OL, assay = "SCT", verbose = TRUE)
  OL_markers <- FindAllMarkers(OL, only.pos = TRUE)


 # Create a new workbook
wb <- createWorkbook()

# Loop through each cluster and add data to a new sheet
for (cluster_value in unique(OL_markers$cluster)) {
  cluster_data <- subset(OL_markers, cluster == cluster_value)
  sheet_name <- paste0("Cluster_", cluster_value)
  addWorksheet(wb, sheet_name)
  writeData(wb, sheet_name, cluster_data)
}

# Save the workbook to a file 
saveWorkbook(wb, "/date/gcb/gcb_MZ/Humanized_mice_OD/Results/cluster_markers/Oligodendroglia_markers.xlsx", overwrite = TRUE)
```

Plotting cluster Markers



```{r,fig.height=20, fig.width=30}



markers <- c(
  # Cluster 0 - Myelinating Oligodendrocytes (MOL)
  "mm10-Ugt8a","mm10-Car2","mm10-Cdh20","mm10-Rtn4","mm10-Rnf220","mm10-Sox2ot","mm10-St18","mm10-Ptprd","mm10-Plp1","mm10-Nkain2",
  # Cluster 1 - Mature Oligodendrocytes (MOL) - homeostatic/metabolic
  "mm10-Garnl3","mm10-Mt1","mm10-Zbtb16","mm10-Polr3e","mm10-Slc6a1","mm10-Slc25a13","mm10-Otud7a","mm10-Mast3","mm10-Elmo1","mm10-Plcl1",
  # Cluster 2 - Mature Oligodendrocytes (MOL) - Apod/Ptgds-high
  "mm10-Apod","mm10-Ptgds","mm10-Phyhd1","mm10-Rsrp1","mm10-Edil3","mm10-Acaca","mm10-Neat1","mm10-St6galnac3","mm10-Anln","mm10-Hif3a",
  # Cluster 3 - Neuronal contamination (Cerebellar Granule Cells)
  "mm10-Kcnip4","mm10-Meg3","mm10-Chn2","mm10-Car10","mm10-Fgf14","mm10-Kcnd2","mm10-Cntnap2","mm10-Rbfox1","mm10-Negr1","mm10-Nrxn1",
  # Cluster 4 - OPCs / Committed OL Precursors (COP)
  "mm10-Lhfpl3","mm10-Xylt1","mm10-Tnr","mm10-Sgcd","mm10-Dscam","mm10-Adarb2","mm10-Ptprz1","mm10-Lrrtm4","mm10-Csmd1","mm10-Kcnip1",
  # Cluster 5 - OPCs (early/proliferative)
  "mm10-Nxph1","mm10-Gpc5","mm10-Sox5","mm10-Sox6","mm10-Slc35f1","mm10-Pcdh7","mm10-Ptprt","mm10-Luzp2","mm10-Nav3","mm10-Pdgfra",
  # Cluster 6 - Neuronal contamination (GABAergic/Glutamatergic)
  "mm10-Grin2b","mm10-Kcnb2","mm10-Atp1b1","mm10-Dclk1","mm10-Tenm3","mm10-Epha6","mm10-Kalrn","mm10-Gria1","mm10-Dlgap2","mm10-Etl4"
)

  


p1 <- DotPlot(OL, features = markers, group.by = "OL_clusters", cols = c("lightgrey", "firebrick")) +
  xlab("")+
  scale_x_discrete(labels = markers) + 
  theme(
    axis.text.x = element_text(hjust = 1, vjust = 0.5, family = "Arial", size = 24, angle = 90),
    axis.text.y = element_text(family = "Arial", size = 20),
    axis.title.y = element_text(family = "Arial", size =24),
    legend.text = element_text(size = 20, family = "Arial"),
    legend.title = element_text(size = 24, family = "Arial")
  ) 
p1
```









1.Compute the GO terms significantly enriched

```{r}
# marker genes computed above

pos_markers <- OL_markers%>%
  dplyr::select(cluster, gene)

# convert symbols to ENTREZID per cluster
marker_list <- split(pos_markers$gene, pos_markers$cluster)

# convert once (union) and map back
all_genes <- unique(unlist(marker_list))

all_genes_clean <- sub("^mm10-", "", all_genes)

gene_map <- bitr(all_genes_clean, fromType = "SYMBOL", toType = "ENTREZID", OrgDb = org.Mm.eg.db)
sym2ent <- setNames(gene_map$ENTREZID, gene_map$SYMBOL)

marker_list_clean <- lapply(marker_list, function(x) sub("^mm10-", "", x))

entrez_by_cluster <- lapply(marker_list_clean, function(x) na.omit(unname(sym2ent[x])))

ego_list <- lapply(names(entrez_by_cluster), function(cl){
  enrichGO(gene          = entrez_by_cluster[[cl]],
           OrgDb         = org.Mm.eg.db,
           keyType       = "ENTREZID",
           ont           = "BP",
           pAdjustMethod = "BH",
           pvalueCutoff  = 0.05,
           qvalueCutoff  = 0.2,
           readable      = TRUE)
}) #list of enriched terms per cluster
names(ego_list) <- names(entrez_by_cluster)

```


2.Create an Excel file to save the results, to be attached as supplementary material

```{r}

# Create a new workbook
wb <- createWorkbook()

# Loop through each cluster and add enrichment results as a sheet
for (cluster_name in names(ego_list)) {
  
  # Extract results from enrichResult object
  if (!is.null(ego_list[[cluster_name]]) && nrow(ego_list[[cluster_name]]@result) > 0) {
    
    # Get the results as a data frame
    go_results <- as.data.frame(ego_list[[cluster_name]])
    
    # Create sheet name (Excel has 31 char limit for sheet names)
    sheet_name <- substr(cluster_name, 1, 31)
    
    # Add worksheet
    addWorksheet(wb, sheet_name)
    
    # Write data
    writeData(wb, sheet_name, go_results)
    
    cat(sprintf("Added sheet for cluster: %s (%d GO terms)\n", 
                cluster_name, nrow(go_results)))
    
  } else {
    cat(sprintf("No significant GO terms for cluster: %s\n", cluster_name))
  }
}

# Save the workbook
saveWorkbook(wb, 
             "/date/gcb/gcb_MZ/Humanized_mice_OD/GO_enrichment_by_cluster_OL.xlsx", 
             overwrite = TRUE)

cat("\n✓ GO enrichment results saved to Excel file!\n")

```

3.Plot the functional enrichment results results for GO terms


```{r,fig.height=20, fig.width=40}
# Collect GO BP results into one data.frame

ego_df <- map2_df(names(ego_list), ego_list, ~{
  if (is.null(.y) || nrow(as.data.frame(.y)) == 0) return(NULL)
  df <- as.data.frame(.y)
  df$cluster <- .x
  df$mLog10Padj <- -log10(df$p.adjust)
  df
})

# Choose top terms per cluster and plot

plot_df <- ego_df %>%
  group_by(cluster) %>%
  slice_max(order_by = mLog10Padj, n = 10, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(Description = fct_reorder(Description, mLog10Padj))

p1 <- ggplot(plot_df, aes(x = cluster, y = Description, size = Count, color = mLog10Padj)) +
  geom_point() +
  scale_color_gradient(low = "gold", high = "red") +
  labs(title = "GO BP enrichment per cluster", x = "Cluster", y = "Term", color = "-log10(FDR)") +
  theme_bw(base_size = 12) + theme_classic()+
  theme(
    axis.text.x = element_text(hjust = 1, vjust = 0.5, family = "Arial", size = 24, angle = 45),
    axis.text.y = element_text(family = "Arial", size = 24),
    axis.title.y = element_text(family = "Arial", size =28, face = "bold"),
    axis.title.x = element_text(family = "Arial", size =28, face = "bold"),
    legend.text = element_text(size = 24, family = "Arial"),
    legend.title = element_text(size = 32, family = "Arial", face = "bold")
  ) 
plots <- list(Func_enrich_GO=p1)
for (nm in names(plots)) {
ggplot2::ggsave(file.path("/date/gcb/gcb_MZ/MOL_FGF_10x/Plots/", paste0(nm, ".png")), plots[[nm]], width =14, height = 16, dpi = 300)
}
p1

```
##8.2 Functional enrichment analysis using KEGG pathways
Functional enrichment using Kegg pathways

1.Kegg pathways analysis

```{r}
ekegg_list <- lapply(names(entrez_by_cluster), function(cl){
  enrichKEGG(gene          = entrez_by_cluster[[cl]],
             organism      = "mmu",
             pvalueCutoff  = 0.05)
})
names(ekegg_list) <- names(entrez_by_cluster)


kegg_df <- map2_df(names(ekegg_list), ekegg_list, ~{
  if (is.null(.y) || nrow(as.data.frame(.y)) == 0) return(NULL)
  df <- as.data.frame(.y)
  df$cluster <- .x
  df$mLog10Padj <- -log10(df$p.adjust)
  df
})
```

2.Create an Excel file to save the results, to be attached as supplementary material

```{r}

# Save to Excel - Option 1: One sheet per cluster
wb_kegg <- createWorkbook()

for (cluster_name in names(ekegg_list)) {
  
  if (!is.null(ekegg_list[[cluster_name]]) && nrow(ekegg_list[[cluster_name]]@result) > 0) {
    
    # Get results as data frame
    kegg_results <- as.data.frame(ekegg_list[[cluster_name]])
    
    # Add mLog10Padj column
    kegg_results$mLog10Padj <- -log10(kegg_results$p.adjust)
    
    # Create sheet name
    sheet_name <- substr(cluster_name, 1, 31)
    
    # Add worksheet and write data
    addWorksheet(wb_kegg, sheet_name)
    writeData(wb_kegg, sheet_name, kegg_results)
    
    cat(sprintf("Added sheet for cluster: %s (%d KEGG pathways)\n", 
                cluster_name, nrow(kegg_results)))
    
  } else {
    cat(sprintf("No significant KEGG pathways for cluster: %s\n", cluster_name))
  }
}

# Save workbook with one sheet per cluster
saveWorkbook(wb_kegg, 
             "/date/gcb/gcb_MZ/Humanized_mice_OD/KEGG_enrichment_by_cluster_OL.xlsx", 
             overwrite = TRUE)

cat("\n✓ KEGG enrichment results saved (one sheet per cluster)!\n")

```

3.Plot the results

```{r,fig.height=20, fig.width=40}
plot_df_kegg <- kegg_df %>%
  group_by(cluster) %>%
  slice_max(order_by = mLog10Padj, n = 10, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(Description = fct_reorder(Description, mLog10Padj))

p1 <- ggplot(plot_df_kegg, aes(x = cluster, y = Description, size = Count, color = mLog10Padj)) +
  geom_point() +
  scale_color_gradient(low = "gold", high = "red") +
  labs(title = "KEGG enrichment per cluster", x = "Cluster", y = "Pathway", color = "-log10(FDR)") +
  theme_bw(base_size = 12)+theme_classic()+
  theme(
    axis.text.x = element_text(hjust = 1, vjust = 0.5, family = "Arial", size = 24, angle = 45),
    axis.text.y = element_text(family = "Arial", size = 24),
    axis.title.y = element_text(family = "Arial", size =28, face = "bold"),
    axis.title.x = element_text(family = "Arial", size =28, face = "bold"),
    legend.text = element_text(size = 24, family = "Arial"),
    legend.title = element_text(size = 32, family = "Arial", face = "bold")
  ) 
plots <- list(Func_enrich_KEGG=p1)
for (nm in names(plots)) {
ggplot2::ggsave(file.path("/date/gcb/gcb_MZ/MOL_FGF_10x/Plots/", paste0(nm, ".png")), plots[[nm]], width =14, height = 16, dpi = 300)
}
p1

```



##4.1 Select only MOLs from the whole dataset

With this line of code we will create a new Seurat object that will contain only MOLs that we will use for the subsequents analysis steps

```{r}


Microglia <- subset(humanized_mouse_brain, idents = c("Microglia"))


```

##4.2 Reclustering of MOLs

1.Renormalize the datasets and recluster using only MOL cells   

```{r}
DefaultAssay(Microglia) <- "RNA"

# Remove the old SCT assay
Microglia[["SCT"]] <- NULL

# Now run SCTransform renormalizing with only the Microglia cells
Microglia <- SCTransform(Microglia, vst.flavor = "v2", verbose = FALSE)

# Continue with custom-named reductions
Microglia <- RunPCA(Microglia, verbose = TRUE, reduction.name = "pca_Microglia_SCT")

Microglia <- FindNeighbors(Microglia, reduction = "pca_Microglia_SCT", dims = 1:15,
                      graph.name = c("Microglia_nn", "Microglia_snn"))

Microglia <- FindClusters(Microglia, resolution = 0.1, 
                     graph.name = "Microglia_snn",
                     cluster.name = "Microglia_clusters")

Microglia <- RunUMAP(Microglia, reduction = "pca_Microglia_SCT", dims = 1:15, 
                reduction.name = "umap_Microglia")


```

```{r}
p1 <- DimPlot(Microglia, reduction = "umap_Microglia", group.by = c("Microglia_clusters")) + labs(x = "UMAP 1", y = "UMAP 2") + #UMAP with new clustering
 theme_minimal() +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      size = 24,
      face = "bold"
    ),
    axis.title.x = element_text(size = 20),
    axis.title.y = element_text(size = 20),
    legend.text = element_text(size = 16),         # Legend labels
    legend.title = element_text(size = 24, face = "bold"),  # Legend title
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank()
  )

p2 <- DimPlot(Microglia, reduction = "umap_Microglia", group.by = c("orig.ident")) + labs(x = "UMAP 1", y = "UMAP 2") + #UMAP with samples
 theme_minimal() +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      size = 24,
      face = "bold"
    ),
    axis.title.x = element_text(size = 20),
    axis.title.y = element_text(size = 20),
    legend.text = element_text(size = 16),         # Legend labels
    legend.title = element_text(size = 24, face = "bold"),  # Legend title
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank()
  )

p3 <- DimPlot(Microglia, reduction = "umap_Microglia", group.by = c("sample")) + labs(x = "UMAP 1", y = "UMAP 2") + #UMAP with samples
 theme_minimal() +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      size = 24,
      face = "bold"
    ),
    axis.title.x = element_text(size = 20),
    axis.title.y = element_text(size = 20),
    legend.text = element_text(size = 16),         # Legend labels
    legend.title = element_text(size = 24, face = "bold"),  # Legend title
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank()
  )

p4 <- DimPlot(Microglia, reduction = "umap_Microglia", group.by = c("broad_clustering")) + labs(x = "UMAP 1", y = "UMAP 2") + #UMAP with samples
 theme_minimal() +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      size = 24,
      face = "bold"
    ),
    axis.title.x = element_text(size = 20),
    axis.title.y = element_text(size = 20),
    legend.text = element_text(size = 16),         # Legend labels
    legend.title = element_text(size = 24, face = "bold"),  # Legend title
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank()
  )

p1
p2
p3
p4
```

Marker features calculations in OLs

```{r}

# find markers for every cluster compared to all remaining cells, report only the positive
# ones

DefaultAssay(Microglia) <- "SCT"
Microglia <- PrepSCTFindMarkers(Microglia, assay = "SCT", verbose = TRUE)
Microglia_markers <- FindAllMarkers(Microglia, only.pos = TRUE)


 # Create a new workbook
wb <- createWorkbook()

# Loop through each cluster and add data to a new sheet
for (cluster_value in unique(Microglia_markers$cluster)) {
  cluster_data <- subset(Microglia_markers, cluster == cluster_value)
  sheet_name <- paste0("Cluster_", cluster_value)
  addWorksheet(wb, sheet_name)
  writeData(wb, sheet_name, cluster_data)
}

# Save the workbook to a file
saveWorkbook(wb, "/date/gcb/gcb_MZ/Humanized_mice_OD/Seurat_clusters_Microglia.xlsx", overwrite = TRUE)
```

Plotting cluster Markers



```{r,fig.height=20, fig.width=30}



markers <- c(
  # Cluster 0 - Neuronal contamination (Cerebellar Granule Cells)
  "mm10-Kcnip4","mm10-Kcnd2","mm10-Fgf14","mm10-Ralyl","mm10-Nrxn1","mm10-Rims1","mm10-Rbfox1","mm10-Tenm1","mm10-Car10","mm10-Negr1",
  # Cluster 1 - Homeostatic Microglia
  "mm10-Plxdc2","mm10-Tanc2","mm10-Csmd3","mm10-Nav2","mm10-Srgap2","mm10-Zfhx3","mm10-Rapgef5","mm10-Tgfbr1","mm10-Elmo1","mm10-Frmd4a",
  # Cluster 2 - Neuronal contamination (GABAergic/Glutamatergic)
  "mm10-Lrrc7","mm10-Grin2b","mm10-Kcnh7","mm10-Tenm3","mm10-Grm5","mm10-Epha6","mm10-Meis2","mm10-Myo16","mm10-Sntg1","mm10-Cpne4",
  # Cluster 3 - Oligodendrocyte contamination
  "mm10-Prr5l","mm10-Sec14l5","mm10-Slc24a2","mm10-Edil3","mm10-Arap2","mm10-Efnb3","mm10-Mog","mm10-D7Ertd443e","mm10-Mag","mm10-Ppp1r16b",
  # Cluster 4 - Activated/Phagocytic Microglia (DAM-like)
  "mm10-Arhgap24","mm10-Fkbp5","mm10-Gcnt2","mm10-C1qa","mm10-Ssh2","mm10-Cd244a","mm10-C1qb","mm10-Man2b1","mm10-Cd300lf","mm10-Npl",
  # Cluster 5 - Microglia - metabolically active / stressed
  "mm10-Cst3","mm10-Psap","mm10-mt-Co3","mm10-mt-Atp6","mm10-mt-Co2","mm10-mt-Co1","mm10-Mef2a","mm10-Snx24","mm10-Inpp5d","mm10-Rps27a"
)
  


p1 <- DotPlot(OL, features = markers, group.by = "OL_clusters", cols = c("lightgrey", "firebrick")) +
  xlab("")+
  scale_x_discrete(labels = markers) + 
  theme(
    axis.text.x = element_text(hjust = 1, vjust = 0.5, family = "Arial", size = 24, angle = 90),
    axis.text.y = element_text(family = "Arial", size = 20),
    axis.title.y = element_text(family = "Arial", size =24),
    legend.text = element_text(size = 20, family = "Arial"),
    legend.title = element_text(size = 24, family = "Arial")
  ) 
p1
```
1.Compute the GO terms significantly enriched

```{r}
# marker genes computed above

pos_markers <- Microglia_markers%>%
  dplyr::select(cluster, gene)

# convert symbMicroglias to ENTREZID per cluster
marker_list <- split(pos_markers$gene, pos_markers$cluster)

# convert once (union) and map back
all_genes <- unique(unlist(marker_list))

all_genes_clean <- sub("^mm10-", "", all_genes)

gene_map <- bitr(all_genes_clean, fromType = "SYMBOL", toType = "ENTREZID", OrgDb = org.Mm.eg.db)
sym2ent <- setNames(gene_map$ENTREZID, gene_map$SYMBOL)

marker_list_clean <- lapply(marker_list, function(x) sub("^mm10-", "", x))

entrez_by_cluster <- lapply(marker_list_clean, function(x) na.omit(unname(sym2ent[x])))

ego_list <- lapply(names(entrez_by_cluster), function(cl){
  enrichGO(gene          = entrez_by_cluster[[cl]],
           OrgDb         = org.Mm.eg.db,
           keyType       = "ENTREZID",
           ont           = "BP",
           pAdjustMethod = "BH",
           pvalueCutoff  = 0.05,
           qvalueCutoff  = 0.2,
           readable      = TRUE)
}) #list of enriched terms per cluster
names(ego_list) <- names(entrez_by_cluster)

```


2.Create an Excel file to save the results, to be attached as supplementary material

```{r}

# Create a new workbook
wb <- createWorkbook()

# Loop through each cluster and add enrichment results as a sheet
for (cluster_name in names(ego_list)) {
  
  # Extract results from enrichResult object
  if (!is.null(ego_list[[cluster_name]]) && nrow(ego_list[[cluster_name]]@result) > 0) {
    
    # Get the results as a data frame
    go_results <- as.data.frame(ego_list[[cluster_name]])
    
    # Create sheet name (Excel has 31 char limit for sheet names)
    sheet_name <- substr(cluster_name, 1, 31)
    
    # Add worksheet
    addWorksheet(wb, sheet_name)
    
    # Write data
    writeData(wb, sheet_name, go_results)
    
    cat(sprintf("Added sheet for cluster: %s (%d GO terms)\n", 
                cluster_name, nrow(go_results)))
    
  } else {
    cat(sprintf("No significant GO terms for cluster: %s\n", cluster_name))
  }
}

# Save the workbook
saveWorkbook(wb, 
             "/date/gcb/gcb_MZ/Humanized_mice_OD/Results/functional_enrichment/GO_enrichment_by_cluster_Microglia.xlsx", 
             overwrite = TRUE)

cat("\n✓ GO enrichment results saved to Excel file!\n")

```


3.Plot the functional enrichment results results for GO terms


```{r,fig.height=20, fig.width=40}
# Collect GO BP results into one data.frame

ego_df <- map2_df(names(ego_list), ego_list, ~{
  if (is.null(.y) || nrow(as.data.frame(.y)) == 0) return(NULL)
  df <- as.data.frame(.y)
  df$cluster <- .x
  df$mLog10Padj <- -log10(df$p.adjust)
  df
})

# Choose top terms per cluster and plot

plot_df <- ego_df %>%
  group_by(cluster) %>%
  slice_max(order_by = mLog10Padj, n = 10, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(Description = fct_reorder(Description, mLog10Padj))

p1 <- ggplot(plot_df, aes(x = cluster, y = Description, size = Count, color = mLog10Padj)) +
  geom_point() +
  scale_color_gradient(low = "gold", high = "red") +
  labs(title = "GO BP enrichment per cluster", x = "Cluster", y = "Term", color = "-log10(FDR)") +
  theme_bw(base_size = 12) + theme_classic()+
  theme(
    axis.text.x = element_text(hjust = 1, vjust = 0.5, family = "Arial", size = 24, angle = 45),
    axis.text.y = element_text(family = "Arial", size = 24),
    axis.title.y = element_text(family = "Arial", size =28, face = "bold"),
    axis.title.x = element_text(family = "Arial", size =28, face = "bold"),
    legend.text = element_text(size = 24, family = "Arial"),
    legend.title = element_text(size = 32, family = "Arial", face = "bold")
  ) 
plots <- list(Func_enrich_GO=p1)
for (nm in names(plots)) {
ggplot2::ggsave(file.path("/date/gcb/gcb_MZ/Humanized_mice_OD/Results/Plots/", paste0(nm, "_Microglia.png")), plots[[nm]], width =14, height = 16, dpi = 300)
}
p1

```

```{r}
# 1) cluster set
#broad_clusters <- unique(humanized_mouse_brain$broad_clustering)
broad_clusters <- "MOL"
# storage
clusters_objects <- list()

for (bc in broad_clusters) {
  bc_clean <- make.names(bc)
  obj_name <- paste0(bc_clean)
  message("Processing broad cluster: ", bc, " --> ", obj_name)

  # subset
  OL <- subset(humanized_mouse_brain, subset = broad_clustering == bc)

  # prep
  DefaultAssay(OL) <- "RNA"
  OL[["SCT"]] <- NULL
  OL <- SCTransform(OL, vst.flavor = "v2", verbose = FALSE)
  OL <- RunPCA(OL, verbose = FALSE, reduction.name = paste0("pca",obj_name,"_SCT"))
  OL <- FindNeighbors(OL, reduction = paste0("pca",obj_name,"_SCT"), dims = 1:15,
                      graph.name = c(paste0(obj_name,"_nn"), paste0(obj_name,"_snn")))
  OL <- FindClusters(OL, resolution = 0.1,
                     graph.name = paste0(obj_name,"_snn"), cluster.name = paste0(obj_name,"_clusters"))
  OL <- RunUMAP(OL, reduction = paste0("pca",obj_name,"_SCT"), dims = 1:15,
                reduction.name = paste0("umap_",obj_name))
  # store into list by key
  clusters_objects[[obj_name]] <- OL
  # save object
  saveRDS(OL, file = file.path("/date/gcb/gcb_MZ/Humanized_mice_OD/Results/objects/",
                               paste0(obj_name, "_SCT.rds")))
  # markers
  DefaultAssay(OL) <- "SCT"
  OL <- PrepSCTFindMarkers(OL, assay = "SCT", verbose = FALSE)
  OL_markers <- FindAllMarkers(OL, only.pos = TRUE)

  # Create a new workbook
  wb <- createWorkbook()

  # Loop through each cluster and add data to a new sheet
  for (cluster_value in unique(OL_markers$cluster)) {
    cluster_data <- subset(OL_markers, cluster == cluster_value)
    sheet_name <- paste0("Cluster_", cluster_value)
    addWorksheet(wb, sheet_name)
    writeData(wb, sheet_name, cluster_data)
}

# Save the workbook to a file
saveWorkbook(wb, paste0("/date/gcb/gcb_MZ/Humanized_mice_OD/Results/cluster_markers/",bc,"_markers.xlsx"), overwrite = TRUE)

  # GO/KEGG
  mark_list <- split(OL_markers$gene, OL_markers$cluster)
  all_genes <- unique(unlist(mark_list))
  all_genes_clean <- sub("^mm10-", "", all_genes)
  gmap <- bitr(all_genes_clean, fromType="SYMBOL", toType="ENTREZID",
               OrgDb=org.Mm.eg.db)
  sym2ent <- setNames(gmap$ENTREZID, gmap$SYMBOL)
  mark_list_clean <- lapply(mark_list, function(z)
      na.omit(unname(sym2ent[sub("^mm10-", "", z)])))
  ego <- lapply(names(mark_list_clean), function(cl) {
    enrichGO(gene=mark_list_clean[[cl]], OrgDb=org.Mm.eg.db,
             keyType="ENTREZID", ont="BP", pAdjustMethod="BH",
             pvalueCutoff=0.05, qvalueCutoff=0.2, readable=TRUE)
  })
  names(ego) <- names(mark_list_clean)

  # optionally save per broad cluster worksheet PB
  for (cl in names(ego)) {
    if (!is.null(ego[[cl]]) && nrow(as.data.frame(ego[[cl]])) > 0) {
      write.xlsx(as.data.frame(ego[[cl]]),
                 file.path("/date/gcb/gcb_MZ/Humanized_mice_OD",
                           paste0(obj_name, "_GO_cluster", cl, ".xlsx")),
                 sheetName=paste0("c",cl), overwrite=TRUE)
    }
  }
  ego_df <- map2_df(names(ego_list), ego_list, ~{
  if (is.null(.y) || nrow(as.data.frame(.y)) == 0) return(NULL)
  df <- as.data.frame(.y)
  df$cluster <- .x
  df$mLog10Padj <- -log10(df$p.adjust)
  df
})

# Choose top terms per cluster and plot

plot_df <- ego_df %>%
  group_by(cluster) %>%
  slice_max(order_by = mLog10Padj, n = 10, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(Description = fct_reorder(Description, mLog10Padj))

p1 <- ggplot(plot_df, aes(x = cluster, y = Description, size = Count, color = mLog10Padj)) +
  geom_point() +
  scale_color_gradient(low = "gold", high = "red") +
  labs(title = "GO BP enrichment per cluster", x = "Cluster", y = "Term", color = "-log10(FDR)") +
  theme_bw(base_size = 12) + theme_classic()+
  theme(
    axis.text.x = element_text(hjust = 1, vjust = 0.5, family = "Arial", size = 24, angle = 45),
    axis.text.y = element_text(family = "Arial", size = 24),
    axis.title.y = element_text(family = "Arial", size =28, face = "bold"),
    axis.title.x = element_text(family = "Arial", size =28, face = "bold"),
    legend.text = element_text(size = 24, family = "Arial"),
    legend.title = element_text(size = 32, family = "Arial", face = "bold")
  ) 
plots <- list(Func_enrich_GO=p1)
for (nm in names(plots)) {
ggplot2::ggsave(file.path("/date/gcb/gcb_MZ/Humanized_mice_OD/Results/Plots/", paste0(nm,"_",bc_clean, ".png")), plots[[nm]], width =14, height = 16, dpi = 300)
}
}

```
resave cluster makers correclty


```{r}

for (bc in broad_clusters) {
  message("Calculating cluster makers: ", bc)
  bc_clean <- make.names(bc)
  OL <- clusters_objects[[bc_clean]]
  # markers
  DefaultAssay(OL) <- "SCT"
  OL <- PrepSCTFindMarkers(OL, assay = "SCT", verbose = FALSE)
  OL_markers <- FindAllMarkers(OL, only.pos = TRUE)
  # Create a new workbook
  wb <- createWorkbook()

  # Loop through each cluster and add data to a new sheet
  for (cluster_value in unique(OL_markers$cluster)) {
    cluster_data <- subset(OL_markers, cluster == cluster_value)
    sheet_name <- paste0("Cluster_", cluster_value)
    addWorksheet(wb, sheet_name)
    writeData(wb, sheet_name, cluster_data)
}

# Save the workbook to a file
saveWorkbook(wb, paste0("/date/gcb/gcb_MZ/Humanized_mice_OD/Results/cluster_markers/",bc,"_markers.xlsx"), overwrite = TRUE)
}
```


UMAP plot of reclustering

```{r}
for (bc in broad_clusters) {
  bc_clean <- make.names(bc)
  obj <- clusters_objects[[bc_clean]]
  umap_name <- paste0("umap_", bc_clean)
  clu_name <- paste0(bc_clean, "_clusters")
  
  p1 <- DimPlot(obj, reduction = umap_name, group.by = c(clu_name)) + labs(x = "UMAP 1", y = "UMAP 2") + #UMAP with new clustering
 theme_minimal() +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      size = 24,
      face = "bold"
    ),
    axis.title.x = element_text(size = 20),
    axis.title.y = element_text(size = 20),
    legend.text = element_text(size = 16),         # Legend labels
    legend.title = element_text(size = 24, face = "bold"),  # Legend title
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank()
  )

p2 <- DimPlot(obj, reduction = umap_name, group.by = c("orig.ident")) + labs(x = "UMAP 1", y = "UMAP 2") + #UMAP with samples
 theme_minimal() +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      size = 24,
      face = "bold"
    ),
    axis.title.x = element_text(size = 20),
    axis.title.y = element_text(size = 20),
    legend.text = element_text(size = 16),         # Legend labels
    legend.title = element_text(size = 24, face = "bold"),  # Legend title
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank()
  )

p3 <- DimPlot(obj, reduction = umap_name, group.by = c("sample")) + labs(x = "UMAP 1", y = "UMAP 2") + #UMAP with samples
 theme_minimal() +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      size = 24,
      face = "bold"
    ),
    axis.title.x = element_text(size = 20),
    axis.title.y = element_text(size = 20),
    legend.text = element_text(size = 16),         # Legend labels
    legend.title = element_text(size = 24, face = "bold"),  # Legend title
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank()
  )

print(p1)
print(p2)
print(p3)
plots <- list(clusters=p1,condition=p2,sample=p3)
for (nm in names(plots)) {
ggplot2::ggsave(file.path("/date/gcb/gcb_MZ/Humanized_mice_OD/Results/Plots/", paste0("umap_",nm,"_",bc_clean, ".png")), plots[[nm]], width =14, height = 16, dpi = 300)
}
}
```

Functiona enrichment and plot per cluster GO

```{r}

for (bc in broad_clusters) {
  
  bc_clean <- make.names(bc)
  umap_name <- paste0("umap_", bc_clean)
  clu_name <- paste0(bc_clean, "_clusters")
  # marker genes computed above
  
  obj <- clusters_objects[[bc_clean]]
  # markers
  DefaultAssay(obj) <- "SCT"
  obj <- PrepSCTFindMarkers(obj, assay = "SCT", verbose = FALSE)
  obj_markers <- FindAllMarkers(obj, only.pos = TRUE)
  cat("Calculating GO for cluster: ", bc)
  pos_markers <- obj_markers%>%
  dplyr::select(cluster, gene)

  # convert symbMicroglias to ENTREZID per cluster
  marker_list <- split(pos_markers$gene, pos_markers$cluster)

  # convert once (union) and map back
  all_genes <- unique(unlist(marker_list))

  all_genes_clean <- sub("^mm10-", "", all_genes)

  gene_map <- bitr(all_genes_clean, fromType = "SYMBOL", toType = "ENTREZID", OrgDb = org.Mm.eg.db)
  sym2ent <- setNames(gene_map$ENTREZID, gene_map$SYMBOL)

  marker_list_clean <- lapply(marker_list, function(x) sub("^mm10-", "", x))

  entrez_by_cluster <- lapply(marker_list_clean, function(x) na.omit(unname(sym2ent[x])))

  ego_list <- lapply(names(entrez_by_cluster), function(cl){
   enrichGO(gene          = entrez_by_cluster[[cl]],
           OrgDb         = org.Mm.eg.db,
           keyType       = "ENTREZID",
           ont           = "BP",
           pAdjustMethod = "BH",
           pvalueCutoff  = 0.05,
           qvalueCutoff  = 0.2,
           readable      = TRUE)
    }) #list of enriched terms per cluster
    names(ego_list) <- names(entrez_by_cluster)

    # Create a new workbook
    wb <- createWorkbook()

    # Loop through each cluster and add enrichment results as a sheet
    for (cluster_name in names(ego_list)) {
  
    # Extract results from enrichResult object
     if (!is.null(ego_list[[cluster_name]]) && nrow(ego_list[[cluster_name]]@result) > 0) {
    
     # Get the results as a data frame
     go_results <- as.data.frame(ego_list[[cluster_name]])
    
     # Create sheet name (Excel has 31 char limit for sheet names)
     sheet_name <- substr(cluster_name, 1, 31)
    
     # Add worksheet
     addWorksheet(wb, sheet_name)
    
     # Write data
     writeData(wb, sheet_name, go_results)
    
     cat(sprintf("Added sheet for cluster: %s (%d GO terms)\n", 
                cluster_name, nrow(go_results)))
    
        } else {
         cat(sprintf("No significant GO terms for cluster: %s\n", cluster_name))
        }
      }

      # Save the workbook
      saveWorkbook(wb, 
             paste0("/date/gcb/gcb_MZ/Humanized_mice_OD/Results/functional_enrichment/GO_enrichment_by_cluster_",bc,".xlsx"), 
             overwrite = TRUE)

       cat("\n✓ GO enrichment results saved to Excel file!\n")


# Collect GO BP results into one data.frame

ego_df <- map2_df(names(ego_list), ego_list, ~{
  if (is.null(.y) || nrow(as.data.frame(.y)) == 0) return(NULL)
  df <- as.data.frame(.y)
  df$cluster <- .x
  df$mLog10Padj <- -log10(df$p.adjust)
  df
})

# Choose top terms per cluster and plot

plot_df <- ego_df %>%
  group_by(cluster) %>%
  slice_max(order_by = mLog10Padj, n = 10, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(Description = fct_reorder(Description, mLog10Padj))

p1 <- ggplot(plot_df, aes(x = cluster, y = Description, size = Count, color = mLog10Padj)) +
  geom_point() +
  scale_color_gradient(low = "gold", high = "red") +
  labs(title = "GO BP enrichment per cluster", x = "Cluster", y = "Term", color = "-log10(FDR)") +
  theme_bw(base_size = 12) + theme_classic()+
  theme(
    axis.text.x = element_text(hjust = 1, vjust = 0.5, family = "Arial", size = 24, angle = 45),
    axis.text.y = element_text(family = "Arial", size = 24),
    axis.title.y = element_text(family = "Arial", size =28, face = "bold"),
    axis.title.x = element_text(family = "Arial", size =28, face = "bold"),
    legend.text = element_text(size = 24, family = "Arial"),
    legend.title = element_text(size = 32, family = "Arial", face = "bold")
  ) 
plots <- list(Func_enrich_GO=p1)
for (nm in names(plots)) {
ggplot2::ggsave(file.path("/date/gcb/gcb_MZ/Humanized_mice_OD/Results/Plots/", paste0(nm,"_",bc, ".png")), plots[[nm]], width =24, height = 26, dpi = 300)
}

}


```






Save objects


```{r}
saveRDS(Microglia,"/date/gcb/gcb_MZ/Humanized_mice_OD/Results/objects/Microglia_SCT.rds")
saveRDS(OL,"/date/gcb/gcb_MZ/Humanized_mice_OD/Results/objects/Oligodendroglia_SCT.rds")
saveRDS(humanized_mouse_brain,"/date/gcb/gcb_MZ/Humanized_mice_OD/Results/objects/humanized_mouse_brain_whole_doublets.rds")
```




##5. Differential Gene Expression Analysis: EAE vs CFA

Pseudobulk DESeq2 is used to compare EAE (n = 3 animals) vs CFA (n = 2 animals)
within each cluster. Raw counts are aggregated per sample per cluster before running
DESeq2, which avoids pseudo-replication and handles unequal group sizes naturally.
CFA is the reference level; positive log2FC values indicate higher expression in EAE.

##5.1 Setup: load libraries and derive condition metadata

```{r}
library(DESeq2)
library(ggrepel)

# Derive condition from the sample label (set during object creation)
humanized_mouse_brain$condition <- ifelse(grepl("^EAE", humanized_mouse_brain$sample), "EAE", "CFA")

table(humanized_mouse_brain$sample, humanized_mouse_brain$condition)
```

##5.2 Helper functions

```{r}
# Pseudobulk DESeq2 for one cluster ----------------------------------------
# Returns a data.frame of DESeq2 results (EAE vs CFA), or NULL if the cluster
# does not meet quality filters.
run_pseudobulk_dge <- function(seurat_obj,
                                cluster_col,
                                cluster_id,
                                sample_col    = "sample",
                                condition_col = "condition",
                                min_cells     = 10,
                                min_counts    = 10) {
  idx <- seurat_obj@meta.data[[cluster_col]] == cluster_id
  if (sum(idx) < min_cells) {
    message("Skipping cluster ", cluster_id, ": only ", sum(idx), " cells")
    return(NULL)
  }

  sub_obj <- seurat_obj[, idx]
  DefaultAssay(sub_obj) <- "RNA"
  sub_obj <- JoinLayers(sub_obj)
  counts_mat <- GetAssayData(sub_obj, assay = "RNA", layer = "counts")

  # Aggregate counts per sample (pseudobulk)
  samples <- unique(sub_obj@meta.data[[sample_col]])
  pb_mat <- vapply(samples, function(s) {
    s_cells <- sub_obj@meta.data[[sample_col]] == s
    as.numeric(Matrix::rowSums(counts_mat[, s_cells, drop = FALSE]))
  }, numeric(nrow(counts_mat)))
  rownames(pb_mat) <- rownames(counts_mat)

  # Sample-level metadata; CFA is the reference level
  col_meta <- data.frame(
    sample    = samples,
    condition = factor(
      sapply(samples, function(s)
        unique(sub_obj@meta.data[[condition_col]][sub_obj@meta.data[[sample_col]] == s])),
      levels = c("CFA", "EAE")
    ),
    row.names = samples
  )

  if (sum(col_meta$condition == "EAE") == 0 || sum(col_meta$condition == "CFA") == 0) {
    message("Skipping cluster ", cluster_id, ": only one condition represented")
    return(NULL)
  }

  # Filter lowly expressed genes
  pb_mat <- pb_mat[rowSums(pb_mat) >= min_counts, , drop = FALSE]
  if (nrow(pb_mat) < 10) {
    message("Skipping cluster ", cluster_id, ": too few expressed genes after filtering")
    return(NULL)
  }

  tryCatch({
    dds <- DESeqDataSetFromMatrix(countData = round(pb_mat),
                                   colData   = col_meta,
                                   design    = ~ condition)
    dds <- DESeq(dds, quiet = TRUE)
    res <- results(dds, contrast = c("condition", "EAE", "CFA"))
    res_df <- as.data.frame(res)
    res_df$gene    <- rownames(res_df)
    res_df$cluster <- as.character(cluster_id)
    res_df[order(res_df$padj, na.last = TRUE), ]
  }, error = function(e) {
    message("DESeq2 error – cluster ", cluster_id, ": ", conditionMessage(e))
    NULL
  })
}

# Volcano plot helper -------------------------------------------------------
plot_volcano <- function(res_df, cluster_label,
                          padj_thresh = 0.05, lfc_thresh = 0.5, top_n = 15) {
  df <- res_df %>%
    filter(!is.na(padj), !is.na(log2FoldChange)) %>%
    mutate(
      sig = case_when(
        padj < padj_thresh & log2FoldChange >  lfc_thresh ~ "Up in EAE",
        padj < padj_thresh & log2FoldChange < -lfc_thresh ~ "Up in CFA",
        TRUE ~ "NS"
      )
    ) %>%
    mutate(label = ifelse(sig != "NS" & rank(padj, ties.method = "first") <= top_n,
                          gene, NA_character_))

  ggplot(df, aes(x = log2FoldChange, y = -log10(padj), color = sig)) +
    geom_point(alpha = 0.6, size = 1.5) +
    ggrepel::geom_text_repel(aes(label = label), size = 3, max.overlaps = 20, na.rm = TRUE) +
    scale_color_manual(
      values = c("Up in EAE" = "firebrick", "Up in CFA" = "steelblue", "NS" = "grey70")
    ) +
    geom_vline(xintercept = c(-lfc_thresh, lfc_thresh), linetype = "dashed", colour = "grey40") +
    geom_hline(yintercept = -log10(padj_thresh),         linetype = "dashed", colour = "grey40") +
    labs(
      title = paste("EAE vs CFA –", cluster_label),
      x     = "log2 Fold Change (EAE / CFA)",
      y     = "-log10(adjusted p-value)",
      color = NULL
    ) +
    theme_classic() +
    theme(
      plot.title  = element_text(hjust = 0.5, size = 16, face = "bold"),
      axis.text   = element_text(size = 12),
      axis.title  = element_text(size = 14),
      legend.text = element_text(size = 12)
    )
}
```

##5.3 DGE within each harmony cluster (whole-brain object)

```{r}
broad_clustering_ids <- levels(humanized_mouse_brain$broad_clustering)

dge_brain_list <- lapply(broad_clustering_ids, function(cl) {
  message("Running DGE for harmony cluster: ", cl)
  run_pseudobulk_dge(
    seurat_obj  = humanized_mouse_brain,
    cluster_col = "broad_clustering",
    cluster_id  = cl
  )
})
names(dge_brain_list) <- broad_clustering_ids

# Combine into one data.frame
dge_brain_all <- do.call(rbind, Filter(Negate(is.null), dge_brain_list))

# Save to Excel (one sheet per cluster)
wb_brain <- createWorkbook()
for (cl in names(dge_brain_list)) {
  if (!is.null(dge_brain_list[[cl]])) {
    sheet_name <- substr(paste0("Cluster_", cl), 1, 31)
    addWorksheet(wb_brain, sheet_name)
    writeData(wb_brain, sheet_name, dge_brain_list[[cl]])
  }
}
saveWorkbook(wb_brain,
             "/date/gcb/gcb_MZ/Humanized_mice_OD/Results/DGE_EAE_vs_CFA_harmony_clusters.xlsx",
             overwrite = TRUE)
cat("Saved whole-brain DGE results.\n")
```

Volcano plots – whole-brain harmony clusters

```{r, fig.height=6, fig.width=8}
for (cl in names(dge_brain_list)) {
  if (!is.null(dge_brain_list[[cl]]) && sum(!is.na(dge_brain_list[[cl]]$padj)) > 0) {
    print(plot_volcano(dge_brain_list[[cl]], paste("broad_clustering", cl)))
  }
}
```

##5.4 Summary: number of significant DEGs per cluster (padj < 0.05, |log2FC| > 0.5)

```{r}
summarise_dge <- function(dge_list) {
  do.call(rbind, lapply(names(dge_list), function(cl) {
    df <- dge_list[[cl]]
    if (is.null(df)) return(data.frame(cluster = cl, n_up_EAE = NA, n_up_CFA = NA))
    df <- df[!is.na(df$padj), ]
    data.frame(
      cluster  = cl,
      n_up_EAE = sum(df$padj < 0.05 & df$log2FoldChange >  0.5),
      n_up_CFA = sum(df$padj < 0.05 & df$log2FoldChange < -0.5)
    )
  }))
}

dge_summary <- summarise_dge(dge_brain_list)
print(dge_summary)

wb_summary <- createWorkbook()
addWorksheet(wb_summary, "DGE_summary")
writeData(wb_summary, "DGE_summary", dge_summary)
saveWorkbook(wb_summary,
             "/date/gcb/gcb_MZ/Humanized_mice_OD/Results/DGE_EAE_vs_CFA_summary.xlsx",
             overwrite = TRUE)
```

##5.5 Functional enrichment of DEGs – GO biological process

Enrichment is run separately on genes up in EAE and genes up in CFA (padj < 0.05,
|log2FC| > 0.5) within each harmony cluster, mirroring the approach used for
cluster markers above.

1. Compute GO enrichment per cluster and direction

```{r}
# Build named gene lists: one entry per cluster x direction
dge_gene_lists <- list()
for (cl in names(dge_brain_list)) {
  df <- dge_brain_list[[cl]]
  if (is.null(df)) next
  df <- df[!is.na(df$padj), ]
  up_eae <- df$gene[df$padj < 0.05 & df$log2FoldChange >  0.5]
  up_cfa <- df$gene[df$padj < 0.05 & df$log2FoldChange < -0.5]
  if (length(up_eae) > 0) dge_gene_lists[[paste0("Cluster_", cl, "_EAE_up")]] <- up_eae
  if (length(up_cfa) > 0) dge_gene_lists[[paste0("Cluster_", cl, "_CFA_up")]] <- up_cfa
}

# Strip mm10- prefix and convert to ENTREZ IDs
all_dge_genes_clean <- sub("^mm10-", "", unique(unlist(dge_gene_lists)))
gene_map_dge <- bitr(all_dge_genes_clean, fromType = "SYMBOL", toType = "ENTREZID",
                     OrgDb = org.Mm.eg.db)
sym2ent_dge <- setNames(gene_map_dge$ENTREZID, gene_map_dge$SYMBOL)

entrez_dge_lists <- lapply(dge_gene_lists, function(x) {
  na.omit(unname(sym2ent_dge[sub("^mm10-", "", x)]))
})

# GO enrichment
ego_dge_list <- lapply(names(entrez_dge_lists), function(nm) {
  enrichGO(gene          = entrez_dge_lists[[nm]],
           OrgDb         = org.Mm.eg.db,
           keyType       = "ENTREZID",
           ont           = "BP",
           pAdjustMethod = "BH",
           pvalueCutoff  = 0.05,
           qvalueCutoff  = 0.2,
           readable      = TRUE)
})
names(ego_dge_list) <- names(entrez_dge_lists)
```

2. Save GO results to Excel (one sheet per cluster x direction)

```{r}
wb_go_dge <- createWorkbook()

for (nm in names(ego_dge_list)) {
  if (!is.null(ego_dge_list[[nm]]) && nrow(ego_dge_list[[nm]]@result) > 0) {
    go_results <- as.data.frame(ego_dge_list[[nm]])
    sheet_name <- substr(nm, 1, 31)
    addWorksheet(wb_go_dge, sheet_name)
    writeData(wb_go_dge, sheet_name, go_results)
    cat(sprintf("Added sheet: %s (%d GO terms)\n", nm, nrow(go_results)))
  } else {
    cat(sprintf("No significant GO terms for: %s\n", nm))
  }
}

saveWorkbook(wb_go_dge,
             "/date/gcb/gcb_MZ/Humanized_mice_OD/Results/DGE_EAE_vs_CFA_GO_enrichment.xlsx",
             overwrite = TRUE)
cat("\n✓ GO enrichment results saved to Excel file!\n")
```

3. Plot GO enrichment results

```{r, fig.height=20, fig.width=40}
ego_dge_df <- map2_df(names(ego_dge_list), ego_dge_list, ~{
  if (is.null(.y) || nrow(as.data.frame(.y)) == 0) return(NULL)
  df <- as.data.frame(.y)
  df$group <- .x
  df$mLog10Padj <- -log10(df$p.adjust)
  df
})

plot_df_go_dge <- ego_dge_df %>%
  group_by(group) %>%
  slice_max(order_by = mLog10Padj, n = 10, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(Description = fct_reorder(Description, mLog10Padj))

p1 <- ggplot(plot_df_go_dge, aes(x = group, y = Description, size = Count, color = mLog10Padj)) +
  geom_point() +
  scale_color_gradient(low = "gold", high = "red") +
  labs(title = "GO BP enrichment – DEGs (EAE vs CFA)",
       x = "Cluster / Direction", y = "Term", color = "-log10(FDR)") +
  theme_bw(base_size = 12) + theme_classic() +
  theme(
    axis.text.x = element_text(hjust = 1, vjust = 0.5, family = "Arial", size = 24, angle = 45),
    axis.text.y = element_text(family = "Arial", size = 24),
    axis.title.y = element_text(family = "Arial", size = 28, face = "bold"),
    axis.title.x = element_text(family = "Arial", size = 28, face = "bold"),
    legend.text = element_text(size = 24, family = "Arial"),
    legend.title = element_text(size = 32, family = "Arial", face = "bold")
  )
plots <- list(DGE_GO_enrichment = p1)
for (nm in names(plots)) {
  ggplot2::ggsave(file.path("/date/gcb/gcb_MZ/Humanized_mice_OD/Results/Plots/",
                             paste0(nm, ".png")),
                  plots[[nm]], width = 24, height = 26, dpi = 300)
}
p1
```

##5.6 Functional enrichment of DEGs – KEGG pathways

1. Compute KEGG enrichment per cluster and direction

```{r}
ekegg_dge_list <- lapply(names(entrez_dge_lists), function(nm) {
  enrichKEGG(gene         = entrez_dge_lists[[nm]],
             organism     = "mmu",
             pvalueCutoff = 0.05)
})
names(ekegg_dge_list) <- names(entrez_dge_lists)

kegg_dge_df <- map2_df(names(ekegg_dge_list), ekegg_dge_list, ~{
  if (is.null(.y) || nrow(as.data.frame(.y)) == 0) return(NULL)
  df <- as.data.frame(.y)
  df$group <- .x
  df$mLog10Padj <- -log10(df$p.adjust)
  df
})
```

2. Save KEGG results to Excel

```{r}
wb_kegg_dge <- createWorkbook()

for (nm in names(ekegg_dge_list)) {
  if (!is.null(ekegg_dge_list[[nm]]) && nrow(ekegg_dge_list[[nm]]@result) > 0) {
    kegg_results <- as.data.frame(ekegg_dge_list[[nm]])
    kegg_results$mLog10Padj <- -log10(kegg_results$p.adjust)
    sheet_name <- substr(nm, 1, 31)
    addWorksheet(wb_kegg_dge, sheet_name)
    writeData(wb_kegg_dge, sheet_name, kegg_results)
    cat(sprintf("Added sheet: %s (%d KEGG pathways)\n", nm, nrow(kegg_results)))
  } else {
    cat(sprintf("No significant KEGG pathways for: %s\n", nm))
  }
}

saveWorkbook(wb_kegg_dge,
             "/date/gcb/gcb_MZ/Humanized_mice_OD/Results/DGE_EAE_vs_CFA_KEGG_enrichment.xlsx",
             overwrite = TRUE)
cat("\n✓ KEGG enrichment results saved to Excel file!\n")
```

3. Plot KEGG enrichment results

```{r, fig.height=20, fig.width=40}
plot_df_kegg_dge <- kegg_dge_df %>%
  group_by(group) %>%
  slice_max(order_by = mLog10Padj, n = 10, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(Description = fct_reorder(Description, mLog10Padj))

p1 <- ggplot(plot_df_kegg_dge, aes(x = group, y = Description, size = Count, color = mLog10Padj)) +
  geom_point() +
  scale_color_gradient(low = "gold", high = "red") +
  labs(title = "KEGG enrichment – DEGs (EAE vs CFA)",
       x = "Cluster / Direction", y = "Pathway", color = "-log10(FDR)") +
  theme_bw(base_size = 12) + theme_classic() +
  theme(
    axis.text.x = element_text(hjust = 1, vjust = 0.5, family = "Arial", size = 24, angle = 45),
    axis.text.y = element_text(family = "Arial", size = 24),
    axis.title.y = element_text(family = "Arial", size = 28, face = "bold"),
    axis.title.x = element_text(family = "Arial", size = 28, face = "bold"),
    legend.text = element_text(size = 24, family = "Arial"),
    legend.title = element_text(size = 32, family = "Arial", face = "bold")
  )
plots <- list(DGE_KEGG_enrichment = p1)
for (nm in names(plots)) {
  ggplot2::ggsave(file.path("/date/gcb/gcb_MZ/Humanized_mice_OD/Results/Plots/",
                             paste0(nm, ".png")),
                  plots[[nm]], width = 24, height = 26, dpi = 300)
}
p1
```





