---
title: "OPC in depth"
author: "Mattia Zaghi"
date: "2026-08-26"
output:
  word_document: default
  html_document: default
  pdf_document: default
---

## What is this vignette for?
This Vignette contains all the necessary data analysis steps followed to analyzed the single cell RNA-seq of Mature humanized_mouse_braindendrocytes treated with FGF and Tomato controls including the code to generate the different figures.


##1.Load packages, configure directories and create the Seurat object

This section of the code is necessary to load all the necessary packages for the analysis, configure the directories and create the Seurat object that is going to be used for the rest of the analysis.

##1.1 Load packages, configure directories

1.First load all the necessary packages in order to run all the subsequent steps & set the working directory where all the output of this analysis will be saved and stored

```{r setup, include=FALSE}
library(Seurat)
library(future)
library(tidyverse)
library(dplyr)
library(ggthemes)
library(scales)
library(cowplot)
library(RColorBrewer)
library(openxlsx)
library(homologene)
library(clusterProfiler)
library(org.Mm.eg.db)
library(org.Hs.eg.db)
library(purrr)
library(forcats)
library(readxl)
library(hdf5r)
library(patchwork)
library(pals)
library(knitr)
library(DoubletFinder)
library(ggpubr)
library(patchwork)
library(rstatix)
library(pheatmap)
opts_knit$set(root.dir = "/date/gcb/gcb_MZ/Humanized_mice_OD/")# change accordingly

#source("/home/mattia/FGF_signaling_induces_disease-associated_humanized_mouse_braindendrocytes/functions.R")
```

##1.1 Load the OPC-OPCs object 


```{r}
OL <- readRDS("/date/gcb/gcb_MZ/Humanized_mice_OD/Results/objects/Oligodendroglia_SCT.rds") 


```


first check clusters on the dimplot

```{r}
my_colors <- c(
  "#1F77B4", "#FF7F0E", "#2CA02C", "#D62728", "#9467BD",
  "#8C564B", "#17BECF", "#BCBD22", "#7F7F7F", "#e377c2",
  "#aec7e8", "#ffbb78", "#98df8a"
)


p1 <- DimPlot(OL, reduction = "umap_OL", group.by = c("OL_clusters"), cols = my_colors) + labs(x = "UMAP 1", y = "UMAP 2") + #UMAP with samples
 theme_minimal() +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      size = 24,
      face = "bold"
    ),
    axis.title.x = element_text(size = 20),
    axis.title.y = element_text(size = 20),
    legend.text = element_text(size = 16),         # Legend labels
    legend.title = element_text(size = 24, face = "bold"),  # Legend title
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank()
  )
p1
```
Remove possible cell doublets contamination from OPCs

```{r}
OPC_EAE <- subset(OL, idents = c("0", "1", "2","4","5"))
```

Renormalize the datasets and recluster using only the selected cells   

```{r}
DefaultAssay(OPC_EAE) <- "RNA"

# Remove the OPC_EAEd SCT assay
OPC_EAE[["SCT"]] <- NULL

# Now run SCTransform renormalizing with only the OPC_EAE_OPC cells
OPC_EAE <- SCTransform(OPC_EAE, vst.flavor = "v2", verbose = FALSE)

# Continue with custom-named reductions
OPC_EAE <- RunPCA(OPC_EAE, verbose = TRUE, reduction.name = "pca_OPC_EAE_SCT")

OPC_EAE <- FindNeighbors(OPC_EAE, reduction = "pca_OPC_EAE_SCT", dims = 1:15,
                      graph.name = c("OPC_EAE_nn", "OPC_EAE_snn"))

OPC_EAE <- FindClusters(OPC_EAE, resOPC_EAEution = 0.1, 
                     graph.name = "OPC_EAE_snn",
                     cluster.name = "OPC_EAE_clusters")

OPC_EAE <- RunUMAP(OPC_EAE, reduction = "pca_OPC_EAE_SCT", dims = 1:15, 
                reduction.name = "umap_OPC_EAE")


```

Reclustering of the EAE OPCs

```{r}
OPC_EAE <- subset(OL, idents = c("0", "1", "2","4","5"))
```

Renormalize the datasets and recluster using only the selected cells   

```{r}
DefaultAssay(OPC_EAE) <- "RNA"

# Remove the OPC_EAEd SCT assay
OPC_EAE[["SCT"]] <- NULL

# Now run SCTransform renormalizing with only the OPC_EAE_OPC cells
OPC_EAE <- SCTransform(OPC_EAE, vst.flavor = "v2", verbose = FALSE)

# Continue with custom-named reductions
OPC_EAE <- RunPCA(OPC_EAE, verbose = TRUE, reduction.name = "pca_OPC_EAE_SCT")

OPC_EAE <- FindNeighbors(OPC_EAE, reduction = "pca_OPC_EAE_SCT", dims = 1:15,
                      graph.name = c("OPC_EAE_nn", "OPC_EAE_snn"))

OPC_EAE <- FindClusters(OPC_EAE, resolution = 0.1, 
                     graph.name = "OPC_EAE_snn",
                     cluster.name = "OPC_EAE_clusters")

OPC_EAE <- RunUMAP(OPC_EAE, reduction = "pca_OPC_EAE_SCT", dims = 1:15, 
                reduction.name = "umap_OPC_EAE")


```





```{r}
my_colors <- c(
  "#1F77B4", "#FF7F0E", "#2CA02C", "#D62728", "#9467BD",
  "#8C564B", "#17BECF", "#BCBD22", "#7F7F7F", "#e377c2",
  "#aec7e8", "#ffbb78", "black", "orange"
)

p1 <- DimPlot(OPC_EAE, reduction = "umap_OPC_EAE", group.by = c("OPC_EAE_clusters"), cols = my_colors) + labs(x = "UMAP 1", y = "UMAP 2") + #UMAP with new clustering
 theme_minimal() +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      size = 24,
      face = "bold"
    ),
    axis.title.x = element_text(size = 20),
    axis.title.y = element_text(size = 20),
    legend.text = element_text(size = 16),         # Legend labels
    legend.title = element_text(size = 24, face = "bold"),  # Legend title
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank()
  )

p2 <- DimPlot(OPC_EAE, reduction = "umap_OPC_EAE", group.by = c("orig.ident")) + labs(x = "UMAP 1", y = "UMAP 2") + #UMAP with samples
 theme_minimal() +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      size = 24,
      face = "bold"
    ),
    axis.title.x = element_text(size = 20),
    axis.title.y = element_text(size = 20),
    legend.text = element_text(size = 16),         # Legend labels
    legend.title = element_text(size = 24, face = "bold"),  # Legend title
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank()
  )

p3 <- DimPlot(OPC_EAE, reduction = "umap_OPC_EAE", group.by = c("sample")) + labs(x = "UMAP 1", y = "UMAP 2") + #UMAP with samples
 theme_minimal() +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      size = 24,
      face = "bold"
    ),
    axis.title.x = element_text(size = 20),
    axis.title.y = element_text(size = 20),
    legend.text = element_text(size = 16),         # Legend labels
    legend.title = element_text(size = 24, face = "bold"),  # Legend title
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank()
  )

p4 <- DimPlot(OPC_EAE, reduction = "umap_OPC_EAE", group.by = c("broad_clustering")) + labs(x = "UMAP 1", y = "UMAP 2") + #UMAP with samples
 theme_minimal() +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      size = 24,
      face = "bold"
    ),
    axis.title.x = element_text(size = 20),
    axis.title.y = element_text(size = 20),
    legend.text = element_text(size = 16),         # Legend labels
    legend.title = element_text(size = 24, face = "bold"),  # Legend title
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank()
  )

p1
p2
p3
p4

```

Marker features calculations in OLs

```{r}

# find markers for every cluster compared to all remaining cells, report only the positive
# ones

  DefaultAssay(OPC_EAE) <- "SCT"
  OPC_EAE <- PrepSCTFindMarkers(OPC_EAE, assay = "SCT", verbose = TRUE)
  OL_markers <- FindAllMarkers(OPC_EAE, only.pos = TRUE)


 # Create a new workbook
wb <- createWorkbook()

# Loop through each cluster and add data to a new sheet
for (cluster_value in unique(OL_markers$cluster)) {
  cluster_data <- subset(OL_markers, cluster == cluster_value)
  sheet_name <- paste0("Cluster_", cluster_value)
  addWorksheet(wb, sheet_name)
  writeData(wb, sheet_name, cluster_data)
}

# Save the workbook to a file 
saveWorkbook(wb, "/date/gcb/gcb_MZ/Humanized_mice_OD/Results/cluster_markers/Oligodendroglia_clean_markers.xlsx", overwrite = TRUE)
```





##1.3 OPC EAE reclustering

Reclustering of the EAE OPCs

```{r}
MOL_EAE <- subset(OL_clean, idents = c("1", "2"))
```

Renormalize the datasets and recluster using only the selected cells   

```{r}
DefaultAssay(MOL_EAE) <- "RNA"

# Remove the MOL_EAEd SCT assay
MOL_EAE[["SCT"]] <- NULL

# Now run SCTransform renormalizing with only the MOL_EAE_OPC cells
MOL_EAE <- SCTransform(MOL_EAE, vst.flavor = "v2", verbose = FALSE)

# Continue with custom-named reductions
MOL_EAE <- RunPCA(MOL_EAE, verbose = TRUE, reduction.name = "pca_MOL_EAE_SCT")

MOL_EAE <- FindNeighbors(MOL_EAE, reduction = "pca_MOL_EAE_SCT", dims = 1:15,
                      graph.name = c("MOL_EAE_nn", "MOL_EAE_snn"))

MOL_EAE <- FindClusters(MOL_EAE, resolution = 0.1, 
                     graph.name = "MOL_EAE_snn",
                     cluster.name = "MOL_EAE_clusters")

MOL_EAE <- RunUMAP(MOL_EAE, reduction = "pca_MOL_EAE_SCT", dims = 1:15, 
                reduction.name = "umap_MOL_EAE")


```


```{r}
my_colors <- c(
  "#1F77B4", "#FF7F0E", "#2CA02C", "#D62728", "#9467BD",
  "#8C564B", "#17BECF", "#BCBD22", "#7F7F7F", "#e377c2",
  "#aec7e8", "#ffbb78", "black", "orange"
)

p1 <- DimPlot(OPC_EAE, reduction = "umap_OPC_EAE", group.by = c("OPC_EAE_clusters"), cols = my_colors) + labs(x = "UMAP 1", y = "UMAP 2") + #UMAP with new clustering
 theme_minimal() +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      size = 24,
      face = "bold"
    ),
    axis.title.x = element_text(size = 20),
    axis.title.y = element_text(size = 20),
    legend.text = element_text(size = 16),         # Legend labels
    legend.title = element_text(size = 24, face = "bold"),  # Legend title
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank()
  )

p2 <- DimPlot(OPC_EAE, reduction = "umap_OPC_EAE", group.by = c("orig.ident")) + labs(x = "UMAP 1", y = "UMAP 2") + #UMAP with samples
 theme_minimal() +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      size = 24,
      face = "bold"
    ),
    axis.title.x = element_text(size = 20),
    axis.title.y = element_text(size = 20),
    legend.text = element_text(size = 16),         # Legend labels
    legend.title = element_text(size = 24, face = "bold"),  # Legend title
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank()
  )

p3 <- DimPlot(OPC_EAE, reduction = "umap_OPC_EAE", group.by = c("sample")) + labs(x = "UMAP 1", y = "UMAP 2") + #UMAP with samples
 theme_minimal() +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      size = 24,
      face = "bold"
    ),
    axis.title.x = element_text(size = 20),
    axis.title.y = element_text(size = 20),
    legend.text = element_text(size = 16),         # Legend labels
    legend.title = element_text(size = 24, face = "bold"),  # Legend title
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank()
  )

p4 <- DimPlot(OPC_EAE, reduction = "umap_OPC_EAE", group.by = c("broad_clustering")) + labs(x = "UMAP 1", y = "UMAP 2") + #UMAP with samples
 theme_minimal() +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      size = 24,
      face = "bold"
    ),
    axis.title.x = element_text(size = 20),
    axis.title.y = element_text(size = 20),
    legend.text = element_text(size = 16),         # Legend labels
    legend.title = element_text(size = 24, face = "bold"),  # Legend title
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank()
  )

p1
p2
p3
p4

```

Marker features calculations in OLs

```{r}

# find markers for every cluster compared to all remaining cells, report only the positive
# ones

  DefaultAssay(MOL_EAE) <- "SCT"
 MOL_EAE <- PrepSCTFindMarkers(MOL_EAE, assay = "SCT", verbose = TRUE)
  OL_markers <- FindAllMarkers(MOL_EAE, only.pos = TRUE)


 # Create a new workbook
wb <- createWorkbook()

# Loop through each cluster and add data to a new sheet
for (cluster_value in unique(OL_markers$cluster)) {
  cluster_data <- subset(OL_markers, cluster == cluster_value)
  sheet_name <- paste0("Cluster_", cluster_value)
  addWorksheet(wb, sheet_name)
  writeData(wb, sheet_name, cluster_data)
}

# Save the workbook to a file 
saveWorkbook(wb, "/date/gcb/gcb_MZ/Humanized_mice_OD/Results/cluster_markers/MOL_EAE_markers.xlsx", overwrite = TRUE)
```

##1.4 OPC OPCs EAE reclustering

eclustering of the EAE OPCs

```{r}
OPC <- subset(OL_clean, idents = c("3", "4"))
```

Renormalize the datasets and recluster using only the selected cells   

```{r}
DefaultAssay(OPC) <- "RNA"

# Remove the OPC_EAEd SCT assay
OPC[["SCT"]] <- NULL

# Now run SCTransform renormalizing with only the OPC_EAE_OPC cells
OPC <- SCTransform(OPC, vst.flavor = "v2", verbose = FALSE)

# Continue with custom-named reductions
OPC <- RunPCA(OPC, verbose = TRUE, reduction.name = "pca_OPC_SCT")

OPC <- FindNeighbors(OPC, reduction = "pca_OPC_SCT", dims = 1:15,
                      graph.name = c("OPC_nn", "OPC_snn"))

OPC <- FindClusters(OPC, resolution = 0.1, 
                     graph.name = "OPC_snn",
                     cluster.name = "OPC_clusters")

OPC <- RunUMAP(OPC, reduction = "pca_OPC_SCT", dims = 1:15, 
                reduction.name = "umap_OPC")


```

Export metadata and UMAP coordinates
```{r}
suppressPackageStartupMessages({ library(Seurat); library(openxlsx) })

out_xlsx <- "/date/gcb/gcb_MZ/Humanized_mice_OD/Results/UMAP_Metadata_count_matrices/Oligo_lineage_metadata.xlsx"


cat("marks:", paste(names(objects), collapse = ", "), "\n")

wb <- createWorkbook()
for (mk in names(objects)) {
  md <- objects[[mk]]@meta.data
  df <- data.frame(cell = rownames(md), md, row.names = NULL,
                   check.names = FALSE, stringsAsFactors = FALSE)
  addWorksheet(wb, mk)
  writeData(wb, mk, df)
  cat(sprintf("  %-9s : %d cells x %d metadata columns\n", mk, nrow(df), ncol(df) - 1))
}
saveWorkbook(wb, out_xlsx, overwrite = TRUE)
cat("Wrote:", out_xlsx, "\n")


out_xlsx <- "/date/gcb/gcb_MZ/Humanized_mice_OD/Results/UMAP_Metadata_count_matrices/Oligo_lineage_UMAP_coordinates.xlsx"

cat("marks:", paste(names(objects), collapse = ", "), "\n")

wb <- createWorkbook()
for (mk in names(objects)) {
  md <- objects[[mk]]@reductions[[paste0("umap_",mk)]]@cell.embeddings
  df <- data.frame(cell = rownames(md), md, row.names = NULL,
                   check.names = FALSE, stringsAsFactors = FALSE)
  addWorksheet(wb, mk)
  writeData(wb, mk, df)
  cat(sprintf("  %-9s : %d cells x %d metadata columns\n", mk, nrow(df), ncol(df) - 1))
}
saveWorkbook(wb, out_xlsx, overwrite = TRUE)
cat("Wrote:", out_xlsx, "\n")

out_xlsx <- "/date/gcb/gcb_MZ/Humanized_mice_OD/Results/UMAP_Metadata_count_matrices/Oligo_lineage_UMAP_coordinates.xlsx"


OL_count_matrices<-data.frame(OL_clean@assays[["RNA"]]@layers[["counts"]])
  
  
  
```



