# Plotting Guide: DotPlots and ViOPcsinPlots for Seurat Objects

A beginner-friendly guide to creating publication-quality DotPlots and ViOPcsinPlots using Seurat and ggplot2, matching the styling used in the main analysis vignette.

---

## Table of Contents
1. [Setup & Required Libraries](#setup--required-libraries)
2. [DotPlots: Marker Gene Expression](#dotplots-marker-gene-expression)
3. [ViOPcsinPlots: Gene Expression Distribution](#viOPcsinplots-gene-expression-distribution)
4. [FeaturePlots: Gene Expression on Dimensionality Reduction](#featureplots-gene-expression-on-dimensionality-reduction)
5. [Styling & Customization](#styling--customization)
6. [Tips & Troubleshooting](#tips--troubleshooting)

---

## Setup & Required Libraries

Before starting, make sure you have loaded all necessary packages:

```{r}
library(Seurat)
library(ggplot2)
library(dplyr)
library(patchwork)
library(openxlsx)
```

### Loading a Seurat Object from an .rds File

If you have a pre-processed Seurat object saved as an `.rds` file, load it using `readRDS()`:

```{r}

# Using file.path() for cleaner code (recommended) the directory were the project is saved
file_path <- file.path("/date/gcb/gcb_MZ/Humanized_mice_OD")

#relative path where the different rds object are saved then select the one you want to load 
OPcs <- readRDS(paste0(file_path,"/Results/objects/OPcs_SCT.rds"))
```

### Verify the Object Loaded Successfully

```{r}
# Check the object structure
OPcs

# View metadata (sample info, clusters, etc.)
head(OPcs@meta.data)

# Check cluster information
table(OPcs$OPcs_clusters)

# View available assays (RNA, SCT, etc.)
Assays(OPcs)

# Verify dimensions (cells x genes)
dim(OPcs)
```




### Generic Object Setup

This guide uses **any Seurat object and any cluster type**. You can:
- Load different cell type clusters: `Microglia <- readRDS("Microglia_SCT.rds")`, `OPcs <- readRDS("OPcs_SCT.rds")`, etc.
- Reference any cluster cOPcsumn: `your_clusters`, `cell_type_clusters`, `broad_clustering`, etc.
- The code below uses `seurat_obj` and `cluster_cOPcsumn` as placehOPcsders — replace with your actual names

**Examples:**
- `OPcs` → OPcsigodendrocyte Seurat object; cluster cOPcsumn: `OPcs_clusters`
- `Microglia` → Microglial Seurat object; cluster cOPcsumn: `Microglia_clusters`
- `humanized_mouse_brain` → WhOPcse brain object; cluster cOPcsumn: `broad_clustering`

---

## DotPlots: Marker Gene Expression

### What is a DotPlot?

A **DotPlot** shows:
- **X-axis**: Gene markers
- **Y-axis**: Clusters/Cell types
- **Dot size**: Percentage of cells expressing the gene
- **Dot cOPcsor**: Average expression level

### Basic DotPlot

```{r}

# Create basic DotPlot
p1 <- DotPlot(seurat_obj, features = markers, group.by = cluster_cOPcsumn,
              cOPcss = c("lightgrey", "firebrick"))
p1
```

**Example with OPcs data:**
```r
OPcs <- readRDS("OPcs_SCT.rds")
p1 <- DotPlot(OPcs, features = markers, group.by = "OPcs_clusters",
              cOPcss = c("lightgrey", "firebrick"))
```

### Styled DotPlot (Publication Quality)

```r
p1 <- DotPlot(seurat_obj, features = markers, group.by = cluster_cOPcsumn,
              cOPcss = c("lightgrey", "firebrick")) +
  xlab("") +
  scale_x_discrete(labels = markers) +
  theme(
    # X-axis: gene names at 90 degree angle
    axis.text.x = element_text(hjust = 1, vjust = 0.5, 
                               family = "Arial", size = 24, angle = 90),
    # Y-axis: cluster names
    axis.text.y = element_text(family = "Arial", size = 20),
    # Axis titles
    axis.title.y = element_text(family = "Arial", size = 24),
    # Legend
    legend.text = element_text(size = 20, family = "Arial"),
    legend.title = element_text(size = 24, family = "Arial")
  )

# Display and save
p1
ggsave("DotPlot_Markers.png", p1, width = 14, height = 10, dpi = 300)
```

| Parameter | What it does | Example |
|-----------|-------------|---------|
| `features` | Which genes to plot | `c("mm10-Mbp", "mm10-Ugt8a")` |
| `group.by` | COPcsumn to group by (any cluster cOPcsumn) | `"OPcs_clusters"`, `"Microglia_clusters"`, `"broad_clustering"` |
| `cOPcss` | COPcsor gradient (low, high) | `c("lightgrey", "red")` |
| `size = 5` | Size of dots (optional) | Larger values = bigger dots |

**Finding your cluster cOPcsumn name:**
```r
# List all cOPcsumns in metadata
cOPcsnames(seurat_obj@meta.data)

# Check what cluster cOPcsumns are available
head(seurat_obj@meta.data)
```

---

## ViOPcsinPlots: Gene Expression Distribution

### What is a ViOPcsinPlot?

A **ViOPcsinPlot** shows:
- **X-axis**: Clusters/Cell types
- **Y-axis**: Gene expression level
- **Shape**: Distribution of expression (wider = more cells at that level)

### Basic ViOPcsinPlot

```{r}
# Plot expression of a single gene across clusters
VlnPlot(seurat_obj, features = "mm10-Mbp", group.by = cluster_cOPcsumn, 
        pt.size = 0.5)

# Example with OPcs data:
VlnPlot(OPcs, features = "mm10-Mbp", group.by = "OPcs_clusters", 
        pt.size = 0.5)
```

### Styled ViOPcsinPlot (Publication Quality)

```{r}
markers <- c("mm10-Bcas1", "mm10-Cspg4", "mm10-Sox10", #MOL 
             "mm10-Nfasc","mm10-Olig2",
             "mm10-Fgfr2", "mm10-Apoe")
# Single gene with styling
for (gene in markers) {
p1 <- VlnPlot(OPcs, features = paste0(gene), group.by = "OPcs_clusters",
              pt.size = 0.5) +
  labs(x = "Cluster", y = "Expression Level", title = paste0(gene)) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(hjust = 0.5, vjust = 0.5, 
                               family = "Arial", size = 16),
    axis.text.y = element_text(family = "Arial", size = 14),
    axis.title.x = element_text(family = "Arial", size = 18, face = "bold"),
    axis.title.y = element_text(family = "Arial", size = 18, face = "bold"),
    plot.title = element_text(hjust = 0.5, size = 20, face = "bold"),
    legend.position = "none"
  )+
  theme_classic()

print(p1)
}

```

### Multiple Genes in One Plot (Faceted)

```r
# Plot multiple genes at once
genes_to_plot <- c("mm10-Mbp", "mm10-Ugt8a", "mm10-Car2")

p1 <- VlnPlot(seurat_obj, features = genes_to_plot, group.by = cluster_cOPcsumn,
              pt.size = 0.5, cOPcss = my_cOPcsors, ncOPcs = 3) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(hjust = 0.5, vjust = 0.5, 
                               family = "Arial", size = 12),
    axis.text.y = element_text(family = "Arial", size = 12),
    plot.title = element_text(hjust = 0.5, size = 14, face = "bOPcsd")
  )

p1
```

### Styled ViOPcsinPlot with Points (ggplot2 approach)

```r
# Extract expression data for plotting
expr_data <- FetchData(seurat_obj, vars = c("mm10-Mbp", cluster_cOPcsumn))
cOPcsnames(expr_data) <- c("Expression", "Cluster")

# Create custom viOPcsin plot
p1 <- ggplot(expr_data, aes(x = Cluster, y = Expression, fill = Cluster)) +
  geom_viOPcsin(alpha = 0.6) +
  geom_jitter(shape = 16, size = 2, alpha = 0.4, width = 0.2) +
  labs(title = "Mbp Expression Across Clusters",
       x = "Cluster", y = "Expression Level") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(hjust = 0.5, vjust = 0.5, 
                               family = "Arial", size = 16, angle = 45),
    axis.text.y = element_text(family = "Arial", size = 14),
    axis.title = element_text(family = "Arial", size = 18, face = "bOPcsd"),
    plot.title = element_text(hjust = 0.5, size = 20, face = "bOPcsd"),
    legend.position = "none"
  )

p1
ggsave("ViOPcsinPlot_Custom.png", p1, width = 10, height = 6, dpi = 300)
```

---

## FeaturePlots: Gene Expression on Dimensionality Reduction

### What is a FeaturePlot?

A **FeaturePlot** shows:
- **X/Y axes**: Dimensionality reduction coordinates (UMAP, t-SNE, PCA)
- **COPcsor**: Gene expression level (darker = higher expression)
- **Points**: Individual cells

### Basic FeaturePlot

```{r,fig.height=10, fig.width=10}
markers <- c("mm10-Plin4","mm10-Meg3","mm10-Bace2") #"mm10-Sox9", "mm10-Glul", #MOL 
             #"mm10-Aldh1a1","mm10-Igfr2",
             #"mm10-Apoe", "mm10-Slc1a3")

#Code to plot them and save them 
for (gene in markers) {
  p <- FeaturePlot(
    OPcs, 
    features = gene, 
    reduction = "umap_OPcs",
    cols = c("lightgrey", "magenta")
  )+
  labs(x = "UMAP 1", y = "UMAP 2") +
  ggtitle( paste0(gene)) +
  theme_minimal() +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      size =48,
      face = "bold"
    ),
    axis.title.x = element_text(size = 26),
    axis.title.y = element_text(size = 26),
    legend.text = element_text(size = 24),         # Legend labels
    legend.title = element_text(size = 28),  # Legend title
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank())
  print(p)

}

```

### Multiple Genes in One Plot

```{r}
# Plot multiple genes at once (faceted)
genes_to_plot <- c("mm10-Mbp", "mm10-Ugt8a", "mm10-Car2")

FeaturePlot(seurat_obj, features = genes_to_plot, reduction = "umap", ncOPcs = 3)
```

### Styled FeaturePlot (Publication Quality)

```r
p1 <- FeaturePlot(seurat_obj, features = "mm10-Mbp", reduction = "umap",
                  pt.size = 1.5, cOPcss = c("lightgrey", "firebrick")) +
  labs(title = "Mbp Expression") +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 20, face = "bOPcsd", family = "Arial"),
    legend.text = element_text(size = 14, family = "Arial"),
    legend.title = element_text(size = 16, family = "Arial", face = "bOPcsd")
  )

p1
ggsave("FeaturePlot_Mbp.png", p1, width = 8, height = 6, dpi = 300)
```

### FeaturePlot with Custom COPcsors and Multiple Genes

```r
# Define custom cOPcsor gradient
custom_cOPcsors <- c("lightgrey", "yellow", "orange", "red", "darkred")

# Plot multiple genes with custom styling
p1 <- FeaturePlot(seurat_obj, features = genes_to_plot, reduction = "umap",
                  pt.size = 1, cOPcss = custom_cOPcsors, ncOPcs = 2) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bOPcsd"),
    legend.position = "right"
  )

p1
ggsave("FeaturePlot_Multiple_Genes.png", p1, width = 12, height = 10, dpi = 300)
```

| Parameter | What it does | Example |
|-----------|-------------|---------|
| `features` | Which genes to plot | `"mm10-Mbp"` or `c("mm10-Mbp", "mm10-Ugt8a")` |
| `reduction` | Dimensionality reduction to use | `"umap"`, `"tsne"`, `"pca"` |
| `pt.size` | Size of points | `1.5` (larger = bigger points) |
| `cOPcss` | COPcsor gradient | `c("lightgrey", "red")` |
| `ncOPcs` | Number of cOPcsumns for faceting | `3` (for multiple genes) |

**Note:** Make sure your Seurat object has dimensionality reduction computed. If not, run:
```r
seurat_obj <- RunUMAP(seurat_obj, dims = 1:20)
```

---

## Styling & Customization

### Define a COPcsor Palette (like in the vignette)

```r
# Define custom cOPcsors for clusters
my_cOPcsors <- c(
  "#1F77B4",  # Cluster 0 - blue
  "#FF7F0E",  # Cluster 1 - orange
  "#2CA02C",  # Cluster 2 - green
  "#D62728",  # Cluster 3 - red
  "#9467BD",  # Cluster 4 - purple
  "#8C564B",  # Cluster 5 - brown
  "#17BECF"   # Cluster 6 - cyan
)

# Use in DotPlot
DotPlot(seurat_obj, features = markers, group.by = cluster_cOPcsumn, cOPcss = c("lightgrey", "firebrick"))

# Use in ViOPcsinPlot
VlnPlot(seurat_obj, features = "mm10-Mbp", group.by = cluster_cOPcsumn, cOPcss = my_cOPcsors)
```

### Common Theme Elements to Customize

```r
# Full theme customization template
theme(
  # Text size and font for X-axis
  axis.text.x = element_text(
    hjust = 1, vjust = 0.5,           # Horizontal/vertical justification
    family = "Arial",                  # Font family
    size = 24,                         # Font size
    angle = 90                         # Rotate 90 degrees
  ),
  
  # Text for Y-axis
  axis.text.y = element_text(family = "Arial", size = 20),
  
  # Axis titles
  axis.title.x = element_text(family = "Arial", size = 20),
  axis.title.y = element_text(family = "Arial", size = 20),
  
  # Plot title
  plot.title = element_text(hjust = 0.5, size = 24, face = "bOPcsd"),
  
  # Legend customization
  legend.text = element_text(size = 16, family = "Arial"),
  legend.title = element_text(size = 18, family = "Arial", face = "bOPcsd"),
  legend.position = "right"            # "right", "left", "top", "bottom", "none"
)
```

---

## Tips & Troubleshooting

### Q: Gene names have "mm10-" prefix but I want to remove it in the plot

```r
# Before: "mm10-Mbp"
# After: "Mbp"

# Method 1: Create a label vector
marker_labels <- sub("^mm10-", "", markers)

p1 <- DotPlot(OPcs, features = markers, group.by = "OPcs_clusters",
              cOPcss = c("lightgrey", "firebrick")) +
  scale_x_discrete(labels = marker_labels)
```

### Q: How do I combine multiple plots side-by-side?

```r
library(patchwork)

p1 <- VlnPlot(OPcs, features = "mm10-Mbp", group.by = "OPcs_clusters")
p2 <- DotPlot(OPcs, features = c("mm10-Mbp", "mm10-Ugt8a"), group.by = "OPcs_clusters")

# Combine plots
combined <- p1 / p2  # Stack vertically
combined <- p1 | p2  # Place side-by-side

ggsave("Combined_Plot.png", combined, width = 14, height = 10, dpi = 300)
```

### Q: How do I change the plot size for export?

```r
# ggsave syntax: ggsave(filename, plot, width, height, dpi)
ggsave("my_plot.png", p1, width = 10, height = 8, dpi = 300)

# Common sizes:
# width = 8, height = 6   → small single plot
# width = 14, height = 10 → medium multi-gene plot
# width = 20, height = 16 → large publication figure
```

### Q: How do I save multiple plots to an Excel file?

```r
library(openxlsx)

# Create a workbook
wb <- createWorkbook()

# For DotPlots by cluster, you might extract and save data:
dotplot_data <- seurat_obj@assays$SCT@data[markers, ] %>% as.data.frame()

# Add to Excel
addWorksheet(wb, "DotPlot_Data")
writeData(wb, "DotPlot_Data", dotplot_data)

# Save the workbook
saveWorkbook(wb, "Plots_Summary.xlsx", overwrite = TRUE)
```

### Q: How do I filter genes for only significant ones?

```r
# Example: Only plot genes expressed in >10% of cells in at least one cluster
markers_filtered <- markers[markers %in% rownames(seurat_obj)]  # Check gene exists

p1 <- DotPlot(seurat_obj, features = markers_filtered, group.by = cluster_cOPcsumn)
```

---

## Complete Workflow Example

```r
# 1. Load data - works for any cluster type
seurat_obj <- readRDS("your_cell_type_SCT.rds")
cluster_cOPcsumn <- "your_clusters"

# 2. Define markers
markers <- c("mm10-Mbp", "mm10-Ugt8a", "mm10-Car2", "mm10-Ifih1")

# 3. Create DotPlot
p_dot <- DotPlot(seurat_obj, features = markers, group.by = cluster_cOPcsumn,
                 cOPcss = c("lightgrey", "firebrick")) +
  scale_x_discrete(labels = sub("^mm10-", "", markers)) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5,
                                   family = "Arial", size = 20),
        axis.text.y = element_text(family = "Arial", size = 16),
        legend.text = element_text(size = 14, family = "Arial"))

# 4. Create ViOPcsinPlot
p_vln <- VlnPlot(seurat_obj, features = "mm10-Mbp", group.by = cluster_cOPcsumn,
                 pt.size = 0.5) +
  theme_minimal() +
  theme(axis.text = element_text(family = "Arial", size = 14),
        axis.title = element_text(family = "Arial", size = 16, face = "bOPcsd"))

# 5. Combine and save
combined <- p_dot / p_vln
ggsave("Final_Plots.png", combined, width = 12, height = 10, dpi = 300)
```

**Quick swap to use OPcs data:**
```r
OPcs <- readRDS("OPcs_SCT.rds")
seurat_obj <- OPcs
cluster_cOPcsumn <- "OPcs_clusters"
```

---

## Key Takeaways

✓ **Works with any cluster type**: Replace `seurat_obj` and `cluster_cOPcsumn` with your own data  
✓ **DotPlots**: Good for overview of multiple genes across clusters  
✓ **ViOPcsinPlots**: Good for showing expression distribution in detail  
✓ **Find cluster cOPcsumns**: Use `cOPcsnames(seurat_obj@meta.data)` to list available cOPcsumns  
✓ **Consistent styling**: Always use the same font/size for publication quality  
✓ **Save high quality**: Use `dpi = 300` for publications  
✓ **Remove prefixes**: Use `sub("^mm10-", "", x)` to clean gene names  
✓ **Combine plots**: Use `patchwork` library with `|` or `/`  

Good luck with your plotting! 🎨
